### 2018-06-21 report.tex
First commit

\documentclass[11pt, a4paper]{report}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
\usepackage{wallpaper}
\usepackage{palatino}
\usepackage{changepage}
% Setup captions
%\captionstyle[\centering]{\centering}
%\changecaptionwidth
%\captionwidth{0.8\linewidth}

% Protect against widows and orphans
%\clubpenalty=10000
%\widowpenalty=10000

%\linespread{1.2}

%\raggedbottom

%\chapterstyle{ger}

%\maxsecnumdepth{subsection}

%%  Setup fancy style quotation
%%  ==================================================================
%\usepackage{tikz}
%\usepackage{framed}

%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font

% Make commands for the quotes
%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}

% select a colour for the shading
%\definecolor{shadecolor}{rgb}{1,1,1}

% wrap everything in its own environment
%\newenvironment{shadequote}% 
%{\begin{snugshade}\begin{quote}\openquote}
%{\hfill\closequote\end{quote}\end{snugshade}}

%% MY STUFF

%% Language and font encodings

%% Sets page size and margins
%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
\usepackage{graphicx}
\graphicspath{{./plots/}{./figures/}}

%\fancypagestyle{title}{%
%  \setlength{\headheight}{120pt}%
%  \fancyhf{}% No header/footer
%  \renewcommand{\headrulewidth}{0pt}% No header rule
%  \renewcommand{\footrulewidth}{0pt}% No footer rule
%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
%}%

%\usepackage{amsmath}
%\usepackage[format=plain]{caption}
\usepackage{subcaption}
\usepackage{float} % force float
\usepackage[nottoc,numbib]{tocbibind} % make bibliography appear in toc
%\usepackage[margin=-50pt]{caption}
%\usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
%\setlength{\parindent}{4em}
%\setlength{\parskip}{1em}
%\usepackage{mathpazo} % Use the Palatino font
\usepackage{parskip}% http://ctan.org/pkg/parskip
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{wrapfig}
\usepackage[export]{adjustbox}
\usepackage{textcomp}
\usepackage[bottom]{footmisc}

\usepackage{titling}
\newcommand{\subtitle}[1]{%
  \posttitle{%
    \par\end{center}
    \begin{center}\large#1\end{center}
    \vskip0.5em}%
}

\renewcommand\thesubfigure{\arabic{subfigure}}  
\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}

\usepackage{listings}
\lstset{basicstyle=\ttfamily,
  showstringspaces=false,
  commentstyle=\color{red},
  keywordstyle=\color{blue}
}

\makeatletter
\renewcommand\p@subfigure{\thefigure\,}
\renewcommand\thesubfigure{\alph{subfigure})}
% If desired for table as well:
% \renewcommand\p@subtable{\thetable\,}
% \renewcommand\thesubtable{\alph{subtable})}
\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
\makeatother

\renewcommand{\baselinestretch}{1.5}
\DeclareCaptionSubType*[arabic]{figure}

%\makeatletter
%\renewcommand\thesection{}
%\renewcommand\thesubsection{\@arabic\c@subsection}
%\makeatother

\usepackage{mathtools}

\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
\DeclarePairedDelimiter\norm{\lVert}{\rVert}%

% Swap the definition of \abs* and \norm*, so that \abs
% and \norm resizes the size of the brackets, and the 
% starred version does not.
\makeatletter
\let\oldabs\abs
\def\abs{\@ifstar{\oldabs}{\oldabs*}}
%
\let\oldnorm\norm
\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
\makeatother

\usepackage{acro}
\acsetup{first-style=short}
\DeclareAcronym{MS1}{
  short = MS1 ,
  long  = first tandem MS analyzer ,
  class = abbrev
}

\DeclareAcronym{MS2}{
  short = MS2 ,
  long  = second tandem MS analyzer ,
  class = abbrev
}

%%  Begin document
%%  ==================================================================
\begin{document}

%%  Begin title page
%%  ==================================================================
    \thispagestyle{empty}
    \ULCornerWallPaper{1}{nat-farve.pdf}
    \ULCornerWallPaper{1}{nat-en.pdf}
    \begin{adjustwidth}{-3cm}{-1.5cm}
    \vspace*{-1cm}
    \textbf{\Huge Bioinformatics Master Thesis} \\
    \vspace*{2.5cm} \\
    \textbf{\Huge Topics in Mass Sectrometry and Proteomics} \\
    \vspace*{.1cm} \\
    %{\huge No subtitle is available} \\
    \begin{tabbing}
    % adjust the hspace below for the longest author name
    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
    %Second Author \> \texttt{<second@author.com>} \\       
    \\[10.5cm]

    \textbf{\Large Supervisors} \\
    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\

    \end{tabbing}
    \end{adjustwidth}
    \newpage
    \ClearWallPaper
%%  ==================================================================
%%  End title page

%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
\tableofcontents % Indholdsfortegnelse
%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\thispagestyle{empty}
\printacronyms[include-classes=abbrev,name=Abbreviations]

\newpage
\setcounter{page}{1}

\section*{Preface}

\chapter{Introduction}

\section{Aminoacids and proteins}

Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.

Proteins are made up of 20 basic units, called aminoacids. All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid (see figure \ref{fig:aminoacids}. The side chain differs between aminaocids and generates them from each other. A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid. Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminaocids}).

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{aminoacids.png}
  \caption{}
  \label{fig:aminaocids}
\end{figure}

Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them. Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another. CITE LEHNINGER PAG 82. The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.

While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures CITE. One frequent instance of such modifications is the oxidation of methionine

\section{The protein-focused biotechnology industry}

Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks. The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans. This potential economic activity is carried out by several biotech companies, including Novozymes.
 Novozymes A/S is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes. The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact. Notorious examples of such applications include waste-water treatment, household care and the baking industry.

In order to place Novozymes ahead of its competitors, the company has utmost interest in developing protein research and several departments in the organization approach the study of proteins and their translation to the market through among other things the application of mass spectrometry.

\section{Mass spectrometry and shotgun proteomics}

Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone. Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up. The layer provided by proteins is one of them, and its study is called proteomics. It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment. The study of the protein-specific complexity is called proteomics. With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.

It may be useful to divide the existing approaches into two types of paradigms: top-down and bottom-up. In the top-down paradigm, intact proteins are directly used for the analysis. In the bottom-up paradigm, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification. These smaller parts are called peptides. CITE 1.6.1 COMPUTATIONAL METHODS. Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly mass spectrometry (MS), which performs the data acquisition. The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry. The interested reader is directed to CITE 1.6.1 COMPUTATIONAL METHODS to learn why. The top-down paradigm will be ignored in the rest of the manuscript. A summary of the bottom-up approach MS analytical pipeline follows.

MS is performed by means of a mass spectrometer, a piece of equipment that can acquire plentiful mass measurements for enough sample components, generating a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample. These inference steps can be condensed to the peptide and protein inference problems. A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment. This is the quantification problem CITE.

\section{The mass spectrometer}

The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector SEE FIGURE.

\subsection{The ion source}

All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components. Thus, ionization of these is required prior to any measurement. This is achieved in the ionization source CITE 5.1 COMPUTATIONAL METHODS. The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (MALDI) and Electro Spray Ionization (ESI) CITE. As a result of ionization, the components acquire positive charge, usually by the addition of 1, 2, or 3 protons, that correspondingly give +1, +2 or +3 charge. The acquired charge yields a mass/charge (m/z) ratio, a property that can be applied in the following component separation and mass measurement of the separated components steps in the analyzer.

\subsection{The mass analyzer}

The plethora of separation methods is reflected upon the range of different analyzers available, mainly quadrupole (Q), time of flight (TOF), Ion trap (IT) and Fourier Transform (FT). These apply different principles to perform the same task: handling of the individual molecular components of the ionized sample for posterior sequential measurements of the components one by one.

\subsection{The detector}

the intensity of a given m/z ratio within an observable window. This analysis returns the MS spectrum. It is up to the MS technician to decide the best protocol, according to the particularities of the equipment, and the dataset.

\subsection{Sample preprocessing}

An MS experiment starts with the generation of protein mixtures samples. They are first separated in order to sort the proteins via physicochemical criteria. This is most frequently carried out via SDS-PAGE based on mass or isoelectric point.  Once the electrophoresis is completed, protein bands can be excised from the gel. Each band will contain a subset, or even only one of the proteins originally available, thus making the downstream analysis simpler REFERENCE COMPUTATIONAL METHODS CHAPTER 2

The reduced complexity protein mix is extracted in a denaturalized state and subjected to enzymatic digestion with specific enzymes, that can cut the chain of aminoacids following a predictable pattern, key for the downstream analysis. The enzyme most frequently sued for this is Trypsin, which cuts peptidic bonds whenever a positively charged residue, either Lysine or Arginine, lies on the carboxyl side of the peptidic bond. Even though enzymes are very specific, they can miss some of their targets, due to steric inaccesability or the presence of specific aminoacids that can weaken their function. This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline. This variability, though limited, needs to be taken care of in downstream analysis.

The result of this process is a mix of peptides following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition. As explained above, this length distribution is fitted to the resolution of the MS analyzer, thus optimizing the throughput of the method. The downstream workflow now diverges based on the simplicity of the original protein sample. When it consisted of a single protein, Peptide Mass Fingerprinting (PMF) is used, otherwise tandem MS (MS/MS) shall be performed.

\subsection{Peptide Mass Fingerprint (PMF)}

If the original sample was known to contain a single protein, PMF, or \textit{protein-centric}proteomics, is conducted. In PMF, the mixture of peptides can be already transfered to the spectrometer, where a a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded. Thus, spectra generated this way can be considered a pattern, or fingerprint, of the peptides making up the original protein. Therefore, the spectra are intended to contain enough information to identify a protein.

\subsection{Tandem MS (MS/MS)}

When presented with the problem of analyzing a mixture of proteins, the capacities of mass spectrometers are easily overcome by a too complex mixture, resulting in the analysis of only a minor part of the total protein complement of the sample. This can be overcome by dividing the initial sample into fractions, and using a series, or tandem, of spectrometers in the analysis. The spectrometers are used to analyze each obtained fraction separately, using different schemes. Fractionation is usually achieved by different methods of separation CITE 1.7, most commonly via HPLC methods.

HPLC methods work by loading the peptide mix in a column containing a stationary and a solid phase. These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases. The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times. Therefore, the input to the machine will consist of a few peptides at a time.
The two most common methods in proteomics are reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on charge)  CITE 4.2 computational methods.

The tandem MS analysis starts when a peptide accesses the analyzer. It is ionized in the ion source and accesses the first mass spectrometer. While different ways of handling the peptides are available, we will focus on the product ion scan. In this protocol, the first analyzer is used to select ionized peptides within a narrow m/z window. The peptides passing this first scan are then subjected to fragmentation, most often via collision-induced dissociation (CID). Briefly, the peptides enter a collision cell containing an inert gas. Given enough kinetic energy, hits of ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units. PAG 123 COMPUTATIONAL METHODS. The most frequently occurring fragment types by far for CID are the b and y ions and partly a ions. pag 134 computational methods. The produced fragmetns then enter the second analyzer, where a m/z spectrum of the fragments is recorded. Thus, unlike in PMF, where the spectrum recorded reflects the m/z ratios acquired by the protein peptides cleaved by the enzyme, tandem MS spectra on product ion scan mode record the m/z ratio of the fragments produced by an ionized peptide with a given m/z ratio. The m/z ratio of this precursor ion is changed during the run, thus, multiple spectra are obtained where PMF would create only one.

\section{Spectra processing: search engines}

A search engine can be used to map the recorded pattern of m/z ratios to a protein entry in a database. This task is performed by \textit{in silico} cleavage of each sequence entry in the database based on the specefic cleavage pattern of the enzyme used, coupled with the simulation of the expected spectrum based on the expected peptides.

Given the stochastic nature of the cleavage and spectra recording process, the resulting spectra exhibit variability manifested in missing peaks or spurious ones. Furthermore, the peptide to spectrum matching (PSM) process against a sufficiently big database can, at random, return wrong matches. This translates to the obtention of multiple matches , of which one, if any, will be correct. Therefore, the lists of matches need to be somehow ranked. The issue is addressed by search engines through the usage of scoring systems that measure the goodness of the match. Assuming the correct protein is present in the database, a good system should give the protein the best score. This way, proteins can be identified.

Two steps in protein identification can be distinguished:

\begin{enumerate}

\item Peptide inference: infer the peptides present in the sample.
\item Protein inferece: based on the infered peptides, infer what proteins generated them.
\end{enumerate}

Both are taken care of by the search engine. Multiple search engines exist that implement different matching and scoring algorithms.

\subsection{X!Tandem}

X!Tandem was one of the first open source search engines in mass spectrometry. It produces a score based on the dot product between the theoretical and the experimental tandem mass spectra CITE Overview of Tandem Mass Spectrometry (MS/MS) Database Search Algorithms. The scores assigned to wrong matches are assumed to follow a hypergeometric distribution, allowing the program to extrapolate $E(s)$ (expected number of wrong matches at a given score) for any score.

\section{Results validation}

The scoring system implemented by the search engines can be used to filter and validate the results. A basic common filter is the false discovery rate (FDR), usually set to 1\%, indicating that 1 out of a hundred filtered matches are expected to be false positives. The most commonly used method to compute the FDR is the target-decoy search. The search engine tries to match the same spectra against a decoy database, usually generated by reversing the sequences present in the original database (target). All matches to the decoy will be regarded as wrong. Thanks to the fact that the basic properties of the decoy (size, composition, etc) are identical to the target, whenever a mistake is made, it is as likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5, thus the number of matches in the decoy provides an accurate estimate of the number of random matches, or false positives, against the target database ($n_{fp}$). Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.

$$FDR = \frac{n_{fp}}{n_{tp} + n_{tp}}$$

The minimal FDR at which a given PSM is considered a positive match constitutes the PSM\textquotesingle s q-value

$$ q(PSM_i) = \min FDR \forall PSM_i \in Positives$$

Finally, a posterior error probability (PEP) can be defined as the probability of the match being random ($P(s_i | T_i = 0)$). This can only be done if a statistical model describing the distribution of scores for correct and wrong matches is fit to the dataset.

\section{Protein quantification}

CONTINUE maybe add reason for the project.

\chapter{Materials and Methods}

\chapter{Results and Discussion}

\chapter{Conclusion}

%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.

\section{Overview of proteomics analysis software}

\subsection{MaxQuant}

One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}. It can be downloaded for free and is developed by the Max Planck institute in Germany. It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.

MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline. A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.

The main processing steps incorporated in MaxQuant consist of:

\begin{itemize}

\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.

\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.

\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.

\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.

\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.

\item Supports label-based and label-free quantification.

\item Its output can be passed to the Perseus software for data visualization. %CITE

\end{itemize}

However, MaxQuant also suffers from some problems:

\begin{enumerate}

\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}. The processing steps cannot easily be fine tuned or exchanged with other pipelines. For example, only the Andromeda search engine is supported. This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.

\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.

\item While it is supported in Linux through Mono, the user experience is best in Windows. As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.

\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.

\end{enumerate}

\subsection{Compomics group}

The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.

\begin{itemize}

\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.

\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.

\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide. Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.

\end{itemize}

SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}. Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}. They easily enable the development of analytical pipelines finely fitted to each use case.

moFF is less documented and integrated in the workflow. Furthermore, it does not offer a GUI yet. The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.

The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool. Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.

\subsection{Other suites}

Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available. They are not open, requiring a license to function. Furthermore, they are GUI oriented and make use of different data formats thus, making exchange across pipelines extremely difficult. The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.

\section{Label-free quantification}

\subsection{MaxLFQ}

The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}. It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.

\subsubsection{Fractionated XIC aggreagation}

First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions. MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity 
 between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation).

The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
%  this commented blank line prevents start of a new paragraph
\begin{equation}\label{eq:peptide_intensity}
I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
\end{equation}

The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
%  this commented blank line prevents start of a new paragraph
\begin{equation}\label{eq:sum_norm_intensity}
H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
\end{equation}
%  this commented blank line prevents start of a new paragraph
\begin{equation}
H(N) = \sum\limits_{p} H_p(N) 
\end{equation}

where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.

The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{xic_normalization}
\caption{Taken from \cite{Cox2014}}
\label{figure:xic_normalization}
\end{figure}

\subsubsection{Protein intensity inference}

Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred. This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).

\begin{enumerate}
\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.

\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b. The median is selected as a summarising statistic to protect from outliers. A minimal number of non-zero intensity peptides are required for the median to be valid, usually 2. Otherwise, the protein ratio for the pair of samples is set to 0.

\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.

\end{enumerate}

\begin{equation}\label{eq:least_squares}
\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
\end{equation}

\begin{figure}[!h]
\centering
\includegraphics[width=0.8\textwidth]{protein_quant_inference}
\caption{Taken from \cite{Cox2014}}
\label{figure:protein_quant_inference}
\end{figure}

For the three most intense tryptic peptides, the signal per mole of protein was shown to be constant within a coefficient of variation of $\pm 10 \%$ \cite{Mikalsen2007}. Thus, protein intensities are linearly proportional to protein quantities and protein quantity ratios can be computed from protein intensity ratios. These values are reported in the MaxQuant output in the \texttt{proteinGroups.txt} file under columns named \textit{LFQ Intensity}.

\subsubsection{Drawbacks of the model}

The LFQ model provides a powerful method to infer protein quantities from a label-free experiment. However, it makes use of point estimates in different steps of the process, namely (1) during peptide intensity aggregation, where the Levenberg-Marquardt minimum is selected (2) during protein ratio estimation, where the peptide median is selected and (3) during protein intensity estimation, where the least squares minimum is selected. These point estimation bottlenecks, specially (2), discard increasing amounts of data that could otherwise be used for more accurate results. Moreover, not just quantities, but also the uncertainty behind them, could also be provided. The only way of MaxQuant to provide uncertainty measurements is the number of peptides supporting the quantification measurement.

\subsection{Bayesian model}

Bayesian statistics provides probability measurements for observed data assuming underlying mathematical models. A model for protein quantification based on the \ac{MS1} intensities or XICs could be developed to assess both quantities and uncertainties behind the estimated quantities.

While \ac{MS1} intensity measurements are dependent upon the underlying protein quantities, other factors influence the final measurement, thus adding noise and distorting the results. Besides the actual quantity, two more factors can be distinguished:

\begin{itemize}

\item \textbf{Sequence derived factors}: some peptides are easier to cleave for the cleaving enzyme (Trypsin, etc) than others. Moreover, the sequence of the peptide could influence the final measurement detected in the analyzer, by having different elution dynamics in the column or the analyzer.

\item \textbf{Random noise}: the stochastic processes intrinsic to \ac{MS1} measurements could also have an impact.

\end{itemize}

The model would take as input:

\begin{enumerate}

\item \textbf{Features extracted from the precursor sequence}, including the surrounding aminoacids in the original protein. They can be used to model the sequenced derived factors.

\item \textbf{Observed \ac{MS1} precursor intensities/XICs}.

\end{enumerate}

and return as output a protein quantification value for each protein, together with a measurement of the uncertainty behind it.

\subsection{Data}

The MaxLFQ paper dataset is available \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}. The authors submitted the RAW files (>50 GB) and the search files containing the final results. They can be used to benchmark any new quantification tool or method, for example, it has been used to benchmark third party quantification tools like StPeter from the TPP \cite{Hoopmann2018}.

The dataset contains the results of the quantification of proteins in 2 different proteomes (\textit{E. coli} and \textit{Homo sapiens}) from 6 samples organized in 2 conditions with 3 replicates each. In the first condition (H), both proteomes were mixed in a 1:1 ratio, while in the second condition (L), the \textit{E. coli} proteome had 3 times more contribution to the mix (3:1). Thus, the fold change of protein quantities across conditions should be 1 for human proteins and 3 for bacterial proteins.

A list of the protein groups identified that could be mapped unambiguously to each species was made available \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}}. Thus, a dataset of thousands of proteins of known ratios between conditions, producing tens of thousands of peptides with known sequence, \ac{MS1} intensity/XIC and parent protein can be built from the publication, and used to train new quantification methods (figure \ref{fig:screenshot_peptide_data}).

% latex table generated in R 3.4.3 by xtable 1.8-2 package
% Fri Jun 01 17:11:23 2018
\begin{table}[ht]
\centering
\begin{tabular}{rlrr}
  \hline
 & Taxonomy & Peptides & Proteins \\ 
  \hline
1 & Escherichia coli (strain K12) & 14483 & 1556 \\ 
  2 & Homo sapiens & 32647 & 3444 \\ 
   \hline
   & Total & 47130 & 5000
\end{tabular}
\caption{Summary table of the dataset}
\label{table:summary_peptide_data}
\end{table}

\begin{figure}[!h]
\includegraphics[width=1\textwidth]{dataset}
\caption{Extract of the available dataset. Total of 47130 entries, one per peptide, for a global of 5000 proteins.}
\label{fig:screenshot_peptide_data}
\end{figure}

This compiled dataset can be downloaded here \href{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}

\section{Extension of pipelines}

The Compomics suite of programmes provides the best documented and accesible set of tools for proteomics analysis, but still lacks proper quantification tools. The implementation of a MS1 intensity-based downstream quantification tool compatible with the output of the Compomics tools would yield a free Linux compatible, CLI supported, complete proteomics pipeline featuring a robust quantification method.

\bibliographystyle{unsrt_etal}
\footnotesize{\bibliography{bibliography}}

\end{document}
%%  ==================================================================
%%  End document

### 2018-06-26 report.tex
Extended chapter 1

@@
 \usepackage{changepage}
+\usepackage{mhchem} % for isotope nomeclature support
 % Setup captions
@@
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
+\usepackage{cleveref} % for enumerate/itemize item referencing
+
 %\setlength{\parindent}{4em}
@@
 \usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{hyperref}
 \usepackage{xspace}
@@
 \acsetup{first-style=short}
+
+\DeclareAcronym{MS}{
+  short = MS ,
+  long  = Mass Spectrometry,
+  class = abbrev
+}
+
 \DeclareAcronym{MS1}{
@@
   long  = second tandem MS analyzer ,
+  class = abbrev
+}
+
+\DeclareAcronym{MS/MS}{
+  short = MS/MS ,
+  long  = Tandem mass spectrometry,
+  class = abbrev
+}
+
+\DeclareAcronym{m/z}{
+  short = m/z ,
+  long  = mass charge ratio,
+  class = abbrev
+}
+
+\DeclareAcronym{MALDI}{
+  short = MALDI ,
+  long  = Matrix-Assisted Laser-Desorption Ionization,
+  class = abbrev
+}
+
+\DeclareAcronym{ESI}{
+  short = ESI ,
+  long  = Electrospray Ionization,
+  class = abbrev
+}
+
+\DeclareAcronym{TOF}{
+  short = TOF ,
+  long  = Time of Flight,
+  class = abbrev
+}
+
+\DeclareAcronym{IT}{
+  short = IT ,
+  long  = Ion Trap,
+  class = abbrev
+}
+
+\DeclareAcronym{Q}{
+  short = Q,
+  long  = Quadrupole,
+  class = abbrev
+}
+
+\DeclareAcronym{FT-ICR}{
+  short = FT-ICR ,
+  long  = Fourier Transform Ion Cyclotron Resonance,
+  class = abbrev
+}
+
+\DeclareAcronym{HPLC}{
+  short = HPLC ,
+  long  = High pressure liquid chromatography,
   class = abbrev
@@
 %%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
+\setcounter{tocdepth}{1}
 \tableofcontents % Indholdsfortegnelse
@@
 \chapter{Introduction}
+\label{chap:introduction}
 
@@
 This potential economic activity is carried out by several biotech companies, including Novozymes.
+
 Novozymes A/S is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
@@
 
-In order to place Novozymes ahead of its competitors, the company has utmost interest in developing protein research and several departments in the organization approach the study of proteins and their translation to the market through among other things the application of mass spectrometry.
+In order to place Novozymes ahead of its competitors, the company has utmost interest in developing protein research and several departments in the organization approach the study of proteins and their translation to the market through among other things the application of Mass Spectrometry (\ac{MS}) and Proteomics.
 
-\section{Mass spectrometry and shotgun proteomics}
+\section{Objectives of the Thesis}
+\label{sec:objectives}
+
+In line with the goal of making Novozymes more competitive, this project aimed at the following objectives:
+
+\begin{enumerate}
+
+\item Develop an open-source, Linux based and straightforward to deploy pipeline for the analysis of mass spectrometry data, starting at the output RAW files and ending in the interpretation of the results.
+
+\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
+
+\item Implement a label-free quantification Bayesian-based method using Pymc3.
+
+\end{enumerate}
+
+\chapter{Mass spectrometry and shotgun proteomics}
+\label{chap:mass_spec}
 
@@
 The top-down paradigm will be ignored in the rest of the manuscript.
+
+MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
+A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
+The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
+These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
+A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
+This is the quantification problem, explained in section \label{sec:quantification} CITE.
+
 A summary of the bottom-up approach MS analytical pipeline follows.
 
-MS is performed by means of a mass spectrometer, a piece of equipment that can acquire plentiful mass measurements for enough sample components, generating a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem CITE.
+\section{Sample processing}
+\label{sec:sample_processing}
 
-\section{The mass spectrometer}
-
-The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector SEE FIGURE.
-
-\subsection{The ion source}
-
-All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Thus, ionization of these is required prior to any measurement.
-This is achieved in the ionization source CITE 5.1 COMPUTATIONAL METHODS.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (MALDI) and Electro Spray Ionization (ESI) CITE.
-As a result of ionization, the components acquire positive charge, usually by the addition of 1, 2, or 3 protons, that correspondingly give +1, +2 or +3 charge.
-The acquired charge yields a mass/charge (m/z) ratio, a property that can be applied in the following component separation and mass measurement of the separated components steps in the analyzer.
-
-\subsection{The mass analyzer}
-
-The plethora of separation methods is reflected upon the range of different analyzers available, mainly quadrupole (Q), time of flight (TOF), Ion trap (IT) and Fourier Transform (FT).
-These apply different principles to perform the same task: handling of the individual molecular components of the ionized sample for posterior sequential measurements of the components one by one.
-
-\subsection{The detector}
-
-the intensity of a given m/z ratio within an observable window.
-This analysis returns the MS spectrum.
-It is up to the MS technician to decide the best protocol, according to the particularities of the equipment, and the dataset.
-
-\subsection{Sample preprocessing}
-
-An MS experiment starts with the generation of protein mixtures samples.
+An \ac{MS} experiment starts with the generation of protein mixture samples.
 They are first separated in order to sort the proteins via physicochemical criteria.
@@
 
-The reduced complexity protein mix is extracted in a denaturalized state and subjected to enzymatic digestion with specific enzymes, that can cut the chain of aminoacids following a predictable pattern, key for the downstream analysis.
-The enzyme most frequently sued for this is Trypsin, which cuts peptidic bonds whenever a positively charged residue, either Lysine or Arginine, lies on the carboxyl side of the peptidic bond.
-Even though enzymes are very specific, they can miss some of their targets, due to steric inaccesability or the presence of specific aminoacids that can weaken their function.
+The reduced complexity protein mix is extracted in a denaturalized state and subjected to enzymatic digestion with specific enzymes, that can cut the chain of aminoacids following a predictable pattern.
+The enzyme most frequently used for this is Trypsin, which cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
+Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: COMPUTATIONAL METHODS 3.2
+
+\begin{enumerate}
+
+\item Missed cleavages \label{item:missed_cleavages}
+
+\item Unsuspected cleavages during the maturation/life cycle of the protein.
+
+\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
+
+\item Naturally occurring, intentionally or unintentionally induced chemical modifications.
+
+\end{enumerate}
+
+\Cref{item:missed_cleavages} can happen due to steric inaccesability or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
 This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-This variability, though limited, needs to be taken care of in downstream analysis.
+This variability, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
+The other
 
 The result of this process is a mix of peptides following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-As explained above, this length distribution is fitted to the resolution of the MS analyzer, thus optimizing the throughput of the method.
-The downstream workflow now diverges based on the simplicity of the original protein sample.
+For Trypsin, the average peptide length is 10 residues, as roughly 1/10 residues are either R or K.
+As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer, thus optimizing the throughput of the method.
+An overview over mass spectrometers follows.
+
+\section{The mass spectrometer}
+\label{sec:thee_mass_spectrometer}
+
+The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector SEE FIGURE.
+
+\subsection{The ion source}
+\label{subsec:the_ion_source}
+
+All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
+Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
+This step is performed in the ion source CITE 5.1 COMPUTATIONAL METHODS.
+The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (MALDI) and Electro Spray Ionization (ESI) CITE.
+Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
+Thus, the charge exhibited by an ion is not obvious when produced via ESI.
+Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
+Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
+Bias in ESI is less predictable.
+This is known as the competitive ionization problem.
+REF ALL THIS
+
+The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that can be applied in the downstream component separation and measurement steps.
+
+\subsection{The mass analyzer}
+\label{subsec:the_mass_analyzer}
+
+The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
+These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.
+
+Moreover, two other analyzers exist which combine mass analysis with intensity measurement.
+These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
+They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
+Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability.
+
+\subsection{The detector}
+\label{subsec:the_detector}
+
+Detectors measure the intensity of an incoming ion signal.
+The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
+Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of an ion signal over an \ac{m/z} range.
+Some topics in signal detection in \ac{MS} need to be discussed.
+
+On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
+The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
+This is one of the reasons why proteins are better fit for analysis when digested into peptides, as \ac{m/z} are reduced, thus increasing the mass resolution.
+
+On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
+They constitute the isotopic envelope of the ion SEE FIGURE, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
+Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
+Even though the natural abundance is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms are \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
+Such event can pe problematic if it entails that the 1 \ce{^{13}_{}C} peak is confused for the completely \ce{^{12}_{}C} peak.
+
+The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
+Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on.
+
+It is up to the \ac{MS} technician to decide on the best pieces of equipment according to the particularities of the dataset.
+
+\section{Mass spectrometry workflows}
+\label{sec:mass_spectrometry_workflows}
+
+The \ac{MS} workflow now diverges based on the simplicity of the original protein sample.
 When it consisted of a single protein, Peptide Mass Fingerprinting (PMF) is used, otherwise tandem MS (MS/MS) shall be performed.
@@
 \subsection{Peptide Mass Fingerprint (PMF)}
+\label{subsec:pmf}
 
-If the original sample was known to contain a single protein, PMF, or \textit{protein-centric}proteomics, is conducted.
-In PMF, the mixture of peptides can be already transfered to the spectrometer, where a a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded.
+If the original sample was known to contain a single protein, PMF, or \textit{protein-centric} proteomics, is conducted.
+In PMF, the mixture of peptides can be already transferred to the spectrometer, where a a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded.
 Thus, spectra generated this way can be considered a pattern, or fingerprint, of the peptides making up the original protein.
-Therefore, the spectra are intended to contain enough information to identify a protein.
+Therefore, the spectra are yield information that can be used to identify the original protein.
 
-\subsection{Tandem MS (MS/MS)}
+\subsection{High pressure liquid chromatography-Tandem MS}
+\label{subsec:ms2}
 
-When presented with the problem of analyzing a mixture of proteins, the capacities of mass spectrometers are easily overcome by a too complex mixture, resulting in the analysis of only a minor part of the total protein complement of the sample.
-This can be overcome by dividing the initial sample into fractions, and using a series, or tandem, of spectrometers in the analysis.
+\subsubsection{HPLC}
+
+If presented with the problem of analyzing a mixture of proteins, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein  of the sample.
+This can be surmounted by splitting the initial sample into fractions, and using a series, or tandem, of spectrometers in the analysis.
 The spectrometers are used to analyze each obtained fraction separately, using different schemes.
-Fractionation is usually achieved by different methods of separation CITE 1.7, most commonly via HPLC methods.
+Fractionation is usually achieved by different methods of separation CITE 1.7, most commonly via High pressure liquid chromatography (\ac{HPLC}) methods.
 
-HPLC methods work by loading the peptide mix in a column containing a stationary and a solid phase.
+\ac{HPLC} methods work by loading the peptide mix in a column containing a stationary and a solid phase.
 These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
 The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times.
-Therefore, the input to the machine will consist of a few peptides at a time.
-The two most common methods in proteomics are reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on charge)  CITE 4.2 computational methods.
+Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
+Notably, the same peptide could elute in several contiguous fractions.
 
-The tandem MS analysis starts when a peptide accesses the analyzer.
-It is ionized in the ion source and accesses the first mass spectrometer.
-While different ways of handling the peptides are available, we will focus on the product ion scan.
+The two most common fractionation methods in proteomics are reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point)  CITE 4.2 computational methods.
+
+The tandem MS \ac{MS/MS}) analysis starts when the peptide mix accesses the analyzer.
+It is ionized in the ion source and enters the first mass spectrometer.
+While different ways of handling the peptides are available, we will focus on the product ion scan method.
 In this protocol, the first analyzer is used to select ionized peptides within a narrow m/z window.
-The peptides passing this first scan are then subjected to fragmentation, most often via collision-induced dissociation (CID).
-Briefly, the peptides enter a collision cell containing an inert gas.
+
+\subsubsection{Fragmentation}
+
+Peptides passing this first scan are then subjected to fragmentation, most often via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
+This way, peptides are further processed into smaller fragments resulting from the breakage of either the peptidic bond, the \ce{C_$\alpha$}-\ce{CO} or the \ce{C_$\alpha$}-\ce{NH} bonds
+
+In CID, peptides enter a collision cell containing an inert gas.
 Given enough kinetic energy, hits of ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
 PAG 123 COMPUTATIONAL METHODS.
-The most frequently occurring fragment types by far for CID are the b and y ions and partly a ions.
+The most frequently occurring fragment types are the b and y ions.
 pag 134 computational methods.
@@
 \section{Spectra processing: search engines}
+\label{sec:search_engines}
 
@@
 \section{Results validation}
+\label{sec:validation}
 

### 2018-06-26 report.tex
mend

@@
 Peptides passing this first scan are then subjected to fragmentation, most often via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-This way, peptides are further processed into smaller fragments resulting from the breakage of either the peptidic bond, the \ce{C_$\alpha$}-\ce{CO} or the \ce{C_$\alpha$}-\ce{NH} bonds
+This way, peptides are further processed into smaller fragments resulting from the breakage of either the peptidic \ce{CO}-\ce{NH}, the \ce{C_$\alpha$}-\ce{CO} or the \ce{C_$\alpha$}-\ce{NH} bonds.
+A clear nomenclature REF was introduced to name these fragments: INSERT FIGURE
 
-In CID, peptides enter a collision cell containing an inert gas.
-Given enough kinetic energy, hits of ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
+CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
+Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
 PAG 123 COMPUTATIONAL METHODS.
-The most frequently occurring fragment types are the b and y ions.
-pag 134 computational methods.
+Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
+This results in the production of mostly b and y ions, as well as the loss of any chemical modifications CITE pag 134 computational methods.
+
+EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
+Thus, the fragmentation process is not ergodic and returns TYPE ions.
+More importantly, chemical modifications are not lost in the process.
+
 The produced fragmetns then enter the second analyzer, where a m/z spectrum of the fragments is recorded.

### 2018-06-28 report.tex
Backup

@@
 
+\DeclareAcronym{NZ}{
+  short = NZ ,
+  long  = Novozymes A/S,
+  class = abbrev
+}
+
 %%  Begin document
@@
 Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid (see figure \ref{fig:aminoacids}.
+All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
 The side chain differs between aminaocids and generates them from each other.
 A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminaocids}).
+Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
 
@@
   \caption{}
-  \label{fig:aminaocids}
+  \label{fig:aminoacids}
 \end{figure}
@@
 The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes.
+This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
 
-Novozymes A/S is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
+\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
 The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
@@
 
-In order to place Novozymes ahead of its competitors, the company has utmost interest in developing protein research and several departments in the organization approach the study of proteins and their translation to the market through among other things the application of Mass Spectrometry (\ac{MS}) and Proteomics.
+The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
+The refinement of the currently used tools and the development of new ones could be of great significance for the company.
+
+Protein research can be approached from different angles.
+This thesis exploited the combination of mass spectrometry \ac{MS} and proteomics workflows (see \ref{chap:mass_spec}) to the qualitative and quantitative characterization of protein samples.
 
@@
 
-In line with the goal of making Novozymes more competitive, this project aimed at the following objectives:
+In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
 
@@
 
-\item Develop an open-source, Linux based and straightforward to deploy pipeline for the analysis of mass spectrometry data, starting at the output RAW files and ending in the interpretation of the results.
+\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
 
@@
 
-\item Implement a label-free quantification Bayesian-based method using Pymc3.
+\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
 
 \end{enumerate}
+
+An overview over the \ac{MS} and downstream computational analysis of its data is presented in \ref{chap:mass_spec}.
+The pipeline development and its benchmark are explained in chapters 3 and 4, while the modelling problem is introduced in chapter 5.
+Finally, a conclusion of the work is given in chapter 6.
 
@@
 
-The reduced complexity protein mix is extracted in a denaturalized state and subjected to enzymatic digestion with specific enzymes, that can cut the chain of aminoacids following a predictable pattern.
-The enzyme most frequently used for this is Trypsin, which cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
+The reduced complexity protein mix is extracted in a denaturalized state so as to remove biases due to the divergent properties acquired by folded proteins.
+Then, proteins are subjected to digestion with specific enzymes, that can cut the aminoacidic chains following a predictable pattern.
+Trypsin is the most frequently used for this task, which cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
 Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: COMPUTATIONAL METHODS 3.2
@@
 The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of an ion signal over an \ac{m/z} range.
+Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
 Some topics in signal detection in \ac{MS} need to be discussed.
@@
 Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
-Even though the natural abundance is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms are \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can pe problematic if it entails that the 1 \ce{^{13}_{}C} peak is confused for the completely \ce{^{12}_{}C} peak.
+Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
+Such event can be problematic if it entails that the 1 \ce{^{13}_{}C} peak is confused for the completely \ce{^{12}_{}C} peak.
 
@@
 
-It is up to the \ac{MS} technician to decide on the best pieces of equipment according to the particularities of the dataset.
+It is up to the \ac{MS} technician to decide on the best pieces of equipment according to their availability and particularities of the dataset.
 
@@
 
-The \ac{MS} workflow now diverges based on the simplicity of the original protein sample.
+The \ac{MS} workflow diverges based on the simplicity of the original protein sample.
 When it consisted of a single protein, Peptide Mass Fingerprinting (PMF) is used, otherwise tandem MS (MS/MS) shall be performed.
@@
 If the original sample was known to contain a single protein, PMF, or \textit{protein-centric} proteomics, is conducted.
-In PMF, the mixture of peptides can be already transferred to the spectrometer, where a a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded.
-Thus, spectra generated this way can be considered a pattern, or fingerprint, of the peptides making up the original protein.
-Therefore, the spectra are yield information that can be used to identify the original protein.
+In PMF, the mixture of peptides can be already transferred to the spectrometer, where a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded.
+The resulting spectra can be considered a pattern, or fingerprint, of the peptides making up the original protein, thus enabling the identification of the original protein.
 
@@
 
-\subsubsection{HPLC}
+\subsubsection{High pressure liquid chromatography (HPLC)}
 
@@
 The spectrometers are used to analyze each obtained fraction separately, using different schemes.
-Fractionation is usually achieved by different methods of separation CITE 1.7, most commonly via High pressure liquid chromatography (\ac{HPLC}) methods.
+
+Fractionation is achieved by different methods of separation CITE 1.7, most commonly via \ac{HPLC} methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point)  CITE 4.2 computational methods.
 
@@
 
-The two most common fractionation methods in proteomics are reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point)  CITE 4.2 computational methods.
+\subsubsection{First MS analysis}
 
-The tandem MS \ac{MS/MS}) analysis starts when the peptide mix accesses the analyzer.
+The tandem MS \ac{MS/MS}) analysis starts at the first \ac{MS} scan (\ac{MS1}), when the peptide mix accesses the analyzer.
 It is ionized in the ion source and enters the first mass spectrometer.
@@
 \subsubsection{Fragmentation}
+\label{subsubsec:fragmentation}
 
-Peptides passing this first scan are then subjected to fragmentation, most often via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-This way, peptides are further processed into smaller fragments resulting from the breakage of either the peptidic \ce{CO}-\ce{NH}, the \ce{C_$\alpha$}-\ce{CO} or the \ce{C_$\alpha$}-\ce{NH} bonds.
-A clear nomenclature REF was introduced to name these fragments: INSERT FIGURE
+As the protein sample complexity increases, the \ac{MS1} information on the peptide fractions will not be enough to correctly map each to the original protein, thus rendering a correct interpretation of the resulting spectra impossible.
+Therefore, more information needs to be extracted from them.
 
+Learning the peptide sequences would provide an additional feature that would make the peptide to protein mapping possible.
+This can be achieved by peptide fragmentation, which triggers the breakage of bonds along the peptidic chain, turning the peptides into smaller fragments.
+A given bond will be more likely to break the less stable it is.
+This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the more likely to break (see figure \ref{fig:abcxyz}).
+A nomenclature REF was introduced to name these fragments:
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{abcxyz}
+\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
+The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
+The a/x, b/y and c/z series are produced by the breakage of bonds I, II and III respectively.
+Specific fragmentation techniques make fragments belonging to one series more likely than others.
+Other fragments are possible but much less likely.
+Taken from CITE COMPOMICS TUTORIAL}
+\label{fig:abcxyz}
+\end{figure}
+
+Provided enough mass resolution, the mass difference between contiguous fragments of the same series i.e a1 and a2, can be related to the mass of the extra residue in the longer fragment.
+If this comparison process is repeated for enough fragment pairs of the same peptide, a sequence can be read, finally providing sufficient information to perform an accurate peptide to protein mapping.
+
+Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
 CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
@@
 Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y ions, as well as the loss of any chemical modifications CITE pag 134 computational methods.
+This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications CITE pag 134 computational methods.
 
@@
 
-The produced fragmetns then enter the second analyzer, where a m/z spectrum of the fragments is recorded.
+\subsubsection{Second MS analysis}
+
+The produced fragments enter the second analyzer \ac{MS2}, where a m/z spectrum of the fragments is recorded.
 Thus, unlike in PMF, where the spectrum recorded reflects the m/z ratios acquired by the protein peptides cleaved by the enzyme, tandem MS spectra on product ion scan mode record the m/z ratio of the fragments produced by an ionized peptide with a given m/z ratio.
 The m/z ratio of this precursor ion is changed during the run, thus, multiple spectra are obtained where PMF would create only one.
+The \ac{MS2} spectra will encode the sequence of each peptide as mentioned in \ref{subsubsec:fragmentation}, and with the right tools, it can be deconvoluted and interpreted as explained in \ref{sec:search_engines}.
 
@@
 
-A search engine can be used to map the recorded pattern of m/z ratios to a protein entry in a database.
-This task is performed by \textit{in silico} cleavage of each sequence entry in the database based on the specefic cleavage pattern of the enzyme used, coupled with the simulation of the expected spectrum based on the expected peptides.
+\ac{MS} search engines are capable of performing the peptide to spectrum matching (PSM).
+They do it by building statistical models that find the peptides within a protein database that best explain the observed spectra.
+\textit{De novo} search engines exist that don't require a reference proteome, but better results are obtained when using one..DEVELOP THAT.
+
+Matching is possible by predicting the cleavage pattern of each protein sequence in the database \textit{in silico} based on the cleavage pattern of the enzyme used, coupled with the simulation of the expected spectrum based on the predicted peptides.
 
 Given the stochastic nature of the cleavage and spectra recording process, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, the peptide to spectrum matching (PSM) process against a sufficiently big database can, at random, return wrong matches.
-This translates to the obtention of multiple matches , of which one, if any, will be correct.
+Furthermore, the PSM process against a sufficiently big database can, at random, return wrong matches.
+This translates to the generation of multiple matches, of which one, if any, will be correct.
 Therefore, the lists of matches need to be somehow ranked.
-The issue is addressed by search engines through the usage of scoring systems that measure the goodness of the match.
-Assuming the correct protein is present in the database, a good system should give the protein the best score.
-This way, proteins can be identified.
+The issue is addressed by search engines through the deployment of statistical models that provide scoring systems measuring the goodness of the match.
+Assuming the correct protein is present in the database, a good scoring system should give the matched peptide which comes from the right protein the best score.
+Under this circumstances, enough evidence for the presence of individual proteins can be collected.
 
@@
 \item Peptide inference: infer the peptides present in the sample.
-\item Protein inferece: based on the infered peptides, infer what proteins generated them.
+\item Protein inferece: based on the inferred peptides, infer what proteins generated them. This is not trivial as the peptide coming from the best PSM could map to more than one protein.
 \end{enumerate}
 
-Both are taken care of by the search engine. Multiple search engines exist that implement different matching and scoring algorithms.
+Both are taken care of by the search engine.
+Multiple search engines exist that implement different matching and scoring algorithms.
+The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
+Notably, the results of each individual search engine can be combined to gather their strenghts, at the expense of an increased computational cost and time REF THIS!!
 
-\subsection{X!Tandem}
-
-X!Tandem was one of the first open source search engines in mass spectrometry.
+%\subsection{X!Tandem}
+%
+%X!Tandem was one of the first open source search engines in mass spectrometry.
 It produces a score based on the dot product between the theoretical and the experimental tandem mass spectra CITE Overview of Tandem Mass Spectrometry (MS/MS) Database Search Algorithms.
@@
 
-The scoring system implemented by the search engines can be used to filter and validate the results.
-A basic common filter is the false discovery rate (FDR), usually set to 1\%, indicating that 1 out of a hundred filtered matches are expected to be false positives.
-The most commonly used method to compute the FDR is the target-decoy search.
-The search engine tries to match the same spectra against a decoy database, usually generated by reversing the sequences present in the original database (target).
-All matches to the decoy will be regarded as wrong.
-Thanks to the fact that the basic properties of the decoy (size, composition, etc) are identical to the target, whenever a mistake is made, it is as likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5, thus the number of matches in the decoy provides an accurate estimate of the number of random matches, or false positives, against the target database ($n_{fp}$).
+The scoring system implemented by search engines provides the best matches, but they are bound to contain false identifications.
+Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
+
+A common filter is the false discovery rate (FDR), usually set to 1\%, indicating tha the application of this filter one out of a hundred filtered matches are expected to be false positives.
+
+The most commonly used method to compute the FDR of a list of matches is the target-decoy search.
+Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
+This decoy database is generated by reversing or in general applying a randomization technique upon the sequences present in the original database (target).
+This guarantess that while the sequences in the decoy database won't overlap with those in the target, their basic properties of the decoy (size, composition, etc) will remain identical to the target.
+
+All matches to the decoy are by definition wrong. Since database properties 
+are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be used to estimate the number of false identifications in the  target results.
+it is as likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5, thus the number of matches in the decoy provides an accurate estimate of the number of random matches, or false positives, against the target database ($n_{fp}$).
 Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.

### 2018-06-29 report.tex
Chapter 1 finished

@@
 
+\DeclareAcronym{FDR}{
+  short = FDR ,
+  long  = False Discovery Rate,
+  class = abbrev
+}
+
+\DeclareAcronym{RT}{
+  short = RT ,
+  long  = Retention time,
+  class = abbrev
+}
+
+\DeclareAcronym{MBR}{
+  short = MBR ,
+  long  = Match Between Runs,
+  class = abbrev
+}
+
+\DeclareAcronym{XIC}{
+  short = XIC ,
+  long  = Extracted Ion Current,
+  class = abbrev
+}
+
+\DeclareAcronym{SC}{
+  short = SC ,
+  long  = Spectral Counting,
+  class = abbrev
+}
+
+\DeclareAcronym{TPP}{
+  short = TPP ,
+  long  = Trans Proteomic Pipeline,
+  class = abbrev
+}
+
 %%  Begin document
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 
@@
 Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry \ac{MS} and proteomics workflows (see \ref{chap:mass_spec}) to the qualitative and quantitative characterization of protein samples.
+This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
 
@@
 
-An overview over the \ac{MS} and downstream computational analysis of its data is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters 3 and 4, while the modelling problem is introduced in chapter 5.
-Finally, a conclusion of the work is given in chapter 6.
+\section{Structure of the Thesis}
 
-\chapter{Mass spectrometry and shotgun proteomics}
+An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
+The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
+Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
+
+\chapter{Mass spectrometry and shotgun proteomics overview}
 \label{chap:mass_spec}
@@
 
-A summary of the bottom-up approach MS analytical pipeline follows.
+A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:
+
+\begin{enumerate}
+
+\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:mass_spectrometry_workflows}.
+
+\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.
+
+\end{enumerate}
 
@@
 \section{The mass spectrometer}
-\label{sec:thee_mass_spectrometer}
+\label{sec:the_mass_spectrometer}
 
-The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector SEE FIGURE.
+The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{mass_spectrometer}
+\caption{Schematic view of a mass spectrometer. Taken from}
+%CITE THIS https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation
+\label{fig:mass_spectrometer}
+\end{figure}
 
@@
 Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can be problematic if it entails that the 1 \ce{^{13}_{}C} peak is confused for the completely \ce{^{12}_{}C} peak.
+Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.
 
@@
 These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times.
+The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
 Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
@@
 
-Given the stochastic nature of the cleavage and spectra recording process, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, the PSM process against a sufficiently big database can, at random, return wrong matches.
+Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
+Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
 This translates to the generation of multiple matches, of which one, if any, will be correct.
@@
 The issue is addressed by search engines through the deployment of statistical models that provide scoring systems measuring the goodness of the match.
-Assuming the correct protein is present in the database, a good scoring system should give the matched peptide which comes from the right protein the best score.
-Under this circumstances, enough evidence for the presence of individual proteins can be collected.
+Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
+Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.
 
-Two steps in protein identification can be distinguished:
+Multiple search engines exist that implement different matching and scoring algorithms.
+The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda CITE THE ENGINES.
+Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time REF THIS!!
 
-\begin{enumerate}
-
-\item Peptide inference: infer the peptides present in the sample.
-\item Protein inferece: based on the inferred peptides, infer what proteins generated them. This is not trivial as the peptide coming from the best PSM could map to more than one protein.
-\end{enumerate}
-
-Both are taken care of by the search engine.
-Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
-Notably, the results of each individual search engine can be combined to gather their strenghts, at the expense of an increased computational cost and time REF THIS!!
-
-%\subsection{X!Tandem}
-%
-%X!Tandem was one of the first open source search engines in mass spectrometry.
-It produces a score based on the dot product between the theoretical and the experimental tandem mass spectra CITE Overview of Tandem Mass Spectrometry (MS/MS) Database Search Algorithms.
-The scores assigned to wrong matches are assumed to follow a hypergeometric distribution, allowing the program to extrapolate $E(s)$ (expected number of wrong matches at a given score) for any score.
-
-\section{Results validation}
+\section{Validation and quality control}
 \label{sec:validation}
@@
 Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-This decoy database is generated by reversing or in general applying a randomization technique upon the sequences present in the original database (target).
+This decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
 This guarantess that while the sequences in the decoy database won't overlap with those in the target, their basic properties of the decoy (size, composition, etc) will remain identical to the target.
@@
 All matches to the decoy are by definition wrong. Since database properties 
-are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be used to estimate the number of false identifications in the  target results.
-it is as likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5, thus the number of matches in the decoy provides an accurate estimate of the number of random matches, or false positives, against the target database ($n_{fp}$).
+are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the equivalent list of target results.
+This is because the existence of shared properties entails that random matches are equally likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5.
 Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.
 
-$$FDR = \frac{n_{fp}}{n_{tp} + n_{tp}}$$
+\begin{equation}\label{eq:fdr}
+FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
+\end{equation}
 
-The minimal FDR at which a given PSM is considered a positive match constitutes the PSM\textquotesingle s q-value
+Equation \ref{eq:fdr} can be used to select the number of positives (the denominator of the fraction) that make the FDR equal to a predefined value, frequently 0.01 or 1 \%.
 
-$$ q(PSM_i) = \min FDR \forall PSM_i \in Positives$$
+%The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value
+%
+%\begin{equation}\label{eq:qvalue}
+%q(PSM_i) = \min FDR \forall PSM_i \in Positive
+%\end{equation}
 
-Finally, a posterior error probability (PEP) can be defined as the probability of the match being random ($P(s_i | T_i = 0)$).
-This can only be done if a statistical model describing the distribution of scores for correct and wrong matches is fit to the dataset.
+\section{Peptide and protein inference}
+\label{sec:inference}
+
+Two steps in protein identification can be distinguished:
+
+\begin{enumerate}
+
+\item \textbf{Peptide inference}: infer the peptides present in the sample.
+\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
+This is not trivial as peptides are degenerate and could map to more than one protein.
+\end{enumerate}
+
+Peptide inference is performed during the PSM process.
+The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
+The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
+Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
 
 \section{Protein quantification}
+\label{sec:quantification}
 
-CONTINUE maybe add reason for the project.
+The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
+However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
+For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
 
-\chapter{Materials and Methods}
+Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free).
+These are explained in subsection \ref{subsec:labelling}.
+If the label-free approach is employed, more distinctions can be made based on:
 
-\chapter{Results and Discussion}
+\begin{itemize}
+\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}
 
-\chapter{Conclusion}
+\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based.
+\end{itemize}
+
+\subsection{Label-based and label-free approaches}
+\label{subsec:labelling}
+
+Two paradigms exist in protein quantification: label-based and label-free.
+In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
+All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability CITE PAGE 237 COMPUTATIONAL METHODS The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified CITE MAXLFQ.
+Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) CITE or Isotope-Coded Affinity Tags (ICAT) CITE, differ in the nature of the label and the way it is introduced.
+Remarkably, peptide labeling costs can be quite high.
+This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
+
+In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
+In order to account for inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing can be carried out for pair-wise transfer of peptide identities from runs where identification was successful to those were it wasn't due to the randomness in the identification process.
+The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF.
+
+\subsection{SC and XIC based quantification}
+\label{subsec:scvsxic}
+
+Quantification can be spectral counting or \ac{XIC} based.
+Spectrum counting based quantification is the simplest quantification method in proteomics.
+It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
+As a result, peptides from abundant proteins are more likely to be
+identified and in more spectra.
+Two approaches were hence followed: count the number of peptides identified for a given protein – like in the emPAI index method – or count the number of spectra ascribed to a protein as in the NSAF method CITE COMPOMICS 4.1.
+These methods have the advantage that they are very simple to implement and don't require any further data processing.
+
+\ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance CITE.
+A wide range of algorithms are available to process these data and output estimates of protein abundance.
+All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
+They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
+
+Although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
+These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra.
+CITE MAXLFQ For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
+
+\subsection{\ac{XIC}-peptide-based models for label-free quantification}
+\label{subsec:peptide_model}
+
+The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample.
+However, the data interpretation requires the transfer of these peptide-level data into the protein level.
+This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.
+
+As stated in  CITE SUMMARY MSQROB: \textit{Peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
+However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
+This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
+
+It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
+This translates into improved results when compared to the other aforementioned methods CITE MSQROB SUMMARY.
+
+%CITE MAXQUANT
+%
+% The MaxLFQ algorithm, provided within the MaxQuant suite, is an \ac{MS1}, summarization method performing two optimisation steps to
+
+\chapter{Assembly of an open-source Linux-based label-free quantification proteomics pipeline}
+\label{chap:pipeline}
+
+\section{Introduction}
+
+Several proteomics pipelines are available on the internet under different licensing conditions.
+Many are released as closed-source software, where information on how the program works is kept from the user.
+This is a serious drawback as it hinders the study of the implemented modelling frameworks and its customisation.
+Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) or openMS  CITE ALL are nevertheless available for the community.
+MaxQuant, a free but closed-source proteomics analysis suite CITE MAXQUANT, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
+
+However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform.
+As a GUI would not be required to run the pipeline, it would also be scalable to big datasets.
+Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
+
+In this chapter, a pipeline making use of the set of tools published by the Compomics  and StatOmics groups CITE THEM, which fulfill all the foregoing requirements, will be presented.
+
+\section{Materials and Methods}
+
+\subsection{Data generation}
+
+\subsection{Data loading}
+
+\subsection{Database search}
+
+\subsection{Quality control and validation}
+
+\subsection{Data refinement}
+
+\subsection{Quantification}
+
+\section{Results and Discussion}
+
+\section{Conclusion}
+
+\chapter{Pipeline benchmarking on a real biological setting}
+\label{chap:benchmark}
+
+\chapter{Bayesian approach to protein quantification}
+\label{chap:model}
 
@@
 
-\section{Overview of proteomics analysis software}
+\chapter{Conclusion of the Thesis}
+\label{chap:conclusion}
 
-\subsection{MaxQuant}
-
-One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}.
+%\section{Overview of proteomics analysis software}
+%
+%\subsection{MaxQuant}
+%
+%One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}.
 It can be downloaded for free and is developed by the Max Planck institute in Germany.
 It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.
-
-MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
+%
+%MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
 A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.
-
-The main processing steps incorporated in MaxQuant consist of:
-
-\begin{itemize}
-
-\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.
-
-\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.
-
-\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.
-
-\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.
-
-\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.
-
-\item Supports label-based and label-free quantification.
-
-\item Its output can be passed to the Perseus software for data visualization. %CITE
-
-\end{itemize}
-
-However, MaxQuant also suffers from some problems:
-
-\begin{enumerate}
-
-\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}.
+%
+%The main processing steps incorporated in MaxQuant consist of:
+%
+%\begin{itemize}
+%
+%\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.
+%
+%\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.
+%
+%\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.
+%
+%\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.
+%
+%\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.
+%
+%\item Supports label-based and label-free quantification.
+%
+%\item Its output can be passed to the Perseus software for data visualization. %CITE
+%
+%\end{itemize}
+%
+%However, MaxQuant also suffers from some problems:
+%
+%\begin{enumerate}
+%
+%\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}.
 The processing steps cannot easily be fine tuned or exchanged with other pipelines.
@@
 This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.
-
-\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.
-
-\item While it is supported in Linux through Mono, the user experience is best in Windows.
+%
+%\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.
+%
+%\item While it is supported in Linux through Mono, the user experience is best in Windows.
 As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.
-
-\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.
-
-\end{enumerate}
-
-\subsection{Compomics group}
-
-The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.
-
-\begin{itemize}
-
-\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.
-
-\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.
-
-\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide.
+%
+%\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.
+%
+%\end{enumerate}
+%
+%\subsection{Compomics group}
+%
+%The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.
+%
+%\begin{itemize}
+%
+%\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.
+%
+%\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.
+%
+%\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide.
 Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.
-
-\end{itemize}
-
-SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}.
+%
+%\end{itemize}
+%
+%SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}.
 Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}.
 They easily enable the development of analytical pipelines finely fitted to each use case.
-
-moFF is less documented and integrated in the workflow.
+%
+%moFF is less documented and integrated in the workflow.
 Furthermore, it does not offer a GUI yet.
 The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.
-
-The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool.
+%
+%The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool.
 Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.
-
-\subsection{Other suites}
-
-Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available.
+%
+%
+%\subsection{Other suites}
+%
+%Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available.
 They are not open, requiring a license to function.
@@
 The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.
-
-\section{Label-free quantification}
-
-\subsection{MaxLFQ}
-
-The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}.
+%
+%\section{Label-free quantification}
+%
+%\subsection{MaxLFQ}
+%
+%The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}.
 It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.
-
-\subsubsection{Fractionated XIC aggreagation}
-
-First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions.
+%
+%\subsubsection{Fractionated XIC aggreagation}
+%
+%First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions.
 MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity
- between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation).
-
-The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
-%  this commented blank line prevents start of a new paragraph
-\begin{equation}\label{eq:peptide_intensity}
-I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
-\end{equation}
-
-The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
-%  this commented blank line prevents start of a new paragraph
-\begin{equation}\label{eq:sum_norm_intensity}
-H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
-\end{equation}
-%  this commented blank line prevents start of a new paragraph
-\begin{equation}
-H(N) = \sum\limits_{p} H_p(N) 
-\end{equation}
-
-where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.
-
-The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{xic_normalization}
-\caption{Taken from \cite{Cox2014}}
-\label{figure:xic_normalization}
-\end{figure}
-
-\subsubsection{Protein intensity inference}
-
-Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred.
+% between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation). 
+%
+%The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
+%%  this commented blank line prevents start of a new paragraph
+%\begin{equation}\label{eq:peptide_intensity}
+%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
+%\end{equation}
+%
+%
+%The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
+%%  this commented blank line prevents start of a new paragraph
+%\begin{equation}\label{eq:sum_norm_intensity}
+%H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
+%\end{equation}
+%%  this commented blank line prevents start of a new paragraph
+%\begin{equation}
+%H(N) = \sum\limits_{p} H_p(N) 
+%\end{equation}
+%
+%
+%where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.
+%
+%The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.
+%
+%\begin{figure}[!h]
+%\centering
+%\includegraphics[width=0.8\textwidth]{xic_normalization}
+%\caption{Taken from \cite{Cox2014}}
+%\label{figure:xic_normalization}
+%\end{figure}
+%
+%
+%\subsubsection{Protein intensity inference}
+%
+%Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred.
 This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).
-
-\begin{enumerate}
-\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.
-
-\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b.
+%
+%\begin{enumerate}
+%\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.
+%
+%\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b.
 The median is selected as a summarising statistic to protect from outliers.
@@
 Otherwise, the protein ratio for the pair of samples is set to 0.
-
-\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.
-
-\end{enumerate}
-
-\begin{equation}\label{eq:least_squares}
-\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
-\end{equation}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{protein_quant_inference}
-\caption{Taken from \cite{Cox2014}}
-\label{figure:protein_quant_inference}
-\end{figure}
-
-For the three most intense tryptic peptides, the signal per mole of protein was shown to be constant within a coefficient of variation of $\pm 10 \%$ \cite{Mikalsen2007}.
+%
+%\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.
+%
+%\end{enumerate}
+%
+%\begin{equation}\label{eq:least_squares}
+%\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
+%\end{equation}
+%
+%\begin{figure}[!h]
+%\centering
+%\includegraphics[width=0.8\textwidth]{protein_quant_inference}
+%\caption{Taken from \cite{Cox2014}}
+%\label{figure:protein_quant_inference}
+%\end{figure}
+%
+%For the three most intense tryptic peptides, the signal per mole of protein was shown to be constant within a coefficient of variation of $\pm 10 \%$ \cite{Mikalsen2007}.
 Thus, protein intensities are linearly proportional to protein quantities and protein quantity ratios can be computed from protein intensity ratios.
 These values are reported in the MaxQuant output in the \texttt{proteinGroups.txt} file under columns named \textit{LFQ Intensity}.
-
-\subsubsection{Drawbacks of the model}
-
-The LFQ model provides a powerful method to infer protein quantities from a label-free experiment.
+%		
+%\subsubsection{Drawbacks of the model}
+%
+%The LFQ model provides a powerful method to infer protein quantities from a label-free experiment.
 However, it makes use of point estimates in different steps of the process, namely (1) during peptide intensity aggregation, where the Levenberg-Marquardt minimum is selected (2) during protein ratio estimation, where the peptide median is selected and (3) during protein intensity estimation, where the least squares minimum is selected.
@@
 The only way of MaxQuant to provide uncertainty measurements is the number of peptides supporting the quantification measurement.
-
-\subsection{Bayesian model}
-
-Bayesian statistics provides probability measurements for observed data assuming underlying mathematical models.
+%
+%
+%\subsection{Bayesian model}
+%
+%Bayesian statistics provides probability measurements for observed data assuming underlying mathematical models.
 A model for protein quantification based on the \ac{MS1} intensities or XICs could be developed to assess both quantities and uncertainties behind the estimated quantities.
-
-While \ac{MS1} intensity measurements are dependent upon the underlying protein quantities, other factors influence the final measurement, thus adding noise and distorting the results.
+%
+%While \ac{MS1} intensity measurements are dependent upon the underlying protein quantities, other factors influence the final measurement, thus adding noise and distorting the results.
 Besides the actual quantity, two more factors can be distinguished:
-
-\begin{itemize}
-
-\item \textbf{Sequence derived factors}: some peptides are easier to cleave for the cleaving enzyme (Trypsin, etc) than others.
+%
+%\begin{itemize}
+%
+%\item \textbf{Sequence derived factors}: some peptides are easier to cleave for the cleaving enzyme (Trypsin, etc) than others.
 Moreover, the sequence of the peptide could influence the final measurement detected in the analyzer, by having different elution dynamics in the column or the analyzer.
-
-\item \textbf{Random noise}: the stochastic processes intrinsic to \ac{MS1} measurements could also have an impact.
-
-\end{itemize}
-
-The model would take as input:
-
-\begin{enumerate}
-
-\item \textbf{Features extracted from the precursor sequence}, including the surrounding aminoacids in the original protein. They can be used to model the sequenced derived factors.
-
-\item \textbf{Observed \ac{MS1} precursor intensities/XICs}.
-
-\end{enumerate}
-
-and return as output a protein quantification value for each protein, together with a measurement of the uncertainty behind it.
-
-\subsection{Data}
-
-The MaxLFQ paper dataset is available \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+%
+%\item \textbf{Random noise}: the stochastic processes intrinsic to \ac{MS1} measurements could also have an impact.
+%
+%\end{itemize}
+%
+%The model would take as input:
+%
+%\begin{enumerate}
+%
+%\item \textbf{Features extracted from the precursor sequence}, including the surrounding aminoacids in the original protein. They can be used to model the sequenced derived factors.
+%
+%\item \textbf{Observed \ac{MS1} precursor intensities/XICs}.
+%
+%\end{enumerate}
+%
+%and return as output a protein quantification value for each protein, together with a measurement of the uncertainty behind it.
+%
+%
+%
+%
+%\subsection{Data}
+%
+%The MaxLFQ paper dataset is available \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
 The authors submitted the RAW files (>50 GB) and the search files containing the final results.
 They can be used to benchmark any new quantification tool or method, for example, it has been used to benchmark third party quantification tools like StPeter from the TPP \cite{Hoopmann2018}.
-
-The dataset contains the results of the quantification of proteins in 2 different proteomes (\textit{E.
+%
+%The dataset contains the results of the quantification of proteins in 2 different proteomes (\textit{E.
 coli} and \textit{Homo sapiens}) from 6 samples organized in 2 conditions with 3 replicates each.
@@
 Thus, the fold change of protein quantities across conditions should be 1 for human proteins and 3 for bacterial proteins.
-
-A list of the protein groups identified that could be mapped unambiguously to each species was made available \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}}.
+%
+%A list of the protein groups identified that could be mapped unambiguously to each species was made available \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}}.
 Thus, a dataset of thousands of proteins of known ratios between conditions, producing tens of thousands of peptides with known sequence, \ac{MS1} intensity/XIC and parent protein can be built from the publication, and used to train new quantification methods (figure \ref{fig:screenshot_peptide_data}).
-
-% latex table generated in R 3.4.3 by xtable 1.8-2 package
-% Fri Jun 01 17:11:23 2018
-\begin{table}[ht]
-\centering
-\begin{tabular}{rlrr}
-  \hline
- & Taxonomy & Peptides & Proteins \\ 
-  \hline
-1 & Escherichia coli (strain K12) & 14483 & 1556 \\ 
-  2 & Homo sapiens & 32647 & 3444 \\ 
-   \hline
-   & Total & 47130 & 5000
-\end{tabular}
-\caption{Summary table of the dataset}
-\label{table:summary_peptide_data}
-\end{table}
-
-\begin{figure}[!h]
-\includegraphics[width=1\textwidth]{dataset}
-\caption{Extract of the available dataset. Total of 47130 entries, one per peptide, for a global of 5000 proteins.}
-\label{fig:screenshot_peptide_data}
-\end{figure}
-
-This compiled dataset can be downloaded here \href{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}
-
-\section{Extension of pipelines}
-
-The Compomics suite of programmes provides the best documented and accesible set of tools for proteomics analysis, but still lacks proper quantification tools.
+%
+%% latex table generated in R 3.4.3 by xtable 1.8-2 package
+%% Fri Jun 01 17:11:23 2018
+%\begin{table}[ht]
+%\centering
+%\begin{tabular}{rlrr}
+%  \hline
+% & Taxonomy & Peptides & Proteins \\ 
+%  \hline
+%1 & Escherichia coli (strain K12) & 14483 & 1556 \\ 
+%  2 & Homo sapiens & 32647 & 3444 \\ 
+%   \hline
+%   & Total & 47130 & 5000
+%\end{tabular}
+%\caption{Summary table of the dataset}
+%\label{table:summary_peptide_data}
+%\end{table}
+%
+%\begin{figure}[!h]
+%\includegraphics[width=1\textwidth]{dataset}
+%\caption{Extract of the available dataset. Total of 47130 entries, one per peptide, for a global of 5000 proteins.}
+%\label{fig:screenshot_peptide_data}
+%\end{figure}
+%
+%This compiled dataset can be downloaded here \href{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}
+%
+%\section{Extension of pipelines}
+%
+%The Compomics suite of programmes provides the best documented and accesible set of tools for proteomics analysis, but still lacks proper quantification tools.
 The implementation of a MS1 intensity-based downstream quantification tool compatible with the output of the Compomics tools would yield a free Linux compatible, CLI supported, complete proteomics pipeline featuring a robust quantification method.

### 2018-07-03 report.tex
Bibliography added
Extended chapter 2

@@
 \usepackage{float} % force float
-\usepackage[nottoc,numbib]{tocbibind} % make bibliography appear in toc
+\usepackage[nottoc]{tocbibind}
 %\usepackage[margin=-50pt]{caption}
@@
 
-\subsection{Data generation}
+\subsection{Data generation and loading}
 
-\subsection{Data loading}
+The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.
 
-\subsection{Database search}
+\subsection{Decoy database preparation and search}
+
+The spectra contained in the MGF files obtained in the previous step were loaded onto the MS-GF+ search engine CITE using the SearchGUI \texttt{SearchCLI} CITE utility.
+The search parameters were set using the \texttt{IdentificationParametersCLI}.
+In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
+Moreover, C carbamidomethylation was set as fixed modification during the search.
+The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
+Up to two missed cleavages were allowed.
+The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
+
+The database was created by combining the Uniprot proteomes for \textit{E.
+coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
+The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the original database.
 
@@
 
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+By default, the FDR is set to 1\%.
+Output was extracted via the Default PSM report, available in the \texttt{ReportCLI} utility
+
 \subsection{Data refinement}
 
+The moFF command line utility CITE was used to perform a match-between-runs and extract apex intensity of each peak cluster.
+This required reinputing the original RAW files, together with the txt files produced by the Default PSM report from PeptideShaker.
+Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
+
 \subsection{Quantification}
+
+Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
 
@@
 
-%\section{Overview of proteomics analysis software}
-%
-%\subsection{MaxQuant}
-%
-%One of the most used program for the quantification of label free proteins is the MaxQuant suite \cite{Cox2008}.
-It can be downloaded for free and is developed by the Max Planck institute in Germany.
-It consists of a user friendly GUI that provides the most needed steps in a proteomics pipeline.
-%
-%MaxQuant has been successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
-A Google group and a tutorial are available  \footnote{\href{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}{http://www.coxdocs.org/doku.php?id=maxquant:viewer:tutorial}} \footnote{\href{https://www.youtube.com/watch?v=\_AJHFHi5CxM}{https://www.youtube.com/watch?v=\_AJHFHi5CxM}}.
-%
-%The main processing steps incorporated in MaxQuant consist of:
-%
-%\begin{itemize}
-%
-%\item Read MS spectra files in the .RAW format, the closed format produced by ThermoScientific MS analyzers.
-%
-%\item Contains the Andromeda search engine \cite{Cox2011} for matching of \ac{MS2} spectra to a proteome database.
-%
-%\item Andromeda supports configuring aminoacid modifications and decoy search for FDR (false discovery rate) estimation.
-%
-%\item Increase peptide identifications by performing match between runs (MBR), which aims at transferring peptide to spectrum matches (PSMs) across replicate runs, based on precursor mass and retention times.
-%
-%\item Filters Andromeda results to provide a list of inferred peptides and proteins with a cutoff FDR value.
-%
-%\item Supports label-based and label-free quantification.
-%
-%\item Its output can be passed to the Perseus software for data visualization. %CITE
-%
-%\end{itemize}
-%
-%However, MaxQuant also suffers from some problems:
-%
-%\begin{enumerate}
-%
-%\item MaxQuant only runs on Microsoft Windows, impeding its integration in automated pipelines on cluster environments \cite{Argentini2016}.
-The processing steps cannot easily be fine tuned or exchanged with other pipelines.
-For example, only the Andromeda search engine is supported.
-This is a serious drawback, as the integration of results from several search engines, like MSGF+ \cite{Kim2014} or MS-Amanda \cite{Dorfer2014}  has been shown to further improve results \cite{Shteynberg2013}.
-%
-%\item It does not provide a command line interface (CLI), thus all analyses must be performed through the GUI. This hinders reproducibility and scalability.
-%
-%\item While it is supported in Linux through Mono, the user experience is best in Windows.
-As Linux is by far the most used OS in Bioinformatics, this implies that an additional OS is required to get the best experience with MaxQuant.
-%
-%\item Lacks a well written official documentation, as most information is made available only through talks published on Youtube or third party tutorials.
-%
-%\end{enumerate}
-%
-%\subsection{Compomics group}
-%
-%The Compomics group at Ghent University \footnote{\href{https://compomics.com/}{https://compomics.com/}} provides high quality  analysis software for the identification and integration of results through the SearchGUI, PeptideShaker and moFF (modest Feature Finder) programs \cite{Barsnes2018} \cite{Vaudel2015} \cite{Argentini2016}.
-%
-%\begin{itemize}
-%
-%\item SearchGUI provides a common interface to a range of search engines so that multiple engines can be used in a straightforward manner.
-%
-%\item PeptideShaker reads the SearchGUI output and performs quality control, gene set enrichment analysis (GSEA), and implements multiple validation filters to provide the best results.
-%
-%\item moFF reads from PeptideShaker output and implements a match between runs (MBR) model analogous to that in MaxQuant, to increase the number of spectra that are matched to a peptide.
-Furthermore, it summarises the \ac{MS1} peaks into refined features, making downstream analyses more sensitive.
-%
-%\end{itemize}
-%
-%SearchGUI and PeptideShaker are very well integrated and documented \footnote{\href{https://compomics.com/bioinformatics-for-proteomics/}{https://compomics.com/bioinformatics-for-proteomics/}}.
-Both provide not just beautiful GUIs, but also a comprehensive and extremely well documented CLI \footnote{\href{https://github.com/compomics/searchgui/wiki/SearchCLI}{https://github.com/compomics/searchgui/wiki/SearchCLI}} \footnote{\href{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}{https://github.com/compomics/peptide-shaker/wiki/PeptideShakerCLI}}.
-They easily enable the development of analytical pipelines finely fitted to each use case.
-%
-%moFF is less documented and integrated in the workflow.
-Furthermore, it does not offer a GUI yet.
-The group is currently working to better integrate moFF and PeptideShaker \footnote{\href{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}{https://groups.google.com/forum/\#!topic/peptide-shaker/Lqe7lYKLcHI}}.
-%
-%The main issue in the Compomics suite of tools is the lack of a robust label-free quantification step, as they currently recommend the export of results to a third party tool.
-Only the spectral count based methods emPAi (Exponentially Modified Protein Abundance Index) and NSAF (Normalized Spectral Abundance Factor), which are not robust \cite{Griffin2010}, are supported.
-%
-%
-%\subsection{Other suites}
-%
-%Other software suites like ProteomeDiscoverer, Progenesis QI, GeneData Expressionist, etc, are available.
-They are not open, requiring a license to function.
-Furthermore, they are GUI oriented and make use of different data formats thus, making exchange across pipelines extremely difficult.
-The OpenMS \cite{Sturm2008} and Trans Proteomic Pipeline (TPP) \cite{Deutsch2011} suites are open, but they also suffer from data exchangeability and lack comprehensive documentations.
-%
-%\section{Label-free quantification}
-%
-%\subsection{MaxLFQ}
-%
-%The label-free quantification engine implemented in MaxQuant is MaxLFQ \cite{Cox2014}.
-It performs 2 minimisation steps to infer protein quantities from extracted ion currents (XIC), as stored in the spectra files.
-%
-%\subsubsection{Fractionated XIC aggreagation}
-%
-%First, due to the prefrationation of samples during the upstream mass spectrometry (MS) analysis, XIC signals for each peptide are distributed along multiple fractions.
-MaxLFQ summarises the XIC from several fractions for each peptide and sample into a single peptide intensity by minimising the sum of the square of the logarithm of the intensity
-% between samples via Levenberg–Marquardt optimization (see figure \ref{figure:xic_normalization} for a graphical explanation). 
-%
-%The intensity of peptide $P$ in sample $A$ is defined as a weighted average of $P$ \textquotesingle s XIC signals across fractions $\{1..k\}$, where the weights are the normalization factors that the minimisation algorithm seeks to find.
-%%  this commented blank line prevents start of a new paragraph
-%\begin{equation}\label{eq:peptide_intensity}
-%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{P,A,j} \times \text{XIC}_{P,A,j}}
-%\end{equation}
-%
-%
-%The minimisation assumes that the best normalization factors minimize the sum of squared logarithm pairwise-ratios $H$.
-%%  this commented blank line prevents start of a new paragraph
-%\begin{equation}\label{eq:sum_norm_intensity}
-%H_P(N) = \sum\limits_{a,b} \abs{\frac{I_{P,a}(N)}{I_{P,b}(N)}}^2
-%\end{equation}
-%%  this commented blank line prevents start of a new paragraph
-%\begin{equation}
-%H(N) = \sum\limits_{p} H_p(N) 
-%\end{equation}
-%
-%
-%where $a,b$ iterates over all possible pair-wise combinations of samples where P was detected.
-%
-%The peptide intensities can be computed from the estimated normalization factors and the observed XIC signals.
-%
-%\begin{figure}[!h]
-%\centering
-%\includegraphics[width=0.8\textwidth]{xic_normalization}
-%\caption{Taken from \cite{Cox2014}}
-%\label{figure:xic_normalization}
-%\end{figure}
-%
-%
-%\subsubsection{Protein intensity inference}
-%
-%Second, once the peptide intensities have been summarised, the intensity of the proteins that the peptides originated from are inferred.
-This is achieved in three steps (see figure \ref{figure:protein_quant_inference} for a graphical explanation).
-%
-%\begin{enumerate}
-%\item Peptide intensity ratios are computed for all possible pairwise combinations by dividing the normalized intensities obtained in the previous step.
-%
-%\item For every inferred protein A, its protein ratio $r_{A, a,b}$ for the pair of samples a,b is set to the median intensity of its children peptide ratio in samples a,b.
-The median is selected as a summarising statistic to protect from outliers.
-A minimal number of non-zero intensity peptides are required for the median to be valid, usually 2.
-Otherwise, the protein ratio for the pair of samples is set to 0.
-%
-%\item The correct protein intensities are assumed to minimise the sum of the squared difference of the logarithm of the protein ratios and logarithm of the protein intensities $I$.
-%
-%\end{enumerate}
-%
-%\begin{equation}\label{eq:least_squares}
-%\sum_{a,b} (\log{(r_{A, a,b})} - \log{(I_{A,a})} + \log{(I_{A,b})})^2
-%\end{equation}
-%
-%\begin{figure}[!h]
-%\centering
-%\includegraphics[width=0.8\textwidth]{protein_quant_inference}
-%\caption{Taken from \cite{Cox2014}}
-%\label{figure:protein_quant_inference}
-%\end{figure}
-%
-%For the three most intense tryptic peptides, the signal per mole of protein was shown to be constant within a coefficient of variation of $\pm 10 \%$ \cite{Mikalsen2007}.
-Thus, protein intensities are linearly proportional to protein quantities and protein quantity ratios can be computed from protein intensity ratios.
-These values are reported in the MaxQuant output in the \texttt{proteinGroups.txt} file under columns named \textit{LFQ Intensity}.
-%		
-%\subsubsection{Drawbacks of the model}
-%
-%The LFQ model provides a powerful method to infer protein quantities from a label-free experiment.
-However, it makes use of point estimates in different steps of the process, namely (1) during peptide intensity aggregation, where the Levenberg-Marquardt minimum is selected (2) during protein ratio estimation, where the peptide median is selected and (3) during protein intensity estimation, where the least squares minimum is selected.
-These point estimation bottlenecks, specially (2), discard increasing amounts of data that could otherwise be used for more accurate results.
-Moreover, not just quantities, but also the uncertainty behind them, could also be provided.
-The only way of MaxQuant to provide uncertainty measurements is the number of peptides supporting the quantification measurement.
-%
-%
-%\subsection{Bayesian model}
-%
-%Bayesian statistics provides probability measurements for observed data assuming underlying mathematical models.
-A model for protein quantification based on the \ac{MS1} intensities or XICs could be developed to assess both quantities and uncertainties behind the estimated quantities.
-%
-%While \ac{MS1} intensity measurements are dependent upon the underlying protein quantities, other factors influence the final measurement, thus adding noise and distorting the results.
-Besides the actual quantity, two more factors can be distinguished:
-%
-%\begin{itemize}
-%
-%\item \textbf{Sequence derived factors}: some peptides are easier to cleave for the cleaving enzyme (Trypsin, etc) than others.
-Moreover, the sequence of the peptide could influence the final measurement detected in the analyzer, by having different elution dynamics in the column or the analyzer.
-%
-%\item \textbf{Random noise}: the stochastic processes intrinsic to \ac{MS1} measurements could also have an impact.
-%
-%\end{itemize}
-%
-%The model would take as input:
-%
-%\begin{enumerate}
-%
-%\item \textbf{Features extracted from the precursor sequence}, including the surrounding aminoacids in the original protein. They can be used to model the sequenced derived factors.
-%
-%\item \textbf{Observed \ac{MS1} precursor intensities/XICs}.
-%
-%\end{enumerate}
-%
-%and return as output a protein quantification value for each protein, together with a measurement of the uncertainty behind it.
-%
-%
-%
-%
-%\subsection{Data}
-%
-%The MaxLFQ paper dataset is available \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-The authors submitted the RAW files (>50 GB) and the search files containing the final results.
-They can be used to benchmark any new quantification tool or method, for example, it has been used to benchmark third party quantification tools like StPeter from the TPP \cite{Hoopmann2018}.
-%
-%The dataset contains the results of the quantification of proteins in 2 different proteomes (\textit{E.
-coli} and \textit{Homo sapiens}) from 6 samples organized in 2 conditions with 3 replicates each.
-In the first condition (H), both proteomes were mixed in a 1:1 ratio, while in the second condition (L), the \textit{E.
-coli} proteome had 3 times more contribution to the mix (3:1).
-Thus, the fold change of protein quantities across conditions should be 1 for human proteins and 3 for bacterial proteins.
-%
-%A list of the protein groups identified that could be mapped unambiguously to each species was made available \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}}.
-Thus, a dataset of thousands of proteins of known ratios between conditions, producing tens of thousands of peptides with known sequence, \ac{MS1} intensity/XIC and parent protein can be built from the publication, and used to train new quantification methods (figure \ref{fig:screenshot_peptide_data}).
-%
-%% latex table generated in R 3.4.3 by xtable 1.8-2 package
-%% Fri Jun 01 17:11:23 2018
-%\begin{table}[ht]
-%\centering
-%\begin{tabular}{rlrr}
-%  \hline
-% & Taxonomy & Peptides & Proteins \\ 
-%  \hline
-%1 & Escherichia coli (strain K12) & 14483 & 1556 \\ 
-%  2 & Homo sapiens & 32647 & 3444 \\ 
-%   \hline
-%   & Total & 47130 & 5000
-%\end{tabular}
-%\caption{Summary table of the dataset}
-%\label{table:summary_peptide_data}
-%\end{table}
-%
-%\begin{figure}[!h]
-%\includegraphics[width=1\textwidth]{dataset}
-%\caption{Extract of the available dataset. Total of 47130 entries, one per peptide, for a global of 5000 proteins.}
-%\label{fig:screenshot_peptide_data}
-%\end{figure}
-%
-%This compiled dataset can be downloaded here \href{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}{https://mega.nz/\#!AgcTgJYa!w9DoAKYRc6u-SaRy\_UIMz3aileUHXoaWgrxf-UycqiQ}
-%
-%\section{Extension of pipelines}
-%
-%The Compomics suite of programmes provides the best documented and accesible set of tools for proteomics analysis, but still lacks proper quantification tools.
-The implementation of a MS1 intensity-based downstream quantification tool compatible with the output of the Compomics tools would yield a free Linux compatible, CLI supported, complete proteomics pipeline featuring a robust quantification method.
-
 \bibliographystyle{unsrt_etal}

### 2018-07-03 report.tex
Finished materials methods chap2

@@
 The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+Briefly, the \textit{Homo sapiens} and \textit{E.
+coli (strain K12)} proteomes were mixed in 1:1 and 1:3 proportions, with 3 replicates for each combination.
+Moreover, each of the 3 replicates of the 2 conditions was analysed in 24 fractions.
+This experimental setting thus generated a total of 2*3*24=144 RAW files.
 The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.
@@
 Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
+Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
+In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group CITE MAIN MSQROB were discarded, and (III) protein groups with only 1 peptide were dropped.
+Intensity normalisation was not performed as a systematic bias in intensities is indeed expected for all \textit{E.
+coli} peptides from samples where the \textit{E.
+coli}proteome was 3 times more abundant.
+
+One preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of the protein variance was fit to every protein individually.
+The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
+
+The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
 

### 2018-07-04 report.tex
Add summaries

--- 2018-07-03 report.tex
@@
 \usepackage[utf8]{inputenc}
+\usepackage{pdfpages}
+
 \DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
@@
 \usepackage{mathtools}
+\usepackage{sectsty}
+
+\chapternumberfont{\LARGE} 
+\chaptertitlefont{\LARGE}
 
@@
 
-\chapter{Introduction}
+\chapter*{Introduction}
 \label{chap:introduction}
@@
   \centering
-  \includegraphics[width=0.5\textwidth]{aminoacids.png}
-  \caption{}
+  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
+  \caption{CAPTION AND REFERENCE}
   \label{fig:aminoacids}
@@
 Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another.
-CITE LEHNINGER PAG 82.
+Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
 The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
 
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures CITE.
-One frequent instance of such modifications is the oxidation of methionine
+While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
+One frequent instance of such modifications is the oxidation of methionine.
 
@@
 These smaller parts are called peptides.
-CITE 1.6.1 COMPUTATIONAL METHODS.
+\cite{Barsnes2008} CITE 1.6.1 COMPUTATIONAL METHODS.
 Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly mass spectrometry (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry.
-The interested reader is directed to CITE 1.6.1 COMPUTATIONAL METHODS to learn why.
+The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
 The top-down paradigm will be ignored in the rest of the manuscript.
@@
 Once the electrophoresis is completed, protein bands can be excised from the gel.
-Each band will contain a subset, or even only one of the proteins originally available, thus making the downstream analysis simpler REFERENCE COMPUTATIONAL METHODS CHAPTER 2
+Each band will contain a subset, or even only one of the proteins originally available, thus making the downstream analysis simpler \cite{Barsnes2008} REFERENCE COMPUTATIONAL METHODS CHAPTER 2
 
@@
 Trypsin is the most frequently used for this task, which cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: COMPUTATIONAL METHODS 3.2
+Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008} COMPUTATIONAL METHODS 3.2
 
@@
 \includegraphics[width=\textwidth]{mass_spectrometer}
-\caption{Schematic view of a mass spectrometer. Taken from}
-%CITE THIS https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation
+\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark}
 \label{fig:mass_spectrometer}
 \end{figure}
+
+\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
 
@@
 Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-This step is performed in the ion source CITE 5.1 COMPUTATIONAL METHODS.
+This step is performed in the ion source \cite{Barsnes2008} CITE 5.1 COMPUTATIONAL METHODS.
 The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (MALDI) and Electro Spray Ionization (ESI) CITE.
@@
 
-Fractionation is achieved by different methods of separation CITE 1.7, most commonly via \ac{HPLC} methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point)  CITE 4.2 computational methods.
+Fractionation is achieved by different methods of separation CITE 1.7, most commonly via \ac{HPLC} methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008} CITE 4.2 computational methods.
 
@@
 Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-PAG 123 COMPUTATIONAL METHODS.
+\cite{Barsnes2008} PAG 123 COMPUTATIONAL METHODS.
 Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications CITE pag 134 computational methods.
+This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008} CITE pag 134 computational methods.
 
@@
 In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability CITE PAGE 237 COMPUTATIONAL METHODS The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified CITE MAXLFQ.
+All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008} CITE PAGE 237 COMPUTATIONAL METHODS The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
 Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) CITE or Isotope-Coded Affinity Tags (ICAT) CITE, differ in the nature of the label and the way it is introduced.
@@
 These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra.
-CITE MAXLFQ For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
+\cite{Cox2014} For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
 
@@
 
-As stated in  CITE SUMMARY MSQROB: \textit{Peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
+As stated in \cite{Goeminne2015}.
+\textit{Peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
 However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
@@
 It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
-This translates into improved results when compared to the other aforementioned methods CITE MSQROB SUMMARY.
-
+This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
 %CITE MAXQUANT
@@
 
-\chapter{Assembly of an open-source Linux-based label-free quantification proteomics pipeline}
+\chapter{A label-free quantification proteomics pipeline}
 \label{chap:pipeline}
+
+\section*{Summary}
+A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
+The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) feature extraction, (IV) protein inference and (V) relative quantification.
+Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
+A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
+The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fraction of this dataset.
 
@@
 
-In this chapter, a pipeline making use of the set of tools published by the Compomics  and StatOmics groups CITE THEM, which fulfill all the foregoing requirements, will be presented.
-
 \section{Materials and Methods}
 
-\subsection{Data generation and loading}
+\subsection{Data generation}
 
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.
+\subsection{Data loading}
 
-\subsection{Decoy database preparation and search}
-
-The spectra contained in the MGF files obtained in the previous step were loaded onto the MS-GF+ search engine CITE using the SearchGUI \texttt{SearchCLI} CITE utility.
-The search parameters were set using the \texttt{IdentificationParametersCLI}.
-In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification during the search.
-The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
-Up to two missed cleavages were allowed.
-The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
-
-The database was created by combining the Uniprot proteomes for \textit{E.
-coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the original database.
+\subsection{Database search}
 
@@
 
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR is set to 1\%.
-Output was extracted via the Default PSM report, available in the \texttt{ReportCLI} utility
-
 \subsection{Data refinement}
 
-The moFF command line utility CITE was used to perform a match-between-runs and extract apex intensity of each peak cluster.
-This required reinputing the original RAW files, together with the txt files produced by the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
-
 \subsection{Quantification}
-
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
 
@@
 
-\chapter{Pipeline benchmarking on a real biological setting}
+\chapter{Pipeline benchmarking on \ac{NZ} data}
 \label{chap:benchmark}
 
-\chapter{Bayesian approach to protein quantification}
+\section*{Summary}
+
+A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
+The experiment consisted of the application of two different treatments to THP-1 cell cultures.
+One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
+The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
+The results confirmed that the computational analysis successfully captured this response.
+It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
+
+\chapter{Bayesian modelling of fold change estimates}
 \label{chap:model}
+
+\section*{Summary}
+
+The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the fold change across conditions.
+However, a Bayesian based approach to this problem is lacking in the literature.
+As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
+The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
 
@@
 
-\chapter{Conclusion of the Thesis}
+\chapter*{Conclusion}
 \label{chap:conclusion}
+
+\chapter*{Appendix}
+\label{chap:appendix}
+
+\includepdf[pages=-]{appendix.pdf}
 

### 2018-07-04 appendix.tex

\documentclass[a4paper]{article}
\usepackage{fontspec}
\usepackage{minted}

%\setsansfont{Calibri}
%\setmonofont{Consolas}

\begin{document}
\renewcommand{\theFancyVerbLine}{
  \sffamily\textcolor[rgb]{0.5,0.5,0.5}{\scriptsize\arabic{FancyVerbLine}}}

\begin{minted}[mathescape,
               linenos,
               xleftmargin=-70pt,
               numbersep=10pt,
               gobble=0,
               frame=lines,
               framesep=2mm]{python}

import pymc3 as pm
import numpy as np
import matplotlib.pyplot as plt
from theano import shared
import shutil
import os

def protein_model(observed_sh, feats_sh, x_treat_sh, x_pep_sh, x_run_sh, x_estimate_sh,
    n_peptides, model_name, n_draws=1000, n_chains=3,
    hierarchical_center=False, remove_backend=True, sequence=False):

    # Check working environment  
    if not os.path.isdir("traces") or not os.path.isdir("plots/traceplots"):
        msg = "Please create a traces dir and a plots/traceplots dir before running this code"
        raise Exception(msg)

    if remove_backend and os.path.isdir(model_name):
        shutil.rmtree(model_name)

    # The number of proteins in this model is always one
    # i.e this model is fitted protein-wise
    n_prots = 1
    # The number of features is set to 9 for now
    # All peptides have 9 features, stored in feats_shared
    n_features = 9

    with pm.Model() as model:

        # Build a hyerarchical linear model of
        # log2(MS1 intensities) by accounting for:

             # Peptide effect
             # Run (batch) effect
             # Treatment effect
             # Remaining random effects

        # The difference in treatment effects is an estimate of the log2FC

        # Set a prior on the intercept
        intercept = pm.Normal("intercept", 22, 1)

        # Set a prior on the remaining random effects
        sigma = pm.HalfNormal('sigma', 1)

        ## Set priors on the peptide effect
        ################################################
        sigma_pep = pm.HalfNormal('sigma_pep', 1)

        # Not using the sequence
        if not sequence:
            mu_pep = pm.Normal('mu_pep', mu=0, sd=sigma_pep, shape=(n_peptides, 1))

        # Using the peptide sequence
        else: 
            # sequence based modelling
            mu_theta = pm.Normal('theta_generic', 0, sigma_pep, shape = 1)
            theta = pm.Normal('theta', mu_theta, sigma_pep, shape = (n_features, 1))    # 9x1
            theta_inter = pm.Normal('theta_inter', mu_theta, sigma_pep, shape = 1)
            mu_pep = pm.Deterministic("mu_pep", theta_inter + feats_sh.dot(theta)) # n_peptidesx1

        ## Set priors on the treatment and run effects
        ################################################    
        sigma_treat = pm.HalfNormal('sigma_treat', 1)
        mu_treat = pm.Normal('mu_treat', 0, sigma_treat)
        sigma_run = pm.HalfNormal('sigma_run', 1)
        mu_run = pm.Normal('mu_run', 0, sigma_run)

        # Standard implementation of the hyerarchies
        if hierarchical_center:
            pep = pm.Normal("pep", mu_pep, sigma_pep) # n_peptidesx1
            treat = pm.Normal('treat', mu_treat, sigma_treat, shape = (n_prots*2, 1))
            run = pm.Normal('run', mu_run, sigma_run, shape = (n_prots*6, 1))

        # Reparametrization to escape funnel of hell as noted in
        # http://twiecki.github.io/blog/2017/02/08/bayesian-hierchical-non-centered/
        else:
            pep_offset = pm.Normal("pep_offset", mu=0, sd=1, shape = (n_peptides, 1))
            pep = pm.Deterministic("pep", mu_pep + pep_offset * sigma_pep)
            treat_offset = pm.Normal("treat_offset", mu=0, sd=1, shape=(n_prots*2, 1))
            treat = pm.Deterministic("treat", mu_treat + treat_offset*sigma_treat)
            run_offset = pm.Normal("run_offset", mu=0, sd=1, shape=(n_prots*6, 1))
            run = pm.Deterministic("run", mu_run + run_offset*sigma_run)

        # Model the effect for all peptides
        # The sh variables consist of -1,0,1 matrices telling pymc3
        # which parameters shall be used with each peptide
        # In practice, the "clone" each parameter to fit the shape of observed_sh
        # observed_sh is a n_peptides*6x1 tensor
        # The first 6 numbers store the MS1 intensities of the first peptide in the 6 runs
        # The next 6 those of the second peptide, and so on

        estimate = pm.Deterministic('estimate', pm.math.sum(x_estimate_sh.dot(treat), axis=1))
        treatment_effect = pm.Deterministic("treatment_effect", pm.math.sum(x_treat_sh.dot(treat), axis=1))
        peptide_effect = pm.Deterministic("peptide_effect", pm.math.sum(x_pep_sh.dot(pep), axis=1))
        run_effect = pm.Deterministic("run_effect", pm.math.sum(x_run_sh.dot(run), axis=1))

        # BIND MODEL TO DATA
        mu = pm.Deterministic("mu", 
            intercept + treatment_effect + peptide_effect + run_effect) #n_peptides*6x1
        if hierarchical_center:
            obs = pm.Normal("obs", mu, sigma, observed=observed_sh)
        else:
            obs_offset = pm.Normal("obs_offset", mu=0, sd=1, shape=(n_peptides*6,1))
            obs = pm.Normal("obs", mu+obs_offset*sigma, sigma, observed=observed_sh)

    print("Success: Model {} compiled".format(model_name))

    with model:
        # Parameters of the simulation:
        # Number of iterations and independent chains.
        n_sim = n_draws*n_chains

        # Save traces to the Text backend i.e a folder called
        # model_name containing csv files for each chain
        trace_name = 'traces/{}'.format(model_name)
        db = pm.backends.Text(trace_name)
        trace = pm.sample(draws=n_draws, njobs=n_chains, trace=db,
                          tune=2000, nuts_kwargs=dict(target_accept=.95))

    # Save a traceplot 
    pm.traceplot(trace, varnames=["estimate"])
    traceplot = "plots/traceplots/{}.png".format(model_name)
    plt.savefig(traceplot)
    plt.close()

    return model

\end{minted}
\end{document}

### 2018-07-04 report.tex

@@
 
-\subsection{Data generation}
+\subsection{Data generation and loading}
 
-\subsection{Data loading}
+The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+Briefly, the \textit{Homo sapiens} and \textit{E.
+coli (strain K12)} proteomes were mixed in 1:1 and 1:3 proportions, with 3 replicates for each combination.
+Moreover, each of the 3 replicates of the 2 conditions was analysed in 24 fractions.
+This experimental setting thus generated a total of 2*3*24=144 RAW files.
+The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.
 
-\subsection{Database search}
+\subsection{Decoy database preparation and search}
+
+The spectra contained in the MGF files obtained in the previous step were loaded onto the MS-GF+ search engine CITE using the SearchGUI \texttt{SearchCLI} CITE utility.
+The search parameters were set using the \texttt{IdentificationParametersCLI}.
+In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
+Moreover, C carbamidomethylation was set as fixed modification during the search.
+The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
+Up to two missed cleavages were allowed.
+The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
+
+The database was created by combining the Uniprot proteomes for \textit{E.
+coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
+The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the original database.
 
@@
 
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+By default, the FDR is set to 1\%.
+Output was extracted via the Default PSM report, available in the \texttt{ReportCLI} utility
+
 \subsection{Data refinement}
 
+The moFF command line utility CITE was used to perform a match-between-runs and extract apex intensity of each peak cluster.
+This required reinputing the original RAW files, together with the txt files produced by the Default PSM report from PeptideShaker.
+Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
+
 \subsection{Quantification}
+
+Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
+Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
+In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group CITE MAIN MSQROB were discarded, and (III) protein groups with only 1 peptide were dropped.
+Intensity normalisation was not performed as a systematic bias in intensities is indeed expected for all \textit{E.
+coli} peptides from samples where the \textit{E.
+coli}proteome was 3 times more abundant.
+
+One preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of the protein variance was fit to every protein individually.
+The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
+
+The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
 

### 2018-07-06 report.tex
Started results chap 2

@@
 A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fraction of this dataset.
+The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
 
@@
 Many are released as closed-source software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented modelling frameworks and its customisation.
+This is a serious drawback as it hinders the study of the implemented models and its customisation.
 Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) or openMS  CITE ALL are nevertheless available for the community.
@@
 Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
+
+The development of a pipeline achieving these goals will be described in this chapter.
 
@@
 This experimental setting thus generated a total of 2*3*24=144 RAW files.
-The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format) that can be used by the tools following the pipeline.
+The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).
 
@@
 
-The spectra contained in the MGF files obtained in the previous step were loaded onto the MS-GF+ search engine CITE using the SearchGUI \texttt{SearchCLI} CITE utility.
+The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine CITE by means of the SearchGUI \texttt{SearchCLI} tool CITE utility.
 The search parameters were set using the \texttt{IdentificationParametersCLI}.
 In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification during the search.
+Moreover, C carbamidomethylation was set as fixed modification.
 The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
@@
 
-The database was created by combining the Uniprot proteomes for \textit{E.
+The target database was created by combining the Uniprot proteomes for \textit{E.
 coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the original database.
+The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
 
@@
 The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR is set to 1\%.
-Output was extracted via the Default PSM report, available in the \texttt{ReportCLI} utility
+By default, the FDR was set to 1\%.
+Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
 
@@
 
-The moFF command line utility CITE was used to perform a match-between-runs and extract apex intensity of each peak cluster.
-This required reinputing the original RAW files, together with the txt files produced by the Default PSM report from PeptideShaker.
+The moFF command line utility CITE was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
+This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
 Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
@@
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group CITE MAIN MSQROB were discarded, and (III) protein groups with only 1 peptide were dropped.
+In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded CITE MAIN MSQROB, and (III) protein groups with only 1 peptide were dropped.
 Intensity normalisation was not performed as a systematic bias in intensities is indeed expected for all \textit{E.
@@
 
-One preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of the protein variance was fit to every protein individually.
+Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of the protein variance was fit to every protein individually.
 The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
 
+\section{Results and Discussion}
+
+\subsection{Peptide to spectrum matching step was successful}
+
+The reanalysis of the proteome benchmark started with the preprocessing of the RAW files produced by the mass spectrometer into an open format that can be handled by downstream tools, such as SearchGUI.
+
+SearchGUI and PeptideShaker su
+
+Identifications decrease with the fraction
+Many spectra cannot be matched -> de novo, modification handling, etc
+Several search engines improve identifications
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{match_percent}
+\caption{}
+\label{fig:match_percent}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{pipeline}
+\caption{Pipeline}
+\label{fig:pipeline}
+\end{figure}
+
 The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
-
-\section{Results and Discussion}
 

### 2018-07-07 report.tex
Extended results of chapter 2

@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
+\captionsetup{font={small},skip=0.25\baselineskip}
+\captionsetup[subfigure]{font={footnotesize}, skip=1pt, singlelinecheck=false}
 \graphicspath{{./plots/}{./figures/}}
@@
 
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{pipeline}
+\caption{Pipeline}
+\label{fig:pipeline}
+\end{figure}
+
 Several proteomics pipelines are available on the internet under different licensing conditions.
@@
 coli (strain K12)} proteomes were mixed in 1:1 and 1:3 proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed in 24 fractions.
-This experimental setting thus generated a total of 2*3*24=144 RAW files.
+Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
+This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
 The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).
@@
 
-\section{Results and Discussion}
+\section{Results}
 
-\subsection{Peptide to spectrum matching step was successful}
+\subsection{Peptide to spectrum matching step identified thousands of spectra}
 
-The reanalysis of the proteome benchmark started with the preprocessing of the RAW files produced by the mass spectrometer into an open format that can be handled by downstream tools, such as SearchGUI.
+The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
+PeptideShaker quality control and filtering capabilities (see figures \ref{figure:qc_validation}) carried out the required search results validation.
+As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
+Likewise, the m/z error was found to be closer to 0 on validated
 
-SearchGUI and PeptideShaker su
+ The combination of both programes enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples (see figure \ref{fig:match_percent}).
+However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14.
 
-Identifications decrease with the fraction
-Many spectra cannot be matched -> de novo, modification handling, etc
-Several search engines improve identifications
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{.4\textwidth}
+  \centering
+  \caption*{A}
+  \includegraphics[width=.9\linewidth]{target_vs_decoy}
+\end{subfigure}
+\begin{subfigure}{.4\textwidth}
+  \centering
+    \caption*{B}
+  \includegraphics[width=.9\linewidth]{score_vs_confidence}
+\end{subfigure}
+\bigskip
+
+\begin{subfigure}{.4\textwidth}
+  \centering
+    \caption*{C}
+  \includegraphics[width=.8\linewidth]{qc}
+\end{subfigure}
+\begin{subfigure}{.4\textwidth}
+  \centering
+    \caption*{D}
+  \includegraphics[width=.9\linewidth]{roc}
+\end{subfigure}
+\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
+\textbf{A} Score distribution for matches to the decoy and the target databases.
+\textbf{B} Evolution of the PSM score with confidence.
+The implemented cutoff at FDR of 1 \% is displayed with a red vertical line.
+\textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status.
+\textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search.
+The cutoff is displayed as a red dot.}
+\label{figure:qc_validation}
+\end{figure}
 
@@
 \includegraphics[width=\textwidth]{match_percent}
-\caption{}
+\caption{Percentage of matched spectra in all samples.
+The total number of spectra per sample ranged between 5894 and 20249.
+\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
+\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.}
 \label{fig:match_percent}
@@
 
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Pipeline}
-\label{fig:pipeline}
-\end{figure}
+\subsection{MBR and Apex intensity extraction capacitate for accurate quantification}
 
 The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
+
+\section{Discussion}
+
+\subsection{Improvement of the PSM process}
+
+The low attained matching rate manifests existing room for improvement in the currently available tools CITE THAT.
+Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats CITE THAT.
+Furthermore, \textit{de novo} search engines are available and supported by SearchGUI.
+Their usage could contribute to further improvements in the identification rate.
+
+The most important reason why many spectra remain unidentified is the presence of post-translational modifications (PTMs), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
+New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications CITE ralf gabriels.
+Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications CITE MS2PIP.
+
+Finally, the extremely low matching rates in the latter fractions translates to decreased contributions to the number of identifications.
+This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
 

### 2018-07-07 report.tex

@@
 \captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={footnotesize}, skip=1pt, singlelinecheck=false}
+\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
 \graphicspath{{./plots/}{./figures/}}
@@
 
- The combination of both programes enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples (see figure \ref{fig:match_percent}).
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14.
+The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples .
+However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
 
@@
 
-The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
-
 \section{Discussion}
@@
 This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{match_percent}
+\caption{}
+\label{fig:match_percent}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{pipeline}
+\caption{Pipeline}
+\label{fig:pipeline}
+\end{figure}
+
+The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
 

### 2018-07-08 report.tex
Extended chapter 2

@@
     \vspace*{2.5cm} \\
-    \textbf{\Huge Topics in Mass Sectrometry and Proteomics} \\
+    \textbf{\Huge Development of a label-free quantification proteomics pipeline} \\
     \vspace*{.1cm} \\
@@
 The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF.
+Furthermore, in order to collect consistent \ac{XIC}, a feature extraction step can be  executed to extract the apex intensity of the identified peak clusters \ref{figure:moff}.
+
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{.45\textwidth}
+    \caption*{A}
+\includegraphics[width=0.9\linewidth]{moff_workflow}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+    \caption*{B}
+\includegraphics[width=0.9\linewidth]{apex_intensity}
+\end{subfigure}
+\caption{Illustration of the \ac{MBR} and apex intensity extraction steps.
+\textbf{A} the information gathered from matches in replicate runs is collected for a reanalysis of the spectra.
+\textbf{B} A retention time window can be used to screen the peak clusters and highest \ac{MS1} intensity of a peak cluster (apex intensity).
+Taken from CITE MOFF}
+\label{figure:moff}
+\end{figure}
 
@@
 It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-As a result, peptides from abundant proteins are more likely to be
-identified and in more spectra.
+As a result, peptides from abundant proteins are more likely to be identified and in more spectra.
 Two approaches were hence followed: count the number of peptides identified for a given protein – like in the emPAI index method – or count the number of spectra ascribed to a protein as in the NSAF method CITE COMPOMICS 4.1.
@@
 A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) feature extraction, (IV) protein inference and (V) relative quantification.
+The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
 Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
@@
 Briefly, the \textit{Homo sapiens} and \textit{E.
-coli (strain K12)} proteomes were mixed in 1:1 and 1:3 proportions, with 3 replicates for each combination.
+coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
 Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
 This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
+One file was missing in the repository.
 The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).
@@
 By default, the FDR was set to 1\%.
+PEP and confidence statistics were computed using the PeptideShaker built-in algorithms.
 Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
@@
 
+\subsection{Code implementation}
+
 \section{Results}
@@
 The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figures \ref{figure:qc_validation}) carried out the required search results validation.
+PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
 As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
-Likewise, the m/z error was found to be closer to 0 on validated
-
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples .
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
+Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
+The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
 
@@
 \centering
-\begin{subfigure}{.4\textwidth}
+\begin{subfigure}{.45\textwidth}
   \centering
@@
 \end{subfigure}
-\begin{subfigure}{.4\textwidth}
+\begin{subfigure}{.45\textwidth}
   \centering
@@
 
-\begin{subfigure}{.4\textwidth}
+\begin{subfigure}{.45\textwidth}
   \centering
@@
 
+The PSM confidence is defined as $1-PEP$, where PEP stands for the Posterior Error Probability.
+Also known as local FDR, the PEP is an estimate of the probability of a given PSM of being an incorrect assignment.
+Thus, the confidence is the probability of the PSM being a correct assigment CITE NEVINZKY.
+The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
+
+The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples .
+However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
+
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{match_percent}
+\includegraphics[width=0.8\textwidth]{match_percent}
 \caption{Percentage of matched spectra in all samples.
@@
 \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.}
+\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
+The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
 \label{fig:match_percent}
@@
 
+\newpage
+
 \subsection{MBR and Apex intensity extraction capacitate for accurate quantification}
+
+The match between runs step allows for increased identifications by transferring successful matches between replicate runs.
+The results of this process for the 13th fraction of the L condition is shown in figures \ref{fig:mbr_summary} and \ref{fig:mbr})
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\textwidth]{mbr_summary}
+\caption{Count of identifications on each run segreagated by source.
+More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
+The MBR acts as an imputation step where extra identifications are performed by gathering the information collected from identifications in other samples and reprocessing the spectra files.
+As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
+In this particular case, hundreds of identifications were accomplished}
+
+\label{fig:mbr_summary}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{mbr}
+\caption{Match Between Runs with moFF.
+Every dot represents a peptide shared across 2 runs.
+The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
+The color depicts whether the identification was carried out during the PSM process, or thanks to a cross-identification achieved by the MBR module.}
+\label{fig:mbr}
+\end{figure}
+
+Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
+The extracted apex intensity can be used as a proxy for peptide abundance in the following quantification step.
+
+\begin{figure}[!h]
+\begin{subfigure}{.9\textwidth}
+  \centering
+    \caption*{A}
+\includegraphics[width=.9\linewidth]{peptide_profile}
+\end{subfigure}
+\bigskip
+
+\begin{subfigure}{.9\textwidth}
+  \centering
+    \caption*{B}
+\includegraphics[width=.9\linewidth]{density_ratio}
+\end{subfigure}
+\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
+coli}.
+The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E.
+coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs.
+\textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
+coli} data in condition H was observed, confirming the good performance of the protocol.}
+\label{fig:apex_intensity}
+\end{figure}
+
+\subsection{Quantification}
+
+The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
 
@@
 This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{match_percent}
-\caption{}
-\label{fig:match_percent}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Pipeline}
-\label{fig:pipeline}
-\end{figure}
-
-The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
 
@@
 
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{graph}
+\end{figure}
+
 %As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.

### 2018-07-15 report.tex
Polish chapter 1 and expand chapter 2

@@
     \vspace*{2.5cm} \\
-    \textbf{\Huge Development of a label-free quantification proteomics pipeline} \\
+    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
     \vspace*{.1cm} \\
@@
     \end{adjustwidth}
-    \newpage
     \ClearWallPaper
@@
 
+%CITE 1.6.1 COMPUTATIONAL METHODS.
 It may be useful to divide the existing approaches into two types of paradigms: top-down and bottom-up.
 In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
+In the bottom-up paradigm (see figure \ref{figure:proteomics_overview}, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
 These smaller parts are called peptides.
-\cite{Barsnes2008} CITE 1.6.1 COMPUTATIONAL METHODS.
-Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly mass spectrometry (MS), which performs the data acquisition.
+\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
 The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
 The top-down paradigm will be ignored in the rest of the manuscript.
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{proteomics_skema_book}
+\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
+\label{fig:proteomics_overview}
+\end{figure}
 
@@
 
+\LARGE
+\textbf{MS analysis and data generation}
+\normalsize
+
 \section{Sample processing}
@@
 
-An \ac{MS} experiment starts with the generation of protein mixture samples.
-They are first separated in order to sort the proteins via physicochemical criteria.
-This is most frequently carried out via SDS-PAGE based on mass or isoelectric point.
-Once the electrophoresis is completed, protein bands can be excised from the gel.
-Each band will contain a subset, or even only one of the proteins originally available, thus making the downstream analysis simpler \cite{Barsnes2008} REFERENCE COMPUTATIONAL METHODS CHAPTER 2
-
-The reduced complexity protein mix is extracted in a denaturalized state so as to remove biases due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, that can cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task, which cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008} COMPUTATIONAL METHODS 3.2
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{sample_processing}
+\caption{Overview of the sample processing step prior to mass spectrometry analysis.
+First, proteins are denatured and digested with a specific protease like Trypsin.
+This yields a peptide mix that is separated into peptide fractions that can be introduced in the mass spectrometer.}
+\label{fig:sample_processing}
+\end{figure}
+
+\subsection{Protein digestion}
+\label{subsec:protein_digestion}
+
+An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
+Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
+Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
+Trypsin is the most frequently used for this task.
+It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
+Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
+As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
+Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
+All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases CITE.
+
+Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
+%COMPUTATIONAL METHODS 3.2
 
@@
 
-\item Missed cleavages \label{item:missed_cleavages}
+\item Missed cleavages %\label{item:missed_cleavages}
 
@@
 
-\Cref{item:missed_cleavages} can happen due to steric inaccesability or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
+%\Cref{item:missed_cleavages}
+
+Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
 This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-This variability, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
-The other
-
-The result of this process is a mix of peptides following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-For Trypsin, the average peptide length is 10 residues, as roughly 1/10 residues are either R or K.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer, thus optimizing the throughput of the method.
-An overview over mass spectrometers follows.
+Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
+
+The result of this process is a complex mix of peptides, made up by hundreds or thousands of different molecules, following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
+A peptide separation step is required before introducing the sample in the spectrometer.
+
+\subsection{Peptide separation}
+\label{subsec:peptide_separation}
+
+If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
+This can be surmounted by independently analyzing different fractions of the sample over time.
+The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
+
+%CITE 4.2 computational methods.
+During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
+These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
+The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
+Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
 
@@
 
-The mass spectrometer consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
+The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
+It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
 
@@
 \includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark}
+\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
 \label{fig:mass_spectrometer}
@@
 \footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
+\stepcounter{footnote}
 
@@
 Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-This step is performed in the ion source \cite{Barsnes2008} CITE 5.1 COMPUTATIONAL METHODS.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (MALDI) and Electro Spray Ionization (ESI) CITE.
+
+%CITE 5.1 COMPUTATIONAL METHODS
+
+This step is performed in the ion source \cite{Barsnes2008}.
+The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
 Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
@@
 Bias in ESI is less predictable.
-This is known as the competitive ionization problem.
-REF ALL THIS
-
-The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that can be applied in the downstream component separation and measurement steps.
+This is known as the competitive ionization problem \cite{Tang2004}.
+
+The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
 
@@
 They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability.
+Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
 
@@
 On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion SEE FIGURE, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
+They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
 Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
@@
 
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{envelope}
+\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
+Each peak in the envelope is separated by 0.5 Th.
+This is explained by the peptide having 2 positive charges that make every extra Da in the ion mass account for 1/2 extra Th.
+Adapted from \cite{Mirzaei2016}.}
+\label{fig:envelope}
+\end{figure}
+
 The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
@@
 
-\section{Mass spectrometry workflows}
-\label{sec:mass_spectrometry_workflows}
-
-The \ac{MS} workflow diverges based on the simplicity of the original protein sample.
-When it consisted of a single protein, Peptide Mass Fingerprinting (PMF) is used, otherwise tandem MS (MS/MS) shall be performed.
-
-\subsection{Peptide Mass Fingerprint (PMF)}
-\label{subsec:pmf}
-
-If the original sample was known to contain a single protein, PMF, or \textit{protein-centric} proteomics, is conducted.
-In PMF, the mixture of peptides can be already transferred to the spectrometer, where a spectrum containing a peak for every m/z ratio present in the ionized peptide mix will be recorded.
-The resulting spectra can be considered a pattern, or fingerprint, of the peptides making up the original protein, thus enabling the identification of the original protein.
-
-\subsection{High pressure liquid chromatography-Tandem MS}
-\label{subsec:ms2}
-
-\subsubsection{High pressure liquid chromatography (HPLC)}
-
-If presented with the problem of analyzing a mixture of proteins, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein  of the sample.
-This can be surmounted by splitting the initial sample into fractions, and using a series, or tandem, of spectrometers in the analysis.
-The spectrometers are used to analyze each obtained fraction separately, using different schemes.
-
-Fractionation is achieved by different methods of separation CITE 1.7, most commonly via \ac{HPLC} methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008} CITE 4.2 computational methods.
-
-\ac{HPLC} methods work by loading the peptide mix in a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
-Notably, the same peptide could elute in several contiguous fractions.
-
-\subsubsection{First MS analysis}
-
-The tandem MS \ac{MS/MS}) analysis starts at the first \ac{MS} scan (\ac{MS1}), when the peptide mix accesses the analyzer.
-It is ionized in the ion source and enters the first mass spectrometer.
-While different ways of handling the peptides are available, we will focus on the product ion scan method.
-In this protocol, the first analyzer is used to select ionized peptides within a narrow m/z window.
-
-\subsubsection{Fragmentation}
-\label{subsubsec:fragmentation}
-
-As the protein sample complexity increases, the \ac{MS1} information on the peptide fractions will not be enough to correctly map each to the original protein, thus rendering a correct interpretation of the resulting spectra impossible.
-Therefore, more information needs to be extracted from them.
-
-Learning the peptide sequences would provide an additional feature that would make the peptide to protein mapping possible.
-This can be achieved by peptide fragmentation, which triggers the breakage of bonds along the peptidic chain, turning the peptides into smaller fragments.
-A given bond will be more likely to break the less stable it is.
-This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the more likely to break (see figure \ref{fig:abcxyz}).
-A nomenclature REF was introduced to name these fragments:
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{abcxyz}
+\section{Tandem MS workflow}
+\label{sec:tandem_ms_workflow}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{tandem_ms}
+\caption[]{Illustration of the tandem MS workflow.
+The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
+The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
+Finally the third spectrometer records fragment mass spectra.
+Taken from \footnotemark{}.}
+\label{fig:tandem_ms}
+\end{figure}
+
+\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
+\stepcounter{footnote}
+
+Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
+In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
+An extra spectrometer, usually a quadrupole, is introduced in between.
+
+\begin{itemize}
+
+\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
+
+\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
+
+\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
+
+\end{itemize}
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{shotgun_adapted}
+\caption{Illustration of the different spectra collected in tandem MS.
+MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
+MS2 spectra record the same magnitudes but the signal is generated by the fragments produced during fragmentation by the ion filtered in the first spectrometer.
+Altogether, they enable peptide identifications.
+Figure adapted from \cite{Verheggen2017}.}
+\label{fig:shotgun}
+\end{figure}
+
+\subsection{Fragmentation}
+\label{subsec:fragmentation}
+
+The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
+Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{abcxyz}
 \caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
 The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
-The a/x, b/y and c/z series are produced by the breakage of bonds I, II and III respectively.
 Specific fragmentation techniques make fragments belonging to one series more likely than others.
 Other fragments are possible but much less likely.
-Taken from CITE COMPOMICS TUTORIAL}
+The fragment nomenclature was introduced in \cite{Roepstorff1984}.}
 \label{fig:abcxyz}
@@
 
-Provided enough mass resolution, the mass difference between contiguous fragments of the same series i.e a1 and a2, can be related to the mass of the extra residue in the longer fragment.
-If this comparison process is repeated for enough fragment pairs of the same peptide, a sequence can be read, finally providing sufficient information to perform an accurate peptide to protein mapping.
-
-Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
+Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
+During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
+These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
+The difference can be exploited to infer which aminoacid is present in the longer fragment.
+If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
+explained in \ref{sec:search_engines}.
+
+%A given bond will be more likely to break the less stable it is.
+This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the most likely to break (see figure \ref{fig:abcxyz}).
+A nomenclature REF was introduced to name these fragments:
+
+%%PAG 123 COMPUTATIONAL METHODS
+%Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
 CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
 Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-\cite{Barsnes2008} PAG 123 COMPUTATIONAL METHODS.
+\cite{Barsnes2008}.
 Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008} CITE pag 134 computational methods.
-
-EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
+This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008}.
+%
+%%CITE pag 134 computational methods
+%
+%EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
 Thus, the fragmentation process is not ergodic and returns TYPE ions.
@@
 
-\subsubsection{Second MS analysis}
-
-The produced fragments enter the second analyzer \ac{MS2}, where a m/z spectrum of the fragments is recorded.
-Thus, unlike in PMF, where the spectrum recorded reflects the m/z ratios acquired by the protein peptides cleaved by the enzyme, tandem MS spectra on product ion scan mode record the m/z ratio of the fragments produced by an ionized peptide with a given m/z ratio.
-The m/z ratio of this precursor ion is changed during the run, thus, multiple spectra are obtained where PMF would create only one.
-The \ac{MS2} spectra will encode the sequence of each peptide as mentioned in \ref{subsubsec:fragmentation}, and with the right tools, it can be deconvoluted and interpreted as explained in \ref{sec:search_engines}.
+\LARGE
+\textbf{Computational analysis}
+\normalsize
 
@@
 
-\ac{MS} search engines are capable of performing the peptide to spectrum matching (PSM).
-They do it by building statistical models that find the peptides within a protein database that best explain the observed spectra.
-\textit{De novo} search engines exist that don't require a reference proteome, but better results are obtained when using one..DEVELOP THAT.
-
-Matching is possible by predicting the cleavage pattern of each protein sequence in the database \textit{in silico} based on the cleavage pattern of the enzyme used, coupled with the simulation of the expected spectrum based on the predicted peptides.
+Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
+\ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
+During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{computational_analysis}
+\caption{Diagram of the MS computational analysis.
+The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
+In other words, the software infers what peptides were introduced in the spectrometer.
+The analysis continues during protein inference and quantification.}
+\label{fig:computational_analysis}
+\end{figure}
+
+%\begin{figure}[!h]
+%\includegraphics[width=\textwidth]{PSM}
+%\caption{}
+%\label{fig:PSM}
+%\end{figure}
 
@@
 This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems measuring the goodness of the match.
+Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
+The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
 Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
@@
 Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda CITE THE ENGINES.
-Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time REF THIS!!
+The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
+Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
 
@@
 
-The scoring system implemented by search engines provides the best matches, but they are bound to contain false identifications.
+The scoring systems implemented in search engines provide the best matches, but they are bound to contain false identifications.
 Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
 
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating tha the application of this filter one out of a hundred filtered matches are expected to be false positives.
-
-The most commonly used method to compute the FDR of a list of matches is the target-decoy search.
+A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
+
+The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
 Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-This decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
-This guarantess that while the sequences in the decoy database won't overlap with those in the target, their basic properties of the decoy (size, composition, etc) will remain identical to the target.
-
-All matches to the decoy are by definition wrong. Since database properties 
-are kept the same, the amount of matches to the decoy exhibiting less than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the equivalent list of target results.
-This is because the existence of shared properties entails that random matches are equally likely to happen in both databases CITE COMPOMICS TUTORIAL 1.5.
-Together with the number of PSMs passing a threshold score ($n_{tp} + n_{tp}$), the FDR can be computed using the formula below.
+The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
+
+All matches to the decoy are by definition wrong.
+Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
+This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
+Together with the number of PSMs passing a given score in the target ($n_{tp} + n_{tp}$), the FDR can be computed using equation \ref{eq:fdr}.
 
@@
 
-Equation \ref{eq:fdr} can be used to select the number of positives (the denominator of the fraction) that make the FDR equal to a predefined value, frequently 0.01 or 1 \%.
-
-%The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value
-%
+The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
+Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
+
+The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s .
+PEP is an estimate of the probability of a given PSM of being an incorrect assignment.
+Thus, the confidence is the probability of the PSM being a correct assigment \cite{Nesvizhskii2010}.
+
 %\begin{equation}\label{eq:qvalue}
-%q(PSM_i) = \min FDR \forall PSM_i \in Positive
+%q(PSM_i) = \min FDR \forall PSM_i \in ositive
 %\end{equation}
+
+The PSM confidence is defined as $1-PEP$, where PEP stands for the Posterior Error Probability. Also known as local FDR,
 
@@
 
-Peptide inference is performed during the PSM process.
+The result of the PSM step returns the inferred list of peptides.
 The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
@@
 
+%CITE PAGE 237 COMPUTATIONAL METHODS
+
 Two paradigms exist in protein quantification: label-based and label-free.
 In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008} CITE PAGE 237 COMPUTATIONAL METHODS The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
-Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) CITE or Isotope-Coded Affinity Tags (ICAT) CITE, differ in the nature of the label and the way it is introduced.
+All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}.
+The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
+Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
 Remarkably, peptide labeling costs can be quite high.
@@
 In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
-In order to account for inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing can be carried out for pair-wise transfer of peptide identities from runs where identification was successful to those were it wasn't due to the randomness in the identification process.
-The transfer can be made based on shared precursor mass and corrected retention times CITE MOFF.
-Furthermore, in order to collect consistent \ac{XIC}, a feature extraction step can be  executed to extract the apex intensity of the identified peak clusters \ref{figure:moff}.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.45\textwidth}
-    \caption*{A}
-\includegraphics[width=0.9\linewidth]{moff_workflow}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-    \caption*{B}
-\includegraphics[width=0.9\linewidth]{apex_intensity}
-\end{subfigure}
-\caption{Illustration of the \ac{MBR} and apex intensity extraction steps.
-\textbf{A} the information gathered from matches in replicate runs is collected for a reanalysis of the spectra.
-\textbf{B} A retention time window can be used to screen the peak clusters and highest \ac{MS1} intensity of a peak cluster (apex intensity).
-Taken from CITE MOFF}
-\label{figure:moff}
-\end{figure}
+In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
+
+\begin{figure}[!h]
+\centering
+%\begin{subfigure}{.45\textwidth}
+%    \caption*{A}
+\includegraphics[width=0.9\linewidth]{mbr_workflow}
+\caption{Illustration of the \ac{MBR} step.
+\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
+Modified from supplementary information in \cite{Argentini2016}.}
+\label{figure:moff_mbr}
+\end{figure}
+
+This way, extra identifications are attained through a reanalysis of the spectra, factoring in the information collected in replicate runs.
+The RT and precursor mass of unidentified spectra in one replicate is matched to that of identified spectra in the remaining replicates.
+As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
 
@@
 Quantification can be spectral counting or \ac{XIC} based.
-Spectrum counting based quantification is the simplest quantification method in proteomics.
+
+On the one hand, spectrum counting based quantification is the simplest method in proteomics.
 It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-As a result, peptides from abundant proteins are more likely to be identified and in more spectra.
-Two approaches were hence followed: count the number of peptides identified for a given protein – like in the emPAI index method – or count the number of spectra ascribed to a protein as in the NSAF method CITE COMPOMICS 4.1.
-These methods have the advantage that they are very simple to implement and don't require any further data processing.
-
-\ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance CITE.
+These methods have the advantage that they are very simple to implement and don't require any further data processing
+
+On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance.
 A wide range of algorithms are available to process these data and output estimates of protein abundance.
@@
 
-Although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra.
-\cite{Cox2014} For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
+As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
+These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
+For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
+
+For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
+
+\begin{figure}[H]
+\centering
+\begin{subfigure}{1\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=0.7\linewidth]{apex_3d}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{1\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=0.7\linewidth]{apex_intensity}
+\end{subfigure}
+\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
+An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
+Taken from \cite{Smith2014}.
+\textbf{B} A refinement analysis of these data attempts to fit a mathematical model of the signal over time and extract a representative measurement, such as the highest (apex) \ac{MS1} intensity.
+Modified from supplementary information in \cite{Argentini2016}.}
+\label{figure:moff_apex}
+\end{figure}
+
+\newpage
 
@@
 
-As stated in \cite{Goeminne2015}.
-\textit{Peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
+As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
 However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
@@
 This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-%CITE MAXQUANT
-%
-% The MaxLFQ algorithm, provided within the MaxQuant suite, is an \ac{MS1}, summarization method performing two optimisation steps to
+This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
 
@@
 \section*{Summary}
+
 A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
@@
 Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) or openMS  CITE ALL are nevertheless available for the community.
-MaxQuant, a free but closed-source proteomics analysis suite CITE MAXQUANT, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
+MaxQuant, a freeware but closed-source proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
 
@@
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded CITE MAIN MSQROB, and (III) protein groups with only 1 peptide were dropped.
+In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (III) protein groups with only 1 peptide were dropped.
 Intensity normalisation was not performed as a systematic bias in intensities is indeed expected for all \textit{E.
@@
 
-\subsection{Peptide to spectrum matching step identified thousands of spectra}
+\subsection{Thousands of spectra were identified and validated}
 
@@
 
-The PSM confidence is defined as $1-PEP$, where PEP stands for the Posterior Error Probability.
-Also known as local FDR, the PEP is an estimate of the probability of a given PSM of being an incorrect assignment.
-Thus, the confidence is the probability of the PSM being a correct assigment CITE NEVINZKY.
  The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
@@
 
-The match between runs step allows for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figures \ref{fig:mbr_summary} and \ref{fig:mbr})
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.5\textwidth]{mbr_summary}
+The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
+The results of this process for the 13th fraction of the L condition is shown in figures \ref{fig:mbr_summary} and \ref{fig:mbr}.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.7\textwidth]{mbr_summary}
 \caption{Count of identifications on each run segreagated by source.
 More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-The MBR acts as an imputation step where extra identifications are performed by gathering the information collected from identifications in other samples and reprocessing the spectra files.
-As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
-In this particular case, hundreds of identifications were accomplished}
+In this particular case, hundreds of new identifications were accomplished.}
 
@@
 The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
-The color depicts whether the identification was carried out during the PSM process, or thanks to a cross-identification achieved by the MBR module.}
+The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
 \label{fig:mbr}
@@
 \textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
-coli} data in condition H was observed, confirming the good performance of the protocol.}
+coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
 \label{fig:apex_intensity}
@@
 New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications CITE ralf gabriels.
-Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications CITE MS2PIP.
+Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
 
@@
 
-\chapter{Bayesian modelling of fold change estimates}
+\chapter{BayesQuant: A Bayesian framework for the computation of pair-wise fold change estimates}
 \label{chap:model}

### 2018-07-15 report.tex
Almost finished chapter 2

@@
 \documentclass[11pt, a4paper]{report}
+% FOR COLORFUL TABLES
+\usepackage{booktabs}
+\usepackage[table]{xcolor}
 \usepackage[english]{babel}
@@
 \usepackage{pdfpages}
+\usepackage{array}
 
@@
 \usepackage{mhchem} % for isotope nomeclature support
+
 % Setup captions
@@
 %CITE 1.6.1 COMPUTATIONAL METHODS.
-It may be useful to divide the existing approaches into two types of paradigms: top-down and bottom-up.
+The existing approaches are divided into two types of paradigms: top-down and bottom-up.
 In the top-down paradigm, intact proteins are directly used for the analysis.
@@
 
-\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:mass_spectrometry_workflows}.
+\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:tandem_ms_workflow}.
 
@@
 Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases CITE.
+All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
 
@@
 
-\newpage
-
 \subsection{\ac{XIC}-peptide-based models for label-free quantification}
@@
 This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) or openMS  CITE ALL are nevertheless available for the community.
+Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
 MaxQuant, a freeware but closed-source proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
@@
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (III) protein groups with only 1 peptide were dropped.
-Intensity normalisation was not performed as a systematic bias in intensities is indeed expected for all \textit{E.
-coli} peptides from samples where the \textit{E.
-coli}proteome was 3 times more abundant.
-
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of the protein variance was fit to every protein individually.
+In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.
+
+Once preprocessing is done, the ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
+The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
 The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
+Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
 
@@
     \caption*{D}
-  \includegraphics[width=.9\linewidth]{roc}
+  \includegraphics[width=.8\linewidth]{roc}
 \end{subfigure}
@@
 
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples .
+The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples.
 However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
@@
 
-\newpage
+\subsection{Protein inference}
+
+\begin{table}[H]
+\begin{tabular}{ccccc}
+  \toprule
+ Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
+  \midrule
+ 46595 & 27384 & 19211 & 25056 & 10039 \\
+ \bottomrule
+\end{tabular}
+\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker is displayed.}
+\label{tab:quantification_table}
+\end{table}
+
+A total of 46595 peptides were inferred to be present in at least one of the fractions.
+Of them more than 27 thousand mapped to a unique protein.
+However, a significant amount mapped to more than one protein.
+Such peptides are called non-unique peptides.
+The 25056 detected proteins were grouped into 10039 protein groups.
+Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
+Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
+Protein grouping decreased the amount of  considered protein entities to 10039.
+
+Protein groups are very frequent due to several factors.
+For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
+PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
+
+However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.
 
@@
 
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.7\textwidth]{mbr_summary}
-\caption{Count of identifications on each run segreagated by source.
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{mbr_combined}
+\caption{Match Between Runs with moFF.
+\textbf{A} Count of identifications on each run segreagated by source.
 More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.}
-
-\label{fig:mbr_summary}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{mbr}
-\caption{Match Between Runs with moFF.
+In this particular case, hundreds of new identifications were accomplished.
+\textbf{B} Match visualization.
 Every dot represents a peptide shared across 2 runs.
@@
 Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
-The extracted apex intensity can be used as a proxy for peptide abundance in the following quantification step.
 
@@
 
-The result of this step was a table where the $log_2(FC)$ estimate, together with test statistics are returned for every protein group.
+The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L.
+During quantification, it is important to take into account as many of the effects influencing protein quantification.
+In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
+The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
+Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
+
+The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).
+
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Sun Jul 15 18:17:23 2018
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll>{\itshape}l}
+  \toprule
+ & log2FC & qval & Protein & Organism \\ 
+  \midrule
+1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
+   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
+  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
+   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
+  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
+   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
+  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
+   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
+  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
+   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
+   \bottomrule
+\end{tabular}
+\caption{Results of the quantification pipeline.
+The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
+Notably, eight correspond to proteins belonging to \textit{E.coli}.
+The two human proteins exhibited in both cases an absolute value of the log2FC below 1.}
+\label{tab:quantification_table}
+\end{table}
+
+The performance of the quantification can be evaluated thanks to the experimental design of the dataset.
+Since the \textit{E.
+coli} proteome was mixed on a 3:1 ratio in condition H, a log2FC of $log2(3)=1.58$ is expected for proteins coming from this organism.
+Likewise, human proteins should have a log2FC of 0.
+As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
+
+Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
+Together, they show the False Positive Rate (FPR), the False Negative Rate (FNR) and the precision exhibited by the algorithm.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{curves_plot}
+\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
+\textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
+\textbf{MQ+MaxLFQ} consists of the standard MaxQuant analysis combined with its LFQ quantification engine.
+Finally, MQ+MSqRob blends the MaxQuant analysis with the MSqRob quantification engine.
+\textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
+\textbf{C, D} PR curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.}
+\label{fig:roc_curves}
+\end{figure}
+
+In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
+Thus, \textit{E.
+coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
+Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
+The result of this analysis is shown in figure \ref{fig:roc_curves}).
+Remarkably, the pipeline presented in this manuscript (Compomics+MSqRob) achieved similar results to those resulting from using MaxQuant and MaxLFQ (MQ+MaxLFQ).
+Nonetheless, the latter performed clearly better when the log2FC filter was applied.
+In all cases, the combination of MaxQuant and MSqRob (Mq+MSqRob) was as good as MQ+MaxLFQ, if not much better.
+
+Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\textwidth]{combined_plot}
+\caption{}
+\label{fig:combined_plot}
+\end{figure}
 
@@
 
-\subsection{Improvement of the PSM process}
-
-The low attained matching rate manifests existing room for improvement in the currently available tools CITE THAT.
-Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats CITE THAT.
+\subsection{Improvement of the PSM step}
+
+The low attained matching rate manifests existing room for improvement in the currently available tools.
+Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
 Furthermore, \textit{de novo} search engines are available and supported by SearchGUI.
@@
 This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
+
+TALK ABOUT FRACTIONS PROBLEM
+HOW PEPTIDES SHOULD APPEAR IN AT MOST 3 FRACTIONS AND ONLY SIGNIFICANTLY A LOT IN THE CENTRAL ONE
+
+\subsection{Improvement of the quantification step}
+
+The deployed relative quantification approach successfully manages to estimate the log2FC or thousands of proteins in the dataset.
+However, other free tools are available to perform the same type of analysis.
 

### 2018-07-16 report.tex
Finish chapter 2. Code still missing

@@
 \usepackage{pdfpages}
-\usepackage{array}
+\usepackage{array} % to make table column italic
+\usepackage{textcomp} % to support the trademark symbol
 
@@
   long  = High pressure liquid chromatography,
+  class = abbrev
+}
+
+\DeclareAcronym{PTM}{
+  short = PTM ,
+  long  = Post-Translational Modification,
   class = abbrev
@@
 %%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 \tableofcontents % Indholdsfortegnelse
@@
 \section*{Preface}
+
+TODO: get nomenclature straight: what is a sample and what is a fraction
 
@@
 A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem, explained in section \label{sec:quantification} CITE.
+This is the quantification problem, explained in section \ref{sec:quantification}.
 
@@
 
-\item Missed cleavages %\label{item:missed_cleavages}
+\item Missed cleavages.
 
@@
 The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on.
+Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
 
@@
 
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s .
-PEP is an estimate of the probability of a given PSM of being an incorrect assignment.
-Thus, the confidence is the probability of the PSM being a correct assigment \cite{Nesvizhskii2010}.
-
-%\begin{equation}\label{eq:qvalue}
-%q(PSM_i) = \min FDR \forall PSM_i \in ositive
-%\end{equation}
-
-The PSM confidence is defined as $1-PEP$, where PEP stands for the Posterior Error Probability. Also known as local FDR,
+The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
+Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment.
+The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
+PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.
 
@@
 \item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and could map to more than one protein.
+This is not trivial as peptides are degenerate and frequently map to more than one protein.
 \end{enumerate}
@@
 
-\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based.
+\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based, explained in \ref{subsec:peptide_model}.
 \end{itemize}
@@
 
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob CITE THEM, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
+A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
 The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
@@
 A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxLFQ \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
+The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
 
@@
 \includegraphics[width=\textwidth]{pipeline}
-\caption{Pipeline}
+\caption{Schema of the presented pipeline.
+First, .RAW files produced by the spectrometer are converted to an open format like MGF.
+After that, spectra are searched by means of a search engine to perform the PSM step.
+Thereafter, PSMs are validated and the most likely set proteins is inferred.
+If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
+The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
 \label{fig:pipeline}
@@
 Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
-MaxQuant, a freeware but closed-source proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and a comprehensive pipeline.
+MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
 
@@
 
-The development of a pipeline achieving these goals will be described in this chapter.
+The development of a pipeline attempting to achieve these goals will be described in this chapter.
 
@@
 One file was missing in the repository.
-The ThermoRawFileParser CITE THERMORAWFILEPARSER program was used to convert RAW files to the open format MGF (Mascot Generic Format).
+The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the open format MGF (Mascot Generic Format).
+
+The MQ+MaxLFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
+The results of the MQ+MSqRob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
+
+%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.
 
@@
 
-Once preprocessing is done, the ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
+Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
 The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
@@
 Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
+
+Quantification in the MQ+MaxLFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
+Intensities were log2 transformed and averaged.
+The difference between conditions H and L was taken as estimate of the log2FC.
+Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
+P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
 
@@
     \caption*{C}
-  \includegraphics[width=.8\linewidth]{qc}
+  \includegraphics[width=.95\linewidth]{qc}
 \end{subfigure}
@@
     \caption*{D}
-  \includegraphics[width=.8\linewidth]{roc}
+  \includegraphics[width=.95\linewidth]{roc}
 \end{subfigure}
@@
 \end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker is displayed.}
+\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
 \label{tab:quantification_table}
@@
 
-\subsection{MBR and Apex intensity extraction capacitate for accurate quantification}
+\subsection{MBR and apex intensity extraction quantification}
 
 The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figures \ref{fig:mbr_summary} and \ref{fig:mbr}.
+The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
 
@@
 \subsection{Quantification}
+\label{subsec:quantification}
 
@@
 Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
-Together, they show the False Positive Rate (FPR), the False Negative Rate (FNR) and the precision exhibited by the algorithm.
-
-\begin{figure}[H]
+Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
+
+\begin{figure}[!h]
 \centering
@@
 \textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
-\textbf{MQ+MaxLFQ} consists of the standard MaxQuant analysis combined with its LFQ quantification engine.
-Finally, MQ+MSqRob blends the MaxQuant analysis with the MSqRob quantification engine.
+\textbf{MQ+MaxLFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
+Finally, MQ+MSqRob blends MaxQuant upstream analysis with the MSqRob quantification engine.
 \textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
@@
 Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
+In order to compare the results of the here presented pipeline (Compomics+MSqRob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
+The pipeline combining MaxQuant and MaxLFQ (MQ+MaxLFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+MSqRob) used MaxQuant up to the fraction normalization step.
+
 The result of this analysis is shown in figure \ref{fig:roc_curves}).
-Remarkably, the pipeline presented in this manuscript (Compomics+MSqRob) achieved similar results to those resulting from using MaxQuant and MaxLFQ (MQ+MaxLFQ).
-Nonetheless, the latter performed clearly better when the log2FC filter was applied.
-In all cases, the combination of MaxQuant and MSqRob (Mq+MSqRob) was as good as MQ+MaxLFQ, if not much better.
+Remarkably, Compomics+MSqRob achieved similar results to those resulting from MQ+MaxLFQ.
+Nonetheless, the latter performed clearly better when the log2FC filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
+In all cases MQ+MSqRob was as good as MQ+MaxLFQ, if not much better.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.8\textwidth]{vennDiagram}
+\caption{Agreement between the different pipelines. The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline.}
+\label{fig:venn}
+\end{figure}
+
+In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn}).
+Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
+The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Compomics+MSqRob or MQ+MaxLFQ being detectd by MQ+MSqRob too.
+Remarkably, 517 proteins were detected by MQ+MSqRob alone, which further confirmed its resolving power.
 
 Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-
-\begin{figure}[H]
+This analysis confirmed that the pipeline providing the best separation between organisms was MQ+MSqRob.
+
+\begin{figure}[!h]
 \centering
 \includegraphics[width=\textwidth]{combined_plot}
-\caption{}
+\caption{Overview over the different pipelines performance.
+Quantification results are segregated by the source organism.
+\textbf{A} Histogram of the log2FC estimates.
+\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its log2FC on x and the minus logarithm of the q-value on y.
+\textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
 \label{fig:combined_plot}
@@
 
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (PTMs), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
-New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications CITE ralf gabriels.
+The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
+New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
 Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
 
-Finally, the extremely low matching rates in the latter fractions translates to decreased contributions to the number of identifications.
+Finally, the extremely low matching rates in the latter fractions translated to decreased contributions to the number of identifications.
 This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
 
-TALK ABOUT FRACTIONS PROBLEM
-HOW PEPTIDES SHOULD APPEAR IN AT MOST 3 FRACTIONS AND ONLY SIGNIFICANTLY A LOT IN THE CENTRAL ONE
+\subsection{Improvement of the feature extraction step}
+
+While the apex intensity is a good estimate of the peptide abundance, other methods are available that could potentially provide more robust data, such as the area of the peak.
+This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
 
@@
 
-The deployed relative quantification approach successfully manages to estimate the log2FC or thousands of proteins in the dataset.
-However, other free tools are available to perform the same type of analysis.
+The deployed relative quantification approach successfully manages to estimate the log2FC of several of proteins in the dataset with high confidence.
+However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
+Several axes of improvement are clear from the results.
+
+\subsubsection{Management of sample fractionation}
+
+Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
+However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
+
+The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
+One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
+A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
+
+MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
+As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
+This is clearly manifested in figure \ref{fig:combined_plot} as the bins at log2FC of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
+While no \textit{E.
+coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Compomics+MSqRob, proving room for improvement in this aspect.
+
+In any case, the way sample fractionation is handled should depend upon the fractionation method.
+For example, membrane fractionation can be exploited when working with membrane proteins \cite{Marmagne2006}.
+
+To sum up, the presently implemented fraction handling is correct and provides good results, but fine-tuning of the pipeline at this step is expected to improve the results of the quantification.
+Thanks to the open-source nature of the software, this is indeed possible.
+
+%which is defined according to the following expressions:
+%
+%\begin{equation}
+%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{A,j} \times \text{XIC}_{P,A,j}}
+%\end{equation}
+%
+%where $\text{N}_{A,j}$ stands for the normalization coefficient for sample $A$ and fraction $j$, and $\text{XIC}_{P,A,j}$ stands for the extracted ion current in the same sample and fraction for peptide P.
+The sum runs over all k fractions.
+%
+%\begin{equation}
+%H_P(N) = \sum\limits_{A,B} \abs{\log(\frac{I_{P,A}(N)}{I_{P,B}(N)})}^2
+%\end{equation}
+%
+%\begin{equation}
+%H(N) = \sum\limits_{p} H_p(N) 
+%\end{equation}
+%
+%where the pair $A$ and $B$ is any of the sample pairs where peptide $P$ is detected. The overall proteome variation is
+
+\subsubsection{The pitfalls of mass spectrometry}
+
+The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
+One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
+Notwithstanding, it reveals the immature state of the technology.
+Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
+Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
+Indeed, the input file for the MQ+MaxLFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn}).
+
+\subsubsection{Support for absolute quantification}
+
+While relative quantification like that performed by MSqRob is enough to infer differential abundance of proteins across conditions, absolute quantification would provide an estimate of protein quantities that would support many other analyses, and facilitate comparison between datasets, in a way similar to what the normalised read counts in FPKM (Fragments Per Kilobase per Million Reads) does in transcriptomics.
+On the other hand, MaxLFQ provides absolute estimates, albeit less robust.
+
+\subsubsection{Quantification of uncertainty}
+
+All the quantification approaches enunciated until now make use of different frequentist approaches to the problem, that evaluate how significantly different the data are from what would be expected under a null hypothesis of a null ratio across the pair of conditions.
+Yet, it does not provide probabilistic interpretations of the uncertainty behind the estimate.
+The development of a Bayesian framework which endeavors at solving this issue is presented in chapter \ref{chap:model}.
 
 \section{Conclusion}
+
+An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
+The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
+Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
+Likewise, the openness of the data in all steps allows for quality control checks along the way.
+However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
+
+\chapter{BayesQuant: A Bayesian framework for the computation of pair-wise fold change estimates}
+\label{chap:model}
+
+\section*{Summary}
+
+The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the fold change across conditions.
+However, a Bayesian based approach to this problem is lacking in the literature.
+As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
+The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{graph}
+\end{figure}
+
+%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
 
@@
 
-\chapter{BayesQuant: A Bayesian framework for the computation of pair-wise fold change estimates}
-\label{chap:model}
-
-\section*{Summary}
-
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the fold change across conditions.
-However, a Bayesian based approach to this problem is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\end{figure}
-
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
 \chapter*{Conclusion}

### 2018-07-16 report.tex
Add pep figure and polish discussion on chap 2

@@
 \def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
+\makeatother
+
+\usepackage{setspace}
+\doublespacing
+
+%\pagestyle{plain} % default for report
+
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
+\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
 \makeatother
@@
 
+% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
+\let\oldthebibliography\thebibliography
+\let\endoldthebibliography\endthebibliography
+\renewenvironment{thebibliography}[1]{
+  \begin{oldthebibliography}{#1}
+    \setlength{\itemsep}{0em}
+    \setlength{\parskip}{0em}
+}
+{
+  \end{oldthebibliography}
+}
+
 \section*{Preface}
@@
 
+\begin{figure}[!h]
+\centering
+%\begin{subfigure}{.45\textwidth}
+%    \caption*{A}
+\includegraphics[width=0.9\linewidth]{pep}
+\caption{Visualizing FDR and PEP.
+The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
+The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
+Figure from \cite{Kall2008}.}
+\label{figure:pep}
+\end{figure}
+
 The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
-Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment.
+Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
 The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
@@
 Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI.
-Their usage could contribute to further improvements in the identification rate.
+Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
+Their usage is predicted tor improve identification rates and quality of the matches.
 
@@
 
-\chapter{BayesQuant: A Bayesian framework for the computation of pair-wise fold change estimates}
+\chapter{BayesQuant: Probabilistic estimation of protein ratios}
 \label{chap:model}
@@
 
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the fold change across conditions.
-However, a Bayesian based approach to this problem is lacking in the literature.
+The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
+However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
 As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
 The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
+
+\section{Introduction}
+
+Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
+Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
+The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (NHST).
+
+\begin{align}
+H_0: & \log_2(FC) = 0 \nonumber \\
+H_1: & \log_2(FC) \neq 0 \nonumber
+\end{align}
+
+The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
+This way, both tools returned point estimates of the $\log_2$(FC) between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
+The null hypothesis usually states that the true value of the parameter is 0.
+Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
+
+The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
+While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
+Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
+This is the so called Bayes-theorem.
 

### 2018-07-16 report.tex
Extend chapter 3 and polish 2

@@
   short = HPLC ,
-  long  = High pressure liquid chromatography,
+  long  = High Pressure Liquid Chromatography,
   class = abbrev
@@
 
+\DeclareAcronym{log2FC}{
+  short = log2FC ,
+  long  = $\log_2$(FC),
+  class = abbrev
+}
+
 \DeclareAcronym{MBR}{
@@
 
+\DeclareAcronym{MGF}{
+  short = MGF ,
+  long  = Mascot Generic Format,
+  class = abbrev
+}
+
 \DeclareAcronym{XIC}{
@@
   long  = Trans Proteomic Pipeline,
+  class = abbrev
+}
+
+\DeclareAcronym{NHST}{
+  short = NHST ,
+  long  = Null-Hypothesis Significance Testing,
   class = abbrev
@@
 
-\chapter{Mass spectrometry and shotgun proteomics overview}
+\chapter{Review on mass spectrometry and shotgun proteomics}
 \label{chap:mass_spec}
@@
 In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm (see figure \ref{figure:proteomics_overview}, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
+In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
 These smaller parts are called peptides.
@@
 Figure from \cite{Kall2008}.}
-\label{figure:pep}
+\label{fig:pep}
 \end{figure}
@@
 \centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
 \includegraphics[width=0.9\linewidth]{mbr_workflow}
@@
 
-Quantification can be spectral counting or \ac{XIC} based.
+Quantification can be \ac{SC} or \ac{XIC} based.
 
 On the one hand, spectrum counting based quantification is the simplest method in proteomics.
+The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance.
 It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing
+These methods have the advantage that they are very simple to implement and don't require any further data processing.
 
@@
 
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
+As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
 These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
@@
 
-\begin{figure}[H]
+\begin{figure}[!h]
 \centering
@@
 
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
+A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016} , was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
+The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) \ac{MBR} and feature extraction, (IV) protein inference and (V) relative quantification.
 Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
@@
 
+\subsection{Background}
+
 \begin{figure}[!h]
@@
 \caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like MGF.
+First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
 After that, spectra are searched by means of a search engine to perform the PSM step.
@@
 Several proteomics pipelines are available on the internet under different licensing conditions.
-Many are released as closed-source software, where information on how the program works is kept from the user.
+Many are released as closed-source commercial software, where information on how the program works is kept from the user.
 This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
+Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability.
 MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
 
-However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform.
-As a GUI would not be required to run the pipeline, it would also be scalable to big datasets.
-Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
-
-The development of a pipeline attempting to achieve these goals will be described in this chapter.
+\subsection{Goals}
+
+The development of a pipeline attempting to achieve the following goals will be described in this chapter.
+
+\begin{enumerate}
+
+\item Fully command line, documented and Linux-supported interface for easy automation and scalability.
+\item Open-source for easy customisation and innovation of analyses.
+\item Free-licensed and cost-free, so anybody with the knowledge can run it, democratizing the analyses.
+\end{enumerate}
 
@@
 One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the open format MGF (Mascot Generic Format).
-
-The MQ+MaxLFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
-The results of the MQ+MSqRob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
+The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.
+
+The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
+The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
 
@@
 
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine CITE by means of the SearchGUI \texttt{SearchCLI} tool CITE utility.
+The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine \cite{Kim2014} by means of the SearchGUI \texttt{SearchCLI} tool \cite{Barsnes2018} utility.
 The search parameters were set using the \texttt{IdentificationParametersCLI}.
@@
 
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
 By default, the FDR was set to 1\%.
@@
 
-The moFF command line utility CITE was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
+The moFF command line utility \cite{Argentini2016} was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
 This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
@@
 
-Quantification in the MQ+MaxLFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
+Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
 Intensities were log2 transformed and averaged.
@@
   \toprule
- & log2FC & qval & Protein & Organism \\ 
+ & \ac{log2FC} & qval & Protein & Organism \\ 
   \midrule
@@
 Notably, eight correspond to proteins belonging to \textit{E.coli}.
-The two human proteins exhibited in both cases an absolute value of the log2FC below 1.}
+The two human proteins exhibited in both cases an absolute value of the \ac{log2FC} below 1.}
 \label{tab:quantification_table}
@@
 Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a log2FC of $log2(3)=1.58$ is expected for proteins coming from this organism.
-Likewise, human proteins should have a log2FC of 0.
+coli} proteome was mixed on a 3:1 ratio in condition H, a \ac{log2FC} of $log2(3)=1.58$ is expected for proteins coming from this organism.
+Likewise, human proteins should have a \ac{log2FC} of 0.
 As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{curves_plot}
+\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
+\textbf{Comp+Rob} is the pipeline presented in the previous sections.
+\textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
+Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine.
+\textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.
+\textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
+\label{fig:roc_curves}
+\end{figure}
 
@@
 Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
-\textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
-\textbf{MQ+MaxLFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+MSqRob blends MaxQuant upstream analysis with the MSqRob quantification engine.
-\textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
-\textbf{C, D} PR curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.}
-\label{fig:roc_curves}
-\end{figure}
 
@@
 coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
-Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
-In order to compare the results of the here presented pipeline (Compomics+MSqRob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+MaxLFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+MSqRob) used MaxQuant up to the fraction normalization step.
-
-The result of this analysis is shown in figure \ref{fig:roc_curves}).
-Remarkably, Compomics+MSqRob achieved similar results to those resulting from MQ+MaxLFQ.
-Nonetheless, the latter performed clearly better when the log2FC filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+MSqRob was as good as MQ+MaxLFQ, if not much better.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{vennDiagram}
-\caption{Agreement between the different pipelines. The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline.}
-\label{fig:venn}
-\end{figure}
-
-In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn}).
+Moreover, a filter based on the \ac{log2FC} can be applied to enrich the data in positives and facilitate the classification task.
+In order to compare the results of the here presented pipeline (Comp+Rob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
+The pipeline combining MaxQuant and MaxLFQ (MQ+LFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+Rob) used MaxQuant up to the fraction normalization step.
+
+The result of this analysis is shown in figure \ref{fig:roc_curves}.
+Remarkably, Comp+Rob achieved similar results to those resulting from MQ+LFQ.
+Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
+In all cases MQ+Rob was as good as MQ+LFQ, if not much better.
+
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{.40\textwidth}
+\caption*{A}
+\includegraphics[width=.9\textwidth]{vennDiagram}
+\end{subfigure}
+\begin{subfigure}{.52\textwidth}
+\caption*{B}
+\includegraphics[width=.95\textwidth]{correlation_plot}
+\end{subfigure}
+\caption{Agreement between the different pipelines.
+\textbf{A} The Venn diagram displays the number of \textit{E.
+coli} proteins found to be differentially abundant by each pipeline.
+\textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
+\label{fig:venn_cor}
+\end{figure}
+
+In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A).
 Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Compomics+MSqRob or MQ+MaxLFQ being detectd by MQ+MSqRob too.
-Remarkably, 517 proteins were detected by MQ+MSqRob alone, which further confirmed its resolving power.
-
-Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-This analysis confirmed that the pipeline providing the best separation between organisms was MQ+MSqRob.
+The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too.
+Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power.
+Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected.
+Even though the observed correlation was low, the estimated log2FC was always greater than 1.
+
+Additionally, the accuracy with which each program assigns a \ac{log2FC} estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
+This analysis confirmed that the pipeline providing the best separation between organisms was MQ+Rob.
 
@@
 Quantification results are segregated by the source organism.
-\textbf{A} Histogram of the log2FC estimates.
-\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its log2FC on x and the minus logarithm of the q-value on y.
+\textbf{A} Histogram of the \ac{log2FC} estimates.
+\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its \ac{log2FC} on x and the minus logarithm of the q-value on y.
 \textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
@@
 
-The deployed relative quantification approach successfully manages to estimate the log2FC of several of proteins in the dataset with high confidence.
+The deployed relative quantification approach successfully manages to estimate the \ac{log2FC} of several of proteins in the dataset with high confidence.
 However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
@@
 As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at log2FC of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
+This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
 While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Compomics+MSqRob, proving room for improvement in this aspect.
+coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.
 
@@
 Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
-Indeed, the input file for the MQ+MaxLFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn}).
+Indeed, the input file for the MQ+LFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn_cor} A).
 
@@
 
+\subsection{Bayesian statistics}
+
 Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
 Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
-The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (NHST).
+The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (\ac{NHST}).
 
@@
 The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the $\log_2$(FC) between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
+This way, both tools returned point estimates of the log2FC between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
 The null hypothesis usually states that the true value of the parameter is 0.
@@
 This is the so called Bayes-theorem.
+
+\begin{equation}
+P(\theta | y) = \cfrac{P(y | \theta) P(\theta)}{P(y)}
+\end{equation}
+
+The Bayes theorem is an extremely versatile tool taking a role analogous to that of statistics like T, F, or $\chi^2$.
+Unlike them, which are tailored to specific scenarios, it can be used compute the probability distribution of almost any parameter in any model.
+4 terms can be distinguished in its expression
+
+\begin{enumerate}
+
+\item $P(\theta)$: the \textbf{prior} probability distribution of the model parameter, introduces previous knowledge about the phenomenon being modelled.
+
+\item $P(y | \theta)$: the probability distribution of the observed data (y) over the possible parameter space.
+It is also known as \textbf{likelihood} of the model, and its duty is updating our beliefs about the phenomenon by capturing the information in the data.
+
+\item $P(y)$: the probability of the data, defined as the marginal probability of the data given a parameter value, for all possible values.
+It acts as a normalizing constant that makes the resulting distribution a true probability distribution adding up to 1.
+It is also known as the data \textbf{evidence}.
+
+\item $P(\theta | y)$: the updated probability distribution of the model parameter, with the information extracted from the observed data.
+Since it reflects the beliefs about the phenomena after observing data, it is called \textbf{posterior} probability distribution,as opposite to the prior.
+
+\end{enumerate}
+
+The Bayesian framework enunciated above can be applied to the quantification problem.
+All is needed is the computation of the posterior probability distribution.
+Unfortunately, this is tractable analytically for simple models only.
+However, Markov Chain-Monte Carlo methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough.
+Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible.
+Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3), all making use of a programming paradigm called \textbf{probabilistic programming}.
+
+\subsection{Probabilistic programming}
 

### 2018-07-17 report.tex
Backup

@@
 \usepackage{textcomp} % to support the trademark symbol
+\usepackage{minted}
 
@@
 
+\DeclareAcronym{MCMC}{
+  short = MCMC,
+  long  = Markov Chain Monte Carlo,
+  class = abbrev
+}
+
 \DeclareAcronym{PTM}{
@@
 
+\DeclareAcronym{DAG}{
+  short = DAG ,
+  long  = Directed Acyclic Graph,
+  class = abbrev
+}
+
 \DeclareAcronym{NZ}{
@@
   long  = Match Between Runs,
+  class = abbrev
+}
+
+\DeclareAcronym{ADVI}{
+  short = ADVI ,
+  long  = Automatic Differentiation Variational Inference,
+  class = abbrev
+}
+
+\DeclareAcronym{NUTS}{
+  short = NUTS ,
+  long  = No-U-Turn-Sampler,
   class = abbrev
@@
 \subsection{Decoy database preparation and search}
+\label{subsec:database_preparation}
 
@@
 \caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
-\label{tab:quantification_table}
+\label{tab:protein_inference}
 \end{table}
 
-A total of 46595 peptides were inferred to be present in at least one of the fractions.
+A total of 46595 peptides were inferred to be present in at least one of the fractions (see table \ref{tab:protein_inference}).
 Of them more than 27 thousand mapped to a unique protein.
@@
 
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L.
+The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L.
 During quantification, it is important to take into account as many of the effects influencing protein quantification.
@@
 Even though the observed correlation was low, the estimated log2FC was always greater than 1.
+The MQ+Rob pipeline systematically estimated log2FV values greater than the other 2.
 
@@
 The low attained matching rate manifests existing room for improvement in the currently available tools.
-Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
-Their usage is predicted tor improve identification rates and quality of the matches.
+As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
+Only MS-GF+ was used in thi work for simplicity.
+Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
+Their usage is predicted to improve identification rates and quality of the matches.
 
@@
 
+Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.
+
 \chapter{BayesQuant: Probabilistic estimation of protein ratios}
@@
 
-\subsection{Bayesian statistics}
+\subsection{Frequentist and Bayesian statistics}
 
@@
 Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes-theorem.
+This is the so called Bayes\textquotesingle s theorem.
 
@@
 
-The Bayesian framework enunciated above can be applied to the quantification problem.
+The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterir probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}
+
+It can be applied to the quantification problem, where $\theta$ turns into the log2FC parameter we try to estimate.
 All is needed is the computation of the posterior probability distribution.
 Unfortunately, this is tractable analytically for simple models only.
-However, Markov Chain-Monte Carlo methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough.
+However, Markov Chain-Monte Carlo (\ac{MCMC}) methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough.
 Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible.
-Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3), all making use of a programming paradigm called \textbf{probabilistic programming}.
+Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3 \cite{Salvatier2016}), all making use of a programming paradigm called \textbf{probabilistic programming}.
 
@@
 
+The management of uncertainty in statistics is done by means of probability distributions, which account for the possible values that a parameter could take, together with how likely they are.
+Probabilistic programming offers the framework to build complex statistical models by storing probability distributions as variables of the program.
+Moreover, probabilistic programming packages supply the tools needed to perform inference with these models by taking experimental evidence (data) and fitting the distributions to the data using Bayes theorem.
+The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using \ac{MCMC} algorithms provided with the probabilistic programming tool.
+
+Pymc3 is a mature Python module dedicated to support probabilistic programming.
+It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (\ac{ADVI}) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
+Therefore, it provides the tools required to build a Bayesian model for the probabilisitc estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
+
+\subsection{Goals}
+
+\begin{enumerate}
+
+\item Develop a Bayesian model to estimate log2FC in proteomics datasets together with the uncertainty of the estimate.
+
+\item Make it scalable and fast for usability in ordinary modern computers.
+
+\end{enumerate}
+
+\section{Materials and Methods}
+
+\subsection{Data input}
+The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant} and \texttt{MSnSet2protdata} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
+
+\subsection{Sequence modelling}
+
+When running with active sequence modelling, the peptide sequences and their neighborhood in the protein was extracted from \textit{E.
+coli} and \textit{Homo sapiens} proteome databases (\ref{subsec:database_preparation}).
+The protein neighborhood was defined by a window spanning 15 aminoacids on both sides of the peptide.
+It was extracted using the seqinr \cite{Charif2007} and GenomicRanges \cite{Lawrence2013} packages implemented in R and Bioconductor.
+
+Features to model the peptide effect based on sequence were extracted using BLABLA
+
+\subsection{Bayesian modelling using Pymc3}
+
+The model was initialized
+
+\begin{minted}[mathescape,
+               linenos,
+               xleftmargin=-70pt,
+               numbersep=10pt,
+               gobble=0,
+               frame=lines,
+               framesep=2mm]{python}
+with pm.Model() as model:
+\end{minted}
+
 \begin{figure}[!h]
@@
 \includegraphics[width=\textwidth]{graph}
-\end{figure}
+\caption{Diagram of the Bayesian model developed in the present paper, without sequence modelling of the peptide effect.
+It is represented as a Directed Acyclic Graph \ac{DAG} where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
+The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to.
+The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.}
+\label{fig:daft_model}
+\end{figure}
+
+\section{Results}
+
+\subsection{Model implementation (I) peptide as a random process}
+
+\subsection{Model implementation (II) peptide effect with sequence features}
+
+\subsection{Posterior probability distribution computation}
+
+\ac{ADVI} returns good results faster than \ac{NUTS}
+
+\subsection{Model checking}
+
+\section{Discussion}
+
+\section{Conclusion}
 

### 2018-07-17 introduction.tex
Backup

--- 2018-07-16 report.tex
@@
-\documentclass[11pt, a4paper]{report}
-% FOR COLORFUL TABLES
-\usepackage{booktabs}
-\usepackage[table]{xcolor}
-\usepackage[english]{babel}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{pdfpages}
-\usepackage{array} % to make table column italic
-\usepackage{textcomp} % to support the trademark symbol
-
-\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
-\usepackage{wallpaper}
-\usepackage{palatino}
-\usepackage{changepage}
-\usepackage{mhchem} % for isotope nomeclature support
-
-% Setup captions
-%\captionstyle[\centering]{\centering}
-%\changecaptionwidth
-%\captionwidth{0.8\linewidth}
-
-% Protect against widows and orphans
-%\clubpenalty=10000
-%\widowpenalty=10000
-
-%\linespread{1.2}
-
-%\raggedbottom
-
-%\chapterstyle{ger}
-
-%\maxsecnumdepth{subsection}
-
-%%  Setup fancy style quotation
-%%  ==================================================================
-%\usepackage{tikz}
-%\usepackage{framed}
-
-%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font
-
-% Make commands for the quotes
-%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
-%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
-%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
-%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}
-
-% select a colour for the shading
-%\definecolor{shadecolor}{rgb}{1,1,1}
-
-% wrap everything in its own environment
-%\newenvironment{shadequote}% 
-%{\begin{snugshade}\begin{quote}\openquote}
-%{\hfill\closequote\end{quote}\end{snugshade}}
-
-%% MY STUFF
-
-%% Language and font encodings
-
-%% Sets page size and margins
-%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
-\graphicspath{{./plots/}{./figures/}}
-
-%\fancypagestyle{title}{%
-%  \setlength{\headheight}{120pt}%
-%  \fancyhf{}% No header/footer
-%  \renewcommand{\headrulewidth}{0pt}% No header rule
-%  \renewcommand{\footrulewidth}{0pt}% No footer rule
-%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
-%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
-%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
-%}%
-
-%\usepackage{amsmath}
-%\usepackage[format=plain]{caption}
-\usepackage{subcaption}
-\usepackage{float} % force float
-\usepackage[nottoc]{tocbibind}
-%\usepackage[margin=-50pt]{caption}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{cleveref} % for enumerate/itemize item referencing
-
-%\setlength{\parindent}{4em}
-%\setlength{\parskip}{1em}
-%\usepackage{mathpazo} % Use the Palatino font
-\usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{xspace}
-\usepackage{wrapfig}
-\usepackage[export]{adjustbox}
-\usepackage{textcomp}
-\usepackage[bottom]{footmisc}
-
-\usepackage{titling}
-\newcommand{\subtitle}[1]{%
-  \posttitle{%
-    \par\end{center}
-    \begin{center}\large#1\end{center}
-    \vskip0.5em}%
-}
-
-\renewcommand\thesubfigure{\arabic{subfigure}}  
-\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
-%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}
-
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily,
-  showstringspaces=false,
-  commentstyle=\color{red},
-  keywordstyle=\color{blue}
-}
-
-\makeatletter
-\renewcommand\p@subfigure{\thefigure\,}
-\renewcommand\thesubfigure{\alph{subfigure})}
-% If desired for table as well:
-% \renewcommand\p@subtable{\thetable\,}
-% \renewcommand\thesubtable{\alph{subtable})}
-\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
-\makeatother
-
-\renewcommand{\baselinestretch}{1.5}
-\DeclareCaptionSubType*[arabic]{figure}
-
-%\makeatletter
-%\renewcommand\thesection{}
-%\renewcommand\thesubsection{\@arabic\c@subsection}
-%\makeatother
-
-\usepackage{mathtools}
-\usepackage{sectsty}
-
-\chapternumberfont{\LARGE} 
-\chaptertitlefont{\LARGE}
-
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
-
-% Swap the definition of \abs* and \norm*, so that \abs
-% and \norm resizes the size of the brackets, and the 
-% starred version does not.
-\makeatletter
-\let\oldabs\abs
-\def\abs{\@ifstar{\oldabs}{\oldabs*}}
-%
-\let\oldnorm\norm
-\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
-
-%\pagestyle{plain} % default for report
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
-\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
-\makeatother
-
-\usepackage{acro}
-\acsetup{first-style=short}
-
-\DeclareAcronym{MS}{
-  short = MS ,
-  long  = Mass Spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{MS1}{
-  short = MS1 ,
-  long  = first tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS2}{
-  short = MS2 ,
-  long  = second tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS/MS}{
-  short = MS/MS ,
-  long  = Tandem mass spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{m/z}{
-  short = m/z ,
-  long  = mass charge ratio,
-  class = abbrev
-}
-
-\DeclareAcronym{MALDI}{
-  short = MALDI ,
-  long  = Matrix-Assisted Laser-Desorption Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{ESI}{
-  short = ESI ,
-  long  = Electrospray Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{TOF}{
-  short = TOF ,
-  long  = Time of Flight,
-  class = abbrev
-}
-
-\DeclareAcronym{IT}{
-  short = IT ,
-  long  = Ion Trap,
-  class = abbrev
-}
-
-\DeclareAcronym{Q}{
-  short = Q,
-  long  = Quadrupole,
-  class = abbrev
-}
-
-\DeclareAcronym{FT-ICR}{
-  short = FT-ICR ,
-  long  = Fourier Transform Ion Cyclotron Resonance,
-  class = abbrev
-}
-
-\DeclareAcronym{HPLC}{
-  short = HPLC ,
-  long  = High pressure liquid chromatography,
-  class = abbrev
-}
-
-\DeclareAcronym{PTM}{
-  short = PTM ,
-  long  = Post-Translational Modification,
-  class = abbrev
-}
-
-\DeclareAcronym{NZ}{
-  short = NZ ,
-  long  = Novozymes A/S,
-  class = abbrev
-}
-
-\DeclareAcronym{FDR}{
-  short = FDR ,
-  long  = False Discovery Rate,
-  class = abbrev
-}
-
-\DeclareAcronym{RT}{
-  short = RT ,
-  long  = Retention time,
-  class = abbrev
-}
-
-\DeclareAcronym{MBR}{
-  short = MBR ,
-  long  = Match Between Runs,
-  class = abbrev
-}
-
-\DeclareAcronym{XIC}{
-  short = XIC ,
-  long  = Extracted Ion Current,
-  class = abbrev
-}
-
-\DeclareAcronym{SC}{
-  short = SC ,
-  long  = Spectral Counting,
-  class = abbrev
-}
-
-\DeclareAcronym{TPP}{
-  short = TPP ,
-  long  = Trans Proteomic Pipeline,
-  class = abbrev
-}
-
-%%  Begin document
-%%  ==================================================================
-\begin{document}
-
-%%  Begin title page
-%%  ==================================================================
-    \thispagestyle{empty}
-    \ULCornerWallPaper{1}{nat-farve.pdf}
-    \ULCornerWallPaper{1}{nat-en.pdf}
-    \begin{adjustwidth}{-3cm}{-1.5cm}
-    \vspace*{-1cm}
-    \textbf{\Huge Bioinformatics Master Thesis} \\
-    \vspace*{2.5cm} \\
-    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
-    \vspace*{.1cm} \\
-    %{\huge No subtitle is available} \\
-    \begin{tabbing}
-    % adjust the hspace below for the longest author name
-    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
-    %Second Author \> \texttt{<second@author.com>} \\       
-    \\[10.5cm]
-
-    \textbf{\Large Supervisors} \\
-    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
-    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\
-
-    \end{tabbing}
-    \end{adjustwidth}
-    \ClearWallPaper
-%%  ==================================================================
-%%  End title page
-
-%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{2}
-\tableofcontents % Indholdsfortegnelse
-%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
-%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\thispagestyle{empty}
-\printacronyms[include-classes=abbrev,name=Abbreviations]
-
-\newpage
-\setcounter{page}{1}
-
-% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
-\let\oldthebibliography\thebibliography
-\let\endoldthebibliography\endthebibliography
-\renewenvironment{thebibliography}[1]{
-  \begin{oldthebibliography}{#1}
-    \setlength{\itemsep}{0em}
-    \setlength{\parskip}{0em}
-}
-{
-  \end{oldthebibliography}
-}
-
-\section*{Preface}
-
-TODO: get nomenclature straight: what is a sample and what is a fraction
-
 \chapter*{Introduction}
@@
 Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
-
-\chapter{Mass spectrometry and shotgun proteomics overview}
-\label{chap:mass_spec}
-
-Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone.
-Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up.
-The layer provided by proteins is one of them, and its study is called proteomics.
-It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment.
-The study of the protein-specific complexity is called proteomics.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
-
-%CITE 1.6.1 COMPUTATIONAL METHODS.
-The existing approaches are divided into two types of paradigms: top-down and bottom-up.
-In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm (see figure \ref{figure:proteomics_overview}, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
-These smaller parts are called peptides.
-\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
-The top-down paradigm will be ignored in the rest of the manuscript.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{proteomics_skema_book}
-\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
-\label{fig:proteomics_overview}
-\end{figure}
-
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
-The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem, explained in section \ref{sec:quantification}.
-
-A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:
-
-\begin{enumerate}
-
-\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:tandem_ms_workflow}.
-
-\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.
-
-\end{enumerate}
-
-\LARGE
-\textbf{MS analysis and data generation}
-\normalsize
-
-\section{Sample processing}
-\label{sec:sample_processing}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{sample_processing}
-\caption{Overview of the sample processing step prior to mass spectrometry analysis.
-First, proteins are denatured and digested with a specific protease like Trypsin.
-This yields a peptide mix that is separated into peptide fractions that can be introduced in the mass spectrometer.}
-\label{fig:sample_processing}
-\end{figure}
-
-\subsection{Protein digestion}
-\label{subsec:protein_digestion}
-
-An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task.
-It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
-
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
-%COMPUTATIONAL METHODS 3.2
-
-\begin{enumerate}
-
-\item Missed cleavages.
-
-\item Unsuspected cleavages during the maturation/life cycle of the protein.
-
-\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
-
-\item Naturally occurring, intentionally or unintentionally induced chemical modifications.
-
-\end{enumerate}
-
-%\Cref{item:missed_cleavages}
-
-Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
-
-The result of this process is a complex mix of peptides, made up by hundreds or thousands of different molecules, following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-A peptide separation step is required before introducing the sample in the spectrometer.
-
-\subsection{Peptide separation}
-\label{subsec:peptide_separation}
-
-If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
-This can be surmounted by independently analyzing different fractions of the sample over time.
-The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
-
-%CITE 4.2 computational methods.
-During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
-
-\section{The mass spectrometer}
-\label{sec:the_mass_spectrometer}
-
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
-It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
-\label{fig:mass_spectrometer}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-\subsection{The ion source}
-\label{subsec:the_ion_source}
-
-All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-
-%CITE 5.1 COMPUTATIONAL METHODS
-
-This step is performed in the ion source \cite{Barsnes2008}.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
-Thus, the charge exhibited by an ion is not obvious when produced via ESI.
-Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
-Bias in ESI is less predictable.
-This is known as the competitive ionization problem \cite{Tang2004}.
-
-The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
-
-\subsection{The mass analyzer}
-\label{subsec:the_mass_analyzer}
-
-The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
-These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.
-
-Moreover, two other analyzers exist which combine mass analysis with intensity measurement.
-These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
-They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
-
-\subsection{The detector}
-\label{subsec:the_detector}
-
-Detectors measure the intensity of an incoming ion signal.
-The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
-Some topics in signal detection in \ac{MS} need to be discussed.
-
-On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
-The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
-This is one of the reasons why proteins are better fit for analysis when digested into peptides, as \ac{m/z} are reduced, thus increasing the mass resolution.
-
-On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
-Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
-Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{envelope}
-\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
-Each peak in the envelope is separated by 0.5 Th.
-This is explained by the peptide having 2 positive charges that make every extra Da in the ion mass account for 1/2 extra Th.
-Adapted from \cite{Mirzaei2016}.}
-\label{fig:envelope}
-\end{figure}
-
-The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
-
-It is up to the \ac{MS} technician to decide on the best pieces of equipment according to their availability and particularities of the dataset.
-
-\section{Tandem MS workflow}
-\label{sec:tandem_ms_workflow}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{tandem_ms}
-\caption[]{Illustration of the tandem MS workflow.
-The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
-The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
-Finally the third spectrometer records fragment mass spectra.
-Taken from \footnotemark{}.}
-\label{fig:tandem_ms}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
-In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
-An extra spectrometer, usually a quadrupole, is introduced in between.
-
-\begin{itemize}
-
-\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
-
-\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
-
-\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
-
-\end{itemize}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{shotgun_adapted}
-\caption{Illustration of the different spectra collected in tandem MS.
-MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
-MS2 spectra record the same magnitudes but the signal is generated by the fragments produced during fragmentation by the ion filtered in the first spectrometer.
-Altogether, they enable peptide identifications.
-Figure adapted from \cite{Verheggen2017}.}
-\label{fig:shotgun}
-\end{figure}
-
-\subsection{Fragmentation}
-\label{subsec:fragmentation}
-
-The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
-Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{abcxyz}
-\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
-The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
-Specific fragmentation techniques make fragments belonging to one series more likely than others.
-Other fragments are possible but much less likely.
-The fragment nomenclature was introduced in \cite{Roepstorff1984}.}
-\label{fig:abcxyz}
-\end{figure}
-
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
-During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
-These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
-The difference can be exploited to infer which aminoacid is present in the longer fragment.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
-explained in \ref{sec:search_engines}.
-
-%A given bond will be more likely to break the less stable it is.
-This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the most likely to break (see figure \ref{fig:abcxyz}).
-A nomenclature REF was introduced to name these fragments:
-
-%%PAG 123 COMPUTATIONAL METHODS
-%Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
-Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-\cite{Barsnes2008}.
-Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008}.
-%
-%%CITE pag 134 computational methods
-%
-%EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
-Thus, the fragmentation process is not ergodic and returns TYPE ions.
-More importantly, chemical modifications are not lost in the process.
-
-\LARGE
-\textbf{Computational analysis}
-\normalsize
-
-\section{Spectra processing: search engines}
-\label{sec:search_engines}
-
-Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
-\ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{computational_analysis}
-\caption{Diagram of the MS computational analysis.
-The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
-In other words, the software infers what peptides were introduced in the spectrometer.
-The analysis continues during protein inference and quantification.}
-\label{fig:computational_analysis}
-\end{figure}
-
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{PSM}
-%\caption{}
-%\label{fig:PSM}
-%\end{figure}
-
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
-This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
-Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
-Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.
-
-Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
-Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
-
-\section{Validation and quality control}
-\label{sec:validation}
-
-The scoring systems implemented in search engines provide the best matches, but they are bound to contain false identifications.
-Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
-
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
-
-The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
-Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
-
-All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
-This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
-Together with the number of PSMs passing a given score in the target ($n_{tp} + n_{tp}$), the FDR can be computed using equation \ref{eq:fdr}.
-
-\begin{equation}\label{eq:fdr}
-FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
-\end{equation}
-
-The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
-Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{pep}
-\caption{Visualizing FDR and PEP.
-The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
-The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
-Figure from \cite{Kall2008}.}
-\label{figure:pep}
-\end{figure}
-
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
-Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
-The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
-PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.
-
-\section{Peptide and protein inference}
-\label{sec:inference}
-
-Two steps in protein identification can be distinguished:
-
-\begin{enumerate}
-
-\item \textbf{Peptide inference}: infer the peptides present in the sample.
-\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and frequently map to more than one protein.
-\end{enumerate}
-
-The result of the PSM step returns the inferred list of peptides.
-The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
-The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
-Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
-
-\section{Protein quantification}
-\label{sec:quantification}
-
-The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
-However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
-For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
-
-Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free).
-These are explained in subsection \ref{subsec:labelling}.
-If the label-free approach is employed, more distinctions can be made based on:
-
-\begin{itemize}
-\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}
-
-\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based, explained in \ref{subsec:peptide_model}.
-\end{itemize}
-
-\subsection{Label-based and label-free approaches}
-\label{subsec:labelling}
-
-%CITE PAGE 237 COMPUTATIONAL METHODS
-
-Two paradigms exist in protein quantification: label-based and label-free.
-In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}.
-The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
-Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
-Remarkably, peptide labeling costs can be quite high.
-This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
-
-In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
-In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption{Illustration of the \ac{MBR} step.
-\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_mbr}
-\end{figure}
-
-This way, extra identifications are attained through a reanalysis of the spectra, factoring in the information collected in replicate runs.
-The RT and precursor mass of unidentified spectra in one replicate is matched to that of identified spectra in the remaining replicates.
-As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
-
-\subsection{SC and XIC based quantification}
-\label{subsec:scvsxic}
-
-Quantification can be spectral counting or \ac{XIC} based.
-
-On the one hand, spectrum counting based quantification is the simplest method in proteomics.
-It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing
-
-On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance.
-A wide range of algorithms are available to process these data and output estimates of protein abundance.
-All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
-They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
-
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
-For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
-
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{A}
-\includegraphics[width=0.7\linewidth]{apex_3d}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{B}
-\includegraphics[width=0.7\linewidth]{apex_intensity}
-\end{subfigure}
-\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
-An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
-Taken from \cite{Smith2014}.
-\textbf{B} A refinement analysis of these data attempts to fit a mathematical model of the signal over time and extract a representative measurement, such as the highest (apex) \ac{MS1} intensity.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_apex}
-\end{figure}
-
-\subsection{\ac{XIC}-peptide-based models for label-free quantification}
-\label{subsec:peptide_model}
-
-The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample.
-However, the data interpretation requires the transfer of these peptide-level data into the protein level.
-This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.
-
-As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
-However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
-This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
-
-It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
-This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
-
-\chapter{A label-free quantification proteomics pipeline}
-\label{chap:pipeline}
-
-\section*{Summary}
-
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
-Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
-A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
-
-\section{Introduction}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like MGF.
-After that, spectra are searched by means of a search engine to perform the PSM step.
-Thereafter, PSMs are validated and the most likely set proteins is inferred.
-If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
-The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
-\label{fig:pipeline}
-\end{figure}
-
-Several proteomics pipelines are available on the internet under different licensing conditions.
-Many are released as closed-source software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
-MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-
-However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform.
-As a GUI would not be required to run the pipeline, it would also be scalable to big datasets.
-Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
-
-The development of a pipeline attempting to achieve these goals will be described in this chapter.
-
-\section{Materials and Methods}
-
-\subsection{Data generation and loading}
-
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-Briefly, the \textit{Homo sapiens} and \textit{E.
-coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
-This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
-One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the open format MGF (Mascot Generic Format).
-
-The MQ+MaxLFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
-The results of the MQ+MSqRob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
-
-%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.
-
-\subsection{Decoy database preparation and search}
-
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine CITE by means of the SearchGUI \texttt{SearchCLI} tool CITE utility.
-The search parameters were set using the \texttt{IdentificationParametersCLI}.
-In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification.
-The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
-Up to two missed cleavages were allowed.
-The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
-
-The target database was created by combining the Uniprot proteomes for \textit{E.
-coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
-
-\subsection{Quality control and validation}
-
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR was set to 1\%.
-PEP and confidence statistics were computed using the PeptideShaker built-in algorithms.
-Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
-
-\subsection{Data refinement}
-
-The moFF command line utility CITE was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
-This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
-
-\subsection{Quantification}
-
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
-Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.
-
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
-The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
-The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
-Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
-
-Quantification in the MQ+MaxLFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
-Intensities were log2 transformed and averaged.
-The difference between conditions H and L was taken as estimate of the log2FC.
-Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
-P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
-
-\subsection{Code implementation}
-
-\section{Results}
-
-\subsection{Thousands of spectra were identified and validated}
-
-The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
-The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \caption*{A}
-  \includegraphics[width=.9\linewidth]{target_vs_decoy}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{B}
-  \includegraphics[width=.9\linewidth]{score_vs_confidence}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{C}
-  \includegraphics[width=.95\linewidth]{qc}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{D}
-  \includegraphics[width=.95\linewidth]{roc}
-\end{subfigure}
-\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
-\textbf{A} Score distribution for matches to the decoy and the target databases.
-\textbf{B} Evolution of the PSM score with confidence.
-The implemented cutoff at FDR of 1 \% is displayed with a red vertical line.
-\textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status.
-\textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search.
-The cutoff is displayed as a red dot.}
-\label{figure:qc_validation}
-\end{figure}
-
- The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
-
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples.
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
-
-\begin{figure}[!h]
-\includegraphics[width=0.8\textwidth]{match_percent}
-\caption{Percentage of matched spectra in all samples.
-The total number of spectra per sample ranged between 5894 and 20249.
-\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
-The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
-\label{fig:match_percent}
-\end{figure}
-
-\subsection{Protein inference}
-
-\begin{table}[H]
-\begin{tabular}{ccccc}
-  \toprule
- Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
-  \midrule
- 46595 & 27384 & 19211 & 25056 & 10039 \\
- \bottomrule
-\end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
-\label{tab:quantification_table}
-\end{table}
-
-A total of 46595 peptides were inferred to be present in at least one of the fractions.
-Of them more than 27 thousand mapped to a unique protein.
-However, a significant amount mapped to more than one protein.
-Such peptides are called non-unique peptides.
-The 25056 detected proteins were grouped into 10039 protein groups.
-Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
-Protein grouping decreased the amount of  considered protein entities to 10039.
-
-Protein groups are very frequent due to several factors.
-For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
-PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
-
-However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.
-
-\subsection{MBR and apex intensity extraction quantification}
-
-The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{mbr_combined}
-\caption{Match Between Runs with moFF.
-\textbf{A} Count of identifications on each run segreagated by source.
-More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.
-\textbf{B} Match visualization.
-Every dot represents a peptide shared across 2 runs.
-The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
-The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
-\label{fig:mbr}
-\end{figure}
-
-Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
-
-\begin{figure}[!h]
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{A}
-\includegraphics[width=.9\linewidth]{peptide_profile}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{B}
-\includegraphics[width=.9\linewidth]{density_ratio}
-\end{subfigure}
-\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
-coli}.
-The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E.
-coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs.
-\textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
-coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
-\label{fig:apex_intensity}
-\end{figure}
-
-\subsection{Quantification}
-\label{subsec:quantification}
-
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L.
-During quantification, it is important to take into account as many of the effects influencing protein quantification.
-In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
-The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
-Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
-
-The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).
-
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Sun Jul 15 18:17:23 2018
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll>{\itshape}l}
-  \toprule
- & log2FC & qval & Protein & Organism \\ 
-  \midrule
-1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
-  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
-   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
-  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
-   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
-  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
-   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
-  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
-   \bottomrule
-\end{tabular}
-\caption{Results of the quantification pipeline.
-The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
-Notably, eight correspond to proteins belonging to \textit{E.coli}.
-The two human proteins exhibited in both cases an absolute value of the log2FC below 1.}
-\label{tab:quantification_table}
-\end{table}
-
-The performance of the quantification can be evaluated thanks to the experimental design of the dataset.
-Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a log2FC of $log2(3)=1.58$ is expected for proteins coming from this organism.
-Likewise, human proteins should have a log2FC of 0.
-As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
-
-Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
-Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
-\textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
-\textbf{MQ+MaxLFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+MSqRob blends MaxQuant upstream analysis with the MSqRob quantification engine.
-\textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
-\textbf{C, D} PR curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.}
-\label{fig:roc_curves}
-\end{figure}
-
-In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
-Thus, \textit{E.
-coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
-Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
-In order to compare the results of the here presented pipeline (Compomics+MSqRob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+MaxLFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+MSqRob) used MaxQuant up to the fraction normalization step.
-
-The result of this analysis is shown in figure \ref{fig:roc_curves}).
-Remarkably, Compomics+MSqRob achieved similar results to those resulting from MQ+MaxLFQ.
-Nonetheless, the latter performed clearly better when the log2FC filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+MSqRob was as good as MQ+MaxLFQ, if not much better.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{vennDiagram}
-\caption{Agreement between the different pipelines. The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline.}
-\label{fig:venn}
-\end{figure}
-
-In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn}).
-Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Compomics+MSqRob or MQ+MaxLFQ being detectd by MQ+MSqRob too.
-Remarkably, 517 proteins were detected by MQ+MSqRob alone, which further confirmed its resolving power.
-
-Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-This analysis confirmed that the pipeline providing the best separation between organisms was MQ+MSqRob.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines performance.
-Quantification results are segregated by the source organism.
-\textbf{A} Histogram of the log2FC estimates.
-\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its log2FC on x and the minus logarithm of the q-value on y.
-\textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
-\label{fig:combined_plot}
-\end{figure}
-
-\section{Discussion}
-
-\subsection{Improvement of the PSM step}
-
-The low attained matching rate manifests existing room for improvement in the currently available tools.
-Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
-Their usage is predicted tor improve identification rates and quality of the matches.
-
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
-New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
-Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
-
-Finally, the extremely low matching rates in the latter fractions translated to decreased contributions to the number of identifications.
-This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
-
-\subsection{Improvement of the feature extraction step}
-
-While the apex intensity is a good estimate of the peptide abundance, other methods are available that could potentially provide more robust data, such as the area of the peak.
-This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
-
-\subsection{Improvement of the quantification step}
-
-The deployed relative quantification approach successfully manages to estimate the log2FC of several of proteins in the dataset with high confidence.
-However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
-Several axes of improvement are clear from the results.
-
-\subsubsection{Management of sample fractionation}
-
-Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
-
-The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
-One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
-A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
-
-MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at log2FC of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
-While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Compomics+MSqRob, proving room for improvement in this aspect.
-
-In any case, the way sample fractionation is handled should depend upon the fractionation method.
-For example, membrane fractionation can be exploited when working with membrane proteins \cite{Marmagne2006}.
-
-To sum up, the presently implemented fraction handling is correct and provides good results, but fine-tuning of the pipeline at this step is expected to improve the results of the quantification.
-Thanks to the open-source nature of the software, this is indeed possible.
-
-%which is defined according to the following expressions:
-%
-%\begin{equation}
-%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{A,j} \times \text{XIC}_{P,A,j}}
-%\end{equation}
-%
-%where $\text{N}_{A,j}$ stands for the normalization coefficient for sample $A$ and fraction $j$, and $\text{XIC}_{P,A,j}$ stands for the extracted ion current in the same sample and fraction for peptide P.
-The sum runs over all k fractions.
-%
-%\begin{equation}
-%H_P(N) = \sum\limits_{A,B} \abs{\log(\frac{I_{P,A}(N)}{I_{P,B}(N)})}^2
-%\end{equation}
-%
-%\begin{equation}
-%H(N) = \sum\limits_{p} H_p(N) 
-%\end{equation}
-%
-%where the pair $A$ and $B$ is any of the sample pairs where peptide $P$ is detected. The overall proteome variation is
-
-\subsubsection{The pitfalls of mass spectrometry}
-
-The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
-One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
-Notwithstanding, it reveals the immature state of the technology.
-Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
-Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
-Indeed, the input file for the MQ+MaxLFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn}).
-
-\subsubsection{Support for absolute quantification}
-
-While relative quantification like that performed by MSqRob is enough to infer differential abundance of proteins across conditions, absolute quantification would provide an estimate of protein quantities that would support many other analyses, and facilitate comparison between datasets, in a way similar to what the normalised read counts in FPKM (Fragments Per Kilobase per Million Reads) does in transcriptomics.
-On the other hand, MaxLFQ provides absolute estimates, albeit less robust.
-
-\subsubsection{Quantification of uncertainty}
-
-All the quantification approaches enunciated until now make use of different frequentist approaches to the problem, that evaluate how significantly different the data are from what would be expected under a null hypothesis of a null ratio across the pair of conditions.
-Yet, it does not provide probabilistic interpretations of the uncertainty behind the estimate.
-The development of a Bayesian framework which endeavors at solving this issue is presented in chapter \ref{chap:model}.
-
-\section{Conclusion}
-
-An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
-The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
-Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
-Likewise, the openness of the data in all steps allows for quality control checks along the way.
-However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
-
-\chapter{BayesQuant: Probabilistic estimation of protein ratios}
-\label{chap:model}
-
-\section*{Summary}
-
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
-However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
-
-\section{Introduction}
-
-Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
-Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
-The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (NHST).
-
-\begin{align}
-H_0: & \log_2(FC) = 0 \nonumber \\
-H_1: & \log_2(FC) \neq 0 \nonumber
-\end{align}
-
-The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the $\log_2$(FC) between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
-The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
-
-The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
-While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes-theorem.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\end{figure}
-
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
-\chapter{Pipeline benchmarking on \ac{NZ} data}
-\label{chap:benchmark}
-
-\section*{Summary}
-
-A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
-The experiment consisted of the application of two different treatments to THP-1 cell cultures.
-One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
-The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
-The results confirmed that the computational analysis successfully captured this response.
-It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
-
-\chapter*{Conclusion}
-\label{chap:conclusion}
-
-\chapter*{Appendix}
-\label{chap:appendix}
-
-\includepdf[pages=-]{appendix.pdf}
-
-\bibliographystyle{unsrt_etal}
-\footnotesize{\bibliography{bibliography}}
-
-\end{document}
-%%  ==================================================================
-%%  End document

### 2018-07-17 modelling.tex
Backup

--- 2018-07-17 report.tex
@@
-\documentclass[11pt, a4paper]{report}
-% FOR COLORFUL TABLES
-\usepackage{booktabs}
-\usepackage[table]{xcolor}
-\usepackage[english]{babel}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{pdfpages}
-\usepackage{array} % to make table column italic
-\usepackage{textcomp} % to support the trademark symbol
-\usepackage{minted}
-
-\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
-\usepackage{wallpaper}
-\usepackage{palatino}
-\usepackage{changepage}
-\usepackage{mhchem} % for isotope nomeclature support
-
-% Setup captions
-%\captionstyle[\centering]{\centering}
-%\changecaptionwidth
-%\captionwidth{0.8\linewidth}
-
-% Protect against widows and orphans
-%\clubpenalty=10000
-%\widowpenalty=10000
-
-%\linespread{1.2}
-
-%\raggedbottom
-
-%\chapterstyle{ger}
-
-%\maxsecnumdepth{subsection}
-
-%%  Setup fancy style quotation
-%%  ==================================================================
-%\usepackage{tikz}
-%\usepackage{framed}
-
-%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font
-
-% Make commands for the quotes
-%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
-%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
-%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
-%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}
-
-% select a colour for the shading
-%\definecolor{shadecolor}{rgb}{1,1,1}
-
-% wrap everything in its own environment
-%\newenvironment{shadequote}% 
-%{\begin{snugshade}\begin{quote}\openquote}
-%{\hfill\closequote\end{quote}\end{snugshade}}
-
-%% MY STUFF
-
-%% Language and font encodings
-
-%% Sets page size and margins
-%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
-\graphicspath{{./plots/}{./figures/}}
-
-%\fancypagestyle{title}{%
-%  \setlength{\headheight}{120pt}%
-%  \fancyhf{}% No header/footer
-%  \renewcommand{\headrulewidth}{0pt}% No header rule
-%  \renewcommand{\footrulewidth}{0pt}% No footer rule
-%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
-%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
-%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
-%}%
-
-%\usepackage{amsmath}
-%\usepackage[format=plain]{caption}
-\usepackage{subcaption}
-\usepackage{float} % force float
-\usepackage[nottoc]{tocbibind}
-%\usepackage[margin=-50pt]{caption}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{cleveref} % for enumerate/itemize item referencing
-
-%\setlength{\parindent}{4em}
-%\setlength{\parskip}{1em}
-%\usepackage{mathpazo} % Use the Palatino font
-\usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{xspace}
-\usepackage{wrapfig}
-\usepackage[export]{adjustbox}
-\usepackage{textcomp}
-\usepackage[bottom]{footmisc}
-
-\usepackage{titling}
-\newcommand{\subtitle}[1]{%
-  \posttitle{%
-    \par\end{center}
-    \begin{center}\large#1\end{center}
-    \vskip0.5em}%
-}
-
-\renewcommand\thesubfigure{\arabic{subfigure}}  
-\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
-%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}
-
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily,
-  showstringspaces=false,
-  commentstyle=\color{red},
-  keywordstyle=\color{blue}
-}
-
-\makeatletter
-\renewcommand\p@subfigure{\thefigure\,}
-\renewcommand\thesubfigure{\alph{subfigure})}
-% If desired for table as well:
-% \renewcommand\p@subtable{\thetable\,}
-% \renewcommand\thesubtable{\alph{subtable})}
-\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
-\makeatother
-
-\renewcommand{\baselinestretch}{1.5}
-\DeclareCaptionSubType*[arabic]{figure}
-
-%\makeatletter
-%\renewcommand\thesection{}
-%\renewcommand\thesubsection{\@arabic\c@subsection}
-%\makeatother
-
-\usepackage{mathtools}
-\usepackage{sectsty}
-
-\chapternumberfont{\LARGE} 
-\chaptertitlefont{\LARGE}
-
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
-
-% Swap the definition of \abs* and \norm*, so that \abs
-% and \norm resizes the size of the brackets, and the 
-% starred version does not.
-\makeatletter
-\let\oldabs\abs
-\def\abs{\@ifstar{\oldabs}{\oldabs*}}
-%
-\let\oldnorm\norm
-\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
-
-%\pagestyle{plain} % default for report
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
-\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
-\makeatother
-
-\usepackage{acro}
-\acsetup{first-style=short}
-
-\DeclareAcronym{MS}{
-  short = MS ,
-  long  = Mass Spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{MS1}{
-  short = MS1 ,
-  long  = first tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS2}{
-  short = MS2 ,
-  long  = second tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS/MS}{
-  short = MS/MS ,
-  long  = Tandem mass spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{m/z}{
-  short = m/z ,
-  long  = mass charge ratio,
-  class = abbrev
-}
-
-\DeclareAcronym{MALDI}{
-  short = MALDI ,
-  long  = Matrix-Assisted Laser-Desorption Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{ESI}{
-  short = ESI ,
-  long  = Electrospray Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{TOF}{
-  short = TOF ,
-  long  = Time of Flight,
-  class = abbrev
-}
-
-\DeclareAcronym{IT}{
-  short = IT ,
-  long  = Ion Trap,
-  class = abbrev
-}
-
-\DeclareAcronym{Q}{
-  short = Q,
-  long  = Quadrupole,
-  class = abbrev
-}
-
-\DeclareAcronym{FT-ICR}{
-  short = FT-ICR ,
-  long  = Fourier Transform Ion Cyclotron Resonance,
-  class = abbrev
-}
-
-\DeclareAcronym{HPLC}{
-  short = HPLC ,
-  long  = High Pressure Liquid Chromatography,
-  class = abbrev
-}
-
-\DeclareAcronym{MCMC}{
-  short = MCMC,
-  long  = Markov Chain Monte Carlo,
-  class = abbrev
-}
-
-\DeclareAcronym{PTM}{
-  short = PTM ,
-  long  = Post-Translational Modification,
-  class = abbrev
-}
-
-\DeclareAcronym{DAG}{
-  short = DAG ,
-  long  = Directed Acyclic Graph,
-  class = abbrev
-}
-
-\DeclareAcronym{NZ}{
-  short = NZ ,
-  long  = Novozymes A/S,
-  class = abbrev
-}
-
-\DeclareAcronym{FDR}{
-  short = FDR ,
-  long  = False Discovery Rate,
-  class = abbrev
-}
-
-\DeclareAcronym{RT}{
-  short = RT ,
-  long  = Retention time,
-  class = abbrev
-}
-
-\DeclareAcronym{log2FC}{
-  short = log2FC ,
-  long  = $\log_2$(FC),
-  class = abbrev
-}
-
-\DeclareAcronym{MBR}{
-  short = MBR ,
-  long  = Match Between Runs,
-  class = abbrev
-}
-
-\DeclareAcronym{ADVI}{
-  short = ADVI ,
-  long  = Automatic Differentiation Variational Inference,
-  class = abbrev
-}
-
-\DeclareAcronym{NUTS}{
-  short = NUTS ,
-  long  = No-U-Turn-Sampler,
-  class = abbrev
-}
-
-\DeclareAcronym{MGF}{
-  short = MGF ,
-  long  = Mascot Generic Format,
-  class = abbrev
-}
-
-\DeclareAcronym{XIC}{
-  short = XIC ,
-  long  = Extracted Ion Current,
-  class = abbrev
-}
-
-\DeclareAcronym{SC}{
-  short = SC ,
-  long  = Spectral Counting,
-  class = abbrev
-}
-
-\DeclareAcronym{TPP}{
-  short = TPP ,
-  long  = Trans Proteomic Pipeline,
-  class = abbrev
-}
-
-\DeclareAcronym{NHST}{
-  short = NHST ,
-  long  = Null-Hypothesis Significance Testing,
-  class = abbrev
-}
-
-%%  Begin document
-%%  ==================================================================
-\begin{document}
-
-%%  Begin title page
-%%  ==================================================================
-    \thispagestyle{empty}
-    \ULCornerWallPaper{1}{nat-farve.pdf}
-    \ULCornerWallPaper{1}{nat-en.pdf}
-    \begin{adjustwidth}{-3cm}{-1.5cm}
-    \vspace*{-1cm}
-    \textbf{\Huge Bioinformatics Master Thesis} \\
-    \vspace*{2.5cm} \\
-    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
-    \vspace*{.1cm} \\
-    %{\huge No subtitle is available} \\
-    \begin{tabbing}
-    % adjust the hspace below for the longest author name
-    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
-    %Second Author \> \texttt{<second@author.com>} \\       
-    \\[10.5cm]
-
-    \textbf{\Large Supervisors} \\
-    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
-    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\
-
-    \end{tabbing}
-    \end{adjustwidth}
-    \ClearWallPaper
-%%  ==================================================================
-%%  End title page
-
-%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{2}
-\tableofcontents % Indholdsfortegnelse
-%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
-%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\thispagestyle{empty}
-\printacronyms[include-classes=abbrev,name=Abbreviations]
-
-\newpage
-\setcounter{page}{1}
-
-% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
-\let\oldthebibliography\thebibliography
-\let\endoldthebibliography\endthebibliography
-\renewenvironment{thebibliography}[1]{
-  \begin{oldthebibliography}{#1}
-    \setlength{\itemsep}{0em}
-    \setlength{\parskip}{0em}
-}
-{
-  \end{oldthebibliography}
-}
-
-\section*{Preface}
-
-TODO: get nomenclature straight: what is a sample and what is a fraction
-
-\chapter*{Introduction}
-\label{chap:introduction}
-
-\section{Aminoacids and proteins}
-
-Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.
-
-Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
-The side chain differs between aminaocids and generates them from each other.
-A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
-  \caption{CAPTION AND REFERENCE}
-  \label{fig:aminoacids}
-\end{figure}
-
-Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
-The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
-
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
-One frequent instance of such modifications is the oxidation of methionine.
-
-\section{The protein-focused biotechnology industry}
-
-Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks.
-The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
-
-\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
-The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
-Notorious examples of such applications include waste-water treatment, household care and the baking industry.
-
-The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
-The refinement of the currently used tools and the development of new ones could be of great significance for the company.
-
-Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
-
-\section{Objectives of the Thesis}
-\label{sec:objectives}
-
-In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
-
-\begin{enumerate}
-
-\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
-
-\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
-
-\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
-
-\end{enumerate}
-
-\section{Structure of the Thesis}
-
-An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
-Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
-
-\chapter{Review on mass spectrometry and shotgun proteomics}
-\label{chap:mass_spec}
-
-Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone.
-Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up.
-The layer provided by proteins is one of them, and its study is called proteomics.
-It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment.
-The study of the protein-specific complexity is called proteomics.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
-
-%CITE 1.6.1 COMPUTATIONAL METHODS.
-The existing approaches are divided into two types of paradigms: top-down and bottom-up.
-In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
-These smaller parts are called peptides.
-\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
-The top-down paradigm will be ignored in the rest of the manuscript.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{proteomics_skema_book}
-\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
-\label{fig:proteomics_overview}
-\end{figure}
-
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
-The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem, explained in section \ref{sec:quantification}.
-
-A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:
-
-\begin{enumerate}
-
-\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:tandem_ms_workflow}.
-
-\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.
-
-\end{enumerate}
-
-\LARGE
-\textbf{MS analysis and data generation}
-\normalsize
-
-\section{Sample processing}
-\label{sec:sample_processing}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{sample_processing}
-\caption{Overview of the sample processing step prior to mass spectrometry analysis.
-First, proteins are denatured and digested with a specific protease like Trypsin.
-This yields a peptide mix that is separated into peptide fractions that can be introduced in the mass spectrometer.}
-\label{fig:sample_processing}
-\end{figure}
-
-\subsection{Protein digestion}
-\label{subsec:protein_digestion}
-
-An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task.
-It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
-
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
-%COMPUTATIONAL METHODS 3.2
-
-\begin{enumerate}
-
-\item Missed cleavages.
-
-\item Unsuspected cleavages during the maturation/life cycle of the protein.
-
-\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
-
-\item Naturally occurring, intentionally or unintentionally induced chemical modifications.
-
-\end{enumerate}
-
-%\Cref{item:missed_cleavages}
-
-Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
-
-The result of this process is a complex mix of peptides, made up by hundreds or thousands of different molecules, following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-A peptide separation step is required before introducing the sample in the spectrometer.
-
-\subsection{Peptide separation}
-\label{subsec:peptide_separation}
-
-If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
-This can be surmounted by independently analyzing different fractions of the sample over time.
-The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
-
-%CITE 4.2 computational methods.
-During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
-
-\section{The mass spectrometer}
-\label{sec:the_mass_spectrometer}
-
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
-It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
-\label{fig:mass_spectrometer}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-\subsection{The ion source}
-\label{subsec:the_ion_source}
-
-All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-
-%CITE 5.1 COMPUTATIONAL METHODS
-
-This step is performed in the ion source \cite{Barsnes2008}.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
-Thus, the charge exhibited by an ion is not obvious when produced via ESI.
-Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
-Bias in ESI is less predictable.
-This is known as the competitive ionization problem \cite{Tang2004}.
-
-The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
-
-\subsection{The mass analyzer}
-\label{subsec:the_mass_analyzer}
-
-The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
-These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.
-
-Moreover, two other analyzers exist which combine mass analysis with intensity measurement.
-These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
-They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
-
-\subsection{The detector}
-\label{subsec:the_detector}
-
-Detectors measure the intensity of an incoming ion signal.
-The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
-Some topics in signal detection in \ac{MS} need to be discussed.
-
-On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
-The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
-This is one of the reasons why proteins are better fit for analysis when digested into peptides, as \ac{m/z} are reduced, thus increasing the mass resolution.
-
-On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
-Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
-Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{envelope}
-\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
-Each peak in the envelope is separated by 0.5 Th.
-This is explained by the peptide having 2 positive charges that make every extra Da in the ion mass account for 1/2 extra Th.
-Adapted from \cite{Mirzaei2016}.}
-\label{fig:envelope}
-\end{figure}
-
-The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
-
-It is up to the \ac{MS} technician to decide on the best pieces of equipment according to their availability and particularities of the dataset.
-
-\section{Tandem MS workflow}
-\label{sec:tandem_ms_workflow}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{tandem_ms}
-\caption[]{Illustration of the tandem MS workflow.
-The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
-The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
-Finally the third spectrometer records fragment mass spectra.
-Taken from \footnotemark{}.}
-\label{fig:tandem_ms}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
-In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
-An extra spectrometer, usually a quadrupole, is introduced in between.
-
-\begin{itemize}
-
-\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
-
-\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
-
-\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
-
-\end{itemize}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{shotgun_adapted}
-\caption{Illustration of the different spectra collected in tandem MS.
-MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
-MS2 spectra record the same magnitudes but the signal is generated by the fragments produced during fragmentation by the ion filtered in the first spectrometer.
-Altogether, they enable peptide identifications.
-Figure adapted from \cite{Verheggen2017}.}
-\label{fig:shotgun}
-\end{figure}
-
-\subsection{Fragmentation}
-\label{subsec:fragmentation}
-
-The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
-Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{abcxyz}
-\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
-The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
-Specific fragmentation techniques make fragments belonging to one series more likely than others.
-Other fragments are possible but much less likely.
-The fragment nomenclature was introduced in \cite{Roepstorff1984}.}
-\label{fig:abcxyz}
-\end{figure}
-
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
-During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
-These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
-The difference can be exploited to infer which aminoacid is present in the longer fragment.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
-explained in \ref{sec:search_engines}.
-
-%A given bond will be more likely to break the less stable it is.
-This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the most likely to break (see figure \ref{fig:abcxyz}).
-A nomenclature REF was introduced to name these fragments:
-
-%%PAG 123 COMPUTATIONAL METHODS
-%Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
-Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-\cite{Barsnes2008}.
-Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008}.
-%
-%%CITE pag 134 computational methods
-%
-%EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
-Thus, the fragmentation process is not ergodic and returns TYPE ions.
-More importantly, chemical modifications are not lost in the process.
-
-\LARGE
-\textbf{Computational analysis}
-\normalsize
-
-\section{Spectra processing: search engines}
-\label{sec:search_engines}
-
-Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
-\ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{computational_analysis}
-\caption{Diagram of the MS computational analysis.
-The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
-In other words, the software infers what peptides were introduced in the spectrometer.
-The analysis continues during protein inference and quantification.}
-\label{fig:computational_analysis}
-\end{figure}
-
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{PSM}
-%\caption{}
-%\label{fig:PSM}
-%\end{figure}
-
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
-This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
-Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
-Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.
-
-Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
-Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
-
-\section{Validation and quality control}
-\label{sec:validation}
-
-The scoring systems implemented in search engines provide the best matches, but they are bound to contain false identifications.
-Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
-
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
-
-The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
-Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
-
-All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
-This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
-Together with the number of PSMs passing a given score in the target ($n_{tp} + n_{tp}$), the FDR can be computed using equation \ref{eq:fdr}.
-
-\begin{equation}\label{eq:fdr}
-FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
-\end{equation}
-
-The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
-Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{pep}
-\caption{Visualizing FDR and PEP.
-The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
-The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
-Figure from \cite{Kall2008}.}
-\label{fig:pep}
-\end{figure}
-
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
-Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
-The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
-PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.
-
-\section{Peptide and protein inference}
-\label{sec:inference}
-
-Two steps in protein identification can be distinguished:
-
-\begin{enumerate}
-
-\item \textbf{Peptide inference}: infer the peptides present in the sample.
-\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and frequently map to more than one protein.
-\end{enumerate}
-
-The result of the PSM step returns the inferred list of peptides.
-The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
-The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
-Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
-
-\section{Protein quantification}
-\label{sec:quantification}
-
-The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
-However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
-For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
-
-Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free).
-These are explained in subsection \ref{subsec:labelling}.
-If the label-free approach is employed, more distinctions can be made based on:
-
-\begin{itemize}
-\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}
-
-\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based, explained in \ref{subsec:peptide_model}.
-\end{itemize}
-
-\subsection{Label-based and label-free approaches}
-\label{subsec:labelling}
-
-%CITE PAGE 237 COMPUTATIONAL METHODS
-
-Two paradigms exist in protein quantification: label-based and label-free.
-In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}.
-The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
-Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
-Remarkably, peptide labeling costs can be quite high.
-This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
-
-In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
-In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption{Illustration of the \ac{MBR} step.
-\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_mbr}
-\end{figure}
-
-This way, extra identifications are attained through a reanalysis of the spectra, factoring in the information collected in replicate runs.
-The RT and precursor mass of unidentified spectra in one replicate is matched to that of identified spectra in the remaining replicates.
-As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
-
-\subsection{SC and XIC based quantification}
-\label{subsec:scvsxic}
-
-Quantification can be \ac{SC} or \ac{XIC} based.
-
-On the one hand, spectrum counting based quantification is the simplest method in proteomics.
-The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance.
-It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing.
-
-On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance.
-A wide range of algorithms are available to process these data and output estimates of protein abundance.
-All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
-They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
-
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
-For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
-
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{A}
-\includegraphics[width=0.7\linewidth]{apex_3d}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{B}
-\includegraphics[width=0.7\linewidth]{apex_intensity}
-\end{subfigure}
-\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
-An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
-Taken from \cite{Smith2014}.
-\textbf{B} A refinement analysis of these data attempts to fit a mathematical model of the signal over time and extract a representative measurement, such as the highest (apex) \ac{MS1} intensity.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_apex}
-\end{figure}
-
-\subsection{\ac{XIC}-peptide-based models for label-free quantification}
-\label{subsec:peptide_model}
-
-The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample.
-However, the data interpretation requires the transfer of these peptide-level data into the protein level.
-This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.
-
-As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
-However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
-This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
-
-It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
-This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
-
-\chapter{A label-free quantification proteomics pipeline}
-\label{chap:pipeline}
-
-\section*{Summary}
-
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016} , was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) \ac{MBR} and feature extraction, (IV) protein inference and (V) relative quantification.
-Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
-A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
-
-\section{Introduction}
-
-\subsection{Background}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
-After that, spectra are searched by means of a search engine to perform the PSM step.
-Thereafter, PSMs are validated and the most likely set proteins is inferred.
-If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
-The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
-\label{fig:pipeline}
-\end{figure}
-
-Several proteomics pipelines are available on the internet under different licensing conditions.
-Many are released as closed-source commercial software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability.
-MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-
-\subsection{Goals}
-
-The development of a pipeline attempting to achieve the following goals will be described in this chapter.
-
-\begin{enumerate}
-
-\item Fully command line, documented and Linux-supported interface for easy automation and scalability.
-\item Open-source for easy customisation and innovation of analyses.
-\item Free-licensed and cost-free, so anybody with the knowledge can run it, democratizing the analyses.
-\end{enumerate}
-
-\section{Materials and Methods}
-
-\subsection{Data generation and loading}
-
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-Briefly, the \textit{Homo sapiens} and \textit{E.
-coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
-This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
-One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.
-
-The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
-The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
-
-%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.
-
-\subsection{Decoy database preparation and search}
-\label{subsec:database_preparation}
-
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine \cite{Kim2014} by means of the SearchGUI \texttt{SearchCLI} tool \cite{Barsnes2018} utility.
-The search parameters were set using the \texttt{IdentificationParametersCLI}.
-In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification.
-The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
-Up to two missed cleavages were allowed.
-The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
-
-The target database was created by combining the Uniprot proteomes for \textit{E.
-coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
-
-\subsection{Quality control and validation}
-
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR was set to 1\%.
-PEP and confidence statistics were computed using the PeptideShaker built-in algorithms.
-Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
-
-\subsection{Data refinement}
-
-The moFF command line utility \cite{Argentini2016} was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
-This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
-
-\subsection{Quantification}
-
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
-Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.
-
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
-The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
-The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
-Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
-
-Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
-Intensities were log2 transformed and averaged.
-The difference between conditions H and L was taken as estimate of the log2FC.
-Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
-P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
-
-\subsection{Code implementation}
-
-\section{Results}
-
-\subsection{Thousands of spectra were identified and validated}
-
-The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
-The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \caption*{A}
-  \includegraphics[width=.9\linewidth]{target_vs_decoy}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{B}
-  \includegraphics[width=.9\linewidth]{score_vs_confidence}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{C}
-  \includegraphics[width=.95\linewidth]{qc}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{D}
-  \includegraphics[width=.95\linewidth]{roc}
-\end{subfigure}
-\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
-\textbf{A} Score distribution for matches to the decoy and the target databases.
-\textbf{B} Evolution of the PSM score with confidence.
-The implemented cutoff at FDR of 1 \% is displayed with a red vertical line.
-\textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status.
-\textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search.
-The cutoff is displayed as a red dot.}
-\label{figure:qc_validation}
-\end{figure}
-
- The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
-
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples.
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
-
-\begin{figure}[!h]
-\includegraphics[width=0.8\textwidth]{match_percent}
-\caption{Percentage of matched spectra in all samples.
-The total number of spectra per sample ranged between 5894 and 20249.
-\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
-The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
-\label{fig:match_percent}
-\end{figure}
-
-\subsection{Protein inference}
-
-\begin{table}[H]
-\begin{tabular}{ccccc}
-  \toprule
- Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
-  \midrule
- 46595 & 27384 & 19211 & 25056 & 10039 \\
- \bottomrule
-\end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
-\label{tab:protein_inference}
-\end{table}
-
-A total of 46595 peptides were inferred to be present in at least one of the fractions (see table \ref{tab:protein_inference}).
-Of them more than 27 thousand mapped to a unique protein.
-However, a significant amount mapped to more than one protein.
-Such peptides are called non-unique peptides.
-The 25056 detected proteins were grouped into 10039 protein groups.
-Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
-Protein grouping decreased the amount of  considered protein entities to 10039.
-
-Protein groups are very frequent due to several factors.
-For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
-PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
-
-However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.
-
-\subsection{MBR and apex intensity extraction quantification}
-
-The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{mbr_combined}
-\caption{Match Between Runs with moFF.
-\textbf{A} Count of identifications on each run segreagated by source.
-More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.
-\textbf{B} Match visualization.
-Every dot represents a peptide shared across 2 runs.
-The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
-The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
-\label{fig:mbr}
-\end{figure}
-
-Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
-
-\begin{figure}[!h]
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{A}
-\includegraphics[width=.9\linewidth]{peptide_profile}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{B}
-\includegraphics[width=.9\linewidth]{density_ratio}
-\end{subfigure}
-\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
-coli}.
-The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E.
-coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs.
-\textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
-coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
-\label{fig:apex_intensity}
-\end{figure}
-
-\subsection{Quantification}
-\label{subsec:quantification}
-
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L.
-During quantification, it is important to take into account as many of the effects influencing protein quantification.
-In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
-The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
-Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
-
-The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).
-
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Sun Jul 15 18:17:23 2018
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll>{\itshape}l}
-  \toprule
- & \ac{log2FC} & qval & Protein & Organism \\ 
-  \midrule
-1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
-  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
-   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
-  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
-   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
-  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
-   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
-  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
-   \bottomrule
-\end{tabular}
-\caption{Results of the quantification pipeline.
-The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
-Notably, eight correspond to proteins belonging to \textit{E.coli}.
-The two human proteins exhibited in both cases an absolute value of the \ac{log2FC} below 1.}
-\label{tab:quantification_table}
-\end{table}
-
-The performance of the quantification can be evaluated thanks to the experimental design of the dataset.
-Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a \ac{log2FC} of $log2(3)=1.58$ is expected for proteins coming from this organism.
-Likewise, human proteins should have a \ac{log2FC} of 0.
-As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
-\textbf{Comp+Rob} is the pipeline presented in the previous sections.
-\textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine.
-\textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.
-\textbf{C, D} PR curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.}
-\label{fig:roc_curves}
-\end{figure}
-
-Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
-Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
-
-In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
-Thus, \textit{E.
-coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
-Moreover, a filter based on the \ac{log2FC} can be applied to enrich the data in positives and facilitate the classification task.
-In order to compare the results of the here presented pipeline (Comp+Rob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+LFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+Rob) used MaxQuant up to the fraction normalization step.
-
-The result of this analysis is shown in figure \ref{fig:roc_curves}.
-Remarkably, Comp+Rob achieved similar results to those resulting from MQ+LFQ.
-Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+Rob was as good as MQ+LFQ, if not much better.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.40\textwidth}
-\caption*{A}
-\includegraphics[width=.9\textwidth]{vennDiagram}
-\end{subfigure}
-\begin{subfigure}{.52\textwidth}
-\caption*{B}
-\includegraphics[width=.95\textwidth]{correlation_plot}
-\end{subfigure}
-\caption{Agreement between the different pipelines.
-\textbf{A} The Venn diagram displays the number of \textit{E.
-coli} proteins found to be differentially abundant by each pipeline.
-\textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
-\label{fig:venn_cor}
-\end{figure}
-
-In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn_cor} A).
-Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too.
-Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power.
-Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected.
-Even though the observed correlation was low, the estimated log2FC was always greater than 1.
-The MQ+Rob pipeline systematically estimated log2FV values greater than the other 2.
-
-Additionally, the accuracy with which each program assigns a \ac{log2FC} estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-This analysis confirmed that the pipeline providing the best separation between organisms was MQ+Rob.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines performance.
-Quantification results are segregated by the source organism.
-\textbf{A} Histogram of the \ac{log2FC} estimates.
-\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its \ac{log2FC} on x and the minus logarithm of the q-value on y.
-\textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
-\label{fig:combined_plot}
-\end{figure}
-
-\section{Discussion}
-
-\subsection{Improvement of the PSM step}
-
-The low attained matching rate manifests existing room for improvement in the currently available tools.
-As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Only MS-GF+ was used in thi work for simplicity.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
-Their usage is predicted to improve identification rates and quality of the matches.
-
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
-New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
-Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
-
-Finally, the extremely low matching rates in the latter fractions translated to decreased contributions to the number of identifications.
-This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
-
-\subsection{Improvement of the feature extraction step}
-
-While the apex intensity is a good estimate of the peptide abundance, other methods are available that could potentially provide more robust data, such as the area of the peak.
-This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
-
-\subsection{Improvement of the quantification step}
-
-The deployed relative quantification approach successfully manages to estimate the \ac{log2FC} of several of proteins in the dataset with high confidence.
-However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
-Several axes of improvement are clear from the results.
-
-\subsubsection{Management of sample fractionation}
-
-Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
-
-The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
-One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
-A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
-
-MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
-While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.
-
-In any case, the way sample fractionation is handled should depend upon the fractionation method.
-For example, membrane fractionation can be exploited when working with membrane proteins \cite{Marmagne2006}.
-
-To sum up, the presently implemented fraction handling is correct and provides good results, but fine-tuning of the pipeline at this step is expected to improve the results of the quantification.
-Thanks to the open-source nature of the software, this is indeed possible.
-
-%which is defined according to the following expressions:
-%
-%\begin{equation}
-%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{A,j} \times \text{XIC}_{P,A,j}}
-%\end{equation}
-%
-%where $\text{N}_{A,j}$ stands for the normalization coefficient for sample $A$ and fraction $j$, and $\text{XIC}_{P,A,j}$ stands for the extracted ion current in the same sample and fraction for peptide P.
-The sum runs over all k fractions.
-%
-%\begin{equation}
-%H_P(N) = \sum\limits_{A,B} \abs{\log(\frac{I_{P,A}(N)}{I_{P,B}(N)})}^2
-%\end{equation}
-%
-%\begin{equation}
-%H(N) = \sum\limits_{p} H_p(N) 
-%\end{equation}
-%
-%where the pair $A$ and $B$ is any of the sample pairs where peptide $P$ is detected. The overall proteome variation is
-
-\subsubsection{The pitfalls of mass spectrometry}
-
-The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
-One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
-Notwithstanding, it reveals the immature state of the technology.
-Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
-Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
-Indeed, the input file for the MQ+LFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn_cor} A).
-
-\subsubsection{Support for absolute quantification}
-
-While relative quantification like that performed by MSqRob is enough to infer differential abundance of proteins across conditions, absolute quantification would provide an estimate of protein quantities that would support many other analyses, and facilitate comparison between datasets, in a way similar to what the normalised read counts in FPKM (Fragments Per Kilobase per Million Reads) does in transcriptomics.
-On the other hand, MaxLFQ provides absolute estimates, albeit less robust.
-
-\subsubsection{Quantification of uncertainty}
-
-All the quantification approaches enunciated until now make use of different frequentist approaches to the problem, that evaluate how significantly different the data are from what would be expected under a null hypothesis of a null ratio across the pair of conditions.
-Yet, it does not provide probabilistic interpretations of the uncertainty behind the estimate.
-The development of a Bayesian framework which endeavors at solving this issue is presented in chapter \ref{chap:model}.
-
-\section{Conclusion}
-
-An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
-The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
-Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
-Likewise, the openness of the data in all steps allows for quality control checks along the way.
-However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
-
-Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.
-
 \chapter{BayesQuant: Probabilistic estimation of protein ratios}
@@
 
-The model was initialized
+In order to get started with Pymc3, we first need to import it.
 
@@
                linenos,
-               xleftmargin=-70pt,
-               numbersep=10pt,
-               gobble=0,
-               frame=lines,
+               numbersep=5pt,
+               autogobble=true,
+               %frame=lines,
                framesep=2mm]{python}
+
+import pymc3 as pm              
+\end{minted}
+
+The model is initialized as a Python context manager.
+Within the context manager, the model is implemented by defining prior distributions for the model parameters and establishing the dependency relationships between them.
+
+The code below formalizes the bias in \ac{MS1} intensity measurements due to a random effect in Pymc3 code.
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               firstnumber=last,
+               framesep=2mm]{python}
+
 with pm.Model() as model:             
+         sigma = pm.HalfNormal('sigma', 1)
+         mu = pm.Normal('mu', mu=0, sd=sigma)
+\end{minted}
+
+Which is equivalent to the following statistical notation:
+
+\begin{align}
+\sigma \sim \mathcal{HN}(1)\,.  \\
+\mu \sim \mathcal{N}(0,\,\sigma^{2})\,.
+\end{align}
+
+and can be read aloud as \textit{the prior probability distribution of the random effect follows a normal distribution with mean 0 and standard deviation $\sigma$.
+In turn, $\sigma$\textquotesingle s prior probability distribution follows a half normal distribution with standard deviation 1.}
+
+\textit{hola}
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               firstnumber=last,
+               framesep=2mm]{python}
+effect = pm.Normal("pep", mu, sigma, n_elements)  
 \end{minted}
@@
 %As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
-\chapter{Pipeline benchmarking on \ac{NZ} data}
-\label{chap:benchmark}
-
-\section*{Summary}
-
-A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
-The experiment consisted of the application of two different treatments to THP-1 cell cultures.
-One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
-The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
-The results confirmed that the computational analysis successfully captured this response.
-It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
-
-\chapter*{Conclusion}
-\label{chap:conclusion}
-
-\chapter*{Appendix}
-\label{chap:appendix}
-
-\includepdf[pages=-]{appendix.pdf}
-
-\bibliographystyle{unsrt_etal}
-\footnotesize{\bibliography{bibliography}}
-
-\end{document}
-%%  ==================================================================
-%%  End document

### 2018-07-17 msreview.tex
Backup

--- 2018-07-16 report.tex
@@
-\documentclass[11pt, a4paper]{report}
-% FOR COLORFUL TABLES
-\usepackage{booktabs}
-\usepackage[table]{xcolor}
-\usepackage[english]{babel}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{pdfpages}
-\usepackage{array} % to make table column italic
-\usepackage{textcomp} % to support the trademark symbol
-
-\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
-\usepackage{wallpaper}
-\usepackage{palatino}
-\usepackage{changepage}
-\usepackage{mhchem} % for isotope nomeclature support
-
-% Setup captions
-%\captionstyle[\centering]{\centering}
-%\changecaptionwidth
-%\captionwidth{0.8\linewidth}
-
-% Protect against widows and orphans
-%\clubpenalty=10000
-%\widowpenalty=10000
-
-%\linespread{1.2}
-
-%\raggedbottom
-
-%\chapterstyle{ger}
-
-%\maxsecnumdepth{subsection}
-
-%%  Setup fancy style quotation
-%%  ==================================================================
-%\usepackage{tikz}
-%\usepackage{framed}
-
-%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font
-
-% Make commands for the quotes
-%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
-%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
-%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
-%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}
-
-% select a colour for the shading
-%\definecolor{shadecolor}{rgb}{1,1,1}
-
-% wrap everything in its own environment
-%\newenvironment{shadequote}% 
-%{\begin{snugshade}\begin{quote}\openquote}
-%{\hfill\closequote\end{quote}\end{snugshade}}
-
-%% MY STUFF
-
-%% Language and font encodings
-
-%% Sets page size and margins
-%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
-\graphicspath{{./plots/}{./figures/}}
-
-%\fancypagestyle{title}{%
-%  \setlength{\headheight}{120pt}%
-%  \fancyhf{}% No header/footer
-%  \renewcommand{\headrulewidth}{0pt}% No header rule
-%  \renewcommand{\footrulewidth}{0pt}% No footer rule
-%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
-%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
-%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
-%}%
-
-%\usepackage{amsmath}
-%\usepackage[format=plain]{caption}
-\usepackage{subcaption}
-\usepackage{float} % force float
-\usepackage[nottoc]{tocbibind}
-%\usepackage[margin=-50pt]{caption}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{cleveref} % for enumerate/itemize item referencing
-
-%\setlength{\parindent}{4em}
-%\setlength{\parskip}{1em}
-%\usepackage{mathpazo} % Use the Palatino font
-\usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{xspace}
-\usepackage{wrapfig}
-\usepackage[export]{adjustbox}
-\usepackage{textcomp}
-\usepackage[bottom]{footmisc}
-
-\usepackage{titling}
-\newcommand{\subtitle}[1]{%
-  \posttitle{%
-    \par\end{center}
-    \begin{center}\large#1\end{center}
-    \vskip0.5em}%
-}
-
-\renewcommand\thesubfigure{\arabic{subfigure}}  
-\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
-%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}
-
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily,
-  showstringspaces=false,
-  commentstyle=\color{red},
-  keywordstyle=\color{blue}
-}
-
-\makeatletter
-\renewcommand\p@subfigure{\thefigure\,}
-\renewcommand\thesubfigure{\alph{subfigure})}
-% If desired for table as well:
-% \renewcommand\p@subtable{\thetable\,}
-% \renewcommand\thesubtable{\alph{subtable})}
-\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
-\makeatother
-
-\renewcommand{\baselinestretch}{1.5}
-\DeclareCaptionSubType*[arabic]{figure}
-
-%\makeatletter
-%\renewcommand\thesection{}
-%\renewcommand\thesubsection{\@arabic\c@subsection}
-%\makeatother
-
-\usepackage{mathtools}
-\usepackage{sectsty}
-
-\chapternumberfont{\LARGE} 
-\chaptertitlefont{\LARGE}
-
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
-
-% Swap the definition of \abs* and \norm*, so that \abs
-% and \norm resizes the size of the brackets, and the 
-% starred version does not.
-\makeatletter
-\let\oldabs\abs
-\def\abs{\@ifstar{\oldabs}{\oldabs*}}
-%
-\let\oldnorm\norm
-\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
-
-%\pagestyle{plain} % default for report
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
-\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
-\makeatother
-
-\usepackage{acro}
-\acsetup{first-style=short}
-
-\DeclareAcronym{MS}{
-  short = MS ,
-  long  = Mass Spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{MS1}{
-  short = MS1 ,
-  long  = first tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS2}{
-  short = MS2 ,
-  long  = second tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS/MS}{
-  short = MS/MS ,
-  long  = Tandem mass spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{m/z}{
-  short = m/z ,
-  long  = mass charge ratio,
-  class = abbrev
-}
-
-\DeclareAcronym{MALDI}{
-  short = MALDI ,
-  long  = Matrix-Assisted Laser-Desorption Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{ESI}{
-  short = ESI ,
-  long  = Electrospray Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{TOF}{
-  short = TOF ,
-  long  = Time of Flight,
-  class = abbrev
-}
-
-\DeclareAcronym{IT}{
-  short = IT ,
-  long  = Ion Trap,
-  class = abbrev
-}
-
-\DeclareAcronym{Q}{
-  short = Q,
-  long  = Quadrupole,
-  class = abbrev
-}
-
-\DeclareAcronym{FT-ICR}{
-  short = FT-ICR ,
-  long  = Fourier Transform Ion Cyclotron Resonance,
-  class = abbrev
-}
-
-\DeclareAcronym{HPLC}{
-  short = HPLC ,
-  long  = High pressure liquid chromatography,
-  class = abbrev
-}
-
-\DeclareAcronym{PTM}{
-  short = PTM ,
-  long  = Post-Translational Modification,
-  class = abbrev
-}
-
-\DeclareAcronym{NZ}{
-  short = NZ ,
-  long  = Novozymes A/S,
-  class = abbrev
-}
-
-\DeclareAcronym{FDR}{
-  short = FDR ,
-  long  = False Discovery Rate,
-  class = abbrev
-}
-
-\DeclareAcronym{RT}{
-  short = RT ,
-  long  = Retention time,
-  class = abbrev
-}
-
-\DeclareAcronym{MBR}{
-  short = MBR ,
-  long  = Match Between Runs,
-  class = abbrev
-}
-
-\DeclareAcronym{XIC}{
-  short = XIC ,
-  long  = Extracted Ion Current,
-  class = abbrev
-}
-
-\DeclareAcronym{SC}{
-  short = SC ,
-  long  = Spectral Counting,
-  class = abbrev
-}
-
-\DeclareAcronym{TPP}{
-  short = TPP ,
-  long  = Trans Proteomic Pipeline,
-  class = abbrev
-}
-
-%%  Begin document
-%%  ==================================================================
-\begin{document}
-
-%%  Begin title page
-%%  ==================================================================
-    \thispagestyle{empty}
-    \ULCornerWallPaper{1}{nat-farve.pdf}
-    \ULCornerWallPaper{1}{nat-en.pdf}
-    \begin{adjustwidth}{-3cm}{-1.5cm}
-    \vspace*{-1cm}
-    \textbf{\Huge Bioinformatics Master Thesis} \\
-    \vspace*{2.5cm} \\
-    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
-    \vspace*{.1cm} \\
-    %{\huge No subtitle is available} \\
-    \begin{tabbing}
-    % adjust the hspace below for the longest author name
-    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
-    %Second Author \> \texttt{<second@author.com>} \\       
-    \\[10.5cm]
-
-    \textbf{\Large Supervisors} \\
-    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
-    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\
-
-    \end{tabbing}
-    \end{adjustwidth}
-    \ClearWallPaper
-%%  ==================================================================
-%%  End title page
-
-%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{2}
-\tableofcontents % Indholdsfortegnelse
-%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
-%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\thispagestyle{empty}
-\printacronyms[include-classes=abbrev,name=Abbreviations]
-
-\newpage
-\setcounter{page}{1}
-
-% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
-\let\oldthebibliography\thebibliography
-\let\endoldthebibliography\endthebibliography
-\renewenvironment{thebibliography}[1]{
-  \begin{oldthebibliography}{#1}
-    \setlength{\itemsep}{0em}
-    \setlength{\parskip}{0em}
-}
-{
-  \end{oldthebibliography}
-}
-
-\section*{Preface}
-
-TODO: get nomenclature straight: what is a sample and what is a fraction
-
-\chapter*{Introduction}
-\label{chap:introduction}
-
-\section{Aminoacids and proteins}
-
-Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.
-
-Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
-The side chain differs between aminaocids and generates them from each other.
-A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
-  \caption{CAPTION AND REFERENCE}
-  \label{fig:aminoacids}
-\end{figure}
-
-Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
-The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
-
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
-One frequent instance of such modifications is the oxidation of methionine.
-
-\section{The protein-focused biotechnology industry}
-
-Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks.
-The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
-
-\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
-The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
-Notorious examples of such applications include waste-water treatment, household care and the baking industry.
-
-The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
-The refinement of the currently used tools and the development of new ones could be of great significance for the company.
-
-Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
-
-\section{Objectives of the Thesis}
-\label{sec:objectives}
-
-In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
-
-\begin{enumerate}
-
-\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
-
-\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
-
-\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
-
-\end{enumerate}
-
-\section{Structure of the Thesis}
-
-An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
-Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
-
-\chapter{Mass spectrometry and shotgun proteomics overview}
+\chapter{Review on mass spectrometry and shotgun proteomics}
 \label{chap:mass_spec}
@@
 A wide range of algorithms are available to process these data and output estimates of protein abundance.
-All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
+All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the distributions symmetrical and thus fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
 They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
 
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
+As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
 These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
@@
 This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
-
-\chapter{A label-free quantification proteomics pipeline}
-\label{chap:pipeline}
-
-\section*{Summary}
-
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
-Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
-A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
-
-\section{Introduction}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like MGF.
-After that, spectra are searched by means of a search engine to perform the PSM step.
-Thereafter, PSMs are validated and the most likely set proteins is inferred.
-If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
-The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
-\label{fig:pipeline}
-\end{figure}
-
-Several proteomics pipelines are available on the internet under different licensing conditions.
-Many are released as closed-source software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
-MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-
-However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform.
-As a GUI would not be required to run the pipeline, it would also be scalable to big datasets.
-Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
-
-The development of a pipeline attempting to achieve these goals will be described in this chapter.
-
-\section{Materials and Methods}
-
-\subsection{Data generation and loading}
-
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-Briefly, the \textit{Homo sapiens} and \textit{E.
-coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
-This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
-One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the open format MGF (Mascot Generic Format).
-
-The MQ+MaxLFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
-The results of the MQ+MSqRob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
-
-%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.
-
-\subsection{Decoy database preparation and search}
-
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine CITE by means of the SearchGUI \texttt{SearchCLI} tool CITE utility.
-The search parameters were set using the \texttt{IdentificationParametersCLI}.
-In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification.
-The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
-Up to two missed cleavages were allowed.
-The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
-
-The target database was created by combining the Uniprot proteomes for \textit{E.
-coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
-
-\subsection{Quality control and validation}
-
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR was set to 1\%.
-PEP and confidence statistics were computed using the PeptideShaker built-in algorithms.
-Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
-
-\subsection{Data refinement}
-
-The moFF command line utility CITE was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
-This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
-
-\subsection{Quantification}
-
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
-Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.
-
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
-The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
-The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
-Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
-
-Quantification in the MQ+MaxLFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
-Intensities were log2 transformed and averaged.
-The difference between conditions H and L was taken as estimate of the log2FC.
-Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
-P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
-
-\subsection{Code implementation}
-
-\section{Results}
-
-\subsection{Thousands of spectra were identified and validated}
-
-The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
-The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \caption*{A}
-  \includegraphics[width=.9\linewidth]{target_vs_decoy}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{B}
-  \includegraphics[width=.9\linewidth]{score_vs_confidence}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{C}
-  \includegraphics[width=.95\linewidth]{qc}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{D}
-  \includegraphics[width=.95\linewidth]{roc}
-\end{subfigure}
-\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
-\textbf{A} Score distribution for matches to the decoy and the target databases.
-\textbf{B} Evolution of the PSM score with confidence.
-The implemented cutoff at FDR of 1 \% is displayed with a red vertical line.
-\textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status.
-\textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search.
-The cutoff is displayed as a red dot.}
-\label{figure:qc_validation}
-\end{figure}
-
- The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
-
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples.
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
-
-\begin{figure}[!h]
-\includegraphics[width=0.8\textwidth]{match_percent}
-\caption{Percentage of matched spectra in all samples.
-The total number of spectra per sample ranged between 5894 and 20249.
-\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
-The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
-\label{fig:match_percent}
-\end{figure}
-
-\subsection{Protein inference}
-
-\begin{table}[H]
-\begin{tabular}{ccccc}
-  \toprule
- Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
-  \midrule
- 46595 & 27384 & 19211 & 25056 & 10039 \\
- \bottomrule
-\end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
-\label{tab:quantification_table}
-\end{table}
-
-A total of 46595 peptides were inferred to be present in at least one of the fractions.
-Of them more than 27 thousand mapped to a unique protein.
-However, a significant amount mapped to more than one protein.
-Such peptides are called non-unique peptides.
-The 25056 detected proteins were grouped into 10039 protein groups.
-Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
-Protein grouping decreased the amount of  considered protein entities to 10039.
-
-Protein groups are very frequent due to several factors.
-For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
-PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
-
-However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.
-
-\subsection{MBR and apex intensity extraction quantification}
-
-The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{mbr_combined}
-\caption{Match Between Runs with moFF.
-\textbf{A} Count of identifications on each run segreagated by source.
-More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.
-\textbf{B} Match visualization.
-Every dot represents a peptide shared across 2 runs.
-The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
-The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
-\label{fig:mbr}
-\end{figure}
-
-Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
-
-\begin{figure}[!h]
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{A}
-\includegraphics[width=.9\linewidth]{peptide_profile}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{B}
-\includegraphics[width=.9\linewidth]{density_ratio}
-\end{subfigure}
-\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
-coli}.
-The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E.
-coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs.
-\textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
-coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
-\label{fig:apex_intensity}
-\end{figure}
-
-\subsection{Quantification}
-\label{subsec:quantification}
-
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L.
-During quantification, it is important to take into account as many of the effects influencing protein quantification.
-In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
-The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
-Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
-
-The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).
-
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Sun Jul 15 18:17:23 2018
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll>{\itshape}l}
-  \toprule
- & log2FC & qval & Protein & Organism \\ 
-  \midrule
-1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
-  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
-   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
-  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
-   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
-  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
-   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
-  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
-   \bottomrule
-\end{tabular}
-\caption{Results of the quantification pipeline.
-The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
-Notably, eight correspond to proteins belonging to \textit{E.coli}.
-The two human proteins exhibited in both cases an absolute value of the log2FC below 1.}
-\label{tab:quantification_table}
-\end{table}
-
-The performance of the quantification can be evaluated thanks to the experimental design of the dataset.
-Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a log2FC of $log2(3)=1.58$ is expected for proteins coming from this organism.
-Likewise, human proteins should have a log2FC of 0.
-As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
-
-Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
-Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
-\textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
-\textbf{MQ+MaxLFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+MSqRob blends MaxQuant upstream analysis with the MSqRob quantification engine.
-\textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
-\textbf{C, D} PR curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.}
-\label{fig:roc_curves}
-\end{figure}
-
-In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
-Thus, \textit{E.
-coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
-Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
-In order to compare the results of the here presented pipeline (Compomics+MSqRob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+MaxLFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+MSqRob) used MaxQuant up to the fraction normalization step.
-
-The result of this analysis is shown in figure \ref{fig:roc_curves}).
-Remarkably, Compomics+MSqRob achieved similar results to those resulting from MQ+MaxLFQ.
-Nonetheless, the latter performed clearly better when the log2FC filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+MSqRob was as good as MQ+MaxLFQ, if not much better.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{vennDiagram}
-\caption{Agreement between the different pipelines. The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline.}
-\label{fig:venn}
-\end{figure}
-
-In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn}).
-Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Compomics+MSqRob or MQ+MaxLFQ being detectd by MQ+MSqRob too.
-Remarkably, 517 proteins were detected by MQ+MSqRob alone, which further confirmed its resolving power.
-
-Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-This analysis confirmed that the pipeline providing the best separation between organisms was MQ+MSqRob.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines performance.
-Quantification results are segregated by the source organism.
-\textbf{A} Histogram of the log2FC estimates.
-\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its log2FC on x and the minus logarithm of the q-value on y.
-\textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
-\label{fig:combined_plot}
-\end{figure}
-
-\section{Discussion}
-
-\subsection{Improvement of the PSM step}
-
-The low attained matching rate manifests existing room for improvement in the currently available tools.
-Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
-Their usage is predicted tor improve identification rates and quality of the matches.
-
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
-New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
-Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
-
-Finally, the extremely low matching rates in the latter fractions translated to decreased contributions to the number of identifications.
-This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
-
-\subsection{Improvement of the feature extraction step}
-
-While the apex intensity is a good estimate of the peptide abundance, other methods are available that could potentially provide more robust data, such as the area of the peak.
-This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
-
-\subsection{Improvement of the quantification step}
-
-The deployed relative quantification approach successfully manages to estimate the log2FC of several of proteins in the dataset with high confidence.
-However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
-Several axes of improvement are clear from the results.
-
-\subsubsection{Management of sample fractionation}
-
-Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
-
-The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
-One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
-A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
-
-MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at log2FC of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
-While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Compomics+MSqRob, proving room for improvement in this aspect.
-
-In any case, the way sample fractionation is handled should depend upon the fractionation method.
-For example, membrane fractionation can be exploited when working with membrane proteins \cite{Marmagne2006}.
-
-To sum up, the presently implemented fraction handling is correct and provides good results, but fine-tuning of the pipeline at this step is expected to improve the results of the quantification.
-Thanks to the open-source nature of the software, this is indeed possible.
-
-%which is defined according to the following expressions:
-%
-%\begin{equation}
-%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{A,j} \times \text{XIC}_{P,A,j}}
-%\end{equation}
-%
-%where $\text{N}_{A,j}$ stands for the normalization coefficient for sample $A$ and fraction $j$, and $\text{XIC}_{P,A,j}$ stands for the extracted ion current in the same sample and fraction for peptide P.
-The sum runs over all k fractions.
-%
-%\begin{equation}
-%H_P(N) = \sum\limits_{A,B} \abs{\log(\frac{I_{P,A}(N)}{I_{P,B}(N)})}^2
-%\end{equation}
-%
-%\begin{equation}
-%H(N) = \sum\limits_{p} H_p(N) 
-%\end{equation}
-%
-%where the pair $A$ and $B$ is any of the sample pairs where peptide $P$ is detected. The overall proteome variation is
-
-\subsubsection{The pitfalls of mass spectrometry}
-
-The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
-One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
-Notwithstanding, it reveals the immature state of the technology.
-Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
-Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
-Indeed, the input file for the MQ+MaxLFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn}).
-
-\subsubsection{Support for absolute quantification}
-
-While relative quantification like that performed by MSqRob is enough to infer differential abundance of proteins across conditions, absolute quantification would provide an estimate of protein quantities that would support many other analyses, and facilitate comparison between datasets, in a way similar to what the normalised read counts in FPKM (Fragments Per Kilobase per Million Reads) does in transcriptomics.
-On the other hand, MaxLFQ provides absolute estimates, albeit less robust.
-
-\subsubsection{Quantification of uncertainty}
-
-All the quantification approaches enunciated until now make use of different frequentist approaches to the problem, that evaluate how significantly different the data are from what would be expected under a null hypothesis of a null ratio across the pair of conditions.
-Yet, it does not provide probabilistic interpretations of the uncertainty behind the estimate.
-The development of a Bayesian framework which endeavors at solving this issue is presented in chapter \ref{chap:model}.
-
-\section{Conclusion}
-
-An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
-The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
-Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
-Likewise, the openness of the data in all steps allows for quality control checks along the way.
-However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
-
-\chapter{BayesQuant: Probabilistic estimation of protein ratios}
-\label{chap:model}
-
-\section*{Summary}
-
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
-However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
-
-\section{Introduction}
-
-Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
-Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
-The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (NHST).
-
-\begin{align}
-H_0: & \log_2(FC) = 0 \nonumber \\
-H_1: & \log_2(FC) \neq 0 \nonumber
-\end{align}
-
-The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the $\log_2$(FC) between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
-The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
-
-The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
-While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes-theorem.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\end{figure}
-
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
-\chapter{Pipeline benchmarking on \ac{NZ} data}
-\label{chap:benchmark}
-
-\section*{Summary}
-
-A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
-The experiment consisted of the application of two different treatments to THP-1 cell cultures.
-One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
-The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
-The results confirmed that the computational analysis successfully captured this response.
-It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
-
-\chapter*{Conclusion}
-\label{chap:conclusion}
-
-\chapter*{Appendix}
-\label{chap:appendix}
-
-\includepdf[pages=-]{appendix.pdf}
-
-\bibliographystyle{unsrt_etal}
-\footnotesize{\bibliography{bibliography}}
-
-\end{document}
-%%  ==================================================================
-%%  End document

### 2018-07-17 pipeline.tex
Backup

--- 2018-07-16 report.tex
@@
-\documentclass[11pt, a4paper]{report}
-% FOR COLORFUL TABLES
-\usepackage{booktabs}
-\usepackage[table]{xcolor}
-\usepackage[english]{babel}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{pdfpages}
-\usepackage{array} % to make table column italic
-\usepackage{textcomp} % to support the trademark symbol
-
-\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
-\usepackage{wallpaper}
-\usepackage{palatino}
-\usepackage{changepage}
-\usepackage{mhchem} % for isotope nomeclature support
-
-% Setup captions
-%\captionstyle[\centering]{\centering}
-%\changecaptionwidth
-%\captionwidth{0.8\linewidth}
-
-% Protect against widows and orphans
-%\clubpenalty=10000
-%\widowpenalty=10000
-
-%\linespread{1.2}
-
-%\raggedbottom
-
-%\chapterstyle{ger}
-
-%\maxsecnumdepth{subsection}
-
-%%  Setup fancy style quotation
-%%  ==================================================================
-%\usepackage{tikz}
-%\usepackage{framed}
-
-%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font
-
-% Make commands for the quotes
-%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
-%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
-%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
-%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}
-
-% select a colour for the shading
-%\definecolor{shadecolor}{rgb}{1,1,1}
-
-% wrap everything in its own environment
-%\newenvironment{shadequote}% 
-%{\begin{snugshade}\begin{quote}\openquote}
-%{\hfill\closequote\end{quote}\end{snugshade}}
-
-%% MY STUFF
-
-%% Language and font encodings
-
-%% Sets page size and margins
-%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
-\graphicspath{{./plots/}{./figures/}}
-
-%\fancypagestyle{title}{%
-%  \setlength{\headheight}{120pt}%
-%  \fancyhf{}% No header/footer
-%  \renewcommand{\headrulewidth}{0pt}% No header rule
-%  \renewcommand{\footrulewidth}{0pt}% No footer rule
-%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
-%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
-%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
-%}%
-
-%\usepackage{amsmath}
-%\usepackage[format=plain]{caption}
-\usepackage{subcaption}
-\usepackage{float} % force float
-\usepackage[nottoc]{tocbibind}
-%\usepackage[margin=-50pt]{caption}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{cleveref} % for enumerate/itemize item referencing
-
-%\setlength{\parindent}{4em}
-%\setlength{\parskip}{1em}
-%\usepackage{mathpazo} % Use the Palatino font
-\usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{xspace}
-\usepackage{wrapfig}
-\usepackage[export]{adjustbox}
-\usepackage{textcomp}
-\usepackage[bottom]{footmisc}
-
-\usepackage{titling}
-\newcommand{\subtitle}[1]{%
-  \posttitle{%
-    \par\end{center}
-    \begin{center}\large#1\end{center}
-    \vskip0.5em}%
-}
-
-\renewcommand\thesubfigure{\arabic{subfigure}}  
-\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
-%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}
-
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily,
-  showstringspaces=false,
-  commentstyle=\color{red},
-  keywordstyle=\color{blue}
-}
-
-\makeatletter
-\renewcommand\p@subfigure{\thefigure\,}
-\renewcommand\thesubfigure{\alph{subfigure})}
-% If desired for table as well:
-% \renewcommand\p@subtable{\thetable\,}
-% \renewcommand\thesubtable{\alph{subtable})}
-\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
-\makeatother
-
-\renewcommand{\baselinestretch}{1.5}
-\DeclareCaptionSubType*[arabic]{figure}
-
-%\makeatletter
-%\renewcommand\thesection{}
-%\renewcommand\thesubsection{\@arabic\c@subsection}
-%\makeatother
-
-\usepackage{mathtools}
-\usepackage{sectsty}
-
-\chapternumberfont{\LARGE} 
-\chaptertitlefont{\LARGE}
-
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
-
-% Swap the definition of \abs* and \norm*, so that \abs
-% and \norm resizes the size of the brackets, and the 
-% starred version does not.
-\makeatletter
-\let\oldabs\abs
-\def\abs{\@ifstar{\oldabs}{\oldabs*}}
-%
-\let\oldnorm\norm
-\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
-
-%\pagestyle{plain} % default for report
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
-\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
-\makeatother
-
-\usepackage{acro}
-\acsetup{first-style=short}
-
-\DeclareAcronym{MS}{
-  short = MS ,
-  long  = Mass Spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{MS1}{
-  short = MS1 ,
-  long  = first tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS2}{
-  short = MS2 ,
-  long  = second tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS/MS}{
-  short = MS/MS ,
-  long  = Tandem mass spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{m/z}{
-  short = m/z ,
-  long  = mass charge ratio,
-  class = abbrev
-}
-
-\DeclareAcronym{MALDI}{
-  short = MALDI ,
-  long  = Matrix-Assisted Laser-Desorption Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{ESI}{
-  short = ESI ,
-  long  = Electrospray Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{TOF}{
-  short = TOF ,
-  long  = Time of Flight,
-  class = abbrev
-}
-
-\DeclareAcronym{IT}{
-  short = IT ,
-  long  = Ion Trap,
-  class = abbrev
-}
-
-\DeclareAcronym{Q}{
-  short = Q,
-  long  = Quadrupole,
-  class = abbrev
-}
-
-\DeclareAcronym{FT-ICR}{
-  short = FT-ICR ,
-  long  = Fourier Transform Ion Cyclotron Resonance,
-  class = abbrev
-}
-
-\DeclareAcronym{HPLC}{
-  short = HPLC ,
-  long  = High Pressure Liquid Chromatography,
-  class = abbrev
-}
-
-\DeclareAcronym{PTM}{
-  short = PTM ,
-  long  = Post-Translational Modification,
-  class = abbrev
-}
-
-\DeclareAcronym{NZ}{
-  short = NZ ,
-  long  = Novozymes A/S,
-  class = abbrev
-}
-
-\DeclareAcronym{FDR}{
-  short = FDR ,
-  long  = False Discovery Rate,
-  class = abbrev
-}
-
-\DeclareAcronym{RT}{
-  short = RT ,
-  long  = Retention time,
-  class = abbrev
-}
-
-\DeclareAcronym{log2FC}{
-  short = log2FC ,
-  long  = $\log_2$(FC),
-  class = abbrev
-}
-
-\DeclareAcronym{MBR}{
-  short = MBR ,
-  long  = Match Between Runs,
-  class = abbrev
-}
-
-\DeclareAcronym{MGF}{
-  short = MGF ,
-  long  = Mascot Generic Format,
-  class = abbrev
-}
-
-\DeclareAcronym{XIC}{
-  short = XIC ,
-  long  = Extracted Ion Current,
-  class = abbrev
-}
-
-\DeclareAcronym{SC}{
-  short = SC ,
-  long  = Spectral Counting,
-  class = abbrev
-}
-
-\DeclareAcronym{TPP}{
-  short = TPP ,
-  long  = Trans Proteomic Pipeline,
-  class = abbrev
-}
-
-\DeclareAcronym{NHST}{
-  short = NHST ,
-  long  = Null-Hypothesis Significance Testing,
-  class = abbrev
-}
-
-%%  Begin document
-%%  ==================================================================
-\begin{document}
-
-%%  Begin title page
-%%  ==================================================================
-    \thispagestyle{empty}
-    \ULCornerWallPaper{1}{nat-farve.pdf}
-    \ULCornerWallPaper{1}{nat-en.pdf}
-    \begin{adjustwidth}{-3cm}{-1.5cm}
-    \vspace*{-1cm}
-    \textbf{\Huge Bioinformatics Master Thesis} \\
-    \vspace*{2.5cm} \\
-    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
-    \vspace*{.1cm} \\
-    %{\huge No subtitle is available} \\
-    \begin{tabbing}
-    % adjust the hspace below for the longest author name
-    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
-    %Second Author \> \texttt{<second@author.com>} \\       
-    \\[10.5cm]
-
-    \textbf{\Large Supervisors} \\
-    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
-    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\
-
-    \end{tabbing}
-    \end{adjustwidth}
-    \ClearWallPaper
-%%  ==================================================================
-%%  End title page
-
-%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{2}
-\tableofcontents % Indholdsfortegnelse
-%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
-%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\thispagestyle{empty}
-\printacronyms[include-classes=abbrev,name=Abbreviations]
-
-\newpage
-\setcounter{page}{1}
-
-% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
-\let\oldthebibliography\thebibliography
-\let\endoldthebibliography\endthebibliography
-\renewenvironment{thebibliography}[1]{
-  \begin{oldthebibliography}{#1}
-    \setlength{\itemsep}{0em}
-    \setlength{\parskip}{0em}
-}
-{
-  \end{oldthebibliography}
-}
-
-\section*{Preface}
-
-TODO: get nomenclature straight: what is a sample and what is a fraction
-
-\chapter*{Introduction}
-\label{chap:introduction}
-
-\section{Aminoacids and proteins}
-
-Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.
-
-Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
-The side chain differs between aminaocids and generates them from each other.
-A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
-  \caption{CAPTION AND REFERENCE}
-  \label{fig:aminoacids}
-\end{figure}
-
-Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
-The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
-
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
-One frequent instance of such modifications is the oxidation of methionine.
-
-\section{The protein-focused biotechnology industry}
-
-Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks.
-The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
-
-\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
-The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
-Notorious examples of such applications include waste-water treatment, household care and the baking industry.
-
-The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
-The refinement of the currently used tools and the development of new ones could be of great significance for the company.
-
-Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
-
-\section{Objectives of the Thesis}
-\label{sec:objectives}
-
-In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
-
-\begin{enumerate}
-
-\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
-
-\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
-
-\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
-
-\end{enumerate}
-
-\section{Structure of the Thesis}
-
-An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
-Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
-
-\chapter{Review on mass spectrometry and shotgun proteomics}
-\label{chap:mass_spec}
-
-Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone.
-Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up.
-The layer provided by proteins is one of them, and its study is called proteomics.
-It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment.
-The study of the protein-specific complexity is called proteomics.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
-
-%CITE 1.6.1 COMPUTATIONAL METHODS.
-The existing approaches are divided into two types of paradigms: top-down and bottom-up.
-In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
-These smaller parts are called peptides.
-\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
-The top-down paradigm will be ignored in the rest of the manuscript.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{proteomics_skema_book}
-\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
-\label{fig:proteomics_overview}
-\end{figure}
-
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
-The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem, explained in section \ref{sec:quantification}.
-
-A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:
-
-\begin{enumerate}
-
-\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:tandem_ms_workflow}.
-
-\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.
-
-\end{enumerate}
-
-\LARGE
-\textbf{MS analysis and data generation}
-\normalsize
-
-\section{Sample processing}
-\label{sec:sample_processing}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{sample_processing}
-\caption{Overview of the sample processing step prior to mass spectrometry analysis.
-First, proteins are denatured and digested with a specific protease like Trypsin.
-This yields a peptide mix that is separated into peptide fractions that can be introduced in the mass spectrometer.}
-\label{fig:sample_processing}
-\end{figure}
-
-\subsection{Protein digestion}
-\label{subsec:protein_digestion}
-
-An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task.
-It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
-
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
-%COMPUTATIONAL METHODS 3.2
-
-\begin{enumerate}
-
-\item Missed cleavages.
-
-\item Unsuspected cleavages during the maturation/life cycle of the protein.
-
-\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
-
-\item Naturally occurring, intentionally or unintentionally induced chemical modifications.
-
-\end{enumerate}
-
-%\Cref{item:missed_cleavages}
-
-Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
-
-The result of this process is a complex mix of peptides, made up by hundreds or thousands of different molecules, following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-A peptide separation step is required before introducing the sample in the spectrometer.
-
-\subsection{Peptide separation}
-\label{subsec:peptide_separation}
-
-If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
-This can be surmounted by independently analyzing different fractions of the sample over time.
-The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
-
-%CITE 4.2 computational methods.
-During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
-
-\section{The mass spectrometer}
-\label{sec:the_mass_spectrometer}
-
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
-It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
-\label{fig:mass_spectrometer}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-\subsection{The ion source}
-\label{subsec:the_ion_source}
-
-All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-
-%CITE 5.1 COMPUTATIONAL METHODS
-
-This step is performed in the ion source \cite{Barsnes2008}.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
-Thus, the charge exhibited by an ion is not obvious when produced via ESI.
-Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
-Bias in ESI is less predictable.
-This is known as the competitive ionization problem \cite{Tang2004}.
-
-The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
-
-\subsection{The mass analyzer}
-\label{subsec:the_mass_analyzer}
-
-The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
-These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.
-
-Moreover, two other analyzers exist which combine mass analysis with intensity measurement.
-These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
-They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
-
-\subsection{The detector}
-\label{subsec:the_detector}
-
-Detectors measure the intensity of an incoming ion signal.
-The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
-Some topics in signal detection in \ac{MS} need to be discussed.
-
-On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
-The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
-This is one of the reasons why proteins are better fit for analysis when digested into peptides, as \ac{m/z} are reduced, thus increasing the mass resolution.
-
-On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
-Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
-Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{envelope}
-\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
-Each peak in the envelope is separated by 0.5 Th.
-This is explained by the peptide having 2 positive charges that make every extra Da in the ion mass account for 1/2 extra Th.
-Adapted from \cite{Mirzaei2016}.}
-\label{fig:envelope}
-\end{figure}
-
-The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
-
-It is up to the \ac{MS} technician to decide on the best pieces of equipment according to their availability and particularities of the dataset.
-
-\section{Tandem MS workflow}
-\label{sec:tandem_ms_workflow}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{tandem_ms}
-\caption[]{Illustration of the tandem MS workflow.
-The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
-The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
-Finally the third spectrometer records fragment mass spectra.
-Taken from \footnotemark{}.}
-\label{fig:tandem_ms}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
-In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
-An extra spectrometer, usually a quadrupole, is introduced in between.
-
-\begin{itemize}
-
-\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
-
-\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
-
-\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
-
-\end{itemize}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{shotgun_adapted}
-\caption{Illustration of the different spectra collected in tandem MS.
-MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
-MS2 spectra record the same magnitudes but the signal is generated by the fragments produced during fragmentation by the ion filtered in the first spectrometer.
-Altogether, they enable peptide identifications.
-Figure adapted from \cite{Verheggen2017}.}
-\label{fig:shotgun}
-\end{figure}
-
-\subsection{Fragmentation}
-\label{subsec:fragmentation}
-
-The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
-Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{abcxyz}
-\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
-The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
-Specific fragmentation techniques make fragments belonging to one series more likely than others.
-Other fragments are possible but much less likely.
-The fragment nomenclature was introduced in \cite{Roepstorff1984}.}
-\label{fig:abcxyz}
-\end{figure}
-
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
-During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
-These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
-The difference can be exploited to infer which aminoacid is present in the longer fragment.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
-explained in \ref{sec:search_engines}.
-
-%A given bond will be more likely to break the less stable it is.
-This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the most likely to break (see figure \ref{fig:abcxyz}).
-A nomenclature REF was introduced to name these fragments:
-
-%%PAG 123 COMPUTATIONAL METHODS
-%Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
-Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-\cite{Barsnes2008}.
-Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008}.
-%
-%%CITE pag 134 computational methods
-%
-%EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
-Thus, the fragmentation process is not ergodic and returns TYPE ions.
-More importantly, chemical modifications are not lost in the process.
-
-\LARGE
-\textbf{Computational analysis}
-\normalsize
-
-\section{Spectra processing: search engines}
-\label{sec:search_engines}
-
-Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
-\ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{computational_analysis}
-\caption{Diagram of the MS computational analysis.
-The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
-In other words, the software infers what peptides were introduced in the spectrometer.
-The analysis continues during protein inference and quantification.}
-\label{fig:computational_analysis}
-\end{figure}
-
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{PSM}
-%\caption{}
-%\label{fig:PSM}
-%\end{figure}
-
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
-This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
-Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
-Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.
-
-Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
-Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
-
-\section{Validation and quality control}
-\label{sec:validation}
-
-The scoring systems implemented in search engines provide the best matches, but they are bound to contain false identifications.
-Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
-
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
-
-The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
-Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
-
-All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
-This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
-Together with the number of PSMs passing a given score in the target ($n_{tp} + n_{tp}$), the FDR can be computed using equation \ref{eq:fdr}.
-
-\begin{equation}\label{eq:fdr}
-FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
-\end{equation}
-
-The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
-Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{pep}
-\caption{Visualizing FDR and PEP.
-The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
-The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
-Figure from \cite{Kall2008}.}
-\label{fig:pep}
-\end{figure}
-
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
-Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
-The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
-PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.
-
-\section{Peptide and protein inference}
-\label{sec:inference}
-
-Two steps in protein identification can be distinguished:
-
-\begin{enumerate}
-
-\item \textbf{Peptide inference}: infer the peptides present in the sample.
-\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and frequently map to more than one protein.
-\end{enumerate}
-
-The result of the PSM step returns the inferred list of peptides.
-The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
-The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
-Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
-
-\section{Protein quantification}
-\label{sec:quantification}
-
-The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
-However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
-For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
-
-Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free).
-These are explained in subsection \ref{subsec:labelling}.
-If the label-free approach is employed, more distinctions can be made based on:
-
-\begin{itemize}
-\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}
-
-\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based, explained in \ref{subsec:peptide_model}.
-\end{itemize}
-
-\subsection{Label-based and label-free approaches}
-\label{subsec:labelling}
-
-%CITE PAGE 237 COMPUTATIONAL METHODS
-
-Two paradigms exist in protein quantification: label-based and label-free.
-In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}.
-The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
-Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
-Remarkably, peptide labeling costs can be quite high.
-This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
-
-In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
-In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption{Illustration of the \ac{MBR} step.
-\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_mbr}
-\end{figure}
-
-This way, extra identifications are attained through a reanalysis of the spectra, factoring in the information collected in replicate runs.
-The RT and precursor mass of unidentified spectra in one replicate is matched to that of identified spectra in the remaining replicates.
-As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
-
-\subsection{SC and XIC based quantification}
-\label{subsec:scvsxic}
-
-Quantification can be \ac{SC} or \ac{XIC} based.
-
-On the one hand, spectrum counting based quantification is the simplest method in proteomics.
-The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance.
-It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing.
-
-On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance.
-A wide range of algorithms are available to process these data and output estimates of protein abundance.
-All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
-They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
-
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
-For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
-
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{A}
-\includegraphics[width=0.7\linewidth]{apex_3d}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{B}
-\includegraphics[width=0.7\linewidth]{apex_intensity}
-\end{subfigure}
-\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
-An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
-Taken from \cite{Smith2014}.
-\textbf{B} A refinement analysis of these data attempts to fit a mathematical model of the signal over time and extract a representative measurement, such as the highest (apex) \ac{MS1} intensity.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_apex}
-\end{figure}
-
-\subsection{\ac{XIC}-peptide-based models for label-free quantification}
-\label{subsec:peptide_model}
-
-The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample.
-However, the data interpretation requires the transfer of these peptide-level data into the protein level.
-This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.
-
-As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
-However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
-This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
-
-It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
-This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
-
 \chapter{A label-free quantification proteomics pipeline}
@@
 Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
-Protein grouping decreased the amount of  considered protein entities to 10039.
+Thus, protein inference algorithms create them when a does not uniquely map to a protein because it is not possible to distinguish which protein it truly comes from.
 
@@
 
-\chapter{BayesQuant: Probabilistic estimation of protein ratios}
-\label{chap:model}
-
-\section*{Summary}
-
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
-However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
-
-\section{Introduction}
-
-\subsection{Bayesian statistics}
-
-Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
-Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
-The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (\ac{NHST}).
-
-\begin{align}
-H_0: & \log_2(FC) = 0 \nonumber \\
-H_1: & \log_2(FC) \neq 0 \nonumber
-\end{align}
-
-The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the log2FC between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
-The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
-
-The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
-While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes-theorem.
-
-\begin{equation}
-P(\theta | y) = \cfrac{P(y | \theta) P(\theta)}{P(y)}
-\end{equation}
-
-The Bayes theorem is an extremely versatile tool taking a role analogous to that of statistics like T, F, or $\chi^2$.
-Unlike them, which are tailored to specific scenarios, it can be used compute the probability distribution of almost any parameter in any model.
-4 terms can be distinguished in its expression
-
-\begin{enumerate}
-
-\item $P(\theta)$: the \textbf{prior} probability distribution of the model parameter, introduces previous knowledge about the phenomenon being modelled.
-
-\item $P(y | \theta)$: the probability distribution of the observed data (y) over the possible parameter space.
-It is also known as \textbf{likelihood} of the model, and its duty is updating our beliefs about the phenomenon by capturing the information in the data.
-
-\item $P(y)$: the probability of the data, defined as the marginal probability of the data given a parameter value, for all possible values.
-It acts as a normalizing constant that makes the resulting distribution a true probability distribution adding up to 1.
-It is also known as the data \textbf{evidence}.
-
-\item $P(\theta | y)$: the updated probability distribution of the model parameter, with the information extracted from the observed data.
-Since it reflects the beliefs about the phenomena after observing data, it is called \textbf{posterior} probability distribution,as opposite to the prior.
-
-\end{enumerate}
-
-The Bayesian framework enunciated above can be applied to the quantification problem.
-All is needed is the computation of the posterior probability distribution.
-Unfortunately, this is tractable analytically for simple models only.
-However, Markov Chain-Monte Carlo methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough.
-Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible.
-Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3), all making use of a programming paradigm called \textbf{probabilistic programming}.
-
-\subsection{Probabilistic programming}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\end{figure}
-
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
-\chapter{Pipeline benchmarking on \ac{NZ} data}
-\label{chap:benchmark}
-
-\section*{Summary}
-
-A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
-The experiment consisted of the application of two different treatments to THP-1 cell cultures.
-One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
-The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
-The results confirmed that the computational analysis successfully captured this response.
-It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
-
-\chapter*{Conclusion}
-\label{chap:conclusion}
-
-\chapter*{Appendix}
-\label{chap:appendix}
-
-\includepdf[pages=-]{appendix.pdf}
-
-\bibliographystyle{unsrt_etal}
-\footnotesize{\bibliography{bibliography}}
-
-\end{document}
-%%  ==================================================================
-%%  End document
+Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.

### 2018-07-17 thp1.tex
Backup

--- 2018-07-16 report.tex
@@
-\documentclass[11pt, a4paper]{report}
-% FOR COLORFUL TABLES
-\usepackage{booktabs}
-\usepackage[table]{xcolor}
-\usepackage[english]{babel}
-\usepackage[T1]{fontenc}
-\usepackage[utf8]{inputenc}
-\usepackage{pdfpages}
-\usepackage{array} % to make table column italic
-\usepackage{textcomp} % to support the trademark symbol
-
-\DeclareUnicodeCharacter{2010}{-}% support older LaTeX versions
-\usepackage{wallpaper}
-\usepackage{palatino}
-\usepackage{changepage}
-\usepackage{mhchem} % for isotope nomeclature support
-
-% Setup captions
-%\captionstyle[\centering]{\centering}
-%\changecaptionwidth
-%\captionwidth{0.8\linewidth}
-
-% Protect against widows and orphans
-%\clubpenalty=10000
-%\widowpenalty=10000
-
-%\linespread{1.2}
-
-%\raggedbottom
-
-%\chapterstyle{ger}
-
-%\maxsecnumdepth{subsection}
-
-%%  Setup fancy style quotation
-%%  ==================================================================
-%\usepackage{tikz}
-%\usepackage{framed}
-
-%\newcommand*\quotefont{\fontfamily{fxl}} % selects Libertine for quote font
-
-% Make commands for the quotes
-%\newcommand*{\openquote}{\tikz[remember picture,overlay,xshift=-15pt,yshift=-10pt]
-%     \node (OQ) {\quotefont\fontsize{60}{60}\selectfont``};\kern0pt}
-%\newcommand*{\closequote}{\tikz[remember picture,overlay,xshift=15pt,yshift=5pt]
-%     \node (CQ) {\quotefont\fontsize{60}{60}\selectfont''};}
-
-% select a colour for the shading
-%\definecolor{shadecolor}{rgb}{1,1,1}
-
-% wrap everything in its own environment
-%\newenvironment{shadequote}% 
-%{\begin{snugshade}\begin{quote}\openquote}
-%{\hfill\closequote\end{quote}\end{snugshade}}
-
-%% MY STUFF
-
-%% Language and font encodings
-
-%% Sets page size and margins
-%\usepackage[a4paper,top=2.5cm,bottom=2cm,left=2.5cm,right=2.5cm,marginparwidth=1.75cm]{geometry}
-\usepackage{graphicx}
-\usepackage{caption}
-\usepackage{subcaption}
-\captionsetup{font={small},skip=0.25\baselineskip}
-\captionsetup[subfigure]{font={bf, footnotesize}, skip=1pt, singlelinecheck=false}
-\graphicspath{{./plots/}{./figures/}}
-
-%\fancypagestyle{title}{%
-%  \setlength{\headheight}{120pt}%
-%  \fancyhf{}% No header/footer
-%  \renewcommand{\headrulewidth}{0pt}% No header rule
-%  \renewcommand{\footrulewidth}{0pt}% No footer rule
-%  %\fancyhead[L]{\small\begin{tabular}{@{}l}\includegraphics[width=0.15\textwidth]{ku}\end{tabular}}
-%   % \fancyhead[C]{\small\begin{tabular}{@{}l}\includegraphics[width=0.6\textwidth]{PROSCION}\end{tabular}}
-%  %\fancyhead[R]{\small\begin{tabular}{@{}r}\includegraphics[width=0.15\textwidth]{naturvidenskab}\end{tabular}}
-%}%
-
-%\usepackage{amsmath}
-%\usepackage[format=plain]{caption}
-\usepackage{subcaption}
-\usepackage{float} % force float
-\usepackage[nottoc]{tocbibind}
-%\usepackage[margin=-50pt]{caption}
-%\usepackage[colorinlistoftodos]{todonotes}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
-\usepackage{cleveref} % for enumerate/itemize item referencing
-
-%\setlength{\parindent}{4em}
-%\setlength{\parskip}{1em}
-%\usepackage{mathpazo} % Use the Palatino font
-\usepackage{parskip}% http://ctan.org/pkg/parskip
-\usepackage{xspace}
-\usepackage{wrapfig}
-\usepackage[export]{adjustbox}
-\usepackage{textcomp}
-\usepackage[bottom]{footmisc}
-
-\usepackage{titling}
-\newcommand{\subtitle}[1]{%
-  \posttitle{%
-    \par\end{center}
-    \begin{center}\large#1\end{center}
-    \vskip0.5em}%
-}
-
-\renewcommand\thesubfigure{\arabic{subfigure}}  
-\captionsetup{format=hang,labelsep=space,indention=-2cm,labelfont=bf,width=.9\textwidth,skip=.5\baselineskip}
-%\captionsetup[sub]{labelfont=bf, labelsep=period,labelformat=simple, subrefformat=brace}
-
-\usepackage{listings}
-\lstset{basicstyle=\ttfamily,
-  showstringspaces=false,
-  commentstyle=\color{red},
-  keywordstyle=\color{blue}
-}
-
-\makeatletter
-\renewcommand\p@subfigure{\thefigure\,}
-\renewcommand\thesubfigure{\alph{subfigure})}
-% If desired for table as well:
-% \renewcommand\p@subtable{\thetable\,}
-% \renewcommand\thesubtable{\alph{subtable})}
-\DeclareCaptionLabelFormat{mysublabelfmt}{\alph{sub\@captype}}
-\makeatother
-
-\renewcommand{\baselinestretch}{1.5}
-\DeclareCaptionSubType*[arabic]{figure}
-
-%\makeatletter
-%\renewcommand\thesection{}
-%\renewcommand\thesubsection{\@arabic\c@subsection}
-%\makeatother
-
-\usepackage{mathtools}
-\usepackage{sectsty}
-
-\chapternumberfont{\LARGE} 
-\chaptertitlefont{\LARGE}
-
-\DeclarePairedDelimiter\abs{\lvert}{\rvert}%
-\DeclarePairedDelimiter\norm{\lVert}{\rVert}%
-
-% Swap the definition of \abs* and \norm*, so that \abs
-% and \norm resizes the size of the brackets, and the 
-% starred version does not.
-\makeatletter
-\let\oldabs\abs
-\def\abs{\@ifstar{\oldabs}{\oldabs*}}
-%
-\let\oldnorm\norm
-\def\norm{\@ifstar{\oldnorm}{\oldnorm*}}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
-
-%\pagestyle{plain} % default for report
-
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd{\@makechapterhead}{50\p@}{0pt}{}{}
-\patchcmd{\@makeschapterhead}{50\p@}{0pt}{}{}
-\makeatother
-
-\usepackage{acro}
-\acsetup{first-style=short}
-
-\DeclareAcronym{MS}{
-  short = MS ,
-  long  = Mass Spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{MS1}{
-  short = MS1 ,
-  long  = first tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS2}{
-  short = MS2 ,
-  long  = second tandem MS analyzer ,
-  class = abbrev
-}
-
-\DeclareAcronym{MS/MS}{
-  short = MS/MS ,
-  long  = Tandem mass spectrometry,
-  class = abbrev
-}
-
-\DeclareAcronym{m/z}{
-  short = m/z ,
-  long  = mass charge ratio,
-  class = abbrev
-}
-
-\DeclareAcronym{MALDI}{
-  short = MALDI ,
-  long  = Matrix-Assisted Laser-Desorption Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{ESI}{
-  short = ESI ,
-  long  = Electrospray Ionization,
-  class = abbrev
-}
-
-\DeclareAcronym{TOF}{
-  short = TOF ,
-  long  = Time of Flight,
-  class = abbrev
-}
-
-\DeclareAcronym{IT}{
-  short = IT ,
-  long  = Ion Trap,
-  class = abbrev
-}
-
-\DeclareAcronym{Q}{
-  short = Q,
-  long  = Quadrupole,
-  class = abbrev
-}
-
-\DeclareAcronym{FT-ICR}{
-  short = FT-ICR ,
-  long  = Fourier Transform Ion Cyclotron Resonance,
-  class = abbrev
-}
-
-\DeclareAcronym{HPLC}{
-  short = HPLC ,
-  long  = High pressure liquid chromatography,
-  class = abbrev
-}
-
-\DeclareAcronym{PTM}{
-  short = PTM ,
-  long  = Post-Translational Modification,
-  class = abbrev
-}
-
-\DeclareAcronym{NZ}{
-  short = NZ ,
-  long  = Novozymes A/S,
-  class = abbrev
-}
-
-\DeclareAcronym{FDR}{
-  short = FDR ,
-  long  = False Discovery Rate,
-  class = abbrev
-}
-
-\DeclareAcronym{RT}{
-  short = RT ,
-  long  = Retention time,
-  class = abbrev
-}
-
-\DeclareAcronym{MBR}{
-  short = MBR ,
-  long  = Match Between Runs,
-  class = abbrev
-}
-
-\DeclareAcronym{XIC}{
-  short = XIC ,
-  long  = Extracted Ion Current,
-  class = abbrev
-}
-
-\DeclareAcronym{SC}{
-  short = SC ,
-  long  = Spectral Counting,
-  class = abbrev
-}
-
-\DeclareAcronym{TPP}{
-  short = TPP ,
-  long  = Trans Proteomic Pipeline,
-  class = abbrev
-}
-
-%%  Begin document
-%%  ==================================================================
-\begin{document}
-
-%%  Begin title page
-%%  ==================================================================
-    \thispagestyle{empty}
-    \ULCornerWallPaper{1}{nat-farve.pdf}
-    \ULCornerWallPaper{1}{nat-en.pdf}
-    \begin{adjustwidth}{-3cm}{-1.5cm}
-    \vspace*{-1cm}
-    \textbf{\Huge Bioinformatics Master Thesis} \\
-    \vspace*{2.5cm} \\
-    \textbf{\Huge Development of label-free quantification \\ methods in proteomics} \\
-    \vspace*{.1cm} \\
-    %{\huge No subtitle is available} \\
-    \begin{tabbing}
-    % adjust the hspace below for the longest author name
-    Antonio Ortega Jiménez \hspace{1cm} \= \texttt{<ntoniohu@gmail.com>}
-    %Second Author \> \texttt{<second@author.com>} \\       
-    \\[10.5cm]
-
-    \textbf{\Large Supervisors} \\
-    Thomas Hamelryck \> \texttt{<thamelry@gmail.com>} \\
-    Mathias F. Gruber \> \texttt{<mafg@novozymes.com>} \\
-
-    \end{tabbing}
-    \end{adjustwidth}
-    \ClearWallPaper
-%%  ==================================================================
-%%  End title page
-
-%%%%%%%%%%%%% Indholdsfortegnelse %%%%%%%%%%%%%
-\setcounter{tocdepth}{2}
-\tableofcontents % Indholdsfortegnelse
-%\listoffigures   % Liste over figurer \begin{figure} ... \end{figure}
-%\listoftables    % Liste over tabeller \begin{table} ... \end{table}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newpage
-
-\thispagestyle{empty}
-\printacronyms[include-classes=abbrev,name=Abbreviations]
-
-\newpage
-\setcounter{page}{1}
-
-% https://tex.stackexchange.com/questions/93859/condense-the-space-between-bibliographic-entries
-\let\oldthebibliography\thebibliography
-\let\endoldthebibliography\endthebibliography
-\renewenvironment{thebibliography}[1]{
-  \begin{oldthebibliography}{#1}
-    \setlength{\itemsep}{0em}
-    \setlength{\parskip}{0em}
-}
-{
-  \end{oldthebibliography}
-}
-
-\section*{Preface}
-
-TODO: get nomenclature straight: what is a sample and what is a fraction
-
-\chapter*{Introduction}
-\label{chap:introduction}
-
-\section{Aminoacids and proteins}
-
-Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.
-
-Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
-The side chain differs between aminaocids and generates them from each other.
-A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.8\textwidth]{aminoacids2.png}
-  \caption{CAPTION AND REFERENCE}
-  \label{fig:aminoacids}
-\end{figure}
-
-Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
-The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
-
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
-One frequent instance of such modifications is the oxidation of methionine.
-
-\section{The protein-focused biotechnology industry}
-
-Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks.
-The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
-
-\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
-The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
-Notorious examples of such applications include waste-water treatment, household care and the baking industry.
-
-The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
-The refinement of the currently used tools and the development of new ones could be of great significance for the company.
-
-Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
-
-\section{Objectives of the Thesis}
-\label{sec:objectives}
-
-In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
-
-\begin{enumerate}
-
-\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
-
-\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
-
-\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
-
-\end{enumerate}
-
-\section{Structure of the Thesis}
-
-An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
-Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.
-
-\chapter{Mass spectrometry and shotgun proteomics overview}
-\label{chap:mass_spec}
-
-Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone.
-Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up.
-The layer provided by proteins is one of them, and its study is called proteomics.
-It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment.
-The study of the protein-specific complexity is called proteomics.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
-
-%CITE 1.6.1 COMPUTATIONAL METHODS.
-The existing approaches are divided into two types of paradigms: top-down and bottom-up.
-In the top-down paradigm, intact proteins are directly used for the analysis.
-In the bottom-up paradigm (see figure \ref{figure:proteomics_overview}, the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
-These smaller parts are called peptides.
-\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
-The top-down paradigm will be ignored in the rest of the manuscript.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{proteomics_skema_book}
-\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
-\label{fig:proteomics_overview}
-\end{figure}
-
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
-The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
-This is the quantification problem, explained in section \ref{sec:quantification}.
-
-A summary of the bottom-up approach MS analytical pipeline is provided in the rest of the chapter. It can be divided into two main steps:
-
-\begin{enumerate}
-
-\item \ac{MS} analysis and data generation. Sections \ref{sec:sample_processing} to \ref{sec:tandem_ms_workflow}.
-
-\item Computational analysis of data. Sections \ref{sec:search_engines} to \ref{sec:quantification}.
-
-\end{enumerate}
-
-\LARGE
-\textbf{MS analysis and data generation}
-\normalsize
-
-\section{Sample processing}
-\label{sec:sample_processing}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{sample_processing}
-\caption{Overview of the sample processing step prior to mass spectrometry analysis.
-First, proteins are denatured and digested with a specific protease like Trypsin.
-This yields a peptide mix that is separated into peptide fractions that can be introduced in the mass spectrometer.}
-\label{fig:sample_processing}
-\end{figure}
-
-\subsection{Protein digestion}
-\label{subsec:protein_digestion}
-
-An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task.
-It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
-
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
-%COMPUTATIONAL METHODS 3.2
-
-\begin{enumerate}
-
-\item Missed cleavages.
-
-\item Unsuspected cleavages during the maturation/life cycle of the protein.
-
-\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
-
-\item Naturally occurring, intentionally or unintentionally induced chemical modifications.
-
-\end{enumerate}
-
-%\Cref{item:missed_cleavages}
-
-Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
-Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
-
-The result of this process is a complex mix of peptides, made up by hundreds or thousands of different molecules, following a length distribution given by the cleavage sites frequency and each protein\textquotesingle s aminoacidic composition.
-A peptide separation step is required before introducing the sample in the spectrometer.
-
-\subsection{Peptide separation}
-\label{subsec:peptide_separation}
-
-If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
-This can be surmounted by independently analyzing different fractions of the sample over time.
-The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
-
-%CITE 4.2 computational methods.
-During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
-
-\section{The mass spectrometer}
-\label{sec:the_mass_spectrometer}
-
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
-It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
-\label{fig:mass_spectrometer}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-\subsection{The ion source}
-\label{subsec:the_ion_source}
-
-All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
-
-%CITE 5.1 COMPUTATIONAL METHODS
-
-This step is performed in the ion source \cite{Barsnes2008}.
-The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
-Thus, the charge exhibited by an ion is not obvious when produced via ESI.
-Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
-Bias in ESI is less predictable.
-This is known as the competitive ionization problem \cite{Tang2004}.
-
-The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
-
-\subsection{The mass analyzer}
-\label{subsec:the_mass_analyzer}
-
-The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
-These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.
-
-Moreover, two other analyzers exist which combine mass analysis with intensity measurement.
-These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
-They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
-
-\subsection{The detector}
-\label{subsec:the_detector}
-
-Detectors measure the intensity of an incoming ion signal.
-The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
-Some topics in signal detection in \ac{MS} need to be discussed.
-
-On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
-The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
-This is one of the reasons why proteins are better fit for analysis when digested into peptides, as \ac{m/z} are reduced, thus increasing the mass resolution.
-
-On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
-Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
-Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
-Such event can be problematic if it entails that the \ce{^{13}_{}C} peak is confused for the \ce{^{12}_{}C} peak.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{envelope}
-\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
-Each peak in the envelope is separated by 0.5 Th.
-This is explained by the peptide having 2 positive charges that make every extra Da in the ion mass account for 1/2 extra Th.
-Adapted from \cite{Mirzaei2016}.}
-\label{fig:envelope}
-\end{figure}
-
-The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
-
-It is up to the \ac{MS} technician to decide on the best pieces of equipment according to their availability and particularities of the dataset.
-
-\section{Tandem MS workflow}
-\label{sec:tandem_ms_workflow}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{tandem_ms}
-\caption[]{Illustration of the tandem MS workflow.
-The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
-The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
-Finally the third spectrometer records fragment mass spectra.
-Taken from \footnotemark{}.}
-\label{fig:tandem_ms}
-\end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
-
-Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
-In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
-An extra spectrometer, usually a quadrupole, is introduced in between.
-
-\begin{itemize}
-
-\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
-
-\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
-
-\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
-
-\end{itemize}
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{shotgun_adapted}
-\caption{Illustration of the different spectra collected in tandem MS.
-MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
-MS2 spectra record the same magnitudes but the signal is generated by the fragments produced during fragmentation by the ion filtered in the first spectrometer.
-Altogether, they enable peptide identifications.
-Figure adapted from \cite{Verheggen2017}.}
-\label{fig:shotgun}
-\end{figure}
-
-\subsection{Fragmentation}
-\label{subsec:fragmentation}
-
-The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
-Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{abcxyz}
-\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
-The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
-Specific fragmentation techniques make fragments belonging to one series more likely than others.
-Other fragments are possible but much less likely.
-The fragment nomenclature was introduced in \cite{Roepstorff1984}.}
-\label{fig:abcxyz}
-\end{figure}
-
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
-During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
-These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
-The difference can be exploited to infer which aminoacid is present in the longer fragment.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
-explained in \ref{sec:search_engines}.
-
-%A given bond will be more likely to break the less stable it is.
-This makes (I) the \ce{C_$\alpha$}-\ce{CO}, (II) the peptidic \ce{CO}-\ce{NH}, and (III) the \ce{C_$\alpha$}-\ce{NH} bonds the most likely to break (see figure \ref{fig:abcxyz}).
-A nomenclature REF was introduced to name these fragments:
-
-%%PAG 123 COMPUTATIONAL METHODS
-%Fragmentation in proteomics is performed via (I) collision-induced (CID) or (II) electron-induced (EID) dissociation.
-CID is an ergodic fragmentation technique where peptides enter a collision cell containing an inert gas.
-Given enough kinetic energy, hits between ionized peptides and the gas will trigger the fragmentation of the peptide into smaller units.
-\cite{Barsnes2008}.
-Since the kinetic energy is randomly distributed across the peptide, the weakest bonds will break first.
-This results in the production of mostly b and y fragments, as well as the loss of any chemical modifications \cite{Barsnes2008}.
-%
-%%CITE pag 134 computational methods
-%
-%EID is produced by the hits against the molecule, which tend to occur on the areas of the peptide more positively charged.
-Thus, the fragmentation process is not ergodic and returns TYPE ions.
-More importantly, chemical modifications are not lost in the process.
-
-\LARGE
-\textbf{Computational analysis}
-\normalsize
-
-\section{Spectra processing: search engines}
-\label{sec:search_engines}
-
-Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
-\ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{computational_analysis}
-\caption{Diagram of the MS computational analysis.
-The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
-In other words, the software infers what peptides were introduced in the spectrometer.
-The analysis continues during protein inference and quantification.}
-\label{fig:computational_analysis}
-\end{figure}
-
-%\begin{figure}[!h]
-%\includegraphics[width=\textwidth]{PSM}
-%\caption{}
-%\label{fig:PSM}
-%\end{figure}
-
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
-Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
-This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
-Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
-Under this circumstances, if repeated for several peptides, enough evidence for the presence of individual proteins can be collected.
-
-Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
-Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
-
-\section{Validation and quality control}
-\label{sec:validation}
-
-The scoring systems implemented in search engines provide the best matches, but they are bound to contain false identifications.
-Nevertheless, these scores can be used to apply a filter that aims at minimizing the amount of errors.
-
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
-
-The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
-Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
-
-All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
-This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
-Together with the number of PSMs passing a given score in the target ($n_{tp} + n_{tp}$), the FDR can be computed using equation \ref{eq:fdr}.
-
-\begin{equation}\label{eq:fdr}
-FDR = \frac{\hat{n}_{fp}}{n_{fp} + n_{tp}}
-\end{equation}
-
-The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
-Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{pep}
-\caption{Visualizing FDR and PEP.
-The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
-The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
-Figure from \cite{Kall2008}.}
-\label{figure:pep}
-\end{figure}
-
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
-Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
-The PSM\textquotesingle s confidence is just defined as $1-PEP$ \cite{Nesvizhskii2010}.
-PEP can be computed from the decoy search results and provides another useful measurement of the uncertainty in the target results.
-
-\section{Peptide and protein inference}
-\label{sec:inference}
-
-Two steps in protein identification can be distinguished:
-
-\begin{enumerate}
-
-\item \textbf{Peptide inference}: infer the peptides present in the sample.
-\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and frequently map to more than one protein.
-\end{enumerate}
-
-The result of the PSM step returns the inferred list of peptides.
-The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
-The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
-Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
-
-\section{Protein quantification}
-\label{sec:quantification}
-
-The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
-However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
-For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
-
-Protein quantification pipelines can be classified based on whether isobaric labelling was used (label-based) or not (label-free).
-These are explained in subsection \ref{subsec:labelling}.
-If the label-free approach is employed, more distinctions can be made based on:
-
-\begin{itemize}
-\item The proxy used for quantification: spectral counting (\ac{SC}) or extracted ion currents (\ac{XIC}). These are explained in \ref{subsec:scvsxic}
-
-\item The way the data are brought to the protein level from the peptide level: summarization-based vs. peptide-based, explained in \ref{subsec:peptide_model}.
-\end{itemize}
-
-\subsection{Label-based and label-free approaches}
-\label{subsec:labelling}
-
-%CITE PAGE 237 COMPUTATIONAL METHODS
-
-Two paradigms exist in protein quantification: label-based and label-free.
-In label-based quantification, originally identical peptides from a number of different samples are made distinguishable by their masses via the incorporation of a label.
-All label-based methods simultaneously analyze several samples in each experiment, removing the difficulties associated with between-run variability \cite{Barsnes2008}.
-The finite number of "plexes" available for a given label sets the limit to how many samples can be differentially quantified \cite{Cox2014}.
-Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
-Remarkably, peptide labeling costs can be quite high.
-This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
-
-In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
-In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
-
-\begin{figure}[!h]
-\centering
-%\begin{subfigure}{.45\textwidth}
-%    \caption*{A}
-\includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption{Illustration of the \ac{MBR} step.
-\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_mbr}
-\end{figure}
-
-This way, extra identifications are attained through a reanalysis of the spectra, factoring in the information collected in replicate runs.
-The RT and precursor mass of unidentified spectra in one replicate is matched to that of identified spectra in the remaining replicates.
-As a consequence, more identifications with a lesser fraction of missing datapoints are achieved.
-
-\subsection{SC and XIC based quantification}
-\label{subsec:scvsxic}
-
-Quantification can be spectral counting or \ac{XIC} based.
-
-On the one hand, spectrum counting based quantification is the simplest method in proteomics.
-It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing
-
-On the other hand, \ac{XIC} based methods rely on intensity measurements at any level of the \ac{MS} workflow as proxies for protein abundance.
-A wide range of algorithms are available to process these data and output estimates of protein abundance.
-All of them require an intensive preprocessing step, usually including (I) taking the $log_2$ intensity to make the data distributions symmetrical and thus make it fit for diverse parametric tests, and (II) quantile normalization to address between-runs variability in the intensity measurements.
-They can be classified in the bases of which MS level is used as proxy for the protein abundances and on whether or not a summarization step is performed to aggregate peptide-level data into protein-level data, or not.
-
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to spectral counting given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
-For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
-
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{A}
-\includegraphics[width=0.7\linewidth]{apex_3d}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{1\textwidth}
-\centering
-\caption*{B}
-\includegraphics[width=0.7\linewidth]{apex_intensity}
-\end{subfigure}
-\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
-An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
-Taken from \cite{Smith2014}.
-\textbf{B} A refinement analysis of these data attempts to fit a mathematical model of the signal over time and extract a representative measurement, such as the highest (apex) \ac{MS1} intensity.
-Modified from supplementary information in \cite{Argentini2016}.}
-\label{figure:moff_apex}
-\end{figure}
-
-\subsection{\ac{XIC}-peptide-based models for label-free quantification}
-\label{subsec:peptide_model}
-
-The data collected in the mass spectrometer refers to peptides originating from a latent protein composition, given by the original sample.
-However, the data interpretation requires the transfer of these peptide-level data into the protein level.
-This can be done by either (I) performing an aggregation of the peptide-level data, where a summary value of the peptide-level data is taken as representative for the protein-level data, or (II) performing the protein quantification directly at the peptide-level by means of linear regression models.
-
-As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
-However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
-This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
-
-It is for this reason that peptide-based models offer the statistical framework required to learn as much from the data as possible.
-This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
-
-\chapter{A label-free quantification proteomics pipeline}
-\label{chap:pipeline}
-
-\section*{Summary}
-
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI, PeptideShaker, moFF and MSqRob, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) MBR and feature extraction, (IV) protein inference and (V) relative quantification.
-Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
-A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
-
-\section{Introduction}
-
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like MGF.
-After that, spectra are searched by means of a search engine to perform the PSM step.
-Thereafter, PSMs are validated and the most likely set proteins is inferred.
-If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
-The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
-\label{fig:pipeline}
-\end{figure}
-
-Several proteomics pipelines are available on the internet under different licensing conditions.
-Many are released as closed-source software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community.
-MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-
-However, still only few of these tools have a fully Linux-supported, well-documented, command-line version available, which would make customised, automatic streamline analyses much easier to perform.
-As a GUI would not be required to run the pipeline, it would also be scalable to big datasets.
-Finally, having free licenses would imply that anyone, provided the technical knowledge, can perform the computational analysis independently, thus speeding the data to knowledge turnover in \ac{NZ} or any organization making use of it.
-
-The development of a pipeline attempting to achieve these goals will be described in this chapter.
-
-\section{Materials and Methods}
-
-\subsection{Data generation and loading}
-
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
-Briefly, the \textit{Homo sapiens} and \textit{E.
-coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
-This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
-One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the open format MGF (Mascot Generic Format).
-
-The MQ+MaxLFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
-The results of the MQ+MSqRob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
-
-%\href{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}{http://www.mcponline.org/content/suppl/2014/06/17/M113.031591.DC1/mcp.M113.031591-1.xlsx}.
-
-\subsection{Decoy database preparation and search}
-
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine CITE by means of the SearchGUI \texttt{SearchCLI} tool CITE utility.
-The search parameters were set using the \texttt{IdentificationParametersCLI}.
-In order to account for potential post-translational modifications, the search was conducted allowing for the following variable modifications: oxidation of M and deamidation of N and Q.
-Moreover, C carbamidomethylation was set as fixed modification.
-The enzyme was set to semispecific Trypsin, allowing for a non-tryptic cleavage on any side of the peptide.
-Up to two missed cleavages were allowed.
-The precursor tolerance was 10 ppm and the fragment tolerance 0.5 Da.
-
-The target database was created by combining the Uniprot proteomes for \textit{E.
-coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
-The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
-
-\subsection{Quality control and validation}
-
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
-By default, the FDR was set to 1\%.
-PEP and confidence statistics were computed using the PeptideShaker built-in algorithms.
-Output was extracted via the Default PSM report txt file, available in the \texttt{ReportCLI} utility.
-
-\subsection{Data refinement}
-
-The moFF command line utility CITE was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
-This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
-
-\subsection{Quantification}
-
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
-Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
-In a nutshell, (I) MS1 apex intensities were $log_2$ transformed, (II) quantile normalized, (III) peptides belonging to protein groups that contained one or more proteins that were also present in a smaller protein group were discarded \cite{Goeminne2016}, and (IV) protein groups with only 1 peptide were dropped.
-
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
-The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
-The significance of the treatment effect differences was assessed through a Student\textquotesingle s T test, implemented in the \texttt{test.contrast()} function.
-Only protein groups that could be unambiguously mapped to one of the organisms were considered for the analysis.
-
-Quantification in the MQ+MaxLFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
-Intensities were log2 transformed and averaged.
-The difference between conditions H and L was taken as estimate of the log2FC.
-Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
-P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
-
-\subsection{Code implementation}
-
-\section{Results}
-
-\subsection{Thousands of spectra were identified and validated}
-
-The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
-The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
-
-\begin{figure}[!h]
-\centering
-\begin{subfigure}{.45\textwidth}
-  \centering
-  \caption*{A}
-  \includegraphics[width=.9\linewidth]{target_vs_decoy}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{B}
-  \includegraphics[width=.9\linewidth]{score_vs_confidence}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{C}
-  \includegraphics[width=.95\linewidth]{qc}
-\end{subfigure}
-\begin{subfigure}{.45\textwidth}
-  \centering
-    \caption*{D}
-  \includegraphics[width=.95\linewidth]{roc}
-\end{subfigure}
-\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
-\textbf{A} Score distribution for matches to the decoy and the target databases.
-\textbf{B} Evolution of the PSM score with confidence.
-The implemented cutoff at FDR of 1 \% is displayed with a red vertical line.
-\textbf{C} Distribution of the difference between the predicted and measured \ac{m/z} values, segregated by validation status.
-\textbf{D} ROC curve built upon the number of false positives and negatives estimated from the decoy search.
-The cutoff is displayed as a red dot.}
-\label{figure:qc_validation}
-\end{figure}
-
- The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
-
-The combination of both programs enabled the identification and validation (matching) of thousands of spectra with high confidence in all samples.
-However, when compared to the total amount of spectra available, the percentage of matched spectra was on average 37.4\%, with a marked decrease starting at fraction 14 (see figure \ref{fig:match_percent}).
-
-\begin{figure}[!h]
-\includegraphics[width=0.8\textwidth]{match_percent}
-\caption{Percentage of matched spectra in all samples.
-The total number of spectra per sample ranged between 5894 and 20249.
-\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
-The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
-\label{fig:match_percent}
-\end{figure}
-
-\subsection{Protein inference}
-
-\begin{table}[H]
-\begin{tabular}{ccccc}
-  \toprule
- Total peptides & Unique peptides & Non-unique & Proteins & Protein groups \\ 
-  \midrule
- 46595 & 27384 & 19211 & 25056 & 10039 \\
- \bottomrule
-\end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
-\label{tab:quantification_table}
-\end{table}
-
-A total of 46595 peptides were inferred to be present in at least one of the fractions.
-Of them more than 27 thousand mapped to a unique protein.
-However, a significant amount mapped to more than one protein.
-Such peptides are called non-unique peptides.
-The 25056 detected proteins were grouped into 10039 protein groups.
-Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inferene algorithms create them when a does not uniquely map to a protein becuase it is not possible to distinguish which protein it truly comes from.
-Protein grouping decreased the amount of  considered protein entities to 10039.
-
-Protein groups are very frequent due to several factors.
-For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
-PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
-
-However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.
-
-\subsection{MBR and apex intensity extraction quantification}
-
-The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
-The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
-
-\begin{figure}[H]
-\centering
-\includegraphics[width=\textwidth]{mbr_combined}
-\caption{Match Between Runs with moFF.
-\textbf{A} Count of identifications on each run segreagated by source.
-More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.
-\textbf{B} Match visualization.
-Every dot represents a peptide shared across 2 runs.
-The coordinate system illustrates its retention time on the first run on the x axis and the difference with the second run on the y axis.
-The color depicts whether the identification was carried out during the PSM process (MS1), or thanks to a cross-identification achieved by the MBR module (MBR).}
-\label{fig:mbr}
-\end{figure}
-
-Once as many identifications as possible were gathered, a refinement of the measured MS1 intensity can be implemented to select the apex of every peak cluster, which yields robust intensity measurements for each sample (see figure \ref{fig:apex_intensity}).
-
-\begin{figure}[!h]
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{A}
-\includegraphics[width=.9\linewidth]{peptide_profile}
-\end{subfigure}
-\bigskip
-
-\begin{subfigure}{.9\textwidth}
-  \centering
-    \caption*{B}
-\includegraphics[width=.9\linewidth]{density_ratio}
-\end{subfigure}
-\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
-coli}.
-The figure illustrates the intrinsic intensity variability between technical replicates, particularly in the case of the \textit{E.
-coli} QLLMMSGFDR peptide, as it was almost non-existent in one of the runs.
-\textbf{B} The expected pattern of overall similar intensities for the \textit{Homo sapiens} data and 3-fold higher intensities for the \textit{E.
-coli} data in condition H was observed, confirming an acceptable performance of the protocol.}
-\label{fig:apex_intensity}
-\end{figure}
-
-\subsection{Quantification}
-\label{subsec:quantification}
-
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} across condition H and condition L.
-During quantification, it is important to take into account as many of the effects influencing protein quantification.
-In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
-The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
-Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
-
-The result of the process was the successful quantification of 5307 protein groups out of 10039 (see table \ref{tab:quantification_table}).
-
-% latex table generated in R 3.4.4 by xtable 1.8-2 package
-% Sun Jul 15 18:17:23 2018
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll>{\itshape}l}
-  \toprule
- & log2FC & qval & Protein & Organism \\ 
-  \midrule
-1 & -4.76E-01 & 1.42E-24 & P78527 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}2 & 6.80E-01 & 6.75E-24 & P0A8V2 & E. coli \\ 
-  3 & 8.50E-01 & 4.39E-22 & P25516 & E. coli \\ 
-   \rowcolor[gray]{0.95}4 & 7.07E-01 & 2.05E-18 & P63284 & E. coli \\ 
-  5 & 1.04E+00 & 1.21E-17 & P37095 & E. coli \\ 
-   \rowcolor[gray]{0.95}6 & 7.78E-01 & 1.51E-16 & P13029 & E. coli \\ 
-  7 & 8.89E-01 & 5.62E-16 & P77804 & E. coli \\ 
-   \rowcolor[gray]{0.95}8 & 8.53E-01 & 1.72E-15 & P23721 & E. coli \\ 
-  9 & -6.97E-01 & 1.41E-14 & P09874 & Homo sapiens \\ 
-   \rowcolor[gray]{0.95}10 & 1.37E+00 & 1.41E-14 & P37666 & E. coli \\ 
-   \bottomrule
-\end{tabular}
-\caption{Results of the quantification pipeline.
-The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
-Notably, eight correspond to proteins belonging to \textit{E.coli}.
-The two human proteins exhibited in both cases an absolute value of the log2FC below 1.}
-\label{tab:quantification_table}
-\end{table}
-
-The performance of the quantification can be evaluated thanks to the experimental design of the dataset.
-Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a log2FC of $log2(3)=1.58$ is expected for proteins coming from this organism.
-Likewise, human proteins should have a log2FC of 0.
-As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
-
-Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
-Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
-\textbf{Compomics+MSqRob} is the pipeline presented in the previous sections.
-\textbf{MQ+MaxLFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+MSqRob blends MaxQuant upstream analysis with the MSqRob quantification engine.
-\textbf{A, B} ROC curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.
-\textbf{C, D} PR curve produced by the three algorithms with no log2FC filter, and keeping proteins with log2FC greater than 1, respectively.}
-\label{fig:roc_curves}
-\end{figure}
-
-In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
-Thus, \textit{E.
-coli} proteins are treated as positives, and those from \textit{Homo sapiens} are treated as negatives.
-Moreover, a filter based on the log2FC can be applied to enrich the data in positives and facilitate the classification task.
-In order to compare the results of the here presented pipeline (Compomics+MSqRob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+MaxLFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+MSqRob) used MaxQuant up to the fraction normalization step.
-
-The result of this analysis is shown in figure \ref{fig:roc_curves}).
-Remarkably, Compomics+MSqRob achieved similar results to those resulting from MQ+MaxLFQ.
-Nonetheless, the latter performed clearly better when the log2FC filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+MSqRob was as good as MQ+MaxLFQ, if not much better.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.8\textwidth]{vennDiagram}
-\caption{Agreement between the different pipelines. The Venn diagram displays the number of \textit{E. coli} proteins found to be differentially abundant by each pipeline.}
-\label{fig:venn}
-\end{figure}
-
-In order to visualize the level of agreement of the pipelines for the assignment of differentially abundant proteins, a Venn diagram was constructed (see figure \ref{fig:venn}).
-Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Compomics+MSqRob or MQ+MaxLFQ being detectd by MQ+MSqRob too.
-Remarkably, 517 proteins were detected by MQ+MSqRob alone, which further confirmed its resolving power.
-
-Additionally, the accuracy with which each program assigns a log2FC estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
-This analysis confirmed that the pipeline providing the best separation between organisms was MQ+MSqRob.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines performance.
-Quantification results are segregated by the source organism.
-\textbf{A} Histogram of the log2FC estimates.
-\textbf{B} Volcano plot, where every protein group is represented by a dot and the coordinates map to its log2FC on x and the minus logarithm of the q-value on y.
-\textbf{C} Boxplots of the minus logarithm of the q-value distribution.}
-\label{fig:combined_plot}
-\end{figure}
-
-\section{Discussion}
-
-\subsection{Improvement of the PSM step}
-
-The low attained matching rate manifests existing room for improvement in the currently available tools.
-Remarkably, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the framework \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
-Their usage is predicted tor improve identification rates and quality of the matches.
-
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
-New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
-Moreover, the prediction of \ac{MS2} peak intensities patterns from peptide sequences promises to increase the amount of evidence available during the PSM process, thus boosting correct identifications \cite{Kirik2018} \cite{Degroeve2013}.
-
-Finally, the extremely low matching rates in the latter fractions translated to decreased contributions to the number of identifications.
-This indicated that decreasing the number of fractions would not have had a major impact in the experiment\textquotesingle s depth.
-
-\subsection{Improvement of the feature extraction step}
-
-While the apex intensity is a good estimate of the peptide abundance, other methods are available that could potentially provide more robust data, such as the area of the peak.
-This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
-
-\subsection{Improvement of the quantification step}
-
-The deployed relative quantification approach successfully manages to estimate the log2FC of several of proteins in the dataset with high confidence.
-However, it was not found to be the best performing of the assayed pipelines in the benchmark dataset.
-Several axes of improvement are clear from the results.
-
-\subsubsection{Management of sample fractionation}
-
-Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
-
-The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
-One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
-A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
-
-MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at log2FC of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
-While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Compomics+MSqRob, proving room for improvement in this aspect.
-
-In any case, the way sample fractionation is handled should depend upon the fractionation method.
-For example, membrane fractionation can be exploited when working with membrane proteins \cite{Marmagne2006}.
-
-To sum up, the presently implemented fraction handling is correct and provides good results, but fine-tuning of the pipeline at this step is expected to improve the results of the quantification.
-Thanks to the open-source nature of the software, this is indeed possible.
-
-%which is defined according to the following expressions:
-%
-%\begin{equation}
-%I_{P,A}(\text{N}) = \sum\limits_{j=1}^{k}{\text{N}_{A,j} \times \text{XIC}_{P,A,j}}
-%\end{equation}
-%
-%where $\text{N}_{A,j}$ stands for the normalization coefficient for sample $A$ and fraction $j$, and $\text{XIC}_{P,A,j}$ stands for the extracted ion current in the same sample and fraction for peptide P.
-The sum runs over all k fractions.
-%
-%\begin{equation}
-%H_P(N) = \sum\limits_{A,B} \abs{\log(\frac{I_{P,A}(N)}{I_{P,B}(N)})}^2
-%\end{equation}
-%
-%\begin{equation}
-%H(N) = \sum\limits_{p} H_p(N) 
-%\end{equation}
-%
-%where the pair $A$ and $B$ is any of the sample pairs where peptide $P$ is detected. The overall proteome variation is
-
-\subsubsection{The pitfalls of mass spectrometry}
-
-The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
-One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
-Notwithstanding, it reveals the immature state of the technology.
-Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
-Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
-Indeed, the input file for the MQ+MaxLFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn}).
-
-\subsubsection{Support for absolute quantification}
-
-While relative quantification like that performed by MSqRob is enough to infer differential abundance of proteins across conditions, absolute quantification would provide an estimate of protein quantities that would support many other analyses, and facilitate comparison between datasets, in a way similar to what the normalised read counts in FPKM (Fragments Per Kilobase per Million Reads) does in transcriptomics.
-On the other hand, MaxLFQ provides absolute estimates, albeit less robust.
-
-\subsubsection{Quantification of uncertainty}
-
-All the quantification approaches enunciated until now make use of different frequentist approaches to the problem, that evaluate how significantly different the data are from what would be expected under a null hypothesis of a null ratio across the pair of conditions.
-Yet, it does not provide probabilistic interpretations of the uncertainty behind the estimate.
-The development of a Bayesian framework which endeavors at solving this issue is presented in chapter \ref{chap:model}.
-
-\section{Conclusion}
-
-An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
-The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
-Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
-Likewise, the openness of the data in all steps allows for quality control checks along the way.
-However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
-
-\chapter{BayesQuant: Probabilistic estimation of protein ratios}
-\label{chap:model}
-
-\section*{Summary}
-
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
-However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
-
-\section{Introduction}
-
-Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
-Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
-The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (NHST).
-
-\begin{align}
-H_0: & \log_2(FC) = 0 \nonumber \\
-H_1: & \log_2(FC) \neq 0 \nonumber
-\end{align}
-
-The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the $\log_2$(FC) between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
-The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
-
-The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
-While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes-theorem.
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\end{figure}
-
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-
 \chapter{Pipeline benchmarking on \ac{NZ} data}
@@
 It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
-
-\chapter*{Conclusion}
-\label{chap:conclusion}
-
-\chapter*{Appendix}
-\label{chap:appendix}
-
-\includepdf[pages=-]{appendix.pdf}
-
-\bibliographystyle{unsrt_etal}
-\footnotesize{\bibliography{bibliography}}
-
-\end{document}
-%%  ==================================================================
-%%  End document

### 2018-07-17 modelling.tex
Organize chapters in separate files and extend methods of chap 3

@@
 
-\subsection{Sequence modelling}
+The final state of the data is the following:
+
+\begin{table}
+\begin{tabular}{lllrrrrrr}
+\toprule
+{} & protein &    organism &     H1 &     H2 &     H3 &     L1 &     L2 &     L3 \\
+\midrule
+     &  P0A8I8 &  E. coli &  25.13 &  25.24 &  24.39 &  21.71 &  22.67 &  22.40 \\
+     &  P0A8I8 &  E. coli &  21.49 &  23.10 &  23.38 &  21.34 &  22.65 &  21.25 \\
+     &  P0A8I8 &  E. coli &  24.10 &  24.54 &  23.81 &  19.88 &  20.87 &  20.30 \\
+\bottomrule
+\end{tabular}
+\caption{Sample data input for the Bayesian model.
+Every row represents a unique peptide.
+The first column refers to the protein it was found to map to in the protein inference step.
+The second column is an annotation field, in this case indicating the protein\textquotesingle s organism.
+The remaining columns indicate the $log_2(Intensity)$ registered for each peptide in the corresponding run.
+In this case, three peptides were observed for the protein with id P0A8I8.}
+\end{table}
+
+\subsection{Sequence feature extraction}
 
@@
 
-\subsection{Bayesian modelling using Pymc3}
+\subsection{Hierarchical model}
 
-In order to get started with Pymc3, we first need to import it.
+The \ac{MS1} intensity measurements are affected by two main sources of variation:
+
+\begin{enumerate}
+\item The fixed effect that the researcher aims at unraveling with proteomics (treatment effect)
+\item Random noise produced by experimental procedures, sequence bias, etc, which lead to a loss of data quality and resolving power.
+\end{enumerate}
+
+The mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect while minimising the remaining unwanted effects.
+The mathematical framework implemented to model these effects was similar to the one used by MSqRob:
+
+\begin{equation}\label{eq:model}
+y_{ijkl} = \beta_{ijkl}^{0} + \beta_{ij}^{treatment} + \beta_{ik}^{peptide} + \beta_{il}^{run} + \epsilon_{i}
+\end{equation}
+
+where $y_{ijkl}$ stands for the $\log_2$-transformed intensity registered in treatment $j$, peptide $k$ and run $l$.
+This way, separate treatment, peptide and run effects are distinguished.
+$i$ stands for the protein index, which is evidently the same for all measurements belonging to peptides mapped to the same protein.
+$\epsilon$ models the noise that cannot be explained with the three effects aforementioned, and is the same for all measurements associated to the same protein.
+
+For example, the measurement $y_{P,A,p,1}$ corresponds to a measurement mapped to protein $P$.
+The measurement is impacted by a fixed effect, produced by treatment $A$, but also the peptide effect in peptide $p$ and the run effect in run $1$.
+An intercept term was also added to account for the starting intensity value.
+
+The data is thus affected by different effects acting in parallel, and in order to model them properly, multilevels need to be defined.
+Multilevel effects can be modelled with three approaches: pooled, unpooled, and with hierarchies (see figure \ref{fig:multilevel}).
+Hierarchical modeling is used when information is available on several different levels of observational units \cite{gelman2013bayesian}.
+The three effects were modeled as independent hierarchies.
+
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Pooled}
+\includegraphics[width=0.85\linewidth]{model_pooled}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Unpooled}
+\includegraphics[width=0.85\linewidth]{model_unpooled}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Hierarchical}
+\includegraphics[width=0.85\linewidth]{model_hierarchical}
+\end{subfigure}
+\caption[]{Alternative multilevel modelling approaches.
+A pooled model assumes that the parameter distribution is the same for all the datapoints.
+The unpooled model represents the opposite case, there every datapoint is assumed to be modelled by a different probability distribution of the parameter for each.
+Finally, a hierarchical model settles for a middle ground where the parameters will not be exactly the same but not completely different either.
+This is achieved by generating a global distribution from which the parameter for each datapoint is sampled \footnotemark{}.}
+\label{fig:multilevel}
+\end{figure}
+
+\footnotetext{\href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
+
+This way, common probability distributions were set as priors for the intercept, treatment, peptide and run effect.
+The particular effect observed on each peptide was then modelled as a value sampled from the corresponding probability distribution.
+A diagram of the resulting model is displayed in figure \ref{fig:daft_model}).
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{graph}
+\caption{Diagram of the Bayesian model developed in the present work, without sequence modelling of the peptide effect.
+It is represented as a Directed Acyclic Graph (\ac{DAG}) where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
+The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to.
+The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.
+The model hierarchies are organized in boxes.
+The one corresponding to the treatment effect can be read as \textit{the treatment effect $\beta_{ij}^{t}$ observed in the data is modelled as a probability distribution conditional on the value of $\mu^t$ and $\sigma^t$, which are in turn probability distributions governed by the hyperparameters 0 and 1.}}
+\label{fig:daft_model}
+\end{figure}
+
+\subsection{Prior probability distribution specification}
+
+Both because of the little knowledge on the value of the parameters governing the data generation process and to provide a prior that everyone can agree upon, non informative priors were provided to the model.
+This was condensed in the specification of Normal and Half Normal distributions for all the random stochastic variables in the model.
+The hyperparameters were selected to be as non informative as possible.
+
+\subsection{Approximation to the posterior probability distribution}
+
+The posterior probability distribution can be approximated by means of \ac{MCMC} methods, which can be run with the power of modern computing.
+The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} have given way to the much more efficient sampler \ac{NUTS} \cite{Hoffman2011}.
+\ac{MCMC} methods simulate sampling from the posterior distribution by randomly exploring the parameter space with some heuristics that guarantee convergence with the true posterior given a big enough sample size.
+Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018} \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
+
+\ac{ADVI} methods, very recently developed and introduced in PyMC3, attempt to solve the time and computational cost requirements characteristic of \ac{MCMC} methods by performing an approximation.
+The approximation consists of the selection of the best performing distribution $q$ within a family of distributions $Q$ that are computationally easy to work with.
+Thus the problem becomes an optimisation problem, where the inference algorithm needs to maximise the performance of the approximate function.
+Several specific methods can do this, but the mean-field implementation  is the default in PyMC3.
+
+Both NUTS and ADVI-powered approximations to the posterior were implemented in the program here presented using PyMC3.
+
+\subsection{PyMC3 implementation}
+
+In order to get started with PyMC3, we first need to import it.
 
@@
                autogobble=true,
-               %frame=lines,
+               frame=lines,
                framesep=2mm]{python}
@@
 
-The code below formalizes the bias in \ac{MS1} intensity measurements due to a random effect in Pymc3 code.
+The code below formalizes the bias in \ac{MS1} intensity measurements due to a random effect in PyMC3 code.
 
@@
                firstnumber=last,
+               frame=lines,
                framesep=2mm]{python}
@@
 \sigma \sim \mathcal{HN}(1)\,.  \\
-\mu \sim \mathcal{N}(0,\,\sigma^{2})\,.
+\mu \sim \mathcal{N}(0,\,\sigma)\,.
 \end{align}
@@
 and can be read aloud as \textit{the prior probability distribution of the random effect follows a normal distribution with mean 0 and standard deviation $\sigma$.
-In turn, $\sigma$\textquotesingle s prior probability distribution follows a half normal distribution with standard deviation 1.}
+In turn, $\sigma$\textquotesingle s prior probability distribution follows a half normal distribution with standard deviation 1}.
+1 and 0 act as hyperparameters of the model, and introduce the state of beliefs or knowledge on the system prior to seeing the data.
 
-\textit{hola}
+The hierarchical structure of the model is set by the definition of a parameter distribution from which the value for each element (peptide) being modelled is sampled.
+This can be done by defining a new normal distribution with $\mu$ and $\sigma$ set to the random variables defined above.
+Thus, the following was implemented,
 
@@
                firstnumber=last,
+               frame=lines,
                framesep=2mm]{python}
-effect = pm.Normal("pep", mu, sigma, n_elements)  
+         beta = pm.Normal("beta", mu, sigma, n_elements)  
 \end{minted}
 
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\caption{Diagram of the Bayesian model developed in the present paper, without sequence modelling of the peptide effect.
-It is represented as a Directed Acyclic Graph \ac{DAG} where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
-The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to.
-The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.}
-\label{fig:daft_model}
-\end{figure}
+which is equivalent to:
+
+\begin{equation}
+\beta_i \sim \mathcal{N}(\mu, \sigma)\,.  \\
+\end{equation}
+
+and can be read as \textit{the bias observed in the $i^{th}$ element due to the effect here modelled is said to follow a normal distribution with mean $\mu$ and standard deviation $\sigma$ both defined as random probability distributions above.}
+
+Finally, the equation \ref{eq:model} defined above closes the model and binds the data to the model parameters. Its PyMC3 implementation is the following:
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               firstnumber=last,
+               frame=lines,
+               framesep=2mm]{python}
+
+         epsilon = pm.HalfNormal('epsilon', 1)                
+         m = pm.Deterministic("m", beta0 + beta)
+         obs = pm.Normal("obs", m, epsilon, observed=y)
+
+\end{minted}
+
+which is equivalent to
+
+\begin{align}
+\nonumber \epsilon \sim \mathcal{HN}(1) \,. \\ 
+m = \beta^0 + \sum_{i=1} \beta_i \,. \\ 
+\nonumber y \sim \mathcal{N}(m, \epsilon)
+\end{align}
+
+and is read as \textit{the observed data is modelled as a normal distribution with mean $m$ and standard deviation $\epsilon$.
+$m$ is set defined as a random deterministic variable resulting from the sum of the intercept and the effects defined above.
+$\epsilon$ follows a new half normal distribution with standard deviation 1}.
+A deterministic variable is a random variable that acquires a fixed value if all random variables it has a dependency on take a fixed value too, i.e its stochasticity disappears if its parameters are fixed.
 

### 2018-07-17 pipeline.tex
Organize chapters in separate files and extend methods of chap 3

@@
 
-\subsection{MBR and apex intensity extraction quantification}
+\subsection{MBR and apex intensity extraction evaluation}
 
@@
 This is the approach taken by the recently published tool RawQuant \cite{Kovalchik2018}, or the feature detection tool in the OpenMS pipeline \cite{Sturm2008}.
+Moreover, in order to fine tune the performance of the \ac{MBR} and Apex intensity extraction steps, several input arguments can be adapted to each dataset.
+These are the spectrometer\textquotesingle s \ac{m/z} resolution, the time window used during the apex search and whether or not to activate outlier filtering and what weights to use.
 

### 2018-07-18 modelling.tex
Extend chapter 3

@@
+\documentclass[thesis]{subfiles}
+\begin{document}
+
 \chapter{BayesQuant: Probabilistic estimation of protein ratios}
@@
 However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model implemented in the probabilistic programming framework Pymc3 was developed and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
+As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
 The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
@@
 Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
-Such approaches, establish a null hypothesis, complementary to the actual hypothesis being tested (alternative hypothesis).
+Such approaches, establish a null hypothesis $H_0$, complementary to the actual hypothesis being tested (alternative hypothesis $H_1$).
 The information about the phenomena being modelled is put to use to measure how likely this null hypothesis is via Null-Hypothesis Significance Testing (\ac{NHST}).
@@
 The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about probability of the alternative (our) hypothesis.
+Thus, p-values do not say anything about the probability of the alternative (our) hypothesis.
 
@@
 
-The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterir probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}
+The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterir probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}.
 
@@
 Unfortunately, this is tractable analytically for simple models only.
-However, Markov Chain-Monte Carlo (\ac{MCMC}) methods can be used to approximate this posterior, and are guaranteed to reach an infinitely exact fit if ran for long enough.
-Their demanding computational cost has only been recently met by the power of modern computers, thus making the approach feasible.
-Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS) and Python (Pymc3 \cite{Salvatier2016}), all making use of a programming paradigm called \textbf{probabilistic programming}.
+However, Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior.
+Both of which are now possible to run thanks to the advent of modern computing.
+
+\subsection{Inference methods: \ac{MCMC} and \ac{VI}}
+
+On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2018}.
+Monte Carlo sampling is performed on the Markov Chain, (hence the name of the technique) to randomly explore the parameter space with some heuristics.
+This heuristic guarantees convergence of the empirical estimate with the true posterior, provided a big enough sample size.
+Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018} \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
+The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} \footnote{A very nice tutorial on how Metropolis-Hastings works \href{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}} \footnote{Likewise, this resource provides very illustrative explanations of Hamiltonian-MC and NUTS \href{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}} have given way to the much more efficient sampler \ac{NUTS} \cite{Hoffman2011}.
+
+On the other hand, \ac{VI} methods, very recently developed and introduced in PyMC3, provide a fast alternative to \ac{MCMC} methods, yet they are not guaranteed to asymptotically approximate the true posterior.
+In \ac{VI}, optimization, instead of sampling, is employed as way to approximate the posterior \cite{Blei2018}.
+More concretely, a family of densities $Q$ over the parameters $\theta$ is defined.
+The goal of \ac{VI} is to find the single density $q$ within the family $Q$ that best approximates $p(\theta|y)$.
+This approximate density should be complex enough to reach a good approximation, and at the same time simple enough to be computationally easy to work with.
+The best candidate is defined as the one minimising the Kullback-Leibler (\ac{KL}) divergence (see equation \ref{eq:KL}).
+
+\begin{gather}\label{eq:KL}
+q^*(\theta) = \argmin \infdiv*{q(\theta)}{p(\theta|x)}
+\end{gather}
+
+where $q(\theta)$ stands for the whole family of densities, $p(\theta|x)$ stands for the true posterior and $KL$ stands for the \ac{KL} divergence. $q^*(\theta)$ is the best candidate density.
+
+From \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}} \textit{the term variational is used because the best $q$ in $Q$ is picked.
+The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation.}
+
+One of the terms yielded by the decomposition of the conditional density embodied by the posterior is the evidence of the data $p(x)$.
+This term is computationally intractable, thus making the computation of the exact \ac{KL} divergence impossible.
+A lower bound up to an additive constant can however be computed thanks to Jensen\textquotesingle s inequality \cite{jensen1906fonctions} \.
+This approximate amount is the Variational or Evidence Lower BOund (\ac{ELBO}).
+\textit{The \ac{ELBO} is the negative \ac{KL} divergence of plus $logp(x)$, which is a constant with respect to $q(z)$.
+Thus maximizing
+the \ac{ELBO} is equivalent to minimizing the \ac{KL} divergence} \cite{Blei2018}.
+
+Finally, several density approximation families are available.
+The mean-field family of densities is one of the most simple, as it ignores all dependencies between latent variables, and treats them as independent.
+From \href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html} the mean-field approximations works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}.
+A generic mean-field family member is shown in equation \ref{eq:mean_field}.
+
+\begin{equation}\label{eq:mean_field}
+q(\theta_1, ..., \theta_m) = \prod_{j=1}^{m}{q(\theta_j)}
+\end{equation}
+
+The demanding computational cost of the \ac{MCMC} and \ac{VI} schemes has only been recently met by the power of modern computers, thus making the approach feasible.
+Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS \cite{Plummer}) and Python (PyMC3 \cite{Salvatier2016}), all making use of a programming paradigm called \textbf{probabilistic programming}.
 
@@
 
-Pymc3 is a mature Python module dedicated to support probabilistic programming.
-It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (\ac{ADVI}) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
+PyMC3 is a mature Python module dedicated to support probabilistic programming.
+It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (\ac{VI}) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
 Therefore, it provides the tools required to build a Bayesian model for the probabilisitc estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
@@
 \subsection{Data input}
+\label{subsec:data_input}
+
 The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant} and \texttt{MSnSet2protdata} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
 
-The final state of the data is the following:
+The final state of the data is shown in table \ref{tab:data_model} for a single protein.
 
 \begin{table}
-\begin{tabular}{lllrrrrrr}
+\begin{tabular}{llrrrrrr}
 \toprule
-{} & protein &    organism &     H1 &     H2 &     H3 &     L1 &     L2 &     L3 \\
+protein &      Organism &     H1 &     H2 &     H3 &     L1 &     L2 &     L3 \\
 \midrule
-     &  P0A8I8 &  E. coli &  25.13 &  25.24 &  24.39 &  21.71 &  22.67 &  22.40 \\
-     &  P0A8I8 &  E. coli &  21.49 &  23.10 &  23.38 &  21.34 &  22.65 &  21.25 \\
-     &  P0A8I8 &  E. coli &  24.10 &  24.54 &  23.81 &  19.88 &  20.87 &  20.30 \\
+ P0A8I8 &       E. coli &  25.13 &  25.24 &  24.39 &  21.71 &  22.67 &  22.40 \\
+ P0A8I8 &       E. coli &  21.49 &  23.10 &  23.38 &  21.34 &  22.65 &  21.25 \\
+ P0A8I8 &       E. coli &  24.10 &  24.54 &  23.81 &  19.88 &  20.87 &  20.30 \\
+ A6NDG6 &  Homo sapiens &  25.08 &  24.98 &  23.83 &  22.54 &  22.52 &  24.96 \\
+ A6NDG6 &  Homo sapiens &  22.70 &  24.32 &  23.00 &  22.29 &  23.27 &  23.91 \\
+ A6NDG6 &  Homo sapiens &  21.18 &  22.51 &  23.25 &  22.30 &  21.75 &  23.61 \\
 \bottomrule
@@
 The remaining columns indicate the $log_2(Intensity)$ registered for each peptide in the corresponding run.
-In this case, three peptides were observed for the protein with id P0A8I8.}
+In this case, three peptides were observed for the proteins with ids \textit{P0A8I8} and \textit{A6NDG6}.}
+\label{tab:data_model}
 \end{table}
@@
 
-\subsection{Hierarchical model}
+\subsection{Hierarchical modelling}
+\label{subsec:hierarchy}
 
@@
 
-The mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect while minimising the remaining unwanted effects.
+A mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
 The mathematical framework implemented to model these effects was similar to the one used by MSqRob:
@@
 
-The data is thus affected by different effects acting in parallel, and in order to model them properly, multilevels need to be defined.
-Multilevel effects can be modelled with three approaches: pooled, unpooled, and with hierarchies (see figure \ref{fig:multilevel}).
-Hierarchical modeling is used when information is available on several different levels of observational units \cite{gelman2013bayesian}.
-The three effects were modeled as independent hierarchies.
+The data is thus affected by different effects acting independently, and in order to model them properly, multilevels need to be defined.
+Multilevel effects can be modelled with three approaches: pooled, unpooled, and hierarchically (see figure \ref{fig:multilevel}).
+Hierarchical modeling is used when the sampling variance is not the only source of variation among the parameters, but they still share a dependency.
+The intercept and the three effects were modeled as independent hierarchies.
 
-\begin{figure}[!h]
+\begin{figure}[H]
 \centering
@@
 
-This way, common probability distributions were set as priors for the intercept, treatment, peptide and run effect.
-The particular effect observed on each peptide was then modelled as a value sampled from the corresponding probability distribution.
+This way, independent and identically distributed (\ac{IID}) priors were set for the intercept, treatment, peptide and run effects.
+The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} prior.
 A diagram of the resulting model is displayed in figure \ref{fig:daft_model}).
@@
 
-\subsection{Approximation to the posterior probability distribution}
+\subsection{Posterior probability distribution computation}
+\label{subsec:posterior_compute}
 
-The posterior probability distribution can be approximated by means of \ac{MCMC} methods, which can be run with the power of modern computing.
-The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} have given way to the much more efficient sampler \ac{NUTS} \cite{Hoffman2011}.
-\ac{MCMC} methods simulate sampling from the posterior distribution by randomly exploring the parameter space with some heuristics that guarantee convergence with the true posterior given a big enough sample size.
-Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018} \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
+Both the \ac{MCMC} \ac{NUTS} sampler and \ac{VI} mean-field strategies were applied to approximate the posterior probability distribution of the \ac{log2FC}.
 
-\ac{ADVI} methods, very recently developed and introduced in PyMC3, attempt to solve the time and computational cost requirements characteristic of \ac{MCMC} methods by performing an approximation.
-The approximation consists of the selection of the best performing distribution $q$ within a family of distributions $Q$ that are computationally easy to work with.
-Thus the problem becomes an optimisation problem, where the inference algorithm needs to maximise the performance of the approximate function.
-Several specific methods can do this, but the mean-field implementation  is the default in PyMC3.
+\subsection{Model checking}
 
-Both NUTS and ADVI-powered approximations to the posterior were implemented in the program here presented using PyMC3.
+Posterior predictive checks of the posterior distribution were carried out using the \texttt{sample\_ppc()} function in PyMC3.
+A visual evaluation of the differences between simulated and observed data was used to assert the correctness of the models.
 
@@
 
-The code below formalizes the bias in \ac{MS1} intensity measurements due to a random effect in PyMC3 code.
+The code below formalizes in PyMC3 code the bias in \ac{MS1} intensity measurements due to a random effect.
 
@@
 
-and can be read aloud as \textit{the prior probability distribution of the random effect follows a normal distribution with mean 0 and standard deviation $\sigma$.
+and can be read as \textit{the prior probability distribution of the random effect follows a normal distribution with mean 0 and standard deviation $\sigma$.
 In turn, $\sigma$\textquotesingle s prior probability distribution follows a half normal distribution with standard deviation 1}.
@@
 The hierarchical structure of the model is set by the definition of a parameter distribution from which the value for each element (peptide) being modelled is sampled.
-This can be done by defining a new normal distribution with $\mu$ and $\sigma$ set to the random variables defined above.
-Thus, the following was implemented,
+This can be done by defining a new normal distribution where its $\mu$ and $\sigma$ are set to the random variables defined above.
+Im PyMC3 code,
 
@@
                framesep=2mm]{python}
-         beta = pm.Normal("beta", mu, sigma, n_elements)  
+         betae = pm.Normal("betae", mu, sigma, n_elements)  
 \end{minted}
@@
 \begin{equation}
-\beta_i \sim \mathcal{N}(\mu, \sigma)\,.  \\
+\beta^e_i \sim \mathcal{N}(\mu, \sigma)\,.  \\
 \end{equation}
 
-and can be read as \textit{the bias observed in the $i^{th}$ element due to the effect here modelled is said to follow a normal distribution with mean $\mu$ and standard deviation $\sigma$ both defined as random probability distributions above.}
+and can be read as \textit{the probability distribution of the bias observed in the $i^{th}$ element due to the effect here modelled is said to follow a normal distribution with mean $\mu$ and standard deviation $\sigma$ both defined as random probability distributions above.}
 
@@
          epsilon = pm.HalfNormal('epsilon', 1)                
-         m = pm.Deterministic("m", beta0 + beta)
+         m = pm.Deterministic("m", beta)
          obs = pm.Normal("obs", m, epsilon, observed=y)
@@
 \nonumber \epsilon \sim \mathcal{HN}(1) \,. \\ 
-m = \beta^0 + \sum_{i=1} \beta_i \,. \\ 
+m = \sum_{i=1} \beta_i \,. \\ 
 \nonumber y \sim \mathcal{N}(m, \epsilon)
@@
 
-and is read as \textit{the observed data is modelled as a normal distribution with mean $m$ and standard deviation $\epsilon$.
-$m$ is set defined as a random deterministic variable resulting from the sum of the intercept and the effects defined above.
+and is read as \textit{the observed data is modelled by a normal distribution with mean $m$ and standard deviation $\epsilon$.
+$m$ is a random deterministic variable resulting from the sum the effects defined above.
 $\epsilon$ follows a new half normal distribution with standard deviation 1}.
@@
 
-\subsection{Model implementation (I) peptide as a random process}
+\subsection{Running BayesQuant}
 
-\subsection{Model implementation (II) peptide effect with sequence features}
+BayesQuant takes a peptide\_summary\_intensity\_moFF\_run.tab (moFF) or a peptides.txt file (MaxQuant) as input, containing a peptide per row.
+Preprocessing using the MSqRob \texttt{preprocess\_MSnSet()} function, as well as other data munging functions, is required.
 
-\subsection{Posterior probability distribution computation}
+%--peptide_file> <--filetype> <--exp_annotation> [--annotation_df] [--organisms] [--protein_file] [--extract_features]
 
-\ac{ADVI} returns good results faster than \ac{NUTS}
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               frame=lines,
+               framesep=2mm]{R}
+
+Rscript prepare_BayesQuant.R --pepf peptides.txt \
+    --filetype MaxQuant --exp exp_annotation.tsv \
+    --output .
+\end{minted}
+
+The R script outputs the input file for BayesQuant. Its data is introduced in BayesQuant via the following code:
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               frame=lines,
+               framesep=2mm]{python}
+
+# Load the module
+from BayesQuant import BayesQuant
+bayesquant = BayesQuant()
+# Read the dataset
+bayesquant.read_data(data="ms1_intensities.tsv")
+# Compile a model for proteins with 3 peptides
+model = bayesquant.compile_model(n_peptides=3)
+# Load a specific protein
+bayesquant.load_data("A6NDG61")
+\end{minted}
+
+The snippet above loads the program, reads in the data, selects the data for a specific protein, and builds a model that matches the number of peptides observed.
+The posterior distribution can be computed using pure MCMC or VI methods, as stated in section \ref{subsec:posterior_compute}.
+The VI procedure will now be explained, due to its significantly better performance.
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               frame=lines,
+               firstnumber=last,
+               framesep=2mm]{python}
+# VI approximation: much faster and pretty accurate
+trace_advi = bayesquant.fit(model_name=p, n_draws=40000)
+\end{minted}
+
+The \texttt{bayesquant.fit()} function carries out two tasks:
+
+\begin{itemize}
+\item Finds the best mean-field approximation $q(\theta)$ to the true posterior $p(\theta|x)$. This approximate distribution is easier to work with.
+\item Sample from the approximate posterior, returning a collection of samples called trace.
+\end{itemize}
+
+The values stored in the trace obtained via VI will follow an approximation to the true posterior, and contain the results of the modelling process.
+Once it\textquotesingle obtained, the inference is complete and model checking ought to be performed to validate the results as explained in section \ref{subsec:model_checking}.
+It is run with
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               frame=lines,
+               firstnumber=last,
+               framesep=2mm]{python}
+# VI approximation: much faster and pretty accurate
+bayesquant.pcc()
+\end{minted}
+
+\subsection{Variational inference optimisation evaluation}
+
+\begin{figure}[H]
+%\begin{adjustbox}{varwidth=\textwidth,fbox,center}
+\begin{subfigure}{.45\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=.9\linewidth]{ELBO/P0A8I8}
+\end{subfigure}
+\begin{subfigure}{.45\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=.9\linewidth]{ELBO/A6NDG6}
+\end{subfigure}
+%\end{adjustbox}
+\caption{Progression of the \ac{ELBO} of the approximation for proteins P0A818 (\textit{E. coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}) over 40k iterations.}
+\label{fig:ELBO}
+\end{figure}
+
+\subsection{Basic model}
+
+The results of the model fitting step are shown in figures \ref{fig:ecoli_res}, and \ref{fig:human_res} for one protein from \textit{E. coli} and \textit{Homo sapiens}, respectively.
+
+\begin{figure}[!h]
+%\begin{adjustbox}{varwidth=\textwidth,fbox,center}
+\begin{subfigure}{\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=\linewidth]{trace_ecoli}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=\linewidth]{trace_human}
+\end{subfigure}
+%\end{adjustbox}
+\caption{Traceplot of the model fit for proteins P0A818 (\textit{E.
+coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}).
+The left panel shows the frequency density over the parameter space for several model parameters: (I) the \ac{log2FC} estimate (difference of treatment effects), (II) the effects in the three observed peptides, and (III) the effects in the six available runs.
+The right panel displays the sampled values from the \ac{VI} approximation stored in the trace.}
+\label{fig:traceplots}
+\end{figure}
+
+\begin{figure}[!h]
+\begin{subfigure}{0.8\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=\linewidth]{posterior_ecoli}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{0.8\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=\linewidth]{posterior_human}
+\end{subfigure}
+\label{fig:posteriors}
+\caption{Annotated posterior probability distribution for the \ac{log2FC} estimate for proteins P0A818 (\textit{E.
+coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}).
+The bar height is mapped to the probability mass in the corresponding interval.
+The 95\% High Probability Density Interval (\ac{HPDI}) is marked with a black line.
+The Region Of Practical Equivalence (\ac{ROPE}), shown in red, is a small range of parameter values that are considered to be practically equivalent to the null value in the present application \cite{Kruschke}, in this case a 0.4 window around 0.
+A green vertical line marks the expected \ac{log2FC} estimate for \textit{E.
+coli} proteins ($\log_2(3)$.}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{performance}
+\caption{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and human proteins with 2,3,4,7,6 and 10 peptides.
+The intervals are represented by horizontal lines.
+The dot represents the mean of the whole distribution.
+Vertical blue lines represent the ROPE defined as in figure \ref{fig:posteriors}, whereas the red line represents the expected estimate for bacterial proteins.}
+\label{fig:dumbbell}
+\end{figure}
+
+\subsection{Extended model: peptide effect with sequence features}
 
 \subsection{Model checking}
+\label{subsec:model_checking}
+
+However the posterior probability is computed, a proper Bayesian analysis is not finished without a model checking step where
+
+\begin{itemize}
+
+\item In \ac{MCMC} methods, evidence for non-convergence must be collected.
+Albeit lack of evidence of non-convergence does not guarantee convergence, the existence of evidence definitely proofs it \cite{Kruschke}.
+
+\item Posterior Predictive Checks are carried out to measure how well the model captures the data.
+
+\end{itemize}
+
+\ac{MCMC} asymptotically approaches the posterior after convergence, however, this condition is sometimes fulfilled by running long chains.
+Too short a chain, reflected by the lack of convergence, will fail to approximate the posterior and return wrong results.
+Most importantly, neither reaching convergence nor the best \ac{VI} approximation guarantee that the fitted model actually captures the data generation process.
+Therefore, the true posterior could be a bad reflection of the natural process being model and the best approximation will not solve it.
 
@@
 %As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
+\end{document}

### 2018-07-18 pipeline.tex
Extend chapter 3

@@
+\documentclass[thesis]{subfiles}
+\begin{document}
+
 \chapter{A label-free quantification proteomics pipeline}
@@
 Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability.
-MaxQuant, a freeware but closed-source and monolithic proteomics analysis suite \cite{Cox2008}, has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
+MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}.
+Nevertheless has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[H]
 \begin{subfigure}{.9\textwidth}
@@
 
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L.
+The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L.
 During quantification, it is important to take into account as many of the effects influencing protein quantification.
@@
 
-\begin{figure}[!h]
+\begin{figure}[H]
 \centering
@@
 Even though the observed correlation was low, the estimated log2FC was always greater than 1.
-The MQ+Rob pipeline systematically estimated log2FV values greater than the other 2.
+The MQ+Rob pipeline systematically estimated \ac{log2FC} values greater than the other 2.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[H]
 \centering
 \includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines performance.
+\caption{Overview over the different pipelines\textquotesingle performance.
 Quantification results are segregated by the source organism.
@@
 MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the resulting inconsistency, and eventually provide null estimates of the ratio.
+As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit2{E.
+coli} proteins.
 This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
@@
 One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
-Notwithstanding, it reveals the immature state of the technology.
+This technology adds  an extra dimension peptide separation process, leading to cleaner spectra.
+Notwithstanding, the developments also reveal the immature state of the technology.
 Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
@@
 Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.
+
+\end{document}

### 2018-07-19 modelling.tex
Finish chapter 3

@@
-\documentclass[thesis]{subfiles}
-\begin{document}
-
 \chapter{BayesQuant: Probabilistic estimation of protein ratios}
@@
 
-The current label-free quantification methods, reviewed in section \ref{sec:quantification} all rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
-However, a Bayesian based approach to this problem, that we know of, is lacking in the literature.
+The current label-free quantification methods reviewed in section \ref{sec:quantification} rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
+However, a Bayesian  approach to this problem, that we know of, is lacking in the literature.
 As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability and its strengths.
+The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.
 
@@
 One of the terms yielded by the decomposition of the conditional density embodied by the posterior is the evidence of the data $p(x)$.
-This term is computationally intractable, thus making the computation of the exact \ac{KL} divergence impossible.
+$p(x)$ is computationally intractable, thus making the computation of the exact \ac{KL} divergence impossible.
 A lower bound up to an additive constant can however be computed thanks to Jensen\textquotesingle s inequality \cite{jensen1906fonctions} \.
@@
 The demanding computational cost of the \ac{MCMC} and \ac{VI} schemes has only been recently met by the power of modern computers, thus making the approach feasible.
-Several Bayesian statistics frameworks are available on popular programming languages like R (JAGS \cite{Plummer}) and Python (PyMC3 \cite{Salvatier2016}), all making use of a programming paradigm called \textbf{probabilistic programming}.
+
+\subsection{Model checking}
+\label{subsec:model_checking}
+
+Whatever the posterior distribution is obtained, a proper Bayesian analysis is not finished without a model checking step that looks at potential problems happenning in the fitting step, and confirms the goodness of fit of the model to the data.
+
+\begin{itemize}
+
+\item Evidence for non-convergence must be collected. Albeit lack of evidence of non-convergence does not guarantee convergence, the existence of evidence definitely proofs it \cite{Kruschke}.
+
+\item Neither reaching convergence nor the best \ac{VI} approximation guarantee that the fitted model actually captures the data generation process.
+Therefore, the true posterior could be a bad reflection of the natural process being model and the best approximation will not solve it.
+Posterior Predictive Checks are carried out to measure how well the model captured the data properties \cite{Kruschke}.
+
+\end{itemize}
 
 \subsection{Probabilistic programming}
+
+The infrastructure providing the inference and model checking algorithms is available under several implementations in different programming languages, like R (JAGS \cite{Plummer}) and Python (PyMC3 \cite{Salvatier2016}).
+All these frameworks make use of a programming paradigm called \textbf{probabilistic programming}.
 
@@
 
-Features to model the peptide effect based on sequence were extracted using BLABLA
+Features to model the peptide effect based on sequence were extracted using the Biopython \cite{Cock2009} built-in  module \texttt{ProtParam.ProteinAnalysis()}.
+The following properties were extracted: amino-acid percentage, peptide length, molecular weight, mass/length ratio, isoelectric point, aromaticity and instability.
 
@@
 
-Both the \ac{MCMC} \ac{NUTS} sampler and \ac{VI} mean-field strategies were applied to approximate the posterior probability distribution of the \ac{log2FC}.
+The \ac{VI} mean-field strategy was applied to approximate the posterior probability distribution of the \ac{log2FC}.
+\ac{ELBO} optimisation was run for 40k iterations.
+Optimisation was validated by visual inspection of \ac{ELBO} traceplots over the 40k iterations.
+Once optimisation was validated, 10k samples were drawn from the approximate distribution to simulate the true posterior.
 
@@
 Posterior predictive checks of the posterior distribution were carried out using the \texttt{sample\_ppc()} function in PyMC3.
-A visual evaluation of the differences between simulated and observed data was used to assert the correctness of the models.
+500 datapoints for each peptide, consisting of 6 numbers, one for each run, were simulated.
+The resulting 6D data was projected onto a 2D plane via Principal Component Analysis.
+The covariance matrix required was computed on the whole dataset.
+The visual evaluation of the differences between simulated and observed data was used to assert the correctness of the models.
 
@@
                framesep=2mm]{python}
-# VI approximation: much faster and pretty accurate
-bayesquant.pcc()
+bayesquant.ppc()
 \end{minted}
@@
 %\end{adjustbox}
-\caption{Progression of the \ac{ELBO} of the approximation for proteins P0A818 (\textit{E. coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}) over 40k iterations.}
+\caption{Progression of \ac{ELBO} maximisation during the \ac{VI} approximation for proteins P0A818 (\textit{E. coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}) over 40k iterations.}
 \label{fig:ELBO}
@@
 
+Even though the ELBO took extremely low (negative) values at start (see figure \ref{fig:ELBO}), it was maximised fast to values around 0, indicating that the mean-field approximation was acceptably good.
+
 \subsection{Basic model}
+\label{subsec:basic_model}
 
-The results of the model fitting step are shown in figures \ref{fig:ecoli_res}, and \ref{fig:human_res} for one protein from \textit{E. coli} and \textit{Homo sapiens}, respectively.
+A protein from each proteome was used to test BayesQuant\textquotesingle s performance P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}).
+
+The traces produced by BayesQuant using \ac{VI} exhibited a clear random walk behaviour, which decreases the posibility that the approximation was wrong.
+Moreover, the  effects caused by the different peptides and runs were estimated and separated from the overall observed noise, allowing for a better estimation of the \ac{log2FC} (see figure \ref{traceplots}).
 
@@
 \caption*{A}
-\includegraphics[width=\linewidth]{trace_ecoli}
+\includegraphics[width=\linewidth]{traceplots/P0A8I8}
 \end{subfigure}
@@
 \caption*{B}
-\includegraphics[width=\linewidth]{trace_human}
+\includegraphics[width=\linewidth]{traceplots/A6NDG6}
 \end{subfigure}
@@
 
+This way, a different posterior distribution was computed for the impact of each peptide and run in all the measurements collected for each protein.
+
+The study of the \ac{log2FC} posterior probability distribution reveals what are the most likely parameter values and allows for the fulfillment of Bayesian Null Hypothesis Testing.
+The 95\% High Probability Density Interval (\ac{HPDI}), which is the narrowest interval containing 95\% of the total probability, indicates the most likely true values of the parameter.
+Presence or absence of overlap between the 95\% \ac{HPDI} and the Region Of Practical Equivalence (\ac{ROPE}) provides a formal way of discarding a point value of the parameter.
+The \ac{ROPE} is a small interval considered to be essentually the same as the null value \cite{Kruschke}.
+In the present application, the null value was set to 0, and the interval was given a width of 0.4 on each side.
+
 \begin{figure}[!h]
+\centering
 \begin{subfigure}{0.8\textwidth}
-\centering
 \caption*{A}
-\includegraphics[width=\linewidth]{posterior_ecoli}
+\includegraphics[width=\linewidth]{posteriors/P0A8I8}
 \end{subfigure}
@@
 \begin{subfigure}{0.8\textwidth}
-\centering
 \caption*{B}
-\includegraphics[width=\linewidth]{posterior_human}
+\includegraphics[width=\linewidth]{posteriors/A6NDG6}
 \end{subfigure}
@@
 The bar height is mapped to the probability mass in the corresponding interval.
-The 95\% High Probability Density Interval (\ac{HPDI}) is marked with a black line.
-The Region Of Practical Equivalence (\ac{ROPE}), shown in red, is a small range of parameter values that are considered to be practically equivalent to the null value in the present application \cite{Kruschke}, in this case a 0.4 window around 0.
+The 95\% \ac{HPDI} is marked with a black line.
+The \ac{ROPE} shown in red, is  \cite{Kruschke}, in this case a 0.4 window around 0.
 A green vertical line marks the expected \ac{log2FC} estimate for \textit{E.
-coli} proteins ($\log_2(3)$.}
+coli} proteins ($\log_2(3)$).}
 \end{figure}
 
+The test analysis indicated that P0Q8I8 most likely has a \ac{log2FC} between 1.4 and 1.7, and it\textquotesingle significantly different from 0, as no overlap was observed between \ac{HPDI} and \ac{ROPE}.
+On the other hand, the human protein was found to acquire a \ac{log2FC} most likely between -0.1 and 0.192.
+This range is fully contained within the \ac{ROPE}, which indicates that the \ac{log2FC} was practically 0 (see figure \ref{fig:posteriors}).
+
+The posterior predictive checks on both proteins pinpoints that the fitted models captured a data generation process approximating what was observed (see figure \ref{fig:ppc}).
+The three observed peptides were found to be undistinguishable from the 500 simulated ones.
+
 \begin{figure}[!h]
-\includegraphics[width=\textwidth]{performance}
-\caption{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and human proteins with 2,3,4,7,6 and 10 peptides.
+\centering
+\begin{subfigure}{0.8\textwidth}
+\caption*{A}
+\includegraphics[width=\linewidth]{PPC/PCA_P0A8I8}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{0.8\textwidth}
+\caption*{B}
+\includegraphics[width=\linewidth]{PPC/PCA_A6NDG6}
+\end{subfigure}
+\label{fig:ppc}
+\caption{Projection of the posterior predictive checks on the 2D plane capturing most variance. Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
+\end{figure}
+
+In order to further validate BayesQuant\textquotesingle s performance, it was tested on pairs of groups of 5 proteins from each proteome.
+Each pair was made by proteins with a different number of observed peptides (2,3,4,6,7,10) (see figure \ref{fig:dumbbell}).
+The more peptides, the more accurate the estimation will be as more data will be available.
+Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed.
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.9\textwidth]{performance.png}
+\caption{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
 The intervals are represented by horizontal lines.
-The dot represents the mean of the whole distribution.
+The dot represents the mean of the whole distribution, and it will be centered in the interval if the distribution is symmetrical.
 Vertical blue lines represent the ROPE defined as in figure \ref{fig:posteriors}, whereas the red line represents the expected estimate for bacterial proteins.}
@@
 
+Indeed, a general trend of width reduction was observed with the increased amount of peptides.
+For example, the widest intervals are observed on proteins with just two peptides available, and the narrowest on proteins with ten.
+Moreover, the \ac{HPDI} tended to align more strongly around an agreement value as well.
+
+The consistent observation that \textit{Homo sapiens} proteins got a \ac{log2FC} estimate around 0, and \textit{E.
+coli} proteins got an estimate which overlapped the predefinned \ac{ROPE} in only three cases (two of which on proteins with two peptides), confirmed the validity of the quantification framework.
+
 \subsection{Extended model: peptide effect with sequence features}
+\label{subsec:extended_model}
 
-\subsection{Model checking}
-\label{subsec:model_checking}
+Given the open-source nature of the program, it is possible to build upon it and extend its functionality.
+The possibility of making use of the peptide sequences to model the peptide effect using a linear model was considered and put to practice.
+The code was thus extended with the following snippets of code:
 
-However the posterior probability is computed, a proper Bayesian analysis is not finished without a model checking step where
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               frame=lines,
+               framesep=2mm]{python}
+
+# Specification of priors
+sigma_theta = pm.HalfNormal('sigma_theta', sd=1)
+theta = pm.Normal('theta', mu=0, sd=sigma_theta, shape = (n_features, 1))
+theta_inter = pm.Normal('theta_inter', mu=0, sd=sigma_theta)
+mu_pep = pm.Normal("mu_pep",
+  mu=theta_inter + pm.math.sum(features.dot(theta)),
+  sd=sigma_pep)   
+\end{minted}
+
+which means that if the peptide effect is modelled as a function of the sequence, $\mu^p$ is redefined to:
+
+%\begin{align}
+%\nonumber \mu^p \sim \mathcal{N}(0, \sigma^p) \, \\
+%\end{align}
+
+\begin{equation}
+\nonumber \sigma^{\theta} \sim \mathcal{HN}(1)
+\end{equation}
+\begin{equation}
+\nonumber \theta_k \sim \mathcal{N}(0, \sigma^{\theta})
+\end{equation}
+\begin{equation}
+\nonumber \mu^p \sim \mathcal{N}(\sum_{k=0}^{k=K}{x_k \theta_k}, \sigma^p)
+\end{equation}
+
+where $\theta_k$ is the weight of the $k^{th}$ feature, and $x_k$ is the numerical value of the $k^{th}$ feature.
+$\sigma^{\theta}$ is the prior for the standard deviation of the Normal all feature weights are sampled from.
+
+The same workflow as in section \ref{subsec:basic_model} was ran with the extended model, and the result is shown in figure \ref{fig:traceplots_seq}.
+The analysis illustrates the null predictive power of the sequence properties extracted, as all converge extremely strongly to a value of 0.
+The result is equivalent to not having defined any weights at all, and instead having run the basic model.
+
+\begin{figure}[H]
+\begin{subfigure}{\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=\linewidth]{traceplots/P0A8I8_seq.png}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=\linewidth]{traceplots/A6NDG6_seq.png}
+\end{subfigure}
+\caption{Traceplot obtained after fitting the extended model on the test proteins mentioned in \ref{subsec:basic_model}. No feature was found to have any predictive power.}
+\label{fig:traceplots_seq}
+\end{figure}
+
+\section{Discussion}
+
+\subsection{Improving usability: parallelization}
+
+While BayesQuant performed well on the majority of proteins, it took around 30 seconds to approximate each protein\textquotesingle posterior distributions using \ac{VI}, and more than two minutes with \ac{MCMC} methods.
+This translates to a dataset of 1k proteins taking 500 minutes (>8h) to be processed.
+However, the average modern computer comes with several processors that could support the parallelization of the program in several threads.
+Several proteins would be procesed simultaneously.
+
+\subsection{Further robustness and validation checks}
+
+A solid way of checking the results are not flawed is the specification of alternative priors.
+Unless precise prior knowledge is available, the selection of one prior over another prior to model the beliefs about the system should not have any impact on the results.
+It is for that reason that models built upon different priors could be run to further validate the program.
+
+\subsection{Advanced model comparison}
+
+Besides the predefinition of a \ac{ROPE} and the check of overlap with the 95 \% \ac{HPDI}, more advanced model comparison methods making use of the Bayes factor are available.
+The Bayes factor is defined as the ratio of the posterior probability of the models given the data \cite{Kruschke}.
+It could be used to measure how much more likely a model taking into account a treatment effect is compared to another one where the treatment effect is considered null i.e the \ac{log2FC} is set to 0.
+The result would provide an alternative criteria for the declaration of a protein being differentially abundant.
+
+\subsection{Improving the sequence modelling of the peptide effect}
+
+Unfortunately, during the development of the project, it was acknowledged that the peptide effect is very difficult phenomenon to capture in a simple Bayesian model as the one presented in this work.
+This is due to the motley and and multilevel nature of the noise caused by the aminoacidic sequences:
 
@@
 
-\item In \ac{MCMC} methods, evidence for non-convergence must be collected.
-Albeit lack of evidence of non-convergence does not guarantee convergence, the existence of evidence definitely proofs it \cite{Kruschke}.
+\item The protein neighborhood, which might affect the protease\textquotesingle s efficiency.
 
-\item Posterior Predictive Checks are carried out to measure how well the model captures the data.
+\item The competitive ionization problem, introduced in chapter \ref{chap:msreview} (\ref{subsec:the_ion_source}) implies that ionization prediction cannot be achieved without models that do not replicate this process mathematically.
+Thus, the ionization of a peptide ought to be predicted taking into account the peptide mix with which it accessed the ion source as a whole.
+
+%which states that the ionization efficiency of a peptide will not just depend of its sequence, but also of that of its competitors in the ion source.
+This is due to the ion source providing a limited amount of charges for an excess peptide input.
+Therefore, a different number of instances of the same peptide acquire charge, depending on how efficient the other peptides simultaneously present are.
+
+\item The mobile proton model \cite{Boyd2010}, which states that the peptide\textquotesingle s fragmentation pattern changes drastically depending on the number of charges and its distribution on the sequence.
+The charge distribution is hypothesised to be in turn determined by the position of positive residues that can allocate the charges.
+This phenomenon implies that search engines have a much harder time as a different spectrum is predicted for molecules with the same peptide sequence but different charges, thus decreasing identifications and the quality of the peaks shapes.
+As a consequence, feature extraction algorithms like the Apex module in moFF would extract worse \ac{MS1} intensities.
+
+\item The characterization of powerful sequence properties that specifically capture the differential ionization efficiency is undone work, as most feature extractors are designed to solve different problems, like protein sructure prediction.
 
@@
 
-\ac{MCMC} asymptotically approaches the posterior after convergence, however, this condition is sometimes fulfilled by running long chains.
-Too short a chain, reflected by the lack of convergence, will fail to approximate the posterior and return wrong results.
-Most importantly, neither reaching convergence nor the best \ac{VI} approximation guarantee that the fitted model actually captures the data generation process.
-Therefore, the true posterior could be a bad reflection of the natural process being model and the best approximation will not solve it.
+Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of proper neural networks, capable of deciphering the intrincate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
 
-\section{Discussion}
+\subsection{Posterior assessment of the effects}
 
-\section{Conclusion}
+The characterization of the uncertainty behind a \ac{log2FC} estimate can be used to better inform downstream analysis programs and eventually help interpret the biological results.
+For example, a functional analysis program could decide to drop proteins declared differentially abundant if the uncertainty is greater than a threshold.
 
-%As such, they are the target of multiple regulatory pathways ensuring their quantities are kept within homeostatic ranges.
-\end{document}
+Moreover, the estimation of the different run effects could be used to help \ac{MS} technicians assess the presence of batch effects in their experiments, and correct them in future analyses.

### 2018-07-19 pipeline.tex
Finish chapter 3

@@
-\documentclass[thesis]{subfiles}
-\begin{document}
-
 \chapter{A label-free quantification proteomics pipeline}
@@
 \includegraphics[width=\textwidth]{pipeline}
-\caption{Schema of the presented pipeline.
-First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
+\mycaption[Schema of the presented pipeline]{First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
 After that, spectra are searched by means of a search engine to perform the PSM step.
@@
 
-\subsection{Thousands of spectra were identified and validated}
+\subsection{Evaluation of the PSM step}
 
@@
 \end{subfigure}
-\caption{Quality control and validation of the 13th fraction of the third replicate in condition L.
-\textbf{A} Score distribution for matches to the decoy and the target databases.
+\mycaption[Quality control and validation of the 13th fraction of the third replicate in condition L]{\textbf{A} Score distribution for matches to the decoy and the target databases.
 \textbf{B} Evolution of the PSM score with confidence.
@@
 \includegraphics[width=0.8\textwidth]{match_percent}
-\caption{Percentage of matched spectra in all samples.
-The total number of spectra per sample ranged between 5894 and 20249.
+\mycaption[Percentage of matched spectra in all samples]{The total number of spectra per sample ranged between 5894 and 20249.
 \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
@@
 
-\subsection{Quantification}
+\subsection{Quantification benchmark}
 \label{subsec:quantification}
@@
 Either the pipeline developed in the presented chapter or those based on MaxQuant are ready to be deployed in a real environment and provide \ac{NZ} with useful insights on its proteomics data for free.
-
-\end{document}

### 2018-07-19 modelling.tex
Add conclusion :P

@@
 Moreover, the estimation of the different run effects could be used to help \ac{MS} technicians assess the presence of batch effects in their experiments, and correct them in future analyses.
+
+\section{Conclusion}
+
+A Bayesian framework for the relative quantification of protein abundance rations (Fold-Change, FC) using \ac{MS1} intensity was presented in this work.
+Unlike the currently prevailing methods in the field, it provides uncertainty estimates that provide a direct interpretation about the accuracy of the quantification.
+Moreover, preliminary results on the sequence modelling of the peptide bias revealed that more complex architectures and sequnce features are required to solve this problem.
+Finally, the method can be used to investigate the presence of batch effects in \ac{MS} experiments for their minimisation in future.

### 2018-07-21 modelling.tex
Extend chap 4

@@
 
-The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant} and \texttt{MSnSet2protdata} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
+The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant()} and \texttt{MSnSet2protdata()} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
+
+Peptides missing in any of the samples available were dropped from the analysis.
 
@@
 
-\subsection{Variational inference optimisation evaluation}
+\subsection{VI optimisation evaluation}
 
@@
 The traces produced by BayesQuant using \ac{VI} exhibited a clear random walk behaviour, which decreases the posibility that the approximation was wrong.
-Moreover, the  effects caused by the different peptides and runs were estimated and separated from the overall observed noise, allowing for a better estimation of the \ac{log2FC} (see figure \ref{traceplots}).
+Moreover, the  effects caused by the different peptides and runs were estimated and separated from the overall observed noise, allowing for a better estimation of the \ac{log2FC} (see figure \ref{fig:traceplots}).
 
@@
 The 95\% \ac{HPDI} is marked with a black line.
-The \ac{ROPE} shown in red, is  \cite{Kruschke}, in this case a 0.4 window around 0.
+The \ac{ROPE} defined as a 0.4 window around 0 is shown in red.
 A green vertical line marks the expected \ac{log2FC} estimate for \textit{E.
@@
 \caption*{A}
-\includegraphics[width=\linewidth]{PPC/PCA_P0A8I8}
+\includegraphics[width=\linewidth]{PPC/eps/PCA_P0A8I8}
 \end{subfigure}
@@
 \caption*{B}
-\includegraphics[width=\linewidth]{PPC/PCA_A6NDG6}
+\includegraphics[width=\linewidth]{PPC/eps/PCA_A6NDG6}
 \end{subfigure}
@@
 \centering
-\includegraphics[width=0.9\textwidth]{performance.png}
+\includegraphics[width=0.9\textwidth]{performance}
 \caption{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
@@
                obeytabs=true,tabsize=2,
+               frame=lines,
+               framesep=2mm]{python}
+
+bayesquant.read_data(
+    # table with ms1 intensity data
+    data_path="data/ms1_intensities.tsv",
+    # table with sequence features
+    features_path="data/protein_analysis.tsv")
+\end{minted}
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               obeytabs=true,tabsize=2,
+               firstnumber=last,
                frame=lines,
@@
 This translates to a dataset of 1k proteins taking 500 minutes (>8h) to be processed.
-However, the average modern computer comes with several processors that could support the parallelization of the program in several threads.
-Several proteins would be procesed simultaneously.
+However, the average modern computer comes with several processors that could support the parallelization of the program in several threads, decreasing the total computing time.
 
@@
 
-\subsection{Improving the sequence modelling of the peptide effect}
+\subsection{Sequence-based modelling of the peptide effect}
 
-Unfortunately, during the development of the project, it was acknowledged that the peptide effect is very difficult phenomenon to capture in a simple Bayesian model as the one presented in this work.
+Unfortunately, during the development of the project, it was acknowledged that the peptide effect is a very difficult phenomenon to capture in a simple Bayesian model as the one presented in this work.
 This is due to the motley and and multilevel nature of the noise caused by the aminoacidic sequences:
@@
 
-\item The competitive ionization problem, introduced in chapter \ref{chap:msreview} (\ref{subsec:the_ion_source}) implies that ionization prediction cannot be achieved without models that do not replicate this process mathematically.
+\item The competitive ionization problem, introduced in chapter \ref{chap:mass_spec} (\ref{subsec:the_ion_source}) implies that ionization prediction cannot be achieved without models that do not replicate this process mathematically.
 Thus, the ionization of a peptide ought to be predicted taking into account the peptide mix with which it accessed the ion source as a whole.
@@
 The charge distribution is hypothesised to be in turn determined by the position of positive residues that can allocate the charges.
-This phenomenon implies that search engines have a much harder time as a different spectrum is predicted for molecules with the same peptide sequence but different charges, thus decreasing identifications and the quality of the peaks shapes.
-As a consequence, feature extraction algorithms like the Apex module in moFF would extract worse \ac{MS1} intensities.
+This phenomenon implies that the \ac{PSM} step is more difficult as a different spectrum is expected for molecules with the same peptide sequence but different charges.
+The decreased identifications and quality of the peaks makes feature extraction algorithms like the Apex module in moFF would extract worse \ac{MS1} intensities, in turn injecting noise in the quantification data.
 
@@
 For example, a functional analysis program could decide to drop proteins declared differentially abundant if the uncertainty is greater than a threshold.
+A Gene Set Enrichment Analysis (\ac{GSEA}) could make use of the probabilistic information to refine the results of the enrichment.
 
@@
 Finally, the method can be used to investigate the presence of batch effects in \ac{MS} experiments for their minimisation in future.
+While the software can be integrated in any pipeline producing peptide-level data, further testing and improvements are probably required for optimal performance.

### 2018-07-21 pipeline.tex
Extend chap 4

@@
 MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}.
-Nevertheless has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
+Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
 
@@
 \item Fully command line, documented and Linux-supported interface for easy automation and scalability.
-\item Open-source for easy customisation and innovation of analyses.
+\item Open-source for easy customisation and extension.
 \item Free-licensed and cost-free, so anybody with the knowledge can run it, democratizing the analyses.
@@
 
-The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine \cite{Kim2014} by means of the SearchGUI \texttt{SearchCLI} tool \cite{Barsnes2018} utility.
+The spectra saved in the MGF files obtained in the previous step were passed to the MS-GF+ search engine \cite{Kim2014} by means of the SearchGUI \texttt{SearchCLI} tool version 3.3.3 \cite{Barsnes2018} utility.
 The search parameters were set using the \texttt{IdentificationParametersCLI}.
@@
 
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker version 1.16.22 utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
 By default, the FDR was set to 1\%.
@@
 Slight inconsistencies in the protein extraction and digestion, bias in the peptide ionization, presence of unexpected PTMs, and detector saturation and insensitivity, remain as problems contributing to the eventual measurement variability characteristic of mass spectrometry \cite{Piehowski2013}.
-Thus, adequate cleaning of the peptide dataset, and care to ignore outliers is preeminent.
+Thus, adequate preprocessing of the peptide dataset, and care to ignore outliers is preeminent.
 Indeed, the input file for the MQ+LFQ consisted of a preprocessed file which removed the most problematic proteins \cite{Cox2014}, which explains why more than 500 extra proteins could be quantified when using MSqRob (see figure \ref{fig:venn_cor} A).

### 2018-07-21 thp1.tex
Extend chapter 3
Extend chap 4

@@
 
-A benchmark dataset generated by \ac{NZ} technicians was run through the pipeline presented in chapter \ref{chap:pipeline} to showcase its performance.
-The experiment consisted of the application of two different treatments to THP-1 cell cultures.
-One group was subjected to an experimental procedure triggering the immunological response, while the other group was subjected to a negative control.
-The analysis of the resulting dataset through the aforementioned pipeline should thus reflect a change in the protein profile corresponding to an activation of the immune system in the first group when compared to the second.
+A benchmark dataset was run through the programmes presented in chapters \ref{chap:pipeline} and \ref{chap:model} to showcase their performance.
+The experimental desgin attempted to capture a protein profile change reflecting a known biological process.
+Thus, a successful computational analysis of the dataset should also reflect the phenomenon.
 The results confirmed that the computational analysis successfully captured this response.
-It was concluded that the software can be used in future experiments where the expected biological phenomena is not known.
+It was concluded that the software can be used in future experiments where the biological phenomena underlying the data are not known.
+
+\section{Introduction}
+
+The protein ratios across conditions are not known in ordinary datasets In order to judge the resolving power of the mass spectrometry and shotgun proteomics workflows presented in this work, an in-house dataset attempting to capture an immunological response was generated.
+Therefore, even though true protein ratios were not known \textit{a priori}, the analysis should find proteins related to the immunological response to be significantly differentially abundant.
+
+Moreover, a function inference step from the estimated protein \ac{log2FC} values is required to arrive to biological interpretations like that.
+These kind of analysis is common to all omics (genomics, transcriptomics, ...) and can be completed with Gene Set Enrichment Analysis \ac{GSEA} and Pathway Analysis, among others.
+
+\subsection{Goals}
+
+\begin{enumerate}
+
+\item Gauge the analytical power of the computational analyses explained in the previous chapters.
+
+%\item Compare its results with the commercial software currently in use.
+
+\item Propose a function inference analysis to complete the proteomics workflow.
+
+\end{enumerate}
+
+\section{Materials and Methods}
+
+\subsection{Sample preparation}
+
+THP-1 cells were grown in RPMI1640 (Sigma) supplemented with 10\% FBS (SeraLab) and 1\% pen/strep (Gibco).
+Stimulation of THP-1 cells was done in 48-well plates (Gibco) with 2x106 cells/well.
+Cells were exposed to 500 ng/mL LPS (E.
+coli O111:B4) (positive control condition, \textit{PC}) or nothing (negative control condition, \textit{NC}) for 48 hours in 37 degrees, 5\% \ce{{CO}_2}.
+After stimulation supernatants were collected and cell pellets were washed in PBS (Gibco).
+Both supernatants and pellets were frozen (-20 degrees).
+Only the pellet samples were considered for the analysis.
+EGWS, Esben Gjerløff Wedebye Schmidt.
+
+\subsection{Mass Spectrometry analysis}
+
+AEGI
+
+\subsection{Computational analysis}
+
+The resulting .RAW files were processed with the two different pipelines fully developed in the present work.
+(I) Compomics+MSqRob and (II) Compomics+BQ (Compomics and BayesQuant).
+The search settings were identical to those used in chapter \ref{chap:pipeline}, except for the proteome databases employed, which consisted of \textit{Homo sapiens} only.
+The validation filters, \ac{MBR}, MS1 Apex intensity, and MSqRob preprocessing parameters were also set to those used previously, for simplicity.
+The two pipelines used thus diverged only in the quantification engine, being identical in all previous steps.
+
+A protein was declared to be differentially abundant (\ac{DAP}) between the 2 conditions assayed if the result of the student\textquotesingle s-T test implemented in MSqRob returned a q-value less than 0.05.
+
+\subsection{Biological inference}
+
+The gProfiler R package \cite{Boyd2010} was used to run automatic \ac{GSEA} and provide a biological interpretation of the results.
+
+Pathways analysis was executed by calling the ComPath Webserver \cite{Domingo2018}.
+UniprotKB IDs were transformed to "Gene name" IDs, as required by the tool using the Retrieve/ID mapping tool from Uniprot.
+The KEGG \cite{Kanehisa2000} and Reactome \cite{Croft2014} databases were used as source.
+
+Protein interactions were explored by querying the STRING database \cite{Szklarczyk2017} with the list of \ac{DAP}s produced by MSqRob.
+
+\section{Results}
+
+\subsection{Compomics+MSqRob}
+
+Data on 6444 peptides, as collected in the peptide summary file from moFF, was passed to MSqRob, resulting in the quantification of 1433 proteins.
+A histogram of the \ac{log2FC} reveals that in most cases, not enough evidence was collected to declare the \ac{log2FC} different from 0 (see figure \ref{fig:compomics_rob} A).
+Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which an non-null estimate was provided were more abundant in the \textit{PC} condition.
+A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of proteins passing the significance criteria ($q-val < 0.05$), hand an absolute value of the \ac{log2FC} less than 1.
+However, all were considered differentially abundant.
+The 32 protein groups composed by a single protein are shown in table \ref{tab:thp1_rob_results}.
+
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Sat Jul 21 16:02:33 2018
+\begin{table}[!h]
+\centering
+\begin{tabular}{rllrr}
+  \hline
+ & Protein & Name & log2FC & qval \\ 
+  \hline
+  1 & P02792 & Ferritin light chain & 9.03E-01 & 2.32E-04 \\ 
+  2 & Q04760 & Lactoylglutathione lyase & -7.24E-01 & 2.32E-04 \\ 
+  3 & P13796 & Plastin-2 & -6.54E-01 & 2.32E-04 \\ 
+  4 & P28066 & Proteasome subunit alpha type-5 & -6.88E-01 & 8.24E-04 \\ 
+  5 & P09874 & Poly [ADP-ribose] polymerase 1 & -3.68E-01 & 2.48E-03 \\ 
+  6 & P53396 & ATP-citrate synthase & -3.79E-01 & 3.25E-03 \\ 
+  7 & P50552 & Vasodilator-stimulated phosphoprotein & -6.95E-01 & 1.43E-02 \\ 
+  8 & P54577 & Tyrosine--tRNA ligase, cytoplasmic & -4.21E-01 & 2.71E-02 \\ 
+  9 & Q15393 & Splicing factor 3B subunit 3 & -1.18E+00 & 2.88E-02 \\ 
+  10 & P08567 & Pleckstrin & -1.26E+00 & 2.96E-02 \\ 
+  11 & P50990 & T-complex protein 1 subunit theta & -3.60E-01 & 2.96E-02 \\ 
+  12 & Q7RTV0 & PHD finger-like domain-containing protein 5A & -1.26E+00 & 2.96E-02 \\ 
+  13 & O75083 & WD repeat-containing protein 1 & -6.55E-01 & 3.12E-02 \\ 
+  14 & P17812 & CTP synthase 1 & -4.60E-01 & 3.12E-02 \\ 
+  15 & Q14498 & RNA-binding protein 39 & -1.50E+00 & 3.29E-02 \\ 
+  16 & P60903 & Protein S100-A10 & 1.13E+00 & 3.29E-02 \\ 
+  17 & P26038 & Moesin & -3.91E-01 & 3.37E-02 \\ 
+  18 & O75368 & SH3 domain-binding glutamic acid-rich-like p & -6.15E-01 & 3.37E-02 \\ 
+  19 & Q8TEM1 & Nuclear pore membrane glycoprotein 210 & -8.24E-01 & 3.54E-02 \\ 
+  20 & Q01518 & Adenylyl cyclase-associated protein 1 & -3.51E-01 & 3.54E-02 \\ 
+  21 & Q08211 & ATP-dependent RNA helicase A & -4.19E-01 & 3.54E-02 \\ 
+  22 & Q14566 & DNA replication licensing factor MCM6 & -4.33E-01 & 3.68E-02 \\ 
+  23 & P30086 & Phosphatidylethanolamine-binding protein 1 & -6.31E-01 & 3.68E-02 \\ 
+  24 & O43143 & RNA helicase DHX15 & -1.30E+00 & 3.80E-02 \\ 
+  25 & O75691 & SMPC 20 homolog & -1.26E+00 & 3.80E-02 \\ 
+  26 & P27797 & Calreticulin & -4.63E-01 & 3.80E-02 \\ 
+  27 & P41218 & Myeloid cell nuclear differentiation antigen & -4.49E-01 & 3.80E-02 \\ 
+  28 & Q15056 & Eukaryotic translation initiation factor 4H & -3.62E+00 & 3.98E-02 \\ 
+  29 & P04406 & Glyceraldehyde-3-phosphate dehydrogenase & -1.90E+00 & 3.98E-02 \\ 
+  30 & P00492 & Hypoxanthine-guanine phosphoribosyltransferase & -4.54E-01 & 4.12E-02 \\ 
+  31 & P50440 & Glycine amidinotransferase, mitochondrial & -1.02E+00 & 4.50E-02 \\ 
+  32 & P62857 & 40S ribosomal protein S28 & -5.50E-01 & 4.98E-02 \\ 
+   \hline
+\end{tabular}
+\caption{THP-1 MSqRob results.
+32 single-protein groups were found to be differentially abundant under significance criteria of q-value less than 0.05.
+The protein id, the gene name, the point estimate of its \ac{log2FC} and its significance is shown}
+\label{tab:thp1_rob_results}
+\end{table}
+
+\begin{figure}[!h]
+\begin{subfigure}{0.45\textwidth}
+\centering
+\caption*{A}
+\includegraphics[width=.9\linewidth]{histogram_msqrob}
+\end{subfigure}
+\begin{subfigure}{0.45\textwidth}
+\centering
+\caption*{B}
+\includegraphics[width=.9\linewidth]{volcano_plot_msqrob}
+\end{subfigure}
+\caption{Compomics+MSqRob results.
+\textbf{A} Histogram of the estimated \ac{log2FC}.
+\textbf{B} Volcano plot, showing the point estimate of the \ac{log2FC} on the x axis, and the minus logarithm of the p-value on the y-axis.}
+\label{fig:compomics_rob}
+\end{figure}
+
+\subsection{Compomics+BayesQuant}
+
+After preprocessing and missing-data handling, a dataset of 1641 peptides was input to BayesQuant, and 269 proteins were quantified.
+A histogram of the mean of the inferred posterior distributions manifests the centrality of its distribution, with most values within the predefined \ac{ROPE} of [-0.4, 0.4].
+In line with what was observed in the results above, the distribution was found to be skewed toward negative values (see figure \ref{fig:compomics_bq}A).
+Only four proteins were assigned a 95\% \ac{HPDI} not touching the \ac{ROPE} (see figure \ref{fig:compomics_bq}B and table \ref{tab:thp1_bq_results}).
+The distribution of the \ac{HPDI} width summarises the overall uncertainty in the estimation process.
+Most proteins had a very narrow HPDI, as shown in figure \ref{fig:compomics_bq}C.
+As expected, the width of the interval exhibited some degree of correlation with   the mean of the posterior, as shown in \ref{fig:compomics_bq}D.
+
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Sat Jul 21 18:02:39 2018
+\begin{table}[!h]
+\centering
+\begin{tabular}{rlrrrr}
+  \hline
+ & Protein & log2FC & HPDI start & HPDI end & Peptides \\ 
+  \hline
+& O75475 & -0.61 & -0.80 & -0.43 & 2 \\ 
+   & O15144 & -0.75 & -0.98 & -0.51 & 3 \\ 
+   & P02792 & 0.50 & 0.41 & 0.58 & 4 \\ 
+   & P04406 & -1.09 & -1.45 & -0.77 & 4 \\ 
+   \hline
+\end{tabular}
+\caption{Compomics+BayesQuant results}
+\label{tab:thp1_bq_results}
+\end{table}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\linewidth]{bq_comb_bq}
+\caption{BayesQuant results.
+\textbf{A} Histogram of the mean of the posterior distributions.
+\textbf{B} Dumbbell plot, representing the 95\% \ac{HPDI} for all proteins.
+Black dots represent the mean, and intervals not overlapping the \ac{ROPE} are shown red.
+\textbf{C} Histogram of the 95\% \ac{HPDI} width.
+\textbf{D} Scatter plot of the 95\% \ac{HPDI} width and the  posterior \ac{log2FC} mean.}
+\label{fig:compomics_bq}
+\end{figure}
+
+Given the scarcity of the results, it was decided to continue the analyses with MSqRob output only.
+
+\subsection{Functional analysis}
+
+A \ac{GSEA} performed on the list of 32 \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most overrepresented biological processes, cellular compartments and molecular functions, compared to the human background.
+The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.
+
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Sat Jul 21 19:00:27 2018
+\small
+\begin{table}[!h]
+\centering
+\begin{tabular}{rllllll}
+  \hline
+ & Term ID & Term Name & Query & Term & Ov & P-value \\ 
+  \hline
+1 & GO:0070062 & extracellular exosome &  32 & 2774 &  19 & 1.17e-05 \\ 
+  2 & GO:1903561 & extracellular vesicle &  32 & 2793 &  19 & 1.31e-05 \\ 
+  3 & GO:0043230 & extracellular organelle &  32 & 2795 &  19 & 1.33e-05 \\ 
+  4 & GO:0005615 & extracellular space &  32 & 3958 &  20 & 6.76e-04 \\ 
+  5 & GO:0035578 & azurophil granule lumen &  32 &  90 &   5 & 8.87e-04 \\ 
+  6 & GO:0044421 & extracellular region part &  32 & 4119 &  20 & 1.34e-03 \\ 
+  7 & GO:0002376 & immune system process &  32 & 2949 &  17 & 1.74e-03 \\ 
+  8 & GO:0060205 & cytoplasmic vesicle lumen &  32 & 335 &   7 & 2.54e-03 \\ 
+  9 & GO:0031983 & vesicle lumen &  32 & 336 &   7 & 2.59e-03 \\ 
+  10 & GO:0005576 & extracellular region &  32 & 4925 &  21 & 4.95e-03 \\ 
+  11 & GO:0003723 & RNA binding &  32 & 1787 &  13 & 5.50e-03 \\ 
+  12 & GO:0065003 & Complex assembly &  32 & 1803 &  13 & 6.08e-03 \\ 
+  13 & GO:0003725 & DS RNA binding &  32 &  65 &   4 & 9.71e-03 \\ 
+  14 & GO:0005766 & primary lysosome &  32 & 153 &   5 & 1.23e-02 \\ 
+  15 & GO:0042582 & azurophil granule &  32 & 153 &   5 & 1.23e-02 \\ 
+  16 & HPA:006020\_13 & caudate; neuronal cells &  14 & 311 &   5 & 1.35e-02 \\ 
+  17 & GO:0031982 & vesicle &  32 & 4271 &  19 & 1.38e-02 \\ 
+  18 & GO:1990904 & ribonucleoprotein complex &  32 & 836 &   9 & 1.44e-02 \\ 
+  19 & GO:0001775 & cell activation &  32 & 1359 &  11 & 1.60e-02 \\ 
+  20 & HPA:022020\_13 & hippocampus; neuronal cells &  14 & 331 &   5 & 1.81e-02 \\ 
+   \hline
+\end{tabular}
+\caption{\ac{GSEA} results.
+The 20 most significant terms are displayed.
+The columns \textit{Query}, \textit{Term} and \textit{Ov} indicate the size of the query, the number of proteins under the term, and the overlap between them.
+The size of the query differs in the GO and HPA terms repositories due to nomenclature mismatches.}
+\label{tab:gsea}
+\end{table}
+\normalsize
+
+On the one hand, nine terms (1, 2, 3, 4, 6, 8, 9, 10 and 17) were cell compartment terms associated with cellular secretion and the extracelullar space.
+On the other hand, five terms (5, 7, 14, 15, 19) were related to the immunological response.
+The remaining terms were connected to RNA functions and the nervous system.
+
+\subsection{Pathway analysis}
+
+The UniprotKB to Gene name mapping returned a list of 51 proteins, which was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were overrepresented compared to the human background.
+In agreement to the \ac{GSEA} results, several directly immunology linked pathways were found to be significantly enriched: (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
+Moreover, the finding of several RNA related pathways exposes the presence of potential cell reprogramming, essential in any change in protein profile.
+
+% latex table generated in R 3.4.4 by xtable 1.8-2 package
+% Sat Jul 21 19:49:28 2018
+\begin{table}[ht]
+\centering
+\begin{tabular}{rllrrl}
+  \hline
+ & Pathway & q-val & Ov & Size & DB \\ 
+  \hline
+1 & Neutrophil degranulation & 0.00e+00 &   9 & 479 & reactome \\ 
+  2 & Immune System & 0.00e+00 &  15 & 2050 & reactome \\ 
+  3 & Innate Immune System & 0.00e+00 &  10 & 1109 & reactome \\ 
+  4 & Metabolism of RNA & 0.00e+00 &  10 & 666 & reactome \\ 
+  5 & Processing of Pre-mRNA & 2.00e-04 &   5 & 240 & reactome \\ 
+  6 & mRNA Splicing & 1.10e-03 &   4 & 180 & reactome \\ 
+  7 & mRNA Splicing & 1.20e-03 &   4 & 188 & reactome \\ 
+  8 & Metabolic pathways & 1.60e-03 &   8 & 1282 & kegg \\ 
+  9 & Carbon metabolism & 2.90e-03 &   3 & 116 & kegg \\ 
+  10 & Spliceosome & 4.00e-03 &   3 & 134 & kegg \\ 
+  11 & Pentose phosphate pathway & 4.20e-03 &   2 &  30 & kegg \\ 
+  12 & Base excision repair & 4.90e-03 &   2 &  33 & kegg \\ 
+  13 & Ribosome & 5.30e-03 &   3 & 153 & kegg \\ 
+  14 & Necroptosis & 6.10e-03 &   3 & 162 & kegg \\ 
+  15 & Apoptosis & 6.60e-03 &   3 & 168 & reactome \\ 
+  16 & Programmed Cell Death & 6.90e-03 &   3 & 171 & reactome \\ 
+  17 & Regulation of Apoptosis & 1.08e-02 &   2 &  53 & reactome \\ 
+  18 & Aminoacyl-tRNA biosynthesis & 1.55e-02 &   2 &  66 & kegg \\ 
+  19 & Glycolysis / Gluconeogenesis & 1.59e-02 &   2 &  68 & kegg \\ 
+  20 & Leukocyte transendothelial migration & 2.96e-02 &   2 & 112 & kegg \\ 
+   \hline
+\end{tabular}
+\caption{Pathway Enrichment results. The 10 most significant pathways found in Reactome and Kegg are shown arranged by decreasing significance. }
+\label{tab:pathway_enrichment}
+\end{table}
+
+\begin{figure}[H]
+\centering
+\includegraphics[width=\linewidth]{string_normal_image}
+\caption{STRING interaction network}
+\label{fig:STRING}
+\end{figure}
+
+Since the cells were indeed treated to create an immunological response, the result supports the hypothesis that the pipeline can capture biological phenomena.
+
+\section{Discussion}

### 2018-07-22 msreview.tex
Extend chapter 3
Finish chapter 3
Finish chap 4 and conclusion

@@
 The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
-The top-down paradigm will be ignored in the rest of the manuscript.
+This review will focus on the data-dependent approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
 
@@
 MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \label{sec:sample processing}, while an overview on mass spectrometers is given in section \label{sec:the mass spectrometer}.
+A detailed explanation of the sample processing required prior to MS is given in section \ref{sec:sample_processing}, while an overview on mass spectrometers is given in section \ref{sec:the_mass_spectrometer}.
 The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
-These inference steps can be condensed to the peptide and protein inference problems, explained in section \label{sec:peptide and protein inference}.
+These inference steps can be condensed to the peptide and protein inference problems, explained in section \ref{sec:inference}.
 A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
@@
 These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
-The difference can be exploited to infer which aminoacid is present in the longer fragment.
+The difference can be exploited to deduce which aminoacid makes up for that difference.
 If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
@@
 \caption{Visualizing FDR and PEP.
-The FDR is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than s.
-The PEP corresponds to the same ratio, but only a specific score, hence its alternative name of local-FDR.
+The FDR at a given score \textit{s} is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than \textit{s}.
+The PEP corresponds to the same ratio, but only at a specific score, hence its alternative name of local-FDR.
 Figure from \cite{Kall2008}.}

### 2018-07-22 pipeline.tex
Finish chap 4 and conclusion

@@
 A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
-The results exhibited an accuracy similar to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
+The results were comparable to those achieved by the MaxQuant \cite{Cox2014} software, excepting a bias produced by the sample fractionation of this dataset.
 
@@
 \textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
-Finally, MQ+Rob blends MaxQuant upstream analysis with the MSqRob quantification engine.
+Finally, \textbf{MQ+Rob} blends MaxQuant upstream analysis with the MSqRob quantification engine.
 \textbf{A, B} ROC curve produced by the three algorithms with no \ac{log2FC} filter, and keeping proteins with \ac{log2FC} greater than 1, respectively.
@@
 
+\subsection{Applicability for Novozymes data}
+
+The SearchGUI and PeptideShaker recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{https://github.com/compomics/searchgui/wiki/DatabaseHelp}}.
+This is due to the software making use of the annotations to improve the protein inference step.
+In any case, a consistent formatting in the FASTA databases is required to used the tools.
+Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with these and many other tools.
+
 \section{Conclusion}
 
-An open-source, free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
+An open-source, cost-free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
 The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.

### 2018-07-22 thp1.tex
Finish chap 4 and conclusion

@@
 
-\item Propose a function inference analysis to complete the proteomics workflow.
+\item Propose a function inference analytical pipeline to complete the proteomics workflow.
 
@@
 
-A protein was declared to be differentially abundant (\ac{DAP}) between the 2 conditions assayed if the result of the student\textquotesingle s-T test implemented in MSqRob returned a q-value less than 0.05.
+In the Compomics+MSqRob pipeline, a protein was declared to be differentially abundant (\ac{DAP}) between the 2 conditions assayed if the result of the student\textquotesingle s-T test implemented in MSqRob returned a q-value less than 0.05.
+The criteria in the Compomics+BayesQuant pipeline was the absence of overlap between the \ac{ROPE} [-0.4, 0.4] and the 95\% \ac{HPDI}.
 
@@
 Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which an non-null estimate was provided were more abundant in the \textit{PC} condition.
-A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of proteins passing the significance criteria ($q-val < 0.05$), hand an absolute value of the \ac{log2FC} less than 1.
+A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria ($qval < 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
 However, all were considered differentially abundant.
-The 32 protein groups composed by a single protein are shown in table \ref{tab:thp1_rob_results}.
+The 32 protein groups composed by a single protein (single protein \ac{DAP}s) are shown in table \ref{tab:thp1_rob_results}.
 
@@
 \end{tabular}
-\caption{Compomics+BayesQuant results}
+\caption{\ac{DAP} declared by the Compomics+BayesQuant pipeline.}
 \label{tab:thp1_bq_results}
@@
 
-A \ac{GSEA} performed on the list of 32 \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most overrepresented biological processes, cellular compartments and molecular functions, compared to the human background.
+A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most overrepresented biological processes, cellular compartments and molecular functions, compared to the human background.
 The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.
@@
 \centering
-\begin{tabular}{rllllll}
-  \hline
- & Term ID & Term Name & Query & Term & Ov & P-value \\ 
-  \hline
-1 & GO:0070062 & extracellular exosome &  32 & 2774 &  19 & 1.17e-05 \\ 
-  2 & GO:1903561 & extracellular vesicle &  32 & 2793 &  19 & 1.31e-05 \\ 
-  3 & GO:0043230 & extracellular organelle &  32 & 2795 &  19 & 1.33e-05 \\ 
-  4 & GO:0005615 & extracellular space &  32 & 3958 &  20 & 6.76e-04 \\ 
-  5 & GO:0035578 & azurophil granule lumen &  32 &  90 &   5 & 8.87e-04 \\ 
-  6 & GO:0044421 & extracellular region part &  32 & 4119 &  20 & 1.34e-03 \\ 
-  7 & GO:0002376 & immune system process &  32 & 2949 &  17 & 1.74e-03 \\ 
-  8 & GO:0060205 & cytoplasmic vesicle lumen &  32 & 335 &   7 & 2.54e-03 \\ 
-  9 & GO:0031983 & vesicle lumen &  32 & 336 &   7 & 2.59e-03 \\ 
-  10 & GO:0005576 & extracellular region &  32 & 4925 &  21 & 4.95e-03 \\ 
-  11 & GO:0003723 & RNA binding &  32 & 1787 &  13 & 5.50e-03 \\ 
-  12 & GO:0065003 & Complex assembly &  32 & 1803 &  13 & 6.08e-03 \\ 
-  13 & GO:0003725 & DS RNA binding &  32 &  65 &   4 & 9.71e-03 \\ 
-  14 & GO:0005766 & primary lysosome &  32 & 153 &   5 & 1.23e-02 \\ 
-  15 & GO:0042582 & azurophil granule &  32 & 153 &   5 & 1.23e-02 \\ 
-  16 & HPA:006020\_13 & caudate; neuronal cells &  14 & 311 &   5 & 1.35e-02 \\ 
-  17 & GO:0031982 & vesicle &  32 & 4271 &  19 & 1.38e-02 \\ 
-  18 & GO:1990904 & ribonucleoprotein complex &  32 & 836 &   9 & 1.44e-02 \\ 
-  19 & GO:0001775 & cell activation &  32 & 1359 &  11 & 1.60e-02 \\ 
-  20 & HPA:022020\_13 & hippocampus; neuronal cells &  14 & 331 &   5 & 1.81e-02 \\ 
-   \hline
-\end{tabular}
+\input{tables/gsea.tex}
+%\begin{tabular}{rllllll}
+%  \hline
+% & Term ID & Term Name & Query & Term & Ov & P-value \\ 
+%  \hline
+%1 & GO:0070062 & extracellular exosome &  32 & 2774 &  19 & 1.17e-05 \\ 
+%  2 & GO:1903561 & extracellular vesicle &  32 & 2793 &  19 & 1.31e-05 \\ 
+%  3 & GO:0043230 & extracellular organelle &  32 & 2795 &  19 & 1.33e-05 \\ 
+%  4 & GO:0005615 & extracellular space &  32 & 3958 &  20 & 6.76e-04 \\ 
+%  5 & GO:0035578 & azurophil granule lumen &  32 &  90 &   5 & 8.87e-04 \\ 
+%  6 & GO:0044421 & extracellular region part &  32 & 4119 &  20 & 1.34e-03 \\ 
+%  7 & GO:0002376 & immune system process &  32 & 2949 &  17 & 1.74e-03 \\ 
+%  8 & GO:0060205 & cytoplasmic vesicle lumen &  32 & 335 &   7 & 2.54e-03 \\ 
+%  9 & GO:0031983 & vesicle lumen &  32 & 336 &   7 & 2.59e-03 \\ 
+%  10 & GO:0005576 & extracellular region &  32 & 4925 &  21 & 4.95e-03 \\ 
+%  11 & GO:0003723 & RNA binding &  32 & 1787 &  13 & 5.50e-03 \\ 
+%  12 & GO:0065003 & Complex assembly &  32 & 1803 &  13 & 6.08e-03 \\ 
+%  13 & GO:0003725 & DS RNA binding &  32 &  65 &   4 & 9.71e-03 \\ 
+%  14 & GO:0005766 & primary lysosome &  32 & 153 &   5 & 1.23e-02 \\ 
+%  15 & GO:0042582 & azurophil granule &  32 & 153 &   5 & 1.23e-02 \\ 
+%  16 & HPA:006020\_13 & caudate; neuronal cells &  14 & 311 &   5 & 1.35e-02 \\ 
+%  17 & GO:0031982 & vesicle &  32 & 4271 &  19 & 1.38e-02 \\ 
+%  18 & GO:1990904 & ribonucleoprotein complex &  32 & 836 &   9 & 1.44e-02 \\ 
+%  19 & GO:0001775 & cell activation &  32 & 1359 &  11 & 1.60e-02 \\ 
+%  20 & HPA:022020\_13 & hippocampus; neuronal cells &  14 & 331 &   5 & 1.81e-02 \\ 
+%   \hline
+%\end{tabular}
 \caption{\ac{GSEA} results.
@@
 
-The UniprotKB to Gene name mapping returned a list of 51 proteins, which was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were overrepresented compared to the human background.
-In agreement to the \ac{GSEA} results, several directly immunology linked pathways were found to be significantly enriched: (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
+The UniprotKB to Gene name mapping of the whole set of \ac{DAP}s (including multiprotein groups) returned by MSqRob returned a list of 51 entities.
+The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were overrepresented compared to the human background.
+In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched: (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
 Moreover, the finding of several RNA related pathways exposes the presence of potential cell reprogramming, essential in any change in protein profile.
@@
 
-\begin{figure}[H]
+\subsection{Protein interaction analysis}
+
+The same list supplied to ComPath was searched in the STRING database \cite{Szklarczyk2017} to screen for protein-protein interactions in the list.
+The database returns a protein association network (see figure \ref{fig:STRING}), that helps interpreting the information contained in gene sets.
+49/51 terms were mapped to the resource and 15 of them were found to be included in the "immune system process" set, placing the biological process as the third most enriched with an FDR of 0.0387.
+27 genes belonged to the "cellular nitrogen compound metabolic process" (FDR 0.0128).
+
+\begin{figure}[!h]
 \centering
 \includegraphics[width=\linewidth]{string_normal_image}
-\caption{STRING interaction network}
+\caption{STRING interaction network.
+Every node represents a protein in the list, and an edge between any pair represents evidence for some kind of association or interaction between them.
+Evidence is either experimentally determined or predicted using diverse methods.
+Red colored nodes are included in the "immune system process" set.}
 \label{fig:STRING}
@@
 
-Since the cells were indeed treated to create an immunological response, the result supports the hypothesis that the pipeline can capture biological phenomena.
+\section{Discussion}
 
-\section{Discussion}
+\subsection{Increasing the experiment data-throughput}
+
+6 samples organized in two different treatments of 3 biological replicates without sample prefractionation each were analyzed.
+The resulting data throughput allowed for the performance of biological inference, as it successfully captured the undergoing biological processes.
+Nevertheless, a more in-depth results report would have been possible with (I) sample prefractionation, which improves peptide separation and allows for the collection and identification of more spectra \cite{Righetti2005}, and (II) the inclusion of technical replicates to differentiate the biological variability from experimental noise.
+
+\subsection{Missing data handling}
+
+One of the most frequent issues found when analyzing proteomics datasets is the handling of missing values \cite{Lazar2016}.
+The presence of missing data in many of the rows in the input moFF file led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups.
+Thanks to the missing data handling implemented in MSqRob quantified 1433 proteins, enough to continue the analysis (see table \ref{tab:balance}).
+The fact that the number of observed groups was 2787 implies that the management of missing values could be improved in both tools, and particularly in BayesQuant.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|c|c|c|}
+\hline 
+Tool & Peptides & Proteins \\ 
+\hline 
+MSqRob & 6444 & 1433 \\ 
+\hline 
+BayesQuant & 1641 & 350 \\ 
+\hline 
+\end{tabular} 
+\caption{Peptide input and quantified proteins for each tool in the THP1 dataset.}
+\label{tab:balance}
+\end{table}
+
+Three different types of missing values have been defined \cite{Wei2018}:
+
+\begin{enumerate}
+
+\item Missing Completely At Random (MCAR): Unpredictable, caused by random errors in the data acquistion step and affecting the whole dataset.
+
+\item Missing At Random (MAR): Caused by other observed variables. Citing \cite{Wei2018}: \textit{ Inaccurate peak detection and deconvolution of co-eluting compounds can be called MAR}.
+
+\item Missing Not At Random (\ac{MNAR}): Caused by signals under the limit of detection (\ac{LOD}) of the spectrometer.
+
+\end{enumerate}
+
+While all three are relevant, \ac{MNAR} is particularly important because it introduces a bias in the dataset, where proteins are more likely to be quantified if their abundance is high enough in all samples for their peptide signals to be above the \ac{LOD}.
+Thus, proteins exhibiting abundances below the \ac{LOD} due to the experimental treatment are bound to generate peptide data featuring missing values that, if not handled properly, are discarded.
+Subsequently, \ac{DAP}s or in general proteins with variant abundance become less likely to be quantified, difficulting the extraction of protein sets.
+
+Missing data should be imputed at the peptide level.
+The implementation of a module performing this task prior to quantification would  improve the accuracy and throughput of the quantification process.
+Several approaches have been proposed, some within a probabilistic framework \cite{Lazar2016}.
+For example, in cases of clear \ac{MNAR}, imputation could be based on an inferred global \ac{LOD}, so that missing values are replaced with it.
+Work on this line could potentially boost the performance of BayesQuant.
+
+\subsection{Applicability for Novozymes data}
+
+\ac{NZ} data is frequently confidential and cannot be uploaded to public servers like STRING.
+However, many of the databases consulted in this work can be built locally, thus removing the need of data leaving the company\textquotesingle s facilities.
+
+\section{Conclusion}
+The results presented in this work demonstrated the usability of the tools presented in the previous chapters in an actual \ac{NZ} setting.
+Moreover, a walk-through over the functional analyses that can be performed with this output clearly showed they can be used to perform computational analyses and biological inference on protein samples with open-source software with no cost for the company.
+However, the absence of proper missng data handling developed into the inability to quantify hundreds of proteins, signaling what are the next steps to improve the tools.

### 2018-07-23 modelling.tex

@@
 
+The posterior distribution can be approximated (inferred) by means of \ac{MCMC} and \ac{VI} methods.
 On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2018}.
@@
 
-Whatever the posterior distribution is obtained, a proper Bayesian analysis is not finished without a model checking step that looks at potential problems happenning in the fitting step, and confirms the goodness of fit of the model to the data.
+However the posterior distribution is obtained, a proper Bayesian analysis is not finished without a model checking step that looks at potential problems happenning in the fitting step, and confirms the goodness of fit of the model to the data.
 
@@
 Probabilistic programming offers the framework to build complex statistical models by storing probability distributions as variables of the program.
-Moreover, probabilistic programming packages supply the tools needed to perform inference with these models by taking experimental evidence (data) and fitting the distributions to the data using Bayes theorem.
-The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using \ac{MCMC} algorithms provided with the probabilistic programming tool.
+Moreover, probabilistic programming packages supply the tools needed to perform inference with these models by taking experimental evidence (data) and fitting the distributions to the data using Bayes\textquotesingle theorem.
+The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using algorithms provided with the probabilistic programming tool.
 
 PyMC3 is a mature Python module dedicated to support probabilistic programming.
-It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (\ac{VI}) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
+It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (ADVI) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
 Therefore, it provides the tools required to build a Bayesian model for the probabilisitc estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
@@
 \end{subfigure}
-\label{fig:posteriors}
 \caption{Annotated posterior probability distribution for the \ac{log2FC} estimate for proteins P0A818 (\textit{E.
@@
 coli} proteins ($\log_2(3)$).}
+\label{fig:posteriors}
 \end{figure}
@@
 \end{subfigure}
+\caption{Projection of the posterior predictive checks on the 2D plane capturing most variance. Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
 \label{fig:ppc}
-\caption{Projection of the posterior predictive checks on the 2D plane capturing most variance. Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
 \end{figure}
@@
 In order to further validate BayesQuant\textquotesingle s performance, it was tested on pairs of groups of 5 proteins from each proteome.
-Each pair was made by proteins with a different number of observed peptides (2,3,4,6,7,10) (see figure \ref{fig:dumbbell}).
+Each pair was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}).
 The more peptides, the more accurate the estimation will be as more data will be available.
@@
     # table with sequence features
-    features_path="data/protein_analysis.tsv")
+    features_path="data/features.tsv")
 \end{minted}

### 2018-07-23 modelling.tex
Fix list of figures

@@
 \end{tabular}
-\caption{Sample data input for the Bayesian model.
+\caption[BayesQuant data input]{Sample data input for the Bayesian model.
 Every row represents a unique peptide.
@@
 \includegraphics[width=\textwidth]{graph}
-\caption{Diagram of the Bayesian model developed in the present work, without sequence modelling of the peptide effect.
+\caption[Bayesian model simplified diagram]{Diagram of the Bayesian model developed in the present work, without sequence modelling of the peptide effect.
 It is represented as a Directed Acyclic Graph (\ac{DAG}) where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
@@
 %\end{adjustbox}
-\caption{Progression of \ac{ELBO} maximisation during the \ac{VI} approximation for proteins P0A818 (\textit{E. coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}) over 40k iterations.}
+\caption[ELBO progression]{Progression of \ac{ELBO} maximisation during the \ac{VI} approximation for proteins P0A818 (\textit{E.
+coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}) over 40k iterations.}
 \label{fig:ELBO}
@@
 %\end{adjustbox}
-\caption{Traceplot of the model fit for proteins P0A818 (\textit{E.
+\caption[Traceplots for 2 proteins]{Traceplot of the model fit for proteins P0A818 (\textit{E.
 coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}).
@@
 \end{subfigure}
-\caption{Annotated posterior probability distribution for the \ac{log2FC} estimate for proteins P0A818 (\textit{E.
+\caption[Posteriors for 2 proteins]{Annotated posterior probability distribution for the \ac{log2FC} estimate for proteins P0A818 (\textit{E.
 coli}, \textbf{A}) and A6NDG6 (\textit{Homo sapiens}, \textbf{B}).
@@
 \end{subfigure}
-\caption{Projection of the posterior predictive checks on the 2D plane capturing most variance. Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
+\caption[Posterior predictive checks]{Projection of the posterior predictive checks on the 2D plane capturing most variance.
+Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
 \label{fig:ppc}
@@
 \includegraphics[width=0.9\textwidth]{performance}
-\caption{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
+\caption[BayesQuant performance]{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
 The intervals are represented by horizontal lines.
@@
 \end{subfigure}
-\caption{Traceplot obtained after fitting the extended model on the test proteins mentioned in \ref{subsec:basic_model}. No feature was found to have any predictive power.}
+\caption[Traceplot from sequence modelling of peptide effect]{Traceplot obtained after fitting the extended model on the test proteins mentioned in \ref{subsec:basic_model}.
+No feature was found to have any predictive power.}
 \label{fig:traceplots_seq}

### 2018-07-23 msreview.tex
Fix list of figures

@@
 \includegraphics[width=\textwidth]{proteomics_skema_book}
-\caption{Overview of a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
+\caption[Bottom-up proteomics analysis]{Diagram over a standard bottom-up proteomics analysis. Figure 1.3 from \cite{Barsnes2008}.}
 \label{fig:proteomics_overview}
@@
 \includegraphics[width=\textwidth]{sample_processing}
-\caption{Overview of the sample processing step prior to mass spectrometry analysis.
+\caption[Sample processing summary]{Diagram of the sample processing step prior to mass spectrometry analysis.
 First, proteins are denatured and digested with a specific protease like Trypsin.
@@
 \includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
+\caption[Mass spectrometer diagram]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
 \label{fig:mass_spectrometer}
@@
 \includegraphics[width=\textwidth]{envelope}
-\caption{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
+\caption[Isotopic envelope]{A doubly charged isotopic envelope with its monoisotopic ion measured at 602.8 Th.
 Each peak in the envelope is separated by 0.5 Th.
@@
 \includegraphics[width=\textwidth]{tandem_ms}
-\caption[]{Illustration of the tandem MS workflow.
+\caption[MS/MS schema]{Illustration of the tandem MS workflow.
 The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
@@
 \includegraphics[width=\textwidth]{shotgun_adapted}
-\caption{Illustration of the different spectra collected in tandem MS.
+\caption[MS/MS spectra]{Illustration of the different spectra collected in tandem MS.
 MS1 spectra record precursor intensity vs \ac{m/z} ratio at different times.
@@
 \includegraphics[width=0.8\textwidth]{abcxyz}
-\caption{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
+\caption[Fragment ions nomenclature]{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
 The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
@@
 \includegraphics[width=\textwidth]{computational_analysis}
-\caption{Diagram of the MS computational analysis.
+\caption[Mass spectrometry computational analysis]{Diagram of the MS computational analysis.
 The PSM process maps the observed spectra to a list of peptides in the reference proteome that could have generated them.
@@
 \includegraphics[width=0.9\linewidth]{pep}
-\caption{Visualizing FDR and PEP.
+\caption[FDR and PEP]{Visualizing FDR and PEP.
 The FDR at a given score \textit{s} is defined as the ratio between False Positives and the sum of True and False Positives found in the list of PSMs of score greater than \textit{s}.
@@
 \includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption{Illustration of the \ac{MBR} step.
+\caption[Match Between Runs]{Illustration of the \ac{MBR} step.
 \textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
@@
 \end{subfigure}
-\caption{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
+\caption[Apex MS1 intensity]{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
 An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.

### 2018-07-23 pipeline.tex
Fix list of figures

@@
 \includegraphics[width=\textwidth]{pipeline}
-\mycaption[Schema of the presented pipeline]{First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
+\caption[Schema of the presented pipeline]{First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
 After that, spectra are searched by means of a search engine to perform the PSM step.
@@
 \end{subfigure}
-\mycaption[Quality control and validation of the 13th fraction of the third replicate in condition L]{\textbf{A} Score distribution for matches to the decoy and the target databases.
+\caption[Quality control and validation of the 13th fraction of the third replicate in condition L]{\textbf{A} Score distribution for matches to the decoy and the target databases.
 \textbf{B} Evolution of the PSM score with confidence.
@@
 \includegraphics[width=0.8\textwidth]{match_percent}
-\mycaption[Percentage of matched spectra in all samples]{The total number of spectra per sample ranged between 5894 and 20249.
+\caption[Percentage of matched spectra in all samples]{The total number of spectra per sample ranged between 5894 and 20249.
 \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
@@
 \end{tabular}
-\caption{\textbf{Protein inference results}. The counts of the different molecular entities detected by peptideShaker are displayed.}
+\caption[rotein inference results]{Protein inference results. The counts of the different molecular entities detected by peptideShaker are displayed.}
 \label{tab:protein_inference}
@@
 \includegraphics[width=\textwidth]{mbr_combined}
-\caption{Match Between Runs with moFF.
+\caption[Proteome benchmark MBR results]{Match Between Runs with moFF.
 \textbf{A} Count of identifications on each run segreagated by source.
@@
 \end{subfigure}
-\caption{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
+\caption[Proteome benchmark apex MS1 intensity results]{\textbf{A} The apex intensity profile across fractions for 2 different peptides, one from \textit{Homo sapiens}, and one from \textit{E.
 coli}.
@@
 \end{tabular}
-\caption{Results of the quantification pipeline.
+\caption[Proteome benchmark dataset quantification results]{Results of the quantification pipeline.
 The 10 protein groups with the lowest \textit{q-val}  as estimated by MSqRob are shown.
@@
 \includegraphics[width=\textwidth]{curves_plot}
-\caption{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
+\caption[Classifier evaluation: ROC and PR curves]{ROC and PR curves displaying the behaviour of 3 different label-free quantification pipelines publicly available.
 \textbf{Comp+Rob} is the pipeline presented in the previous sections.
@@
 \end{subfigure}
-\caption{Agreement between the different pipelines.
+\caption[Pipeline agreement]{Agreement between the different pipelines.
 \textbf{A} The Venn diagram displays the number of \textit{E.
@@
 \includegraphics[width=\textwidth]{combined_plot}
-\caption{Overview over the different pipelines\textquotesingle performance.
+\caption[Pipelines performance]{Overview over the different pipelines\textquotesingle performance.
 Quantification results are segregated by the source organism.

### 2018-07-23 modelling.tex
Backup

@@
 \includegraphics[width=0.9\textwidth]{performance}
-\caption[BayesQuant performance]{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
+\caption[HPDI inferred by BayesQuant]{Visualization of the 95\% \ac{HPDI} inferred from 5 bacterial and 5 human proteins with 2, 3, 4, 6, 7 and 10 peptides.
 The intervals are represented by horizontal lines.

### 2018-07-23 pipeline.tex
Backup

@@
 \includegraphics[width=\textwidth]{pipeline}
-\caption[Schema of the presented pipeline]{First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
+\caption[Schema of the presented pipeline]{\textbf{Schema of the presented pipeline}.
+First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
 After that, spectra are searched by means of a search engine to perform the PSM step.

### 2018-07-23 conclusion.tex
Make conclusion separate file and polish introduction

\chapter{Conclusion of the Thesis}
\label{chap:conclusion}

The pipelines presented in chapter \ref{chap:pipeline} and the MaxQuant suite offered the easiest to deploy and most output rich of all the cost-free pieces of software available for the proteomics community. All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools. More functionalities can be added, including \textit{de novo} and \ac{PTM} search, and missing data management. BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model} can be passed any peptide level dataset and provide estimates of the uncertainty behind traditional \ac{log2FC} estimates and nuisance effects. It can be customised to expand its currently implemented model. The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of the competitive ionization problem. Recently developed tools like RawQuant could be applied in future work to extract from RAW files a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach. The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software, and it is encouraged to continue on that line.

### 2018-07-23 introduction.tex
Extend chapter 3
Finish chapter 3
Fix list of figures
Make conclusion separate file and polish introduction

@@
-\chapter*{Introduction}
+\chapter{Introduction}
 \label{chap:introduction}
 
-\section{Aminoacids and proteins}
+\section{Novozymes and biotechnology}
 
-Proteins represent the last link in the central dogma of biology, where information encoded in DNA, is transcribed to RNA for posterior translation into proteins at the ribosome.
+Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes.
+The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
+Notorious examples of such applications include waste-water treatment, household care or fermentation.
+For example, a product called Fermax was recently released to prevent foam development during the sugarcane ethanol production (see figure \ref{fig:ethanol}).
 
-Proteins are made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group, an amino group, and last but not least, a radical, also called side chain of the aminoacid.
-The side chain differs between aminaocids and generates them from each other.
+\begin{figure}[!h]
+\centering
+  \includegraphics[width=\textwidth]{fermax}
+\caption[Example of bioprocess optimised by Novozymes]{Schematic representation of a bioindustrial process optimised with a Novozymes product, marked as a green circle. Taken from Novozymes intranet.}
+\label{fig:ethanol}
+\end{figure}
+
+The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}.
+Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans.
+
+\ac{NZ} takes advantage of these molecules by matching processes where a chemical transformation takes place, with an enzyme that could improve its efficiency, defined in terms of resource consumption, speed or waste subproduction.
+Moreover, \ac{NZ} thrives to improve the performance of the selected enzymes in the target process by modifying its Wild Type (WT) form.
+Proteomics plays a key role on the research process required, thus \ac{NZ} has a keen interested in being at the vanguard of research and tool development in this field.
+
+\section{Bioinformatics and proteomics}
+
+The majority of enzymes consist of proteins.
+But what are proteins?
+Proteins are molecules made up of 20 basic units, called aminoacids.
+All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group (\ce{COOH}), an amino group (\ce{NH_2}), and last but not least, a radical, also called side chain of the aminoacid.
 A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
-Even though the side chains are all different, they can be classified into four different groups: aliphatic, polar, positively charged and negatively charged (see figure \ref{fig:aminoacids}).
+The side chain differs between aminaocids and creates the aminoacidic diversity   (see figure \ref{fig:aminoacids}).
 
@@
   \includegraphics[width=0.8\textwidth]{aminoacids2.png}
-  \caption{CAPTION AND REFERENCE}
+  \caption[The 20 aminoacids]{Molecular structure of the 20 aminoacids, organized by chemical properties into five groups.
+Taken from \footnote{\href{https://cnx.org/contents/vxf032Bc@4/Proteins}{https://cnx.org/contents/vxf032Bc@4/Proteins}}.}
   \label{fig:aminoacids}
@@
 
-Two aminoacids are joined together through the formation of a peptidic (covalent) bond between them.
-Such a linkage is formed by removal of the elements of water (dehydration) from the $\alpha$-carboxyl group of one amino acid and the $\alpha$-amino group of another \cite{Nelson2008}.
-The remaining $\alpha$-amino and $\alpha$-carboxyl groups are available for linkage to other aminoacids, and in this way peptidic chains or peptides can be created.
+The carboxyl group and amino group of different aminoacids can undergo a dehydration that generates a covalent bond, called peptidic bond, subsequently joining two aminoacids (see figure \ref{fig:peptidic_bond}).
+The remaining amino and carboxyl groups can in turn be joined to new aminoacids, giving way for the formation of a biopolymer, where each subunit is an aminoacid.
+A peptide consists of a small chain of aminoacids joined this way.
+If made long enough, it acquires complex structures and becomes a protein proper \cite{Nelson2008}.
 
-While there are 20 basic units that constitute the majority of naturally observable proteins, their side chains can be modified both by physiological processes and by experimental procedures cite.
-One frequent instance of such modifications is the oxidation of methionine.
+\begin{figure}
+  \centering
+  \includegraphics[width=0.8\textwidth]{peptidic_bond.png}
+  \caption[The peptidic bond]{The peptidic bond brings together different aminoacids into a single protein. Taken from \cite{Nelson2008}.}
+  \label{fig:peptidic_bond}
+\end{figure}
 
-\section{The protein-focused biotechnology industry}
+Proteomics is the branch of bioinformatics whose purpose is the application of algorithms and data analysis to biological studies dealing with proteins.
+With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
 
-Proteins carry out most of the cell\textquotesingle s molecular functions, they work as molecular agents that can perform an extremely wide range of tasks.
-The advent of biotechnology has sought to take advantage of this power, either by using proteins as present in natural conditions (wild type) or engineered by humans.
-This potential economic activity is carried out by several biotech companies, including Novozymes A/S (NZ).
-
-\ac{NZ} is a company whose line of business consists of the development of enzymatic products performing chemical transformations in different industrial processes.
-The application of these products, instead of conventional chemical-based solutions, has the advantage that they require less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
-Notorious examples of such applications include waste-water treatment, household care and the baking industry.
-
-The advancement of the way \ac{NZ} does protein research is thus key to place the organization ahead of its competitors.
-The refinement of the currently used tools and the development of new ones could be of great significance for the company.
-
-Protein research can be approached from different angles.
-This thesis exploited the combination of mass spectrometry (\ac{MS}) and proteomics workflows (see chapter \ref{chap:mass_spec}) for the qualitative and quantitative characterization of protein samples.
+More concretely, proteomics studies how proteins are capable of coordinating their functions by means of complex systems of regulation.
+Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components CITE NETWORK THEORY.
+For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of a frequent disturbance of these regulations: cancer \cite{Saraon2012}.
+The application of proteomics to biological problems is not restricted to medical problems like that, but also biotechnological processes like those mentioned above.
 
@@
 
-In line with the goal of making \ac{NZ} more competitive, this project aimed at the following objectives:
+In line with the goal of advancing \ac{NZ} capabilities in proteomics research, with the final ambition of speeding the research turnover in the company, this project aimed at the following:
 
@@
 
-\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of \ac{MS} data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
+\item Develop an open-source, Linux based and easily deployable pipeline for the analysis of proteomics data, starting at the raw high-throughput data files and ending in the  biological interpretation of the results.
 
-\item Evaluate this pipeline with a benchmark dataset to assess if the pipeline is able to reflect the biological phenomena captured in the data.
+\item Propose a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
 
-\item Establish a label-free quantification probabilistic model that provides relative abundance estimates and a measurement of their uncertainty based on the available data.
+\item Evaluate the tools with benchmark datasets to assess their performance and check they convey the biological phenomena captured in the data.
 
@@
 
-An overview over the \ac{MS} and following computational data analysis steps is presented in \ref{chap:mass_spec}.
-The pipeline development and its benchmark are explained in chapters \ref{chap:pipeline} and \ref{chap:benchmark}, while the modelling problem is introduced in chapter \ref{chap:model}.
+An overview over the main data collection technique used in proteomics, Mass Spectrometry (\ac{MS}), and the ensuing computational data analysis steps is presented in chapter \ref{chap:mass_spec}.
+An inspection of the developed pipeline is provided in chapter \ref{chap:pipeline}, while the probabilistic model is introduced in chapter \ref{chap:model}.
+A benchmark of both tools on a \ac{NZ} dataset is given in chapter \ref{chap:benchmark}.
 Finally, a conclusion of the work is given in chapter \ref{chap:conclusion}.

### 2018-07-23 modelling.tex
Make conclusion separate file and polish introduction

@@
 The posterior distribution can be approximated (inferred) by means of \ac{MCMC} and \ac{VI} methods.
-On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2018}.
+On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2017}.
 Monte Carlo sampling is performed on the Markov Chain, (hence the name of the technique) to randomly explore the parameter space with some heuristics.
@@
 On the other hand, \ac{VI} methods, very recently developed and introduced in PyMC3, provide a fast alternative to \ac{MCMC} methods, yet they are not guaranteed to asymptotically approximate the true posterior.
-In \ac{VI}, optimization, instead of sampling, is employed as way to approximate the posterior \cite{Blei2018}.
+In \ac{VI}, optimization, instead of sampling, is employed as way to approximate the posterior \cite{Blei2017}.
 More concretely, a family of densities $Q$ over the parameters $\theta$ is defined.
@@
 Thus maximizing
-the \ac{ELBO} is equivalent to minimizing the \ac{KL} divergence} \cite{Blei2018}.
+the \ac{ELBO} is equivalent to minimizing the \ac{KL} divergence} \cite{Blei2017}.
 
@@
 The values stored in the trace obtained via VI will follow an approximation to the true posterior, and contain the results of the modelling process.
-Once it\textquotesingle obtained, the inference is complete and model checking ought to be performed to validate the results as explained in section \ref{subsec:model_checking}.
-It is run with
+Once it is obtained, the inference is complete and model checking ought to be performed to validate the results as explained in section \ref{subsec:model_checking}.
+It is run with:
 
@@
                framesep=2mm]{python}
+# Run Posterior Predictive Checks
 bayesquant.ppc()
 \end{minted}
+
+A step by step exposition of the plots and results produced by the program when ran on the proteins  P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}) will follow.
 
@@
 
+After the \texttt{.fit()} method is run, a plot of the maximisation of the \ac{ELBO} is produced automatically.
 Even though the ELBO took extremely low (negative) values at start (see figure \ref{fig:ELBO}), it was maximised fast to values around 0, indicating that the mean-field approximation was acceptably good.
 
-\subsection{Basic model}
+\subsection{Sampling from the approximation to the posterior}
 \label{subsec:basic_model}
 
-A protein from each proteome was used to test BayesQuant\textquotesingle s performance P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}).
-
-The traces produced by BayesQuant using \ac{VI} exhibited a clear random walk behaviour, which decreases the posibility that the approximation was wrong.
+The 10k traces produced by BayesQuant from the \ac{VI} approximation function exhibited a clear random walk behaviour, which decreases the posibility that the approximation was wrong.
 Moreover, the  effects caused by the different peptides and runs were estimated and separated from the overall observed noise, allowing for a better estimation of the \ac{log2FC} (see figure \ref{fig:traceplots}).
@@
 
-This way, a different posterior distribution was computed for the impact of each peptide and run in all the measurements collected for each protein.
+This way, a different posterior distribution was computed for the impact of each "peptide" and "run" in all the measurements collected for each protein.
 
-The study of the \ac{log2FC} posterior probability distribution reveals what are the most likely parameter values and allows for the fulfillment of Bayesian Null Hypothesis Testing.
+The study of the \ac{log2FC} posterior probability distribution (see figure \ref{fig:posteriors}) shows the most likely parameter values and empowers Bayesian Null Hypothesis Testing.
 The 95\% High Probability Density Interval (\ac{HPDI}), which is the narrowest interval containing 95\% of the total probability, indicates the most likely true values of the parameter.
@@
 The \ac{ROPE} is a small interval considered to be essentually the same as the null value \cite{Kruschke}.
-In the present application, the null value was set to 0, and the interval was given a width of 0.4 on each side.
+In this and all future applications in the thesis, the null value is set to 0, and the interval is given a width of 0.4 on each side.
 
@@
 
-The test analysis indicated that P0Q8I8 most likely has a \ac{log2FC} between 1.4 and 1.7, and it\textquotesingle significantly different from 0, as no overlap was observed between \ac{HPDI} and \ac{ROPE}.
+The analysis indicated that P0Q8I8 (\textit{E.
+coli}) most likely has a \ac{log2FC} between 1.4 and 1.7, and it\textquotesingle significantly different from 0, as no overlap was observed between \ac{HPDI} and \ac{ROPE}.
 On the other hand, the human protein was found to acquire a \ac{log2FC} most likely between -0.1 and 0.192.
@@
 
-In order to further validate BayesQuant\textquotesingle s performance, it was tested on pairs of groups of 5 proteins from each proteome.
-Each pair was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}).
+In order to further validate BayesQuant\textquotesingle s performance, it was tested on groups of 5 proteins, one group from each proteome.
+Each pair of groups was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}).
 The more peptides, the more accurate the estimation will be as more data will be available.
-Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed.
+Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed (see figure \ref{fig:dumbbell}).
 

### 2018-07-23 msreview.tex
Make conclusion separate file and polish introduction

@@
-\chapter{Review on mass spectrometry and shotgun proteomics}
+\chapter{Review on mass spectrometry (\ac{MS}) and shotgun proteomics}
 \label{chap:mass_spec}
 
-Life systems consist of complex systems, meaning their behaviour cannot be easily explained by analyzing the individual elements alone.
-Moreover, they present multiple layers of complexity, given by the nature of the elements that make it up.
-The layer provided by proteins is one of them, and its study is called proteomics.
-It is a complex layer because thousands of different proteins can be present in a single cell at any time, and their exact composition and quantities constantly change, responding to the stimuli of the surrounding environment.
-The study of the protein-specific complexity is called proteomics.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
-
-%CITE 1.6.1 COMPUTATIONAL METHODS.
-The existing approaches are divided into two types of paradigms: top-down and bottom-up.
+The main source of data in proteomics is \ac{MS}.
+However, different approaches to how these technique is used make two paradigms in proteomics analysis: top-down and bottom-up.
 In the top-down paradigm, intact proteins are directly used for the analysis.
 In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
-These smaller parts are called peptides.
-\cite{Barsnes2008}  Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
-The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as explained in \ref{subsec:the_detector}.
+These smaller parts are peptides \cite{Barsnes2008}, consisting of around 10 chained aminoacids.
+Such peptides acquire physicochemical properties fitting the requirements of the downstream analytical methods, mainly the mass spectrometer (MS), which performs the data acquisition.
+The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as illustrated in section \ref{subsec:the_detector}.
 This review will focus on the data-dependent approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
+In this approach, the equipment is configured for the analysis of one peptide at a time, with the drawback of being biased for the most abundant peptides.
 
@@
 
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for plenty of sample components.
+MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for eventually thousands of sample components.
 A detailed explanation of the sample processing required prior to MS is given in section \ref{sec:sample_processing}, while an overview on mass spectrometers is given in section \ref{sec:the_mass_spectrometer}.
@@
 
-\LARGE
-\textbf{MS analysis and data generation}
-\normalsize
+\section{Sample processing}
 
-\section{Sample processing}
+Prior to its introduction in the mass spectrometer for shotgun proteomics studies, a protein sample (I) is cleaved into peptides and (II) these peptides are separated by means of some physicochemical properties.
+This way, the analytical equipment works with one peptide at a time.
+
 \label{sec:sample_processing}
@@
 Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
-Then, proteins are subjected to digestion with specific enzymes, which cut the aminoacidic chains following a predictable pattern.
-Trypsin is the most frequently used for this task.
+Then, proteins are subjected to digestion with specific proteases i.e protein-cutting molecules, which cut the aminoacidic chains following a predictable pattern.
+Trypsin is the most frequently used protease.
 It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues.
-As explained in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, since R and K are positively charged aminoacids, the resulting peptide is guaranteed in most cases to be able to capture at least one charge, thus making it visible in the spectrometer.
-All of these properties combined, together with its low prize, makes Trypsin the enzyme of choice in this step for most cases.
+Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before.
+As demonstrated in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
+Moreover, in as much R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed in most cases to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}.
+All of these properties combined, together with its low prize, makes Trypsin the protease of choice in this step for most cases.
 
-Even though enzymes are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
+Even though proteases are very specific, the cleaving process is far from perfect, as there could be: \cite{Barsnes2008}
 %COMPUTATIONAL METHODS 3.2
@@
 
-%\Cref{item:missed_cleavages}
-
 Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
@@
 If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
-This can be surmounted by independently analyzing different fractions of the sample over time.
+This can be surmounted by analyzing one peptide in the sample at a time.
 The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
@@
 %CITE 4.2 computational methods.
-During \ac{HPLC}, the peptide mix loaded into a column containing a stationary and a solid phase.
+During \ac{HPLC}, the peptide mix is loaded into a column containing a stationary and a solid phase.
 These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
 The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the machine will consist of a simplified mix of peptides at a time.
+Therefore, the input to the spectrometer will consist of one peptide at any given time.
 
@@
 
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide mix such as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
+The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
 It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
@@
 Bias in ESI is less predictable.
-This is known as the competitive ionization problem \cite{Tang2004}.
+All the sources of bias introduced during ionization are grouped into the competitive ionization problem \cite{Tang2004}.
 
@@
 \caption[MS/MS schema]{Illustration of the tandem MS workflow.
-The first spectrometer acts as an ion selector, that lets through ions with a given \ac{m/z} ratio.
-The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation occurs.
+The first spectrometer acts as an ion selector, that not only registers spectra, but also lets through ions with a given \ac{m/z} ratio.
+The second spectrometer does not perform mass analysis, but instead provides the medium where peptide fragmentation (see section \ref{subsec:fragmentation}) occurs.
 Finally the third spectrometer records fragment mass spectra.
@@
 The difference can be exploited to deduce which aminoacid makes up for that difference.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum.
-explained in \ref{sec:search_engines}.
+If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum, as  explained in section \ref{sec:search_engines}.
 
@@
 More importantly, chemical modifications are not lost in the process.
-
-\LARGE
-\textbf{Computational analysis}
-\normalsize
 
@@
 
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting spectra exhibit variability manifested in missing peaks or spurious ones.
+Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting data exhibit variability manifested in both missing and spurious peaks.
 Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
@@
 \includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption[Match Between Runs]{Illustration of the \ac{MBR} step.
+\caption[Match Between Runs module]{Illustration of the \ac{MBR} step.
 \textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
@@
 \end{subfigure}
-\caption[Apex MS1 intensity]{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
+\caption[Apex MS1 intensity module]{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
 An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.

### 2018-07-23 pipeline.tex
Make conclusion separate file and polish introduction

@@
 \end{subfigure}
-\caption[Quality control and validation of the 13th fraction of the third replicate in condition L]{\textbf{A} Score distribution for matches to the decoy and the target databases.
+\caption[]{\textbf{Quality control and validation of the 13th fraction of the third replicate in condition L} \textbf{A} Score distribution for matches to the decoy and the target databases.
 \textbf{B} Evolution of the PSM score with confidence.
@@
 \end{subfigure}
-\caption[Pipeline agreement]{Agreement between the different pipelines.
+\caption[Pipeline agreement plots]{Agreement between the different pipelines.
 \textbf{A} The Venn diagram displays the number of \textit{E.

### 2018-07-23 thp1.tex
Fix list of figures
Backup
Make conclusion separate file and polish introduction

@@
 \end{subfigure}
-\caption{Compomics+MSqRob results.
+\caption[Compomics+MSqRob results on THP-1 dataset]{\textbf{Compomics+MSqRob results}.
 \textbf{A} Histogram of the estimated \ac{log2FC}.
@@
 \includegraphics[width=\linewidth]{bq_comb_bq}
-\caption{BayesQuant results.
+\caption[Compomics+BayesQuant results on the THP-1 dataset]{\textbf{BayesQuant results}.
 \textbf{A} Histogram of the mean of the posterior distributions.
@@
 Black dots represent the mean, and intervals not overlapping the \ac{ROPE} are shown red.
-\textbf{C} Histogram of the 95\% \ac{HPDI} width.
+\textbf{C} Histogram of the 95\% \ac{HPDI} width, which can be used as proxy for the overall uncertainty in the quantified data.
 \textbf{D} Scatter plot of the 95\% \ac{HPDI} width and the  posterior \ac{log2FC} mean.}
@@
 
-The UniprotKB to Gene name mapping of the whole set of \ac{DAP}s (including multiprotein groups) returned by MSqRob returned a list of 51 entities.
+The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multiprotein groups) returned by MSqRob returned a list of 51 entities.
 The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were overrepresented compared to the human background.
-In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched: (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
+50 entities were mapped to the database (only \textit{C12orf42} was unmapped).
+In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched (see table \ref{tab:pathway_enrichment}) (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
 Moreover, the finding of several RNA related pathways exposes the presence of potential cell reprogramming, essential in any change in protein profile.
@@
 \end{tabular}
-\caption{Pathway Enrichment results. The 10 most significant pathways found in Reactome and Kegg are shown arranged by decreasing significance. }
+\caption{Pathway Enrichment results.
+50 proteins were mapped to pathway databses and an enrichment was performed.
+The 10 most significant pathways found in Reactome and Kegg are shown arranged by decreasing significance.
+The columns \textit{Ov} and \textit{Size} indicate the number of proteins shared between pathway and query, and the number of proteins included in the pathway, respectively.
+The input was always 49 set to protein groups.}
 \label{tab:pathway_enrichment}
@@
 The database returns a protein association network (see figure \ref{fig:STRING}), that helps interpreting the information contained in gene sets.
-49/51 terms were mapped to the resource and 15 of them were found to be included in the "immune system process" set, placing the biological process as the third most enriched with an FDR of 0.0387.
+49/51 terms were mapped to the resource and 15 of them were found to be included in the "immune system process" set, placing this biological process as the third most enriched with an FDR of 0.0387.
 27 genes belonged to the "cellular nitrogen compound metabolic process" (FDR 0.0128).
@@
 \includegraphics[width=\linewidth]{string_normal_image}
-\caption{STRING interaction network.
+\caption[Protein interaction network]{\textbf{STRING interaction network}.
 Every node represents a protein in the list, and an edge between any pair represents evidence for some kind of association or interaction between them.
@@
 
+A discussion over the major issues and strenghts detected while obtaining these results follows.
+
 \subsection{Increasing the experiment data-throughput}
 
-6 samples organized in two different treatments of 3 biological replicates without sample prefractionation each were analyzed.
+Six samples organized in two different treatments of three biological replicates without sample prefractionation each were analyzed.
 The resulting data throughput allowed for the performance of biological inference, as it successfully captured the undergoing biological processes.
-Nevertheless, a more in-depth results report would have been possible with (I) sample prefractionation, which improves peptide separation and allows for the collection and identification of more spectra \cite{Righetti2005}, and (II) the inclusion of technical replicates to differentiate the biological variability from experimental noise.
+Nevertheless, more in-depth results would have been possible with (I) sample prefractionation, which improves peptide separation and allows for the collection and identification of more spectra \cite{Righetti2005}, and (II) the inclusion of technical replicates to differentiate the biological variability from experimental noise.
 
@@
 One of the most frequent issues found when analyzing proteomics datasets is the handling of missing values \cite{Lazar2016}.
-The presence of missing data in many of the rows in the input moFF file led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups.
-Thanks to the missing data handling implemented in MSqRob quantified 1433 proteins, enough to continue the analysis (see table \ref{tab:balance}).
-The fact that the number of observed groups was 2787 implies that the management of missing values could be improved in both tools, and particularly in BayesQuant.
+The presence of missing data in many of the rows in the data passed to BayesQuant led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups.
+Thanks to the missing data handling implemented in MSqRob, this tool quantified 1433 proteins, enough to continue the analysis (see table \ref{tab:balance}).
+However, the fact that the number of observed groups was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.
 
@@
 \section{Conclusion}
-The results presented in this work demonstrated the usability of the tools presented in the previous chapters in an actual \ac{NZ} setting.
-Moreover, a walk-through over the functional analyses that can be performed with this output clearly showed they can be used to perform computational analyses and biological inference on protein samples with open-source software with no cost for the company.
-However, the absence of proper missng data handling developed into the inability to quantify hundreds of proteins, signaling what are the next steps to improve the tools.
+A fine biological interpretation of the results depends on the original question that motivated the study, and thus the work on this chapter did not focus on that, but instead showed how it can be started with the outputsability of the tools presented in the previous chapters.
+The results indicated that these tools can be deployed to perform computational analyses and biological inference on protein samples with open-source software with no cost for \ac{NZ}.
+However, the absence of proper missing data handling developed into the inability to quantify hundreds of proteins, signaling what are the next steps to improve the tools.

### 2018-07-24 hierarch_diagram.tex
Add Kruschke-like diagram and fix some section referencing

% Hierarchical diagrams, DBDA
% ---------------------------------
% Tinu Schneider, October 2013

\documentclass[12pt]{article}

\usepackage[T1]{fontenc} 
\usepackage[utf8]{inputenc}

\usepackage[margin = 20mm, nohead, nofoot]{geometry}

\usepackage{tikz}
    \usetikzlibrary{positioning, arrows, backgrounds, decorations, 
    				decorations.pathmorphing, decorations.markings, calc}

\graphicspath{{./distribution_diagrams/svg_without_names}}

\pagestyle{empty}

\usepackage{tgheros} 
\renewcommand*\familydefault{\sfdefault}

% Macros for the distributions
% greek
\newcommand{\normalGreek}[1]{\greek{#1}{normal} } 
\newcommand{\gammaGreek}[1]{\greek{#1}{gamma} }
\newcommand{\halfnormalGreek}[1]{\greek{#1}{halfnormal} }

\newcommand{\greek}[2]{\large$#1$\\ \includegraphics*[scale = 0.25]{distribution_diagrams/png_with_params/#2.png} }

% letters (constants)  
\newcommand{\normalLetters}[1]{\letters{#1}{normal}} 
\newcommand{\gammaLetters} [1]{\letters{#1}{gamma}}   
\newcommand{\halfnormalLetters} [1]{\letters{#1}{halfnormal}}

\newcommand{\letters}[2]{$#1$\\ \includegraphics*[scale = 0.25]{distribution_diagrams/png/#2.png} }

\usepackage{stackengine, scalerel}
\newcommand\Tunderbrace[3][]{%
  \def\tmp{#2}%
  \setbox0=\hbox{\tmp}%
  \stackunder[1pt]{%
    \stackunder[0pt]{\tmp}{\rotatebox{90}{\scaleto[2ex]{#1\{}{\wd0}}}%
  }{%
    \scriptsize #3%
  }%
}
\newcommand\Toverbrace[3][]{%
  \def\tmp{#2}%
  \setbox0=\hbox{\tmp}%
  \stackon[1pt]{%
    \stackon[0pt]{\tmp}{\rotatebox{90}{\scaleto[2ex]{#1\}}{\wd0}}}%
  }{%
    \scriptsize #3%
  }%
}

\begin{document} % ------------------------------ %

% distances
\def\smallY{0.25}
\def\midY{2.25}
\def\bigY{4.5}

\def\smallX{1.2}
\def\midX{1.5}
\def\bigX{2.7}

\begin{tikzpicture}[ framed, on grid,
   	node distance = \midY cm, 
   	inner frame xsep = 2.5ex,
   	inner frame ysep = 3.0ex,
    	box/.style = { rectangle,
	       		color = lightgray, 
	       		text = black,  
	       		text width = 2.0cm, 
	       		align = center},
	    boxFramed/.style = {box, 
	    		% draw, % uncomment this line, if you want the frames
	    		inner sep = 0pt},
	    boxFramedRound/.style = {boxFramed, rounded corners = 8pt},
	    boxFormula/.style = {box, inner sep = 1pt},
       	boxLegend/.style = {align = left, 
       			text width = 4cm, 
       			xshift = 2.2cm, 
       			yshift = 1.2cm,  
       			font = \scriptsize, 
       			color = black!70},
    	allArrows/.style = {font = \footnotesize, 
    			color = black!80, text = black},
    	arrSnake/.style = {allArrows, 
    			>=stealth', 
    			->, 
    			shorten >= 2pt, 
    			line join = round, 
    			decorate, 
  				decoration={snake, segment length = 10, 
  							amplitude = 1.2, post = lineto, 
  							post length = 4pt} },
     	implies/.style  = {allArrows, -implies, double, 
     			double equal sign distance, shorten >= 1pt}, 
     	labArrow/.style = {pos = 0.39},
     	labArrowLegend/.style = {labArrow, above, font = \tiny, inner sep = 3pt}
  	]

   	% nodes for the distributions and equation(s)
   	% starting at the bottom				
	\node [box, xshift = 2cm] (yi) {\Large $y_{i}$};

   	\node [boxFramedRound, above = of yi, yshift = -.2cm] (final) {\normalGreek{\mu_{i}, \epsilon} } ;

   	\node [boxFormula, above = of final, xshift=-\bigX cm, text width = 2.5 cm] (linMod)   
   			{ \large $\beta^0 +$\Toverbrace{$\beta^p_i + \beta^t_i + \beta^r_i$}{}} ;

   			   	\node [boxFormula, above = of linMod, xshift= 0 cm, text width = 2.5 cm] (selectPar)   
   			{ \large \Tunderbrace{$\beta^p \cdot x^p_i + \beta^t \cdot x^t_i + \beta^r \cdot x^r_i$}{} } ;

   	\node [boxFramed, above = of final,  xshift =  \bigX cm, yshift = \smallY cm] (proEpsilon) {\halfnormalGreek{S} } ;

   	\node [boxFramedRound, above = of selectPar, yshift=.2cm, xshift = -\bigX-2.4 cm]   (proIntercept)   {\normalGreek{\mu_0, \sigma_0} };
   	\node [boxFramedRound, above = of selectPar, yshift=\smallY cm, xshift = \smallX cm]   (proBeta)    {\normalGreek{\mu_j, \sigma_j} } ;

   	\node [boxFramed, above = of proBeta,  xshift = -\smallX-1.2 cm] (promu)  {\normalGreek{M_j, \sigma_j} } ;
   	\node [boxFramed, above = of proBeta,  xshift =  \smallX+1.2 cm] (prosigma) {\halfnormalGreek{S_j} } ;

   	% Arrows	    	
   	\draw[arrSnake] (final.south) --  node[right, labArrow] {$\ i$} (yi);

    \draw[implies]  ($(linMod.south)+(1,-.15)$) -- node[right, labArrow] {$\ \ \ i$} (final.125);
    \draw[implies]  ($(selectPar.south)+(1.25,0)$) -- node[right, labArrow] {$\ i|p,t,r$} ($(linMod.100)+(1.1,0)$);

    \draw[arrSnake] (proEpsilon.south) -- (final.55);

    \draw[arrSnake] (proIntercept.south) -- node[left, labArrow] {$\ \ i$} ($(linMod)+(-1,0)$);       	 	
    \draw[arrSnake] (proBeta.south) --  node[left,  labArrow] {$j=p\quad $} ($(selectPar)+(-1,.8)$);
    \draw[arrSnake] (proBeta.south) --  node[below,  labArrow] {$j=t\quad $} ($(selectPar)+(1,.8)$); 
    \draw[arrSnake] (proBeta.south) --  node[right,  labArrow] {$j=r\quad $} ($(selectPar)+(2.5,.8)$);

    \draw[arrSnake] (promu.south) -- (proBeta.122);       	 	
    \draw[arrSnake] (prosigma.south) -- (proBeta.57);  
    \draw[arrSnake] (prosigma.west) -- (promu.east);

  	% Legend; needs improvement...
   	\node [boxLegend] at ( current bounding box.south west) (legendTitle) 
   		{%
   		\begin{tabular}{ll}
   			$A, \dots, Z$  & Constants \\
   			$a, \dots, z$ & Variables \\
   			$\alpha, \dots, \epsilon$ & Parameters \\ %[1ex]

			 & deterministic \\[-0.4ex]
			 & relation \\ %[1ex]

			 & stochastic \\[-0.4ex]
			 & relation \\ %[1ex]

			\multicolumn{2}{l}{$i, j = 1, 2, 3, \dots $} \\
   		\end{tabular}
   	};

   	\def\xLegendArrowLeft{-3.75}
   	\def\xLegendArrowRight{-2.65}

   	\def\yLegendDouble{0.83}
   	\def\yLegendSnake{0.25}

   	\draw[implies, black!70]  (\xLegendArrowLeft, \yLegendDouble) -- (\xLegendArrowRight, \yLegendDouble);    	   	
    \draw[arrSnake, black!70] (\xLegendArrowLeft, \yLegendSnake) -- (\xLegendArrowRight, \yLegendSnake);

\end{tikzpicture}

\end{document} % ------------------------------ %

### 2018-07-24 modelling.tex
Add Kruschke-like diagram and fix some section referencing

@@
 The Bayes theorem is an extremely versatile tool taking a role analogous to that of statistics like T, F, or $\chi^2$.
-Unlike them, which are tailored to specific scenarios, it can be used compute the probability distribution of almost any parameter in any model.
+Unlike them, which are tailored to specific scenarios, it can be used to compute the probability distribution of almost any parameter in any model.
 4 terms can be distinguished in its expression
@@
 
-The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterir probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}.
+The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterior probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}.
 
@@
 \section{Results}
+
+\begin{figure}[!h]
+ \centering 
+ \includepdf[pages=-]{hierarch_diagram.pdf}
+\end{figure}
 

### 2018-07-24 pipeline.tex
Add Kruschke-like diagram and fix some section referencing

@@
 \begin{figure}[!h]
+\centering
 \includegraphics[width=0.8\textwidth]{match_percent}
-\caption[Percentage of matched spectra in all samples]{The total number of spectra per sample ranged between 5894 and 20249.
+\caption[Percentage of matched spectra in all samples]{\textbf{Percentage of matched spectra in all samples}.
+The total number of spectra per sample ranged between 5894 and 20249.
 \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.

### 2018-07-24 thp1.tex
Add Kruschke-like diagram and fix some section referencing

@@
 
-The protein ratios across conditions are not known in ordinary datasets In order to judge the resolving power of the mass spectrometry and shotgun proteomics workflows presented in this work, an in-house dataset attempting to capture an immunological response was generated.
+The protein ratios across conditions are not known in ordinary datasets.
+In order to judge the resolving power of the mass spectrometry and shotgun proteomics workflows presented in this work, an in-house dataset attempting to capture an immunological response was generated.
 Therefore, even though true protein ratios were not known \textit{a priori}, the analysis should find proteins related to the immunological response to be significantly differentially abundant.
@@
 Most proteins had a very narrow HPDI, as shown in figure \ref{fig:compomics_bq}C.
-As expected, the width of the interval exhibited some degree of correlation with   the mean of the posterior, as shown in \ref{fig:compomics_bq}D.
+As expected, the width of the interval exhibited some degree of correlation with the mean of the posterior, as shown in figure \ref{fig:compomics_bq}D.
 

### 2018-07-25 introduction.tex
Extend references

@@
 Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes.
-The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact.
+The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact \cite{madigan2000brock}.
 Notorious examples of such applications include waste-water treatment, household care or fermentation.
@@
 The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}.
-Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans.
+Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.
 
@@
 Proteins are molecules made up of 20 basic units, called aminoacids.
-All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group (\ce{COOH}), an amino group (\ce{NH_2}), and last but not least, a radical, also called side chain of the aminoacid.
+All aminoacids share a common chemical structure, where a carbon atom ($C_\alpha$) is covalently bonded to a hydrogen atom, a carboxyl group (\ce{-COOH}), an amino group (\ce{-NH_2}), and last but not least, a radical, also called side chain of the aminoacid.
 A slight deviation from this pattern exists in proline, where the radical is bound to the nitrogen atom, making it an iminoacid.
@@
 A peptide consists of a small chain of aminoacids joined this way.
-If made long enough, it acquires complex structures and becomes a protein proper \cite{Nelson2008}.
+If made long enough, it acquires complex structures and becomes a protein \cite{Nelson2008}.
 
@@
 More concretely, proteomics studies how proteins are capable of coordinating their functions by means of complex systems of regulation.
-Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components CITE NETWORK THEORY.
+Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components \cite{Milo2002}.
 For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of a frequent disturbance of these regulations: cancer \cite{Saraon2012}.
-The application of proteomics to biological problems is not restricted to medical problems like that, but also biotechnological processes like those mentioned above.
+The application of proteomics to biological problems is not restricted to medical problems such as cancer, but also biotechnological processes like those mentioned above.
 

### 2018-07-25 msreview.tex
Add Kruschke-like diagram and fix some section referencing
Extend references

@@
 The main source of data in proteomics is \ac{MS}.
-However, different approaches to how these technique is used make two paradigms in proteomics analysis: top-down and bottom-up.
-In the top-down paradigm, intact proteins are directly used for the analysis.
+However, different approaches to how this technique is used make for two paradigms in proteomics analysis: top-down and bottom-up \cite{Joshi2017}.
+In the top-down paradigm, intact proteins are directly used for the analysis \cite{Sinitcyn2018}.
 In the bottom-up paradigm (see figure \ref{fig:proteomics_overview}), the proteins are first cleaved into smaller parts, and these parts are then used for identification, characterization, and quantification.
@@
 The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as illustrated in section \ref{subsec:the_detector}.
-This review will focus on the data-dependent approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
+This review will focus on the Data-Dependent Aquisition (\ac{DDA} approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
 In this approach, the equipment is configured for the analysis of one peptide at a time, with the drawback of being biased for the most abundant peptides.
@@
 Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before.
-As demonstrated in \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
+As demonstrated in section \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
 Moreover, in as much R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed in most cases to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}.
@@
 
-Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function.
+Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function \cite{Siepen2007}.
 This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
@@
 
-If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample.
+If presented with the problem of analyzing a mixture of peptides, the capacities of mass spectrometers are easily overwhelmed by a too complex mixture, resulting in the analysis of only a minor part of the total protein of the sample \cite{Barsnes2008}.
 This can be surmounted by analyzing one peptide in the sample at a time.
@@
 The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
-Therefore, the input to the spectrometer will consist of one peptide at any given time.
+Therefore, the input to the spectrometer will consist of few peptides at any given time.
 
@@
 The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
-It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}).
+It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.
 
@@
 The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too.
+Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too \cite{Joshi2016}.
 Thus, the charge exhibited by an ion is not obvious when produced via ESI.
 Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW), thus introducing a bias.
+Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW) \cite{Hessling2013}, thus introducing a bias \cite{Hessling2013}.
 Bias in ESI is less predictable.
-All the sources of bias introduced during ionization are grouped into the competitive ionization problem \cite{Tang2004}.
+All the sources of bias introduced during ionization cause the competitive ionization problem \cite{Zhang2009} \cite{Tang2004}.
 
 The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
+It is measured in $\frac{Da}{e}$, sometimes referred to as Th (Thomson).
 
@@
 On the other hand, due to the natural occurrence of isotopes, particularly \ce{^{13}_{}C}, the same peptide will induce the measurement of several signals with very close \ac{m/z} values.
-They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}), and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms.
+They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}) \cite{Mirzaei2016}, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms or other naturally occurring isotopes.
 Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
@@
 The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide, as increases of 1 Da at charge 1 will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
+Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide with the following expression.
+
+\begin{equation}
+m/z = \frac{(m + z × H^+)}{z} \text{for \textit{z}} \in {1, 2,...}
+\end{equation}
+
+where $H^+$ is the mass of a single charge (1 Da) and $z$ is the number of charges acquired.
+A single charge will induce a separation of 1 \ac{m/z}, while at charge 2 it will be $1/2 = 0.5$ \ac{m/z}, at 3 $1/3 = 0.33$ \ac{m/z}, and so on (see figure \ref{fig:envelope}).
 
@@
 Shotgun proteomics analyses make use of two or more mass spectrometers connected in series, giving rise to the so-called Tandem MS (\ac{MS/MS}) workflow.
-In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}.
+In this setting, each mass spectrometer collects a different type of spectra and thus different information (see figure \ref{fig:shotgun}).
 An extra spectrometer, usually a quadrupole, is introduced in between.
@@
 
-\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio.
+\item The first spectrometer records the intensity versus \ac{m/z} ratio of the peptides eluting from the column at a given time and is used to filter ions exhibiting a selected \ac{m/z} ratio (in the \ac{DDA} protocol).
 
@@
 
-\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The produced spectrum can be use to read out the peptide sequence.
+\item The last spectrometer records the intensity versus \ac{m/z} ratio of the fragments produced in the previous step. The resulting spectrum can be use to read out the peptide sequence.
 
@@
 
+Fragmentation occurs in the second spectrometer.
+But what is it, and why is it done?
 The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
 Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
+Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
 
@@
 
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
 During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
@@
 The difference can be exploited to deduce which aminoacid makes up for that difference.
-If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS2} spectrum, as  explained in section \ref{sec:search_engines}.
+If this process is repeated for enough pairs of contiguous fragments, with the right software, a sequence can be read from the \ac{MS/MS} spectrum, as  explained in section \ref{sec:search_engines}.
 
@@
 \ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform \textit{in silico} prediction of the peptides produced in the protein digestion step and their expected spectrum.
+During this step, search engines perform (I) \textit{in silico} prediction of the peptides produced in the protein digestion step and (II) their expected spectrum.
 
@@
 
-Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting data exhibit variability manifested in both missing and spurious peaks.
+Given the stochastic nature of the protein cleavage and spectra recording processes, the resulting data exhibit variability manifested in both missing and spurious peaks \cite{Stein1999}.
 Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
@@
 Multiple search engines exist that implement different matching and scoring algorithms.
-The most modern ones include MS-GF+, MS-Amanda, Comet, X!Tandem or Andromeda.
+The most modern ones include MS-GF+ \cite{Kim2014}, MS-Amanda \cite{Dorfer2014}, Comet \cite{Eng2013} and Andromeda \cite{Cox2011}.
 Notably, the results of each individual search engine can be combined to gather their strengths, at the expense of an increased computational cost and time \cite{Shteynberg2013}.
@@
 
-A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after the its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
+A common filter is the false discovery rate (FDR), usually set to 1\%, indicating that after its application, only one out of a hundred filtered matches are expected to be false positives (wrong matches).
 
 The most commonly used method to compute the \ac{FDR} of a list of matches is the target-decoy search.
-Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target).
+Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database i.e it simulates random matches.
+The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target) \cite{Elias2010}.
 
@@
 
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM.
+The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM \cite{Nesvizhskii2010}.
 Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
@@
 The combination of all the aforementioned computational analyses yields a list of protein identities that reports the protein composition i.e qualitative information of the original sample.
-However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, rather than protein presence alone.
+However, in most proteomics applications, quantitative data can be of great interest, as many biological phenomena are manifested mainly through changes in the protein abundances, cell compartments, rather than the simple dichotomy of the protein\textquotesingle s existence \cite{Barsnes2008}.
 For instance, cancer cells in response to a drug could modulate the abundance of several proteins without removing them from the cytosol or introducing new ones.
@@
 Different techniques, like Stable Isotope Labeling by Amino acids in Cell culture (SILAC) or Isotope-Coded Affinity Tags (ICAT), differ in the nature of the label and the way it is introduced.
-Remarkably, peptide labeling costs can be quite high.
+Most of them have potential limitations \cite{Patel2009}.
 This, together with the limited amount of samples that can be compared makes a case for label-free quantification.
@@
 It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing.
+These methods have the advantage that they are very simple to implement and don't require any further data processing \cite{Barsnes2008}.
 
@@
 
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract the apex intensity of the identified peak clusters, as shown in figure \ref{figure:moff_apex}.
+For a more robust XIC based quantification, a feature extraction step is usually executed to extract consistent descriptors of identified peak clusters in the RT-m/z plance, as shown in figure \ref{figure:moff_apex}.
 
@@
 This translates into improved results when compared to the other aforementioned methods \cite{Goeminne2015}.
-This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.
+%This hypothesis is the motivation for the method explained in chapter \ref{chap:model}.

### 2018-07-25 pipeline.tex
Extend references

@@
 
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016} , was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
-The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) \ac{MBR} and feature extraction, (IV) protein inference and (V) relative quantification.
-Its output can be passed to follow-up analyses in R or Python to get a biological interpretation of the results.
+A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
+The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) protein inference, (IV) \ac{MBR} and feature extraction, and (V) relative quantification.
+Its output can be passed to follow-up analyses in online databases or Bioconductor to get a biological interpretation of the results.
 A benchmark of its performance was accomplished using the proteome benchmark dataset published in \cite{Cox2014}.
@@
 Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
+The organization of a pipeline with comprehensive documentation, open-source nature, and easy implemetation and customisation was asserted to be lacking in the literature.
 
@@
 The preprocessing of the RAW files produced by the mass spectrometer into the MGF open format enabled searchGUI to dispose of the registered spectra.
-PeptideShaker quality control and filtering capabilities (see figure \ref{figure:qc_validation}) carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values.
+PeptideShaker quality control and filtering capabilities carried out the required search results validation.
+As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation}).
 Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.

### 2018-07-30 pipeline.tex
Proofread chapt 1

@@
 \caption[Schema of the presented pipeline]{\textbf{Schema of the presented pipeline}.
-First, .RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
+.RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
 After that, spectra are searched by means of a search engine to perform the PSM step.
@@
 
-Several proteomics pipelines are available on the internet under different licensing conditions.
+Several proteomics pipelines are available under different licensing conditions.
 Many are released as closed-source commercial software, where information on how the program works is kept from the user.
 This is a serious drawback as it hinders the study of the implemented models and its customisation.
-Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, good broad base or format exchangeability.
+Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, a broad user base or format exchangeability.
 MaxQuant, a freeware but closed-source, monolithic proteomics analysis suite \cite{Cox2008}, is developed for Windows and has only very recently been ported to Linux \cite{Sinitcyn2018}.
 Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-The organization of a pipeline with comprehensive documentation, open-source nature, and easy implemetation and customisation was asserted to be lacking in the literature.
+The organization of a pipeline with comprehensive documentation, open-source nature, and easy implementation and customisation was asserted to be lacking in the literature.
 
@@
 The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.
+These files were used as input for the Compomics+Rob pipeline.
 
-The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which is available as an excel file.
+The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
 The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
@@
 coli (strain K12)} (UP000000625) and \textit{Homo sapiens} (UP000005640), downloaded in June 2018.
+A database accounting for common contaminants was also factored in.
 The decoy database was created using the \texttt{FastaCLI} utility in SearchGUI by reversing all sequences in the target.
@@
 
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker version 1.16.22 utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker version 1.16.22 utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{A summary of the validation filters in PeptideShaker https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
 By default, the FDR was set to 1\%.
@@
 
-\subsection{Data refinement}
+\subsection{\ac{MBR} and Apex MS1 intensity extraction}
 
-The moFF command line utility \cite{Argentini2016} was applied to perform (I) match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
+The moFF command line utility \cite{Argentini2016} was applied to (I) perform match-between-runs and (II) extract MS1 apex intensity of each peak cluster.
 This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
@@
 
-Relative quantification was performed using the MSqRob utility by passing the peptide summary file from moFF.
+Relative quantification was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob).
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
@@
 
-Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the MSqRob package \cite{Goeminne2016}, was fit to every protein individually.
+Once preprocessing is done, a ridge regression model with Huber weights and empirical Bayes estimation of protein variance, implemented in the \texttt{fit.model()} function, was fit to every protein individually.
 The peptide and fraction effects were considered random, while the condition was treated as a fixed effect.
@@
 Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
-Intensities were log2 transformed and averaged.
+Intensities were $\log_2$ transformed and averaged.
 The difference between conditions H and L was taken as estimate of the log2FC.
@@
 PeptideShaker quality control and filtering capabilities carried out the required search results validation.
-As expected, matches to the target and decoy exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation}).
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1 \% FDR filter.
+As expected, matches to the target and decoy databases exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation} A).
+Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1\% FDR filter.
 The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
@@
 \end{subfigure}
-\caption[]{\textbf{Quality control and validation of the 13th fraction of the third replicate in condition L} \textbf{A} Score distribution for matches to the decoy and the target databases.
+\caption[Quality control and validation]{\textbf{QC and validation of the 13th fraction of the third replicate in condition L} \textbf{A} Score distribution for matches to the decoy and the target databases.
 \textbf{B} Evolution of the PSM score with confidence.
@@
 
- The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\%.
+The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\% (see figure \ref{figure:qc_validation} B).
 
@@
 \textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
-\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.
-The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.}
+The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.
+\textbf{B} The mean for each analysed fraction across conditions and replicates is displayed together with error bars to represent the standard deviation.}
 \label{fig:match_percent}
@@
 \end{tabular}
-\caption[rotein inference results]{Protein inference results. The counts of the different molecular entities detected by peptideShaker are displayed.}
+\caption[Protein inference results]{Protein inference results. The counts of the different molecular entities detected by peptideShaker are displayed.}
 \label{tab:protein_inference}
@@
 
-A total of 46595 peptides were inferred to be present in at least one of the fractions (see table \ref{tab:protein_inference}).
-Of them more than 27 thousand mapped to a unique protein.
-However, a significant amount mapped to more than one protein.
+A total of 46595 peptides were identified in at least one of the fractions (see table \ref{tab:protein_inference}).
+Of them more than 27 thousand mapped to a unique protein (\textit{unique peptides}).
+However, a significant amount mapped to more than one protein (\textit{degenerate peptides}).
 Such peptides are called non-unique peptides.
 The 25056 detected proteins were grouped into 10039 protein groups.
-Protein groups are peptide generating entities for which enough data to confirm the presence of at least one of them is available, but not to exactly assess which of them.
-Thus, protein inference algorithms create them when a does not uniquely map to a protein because it is not possible to distinguish which protein it truly comes from.
+Protein groups are peptide generating entities for which enough data is available to confirm the presence of at least one of them, but not to precisely assess which of them.
 
@@
 For example, protein isoforms, consisting of protein sequences differing in potentially only one aminoacid, are very difficult to resolve.
-PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group.
-
+PeptideShaker executes a smart protein grouping by harnessing the annotation of proteins and classifying protein groups based on how well the annotation backs the protein group \cite{Vaudel2015}.
 However, in the present analysis the quality of the protein group was not taken into account, as explained in the Materials and Methods.

### 2018-07-30 hierarch_diagram.tex

@@
 
-\usepackage[T1]{fontenc} 
-\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc} 
+%\usepackage[utf8]{inputenc}
 
@@
 
-   	\node [boxFramedRound, above = of selectPar, yshift=.2cm, xshift = -\bigX-2.4 cm]   (proIntercept)   {\normalGreek{\mu_0, \sigma_0} };
+   	\node [boxFramedRound, above = of selectPar, yshift=.2cm, xshift = -\bigX-2.4 cm]   (proIntercept)   {\normalGreek{M_0, S_0} };
    	\node [boxFramedRound, above = of selectPar, yshift=\smallY cm, xshift = \smallX cm]   (proBeta)    {\normalGreek{\mu_j, \sigma_j} } ;
@@
     \draw[arrSnake] (proBeta.south) --  node[left,  labArrow] {$j=p\quad $} ($(selectPar)+(-1,.8)$);
-    \draw[arrSnake] (proBeta.south) --  node[below,  labArrow] {$j=t\quad $} ($(selectPar)+(1,.8)$); 
+    \draw[arrSnake] (proBeta.south) --  node[xshift=0cm,  labArrow] {$j=t\quad $} ($(selectPar)+(1,.8)$); 
     \draw[arrSnake] (proBeta.south) --  node[right,  labArrow] {$j=r\quad $} ($(selectPar)+(2.5,.8)$);
@@
 
-			\multicolumn{2}{l}{$i, j = 1, 2, 3, \dots $} \\
+			\multicolumn{2}{l}{$i = 1, \dots, N$} \\
+			\multicolumn{2}{l}{$j \in$ peptide, treatment, run} \\
+
    		\end{tabular}

### 2018-07-30 msreview.tex

@@
 The main source of data in proteomics is \ac{MS}.
-However, different approaches to how this technique is used make for two paradigms in proteomics analysis: top-down and bottom-up \cite{Joshi2017}.
+However, different approaches to how this technique is used make for two paradigms in proteomics analysis: top-down and bottom-up \cite{Joshi2016}.
 In the top-down paradigm, intact proteins are directly used for the analysis \cite{Sinitcyn2018}.
@@
 These inference steps can be condensed to the peptide and protein inference problems, explained in section \ref{sec:inference}.
-A third computational problem needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
+A third computational challenge needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
 This is the quantification problem, explained in section \ref{sec:quantification}.
@@
 Finally the third spectrometer records fragment mass spectra.
-Taken from \footnotemark{}.}
+Taken from \href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}.}
 \label{fig:tandem_ms}
@@
 
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
 \stepcounter{footnote}
@@
 \includegraphics[width=0.8\textwidth]{abcxyz}
-\caption[Fragment ions nomenclature]{The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
+\caption[Fragment ions nomenclature]{\textbf{Fragment ions nomenclature}.
+The common fragments and their relation to the peptide sequence can be organized into 2 groups of 3 series each.
 The abc fragments keep the N-terminal residue, while xyz keep the C-terminal one.
@@
 During peptide fragmentation, bonds along the peptidic chain are broken, turning the peptide into smaller fragments.
-These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions.
+These fragments will consist of truncated versions of the original peptide at different positions, thus making it possible to read an \ac{m/z} ratio difference between any pair of fragment ions \cite{Barsnes2008}.
 The difference can be exploited to deduce which aminoacid makes up for that difference.
@@
 
-Computational analysis of MS data starts with the matching of the spectra to a referene proteome (see figure \ref{fig:computational_analysis}).
+Computational analysis of MS data starts with the matching of the spectra to a reference proteome (see figure \ref{fig:computational_analysis}).
 \ac{MS} search engines are capable of performing the crucial step of peptide to spectrum matching (PSM).
-During this step, search engines perform (I) \textit{in silico} prediction of the peptides produced in the protein digestion step and (II) their expected spectrum.
+During this step, search engines attempt to match each observed spectrum to theoretical spectra based on the reference proteome.
+Theoretical spectra are obtained via (I) \textit{in silico} simulation of the spectra produced by the peptides contained in the proteome, in turn obtained by performing a virtual cleaveage of the sequences in the proteome, or (II) spectral libraries, consisting of annotated observed spectra \cite{Li2012}.
 
@@
 Furthermore, random (wrong) matches can be returned by the PSM process when running against a sufficiently big database.
-This translates to the generation of multiple matches, of which one, if any, will be correct.
-Therefore, the lists of matches need to be somehow ranked by goodness-of-fit.
+This translates to the generation of multiple matches for a single spectrum, of which one, if any, will be correct.
+Therefore, the lists of matches needs to be somehow ranked by goodness-of-fit.
 The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
@@
 Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database i.e it simulates random matches.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original database (target) \cite{Elias2010}.
+The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original (target) database \cite{Elias2010}.
 
 All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
+Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target, the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
 This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
@@
 
-The equation tells us that we can compute the FDR at any score by counting how many decoy hits have a greater score ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
+The equation tells us that the FDR at any score $s$ can be computed by counting how many decoy hits have a score greater than $s$ ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
 Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
@@
 \item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
-This is not trivial as peptides are degenerate and frequently map to more than one protein.
+This is not trivial as peptides are degenerate and frequently map to more than one protein \cite{Li2012}.
 \end{enumerate}
 
-The result of the PSM step returns the inferred list of peptides.
-The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be inferred using different algorithms.
-The degenerate nature of peptides is dealt with the Occam\textquotesingle s razor principle, which states that the most likely solution is the simplest one.
-Thus, protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
+The result of the PSM step returns an inferred list of peptides.
+The ensemble of proteins most likely to have generated the list of peptides stemming from the filtered PSMs can be identified using different algorithms \cite{Li2012} \cite{Huang2012}.
+Protein inference algorithms aim at explaining the maximum amount of peptides using the least amount of proteins.
 
@@
 
-In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs.
+In the label-free quantification approach, peptides from different samples are not labelled differently and are thus distinguished by their presence in different, independent \ac{MS} runs \cite{Barsnes2008}.
 In order to account for the introduced inter-run variability in peptide identifications and \ac{RT}, a match-between-runs (MBR) processing step can be carried out (see figure \ref{figure:moff_mbr}).
@@
 \includegraphics[width=0.9\linewidth]{mbr_workflow}
-\caption[Match Between Runs module]{Illustration of the \ac{MBR} step.
-\textbf{A} the information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
+\caption[Match Between Runs module]{\textbf{\ac{MBR} step}.
+The information gathered from matches in replicate runs is put to use during a reanalysis of the spectra.
 Modified from supplementary information in \cite{Argentini2016}.}
@@
 
-Quantification can be \ac{SC} or \ac{XIC} based.
+Quantification can be spectrum counting (\ac{SC}) or extractde ion currents (\ac{XIC}) based.
 
 On the one hand, spectrum counting based quantification is the simplest method in proteomics.
-The number of peptides a peptide is detected in the spectrometer is detected as proxy for its abundance.
-It relies on the rationale that highly abundant peptides will have a higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
-These methods have the advantage that they are very simple to implement and don't require any further data processing \cite{Barsnes2008}.
+The number of times a peptide is detected in the spectrometer is used as proxy for its abundance.
+It relies on the rationale that highly abundant peptides will have an accordingly higher intensity and are thus more likely to trigger the acquisition of MS/MS spectra.
+These methods have the advantage that they are very simple to implement and don\textquotesingle t require any further data processing \cite{Barsnes2008}.
 
@@
 
-For a more robust XIC based quantification, a feature extraction step is usually executed to extract consistent descriptors of identified peak clusters in the RT-m/z plance, as shown in figure \ref{figure:moff_apex}.
+For a more robust XIC based quantification, a feature extraction step is usually executed to extract consistent descriptors of identified peak clusters in the RT-m/z plane, as shown in figure \ref{figure:moff_apex}.
 
@@
 \end{subfigure}
-\caption[Apex MS1 intensity module]{Illustration of the apex intensity extraction step.\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
-An ion current at the same m/z ratio, thus corresponding to the same precursor ion, is detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
+\caption[Apex MS1 intensity module]{\textbf{The apex intensity extraction step}.
+\textbf{A} Visualization of a 3D peak cluster on the RT / m/z plane.
+Ion currents are detected with varying degrees of intensity over a more or less narrow time window, spreading the signal over time.
 Taken from \cite{Smith2014}.

### 2018-08-01 modelling.tex

@@
 
-The current label-free quantification methods reviewed in section \ref{sec:quantification} rely on frequentist statistics, which return point estimates of model parameters such as the estimate of the abundance ratio (fold change) across conditions.
+The current label-free quantification methods, reviewed in section \ref{sec:quantification}, rely on frequentist statistics, which return point estimates of model parameters such as the abundance ratio (fold change) across conditions.
 However, a Bayesian  approach to this problem, that we know of, is lacking in the literature.
@@
 
-Both the LFQ and MSqRob quantitication engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
+Both the LFQ and MSqRob quantification engines presented in chapter \ref{chap:pipeline} took a frequentist approach to the problem of protein quantification.
 Such approaches, establish a null hypothesis $H_0$, complementary to the actual hypothesis being tested (alternative hypothesis $H_1$).
@@
 The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools returned point estimates of the log2FC between a pair of conditions and a p-value, which provides a measurement of the probability of the null hypothesis.
+This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a p-value, which measures the probability of the null hypothesis.
 The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about the probability of the alternative (our) hypothesis.
+Thus, p-values do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.
 
@@
 \item $P(\theta | y)$: the updated probability distribution of the model parameter, with the information extracted from the observed data.
-Since it reflects the beliefs about the phenomena after observing data, it is called \textbf{posterior} probability distribution,as opposite to the prior.
+Since it reflects the beliefs about the phenomena after observing data, it is called \textbf{posterior} probability distribution, as opposite to the prior.
 
@@
 
-It can be applied to the quantification problem, where $\theta$ turns into the log2FC parameter we try to estimate.
+It can be applied to the quantification problem, where $\theta$ turns into the \ac{log2FC} parameter we try to estimate.
 All is needed is the computation of the posterior probability distribution.
@@
 Monte Carlo sampling is performed on the Markov Chain, (hence the name of the technique) to randomly explore the parameter space with some heuristics.
-This heuristic guarantees convergence of the empirical estimate with the true posterior, provided a big enough sample size.
+The heuristics guarantee convergence of the empirical estimate on the true posterior, provided a big enough sample size.
 Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018} \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
-The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} \footnote{A very nice tutorial on how Metropolis-Hastings works \href{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}} \footnote{Likewise, this resource provides very illustrative explanations of Hamiltonian-MC and NUTS \href{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}} have given way to the much more efficient sampler \ac{NUTS} \cite{Hoffman2011}.
+The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} \footnote{A very nice tutorial on the Metropolis-Hastings algorithm \href{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}} have given way to the much more efficient sampler \ac{NUTS} \footnote{Likewise, this resource provides very illustrative explanations of Hamiltonian-MC and NUTS \href{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}} \cite{Hoffman2011}.
 
 On the other hand, \ac{VI} methods, very recently developed and introduced in PyMC3, provide a fast alternative to \ac{MCMC} methods, yet they are not guaranteed to asymptotically approximate the true posterior.
-In \ac{VI}, optimization, instead of sampling, is employed as way to approximate the posterior \cite{Blei2017}.
+In \ac{VI}, optimization, instead of sampling, is employed as the way to approximate the posterior \cite{Blei2017}.
 More concretely, a family of densities $Q$ over the parameters $\theta$ is defined.
@@
 
-From \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}} \textit{the term variational is used because the best $q$ in $Q$ is picked.
-The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation.}
+\textit{The term variational is used because the best $q$ in $Q$ is picked.
+The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation}\footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
 
@@
 The mean-field family of densities is one of the most simple, as it ignores all dependencies between latent variables, and treats them as independent.
-From \href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html} the mean-field approximations works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}.
+The mean-field approximation works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}\footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
 A generic mean-field family member is shown in equation \ref{eq:mean_field}.
@@
 The demanding computational cost of the \ac{MCMC} and \ac{VI} schemes has only been recently met by the power of modern computers, thus making the approach feasible.
+
+\subsection{Hierarchical modelling}
+\label{subsec:hierarchy}
+
+The data recorded in this particular problem consist of \ac{MS1} intensities. They are affected by two main sources of variation:
+
+\begin{enumerate}
+\item The fixed effect that the researcher aims at unraveling with proteomics (treatment effect).
+\item Random noise produced by experimental procedures, sequence bias, etc, which lead to a loss of data quality and resolving power.
+\end{enumerate}
+
+A mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
+A mathematical framework implemented to model these effects, similar to the one used by MSqRob:
+
+\begin{equation}\label{eq:model}
+y_{ijkl} = \beta_{i}^{0} + \beta_{ij}^{treatment} + \beta_{ik}^{peptide} + \beta_{il}^{run} + \epsilon_{i}
+\end{equation}
+
+where $y_{ijkl}$ stands for the $\log_2$-transformed intensity registered in treatment $j$, peptide $k$ and run $l$.
+This way, separate treatment, peptide and run effects are distinguished.
+$i$ stands for the protein index, which is evidently the same for all measurements belonging to peptides mapped to the same protein.
+$\epsilon$ models the noise that cannot be explained with the three effects aforementioned, and is the same for all measurements associated to the same protein.
+
+For example, the measurement $y_{P,A,p,1}$ corresponds to a measurement mapped to protein $P$.
+The measurement is impacted by a fixed effect, produced by treatment $A$, but also the peptide effect in peptide $p$ and the run effect in run $1$.
+An intercept term was also added to account for the background protein intensity value.
+
+The data is thus said to be affected by different effects acting independently, and in order to model them properly, multilevels need to be defined.
+It can be done via three approaches: pooled, unpooled, and hierarchically (see figure \ref{fig:multilevel}).
+Hierarchical modelling is used when the sampling variance is not the only source of variation among the parameters, but they still share a dependency.
+The intercept and the three effects were modelled as independent hierarchies.
+
+\begin{figure}[!h]
+\centering
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Pooled}
+\includegraphics[width=0.85\linewidth]{model_pooled}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Unpooled}
+\includegraphics[width=0.85\linewidth]{model_unpooled}
+\end{subfigure}
+\bigskip
+\begin{subfigure}{.9\textwidth}
+\centering
+\caption*{Hierarchical}
+\includegraphics[width=0.85\linewidth]{model_hierarchical}
+\end{subfigure}
+\caption[]{Alternative multilevel modelling approaches.
+A pooled model assumes that the parameter distribution is the same for all the datapoints.
+The unpooled model represents the opposite case, every datapoint is modelled by a different probability distribution.
+Finally, a hierarchical model settles for a middle ground where the parameters will not be exactly the same but not completely different either.
+This is achieved by generating a global distribution from which the parameter for each datapoint is sampled \footnotemark{}.}
+\label{fig:multilevel}
+\end{figure}
+
+\footnotetext{\href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
+
+This way, independent and identically distributed (\ac{IID}) priors were set for the intercept, treatment, peptide and run effects.
+The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} prior.
+A diagram of the resulting model is displayed in figure \ref{fig:daft_model}).
+
+\begin{figure}[!h]
+\centering
+\includegraphics[width=\textwidth]{graph}
+\caption[Bayesian model simplified diagram]{Diagram of the Bayesian model developed in the present work.
+It is represented as a Directed Acyclic Graph (\ac{DAG}) where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
+The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to.
+The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.
+The model hierarchies are organized in boxes.
+The one corresponding to the treatment effect can be read as \textit{the treatment effect $\beta_{ij}^{t}$ observed in the data is modelled as a probability distribution conditional on the value of $\mu^t$ and $\sigma^t$, which are in turn probability distributions governed by the hyperparameters 0 and 1.}}
+\label{fig:daft_model}
+\end{figure}
 
@@
 \item Neither reaching convergence nor the best \ac{VI} approximation guarantee that the fitted model actually captures the data generation process.
-Therefore, the true posterior could be a bad reflection of the natural process being model and the best approximation will not solve it.
+Therefore, the true posterior could be a bad reflection of the natural process being modelled and the best approximation will not fix it.
 Posterior Predictive Checks are carried out to measure how well the model captured the data properties \cite{Kruschke}.
@@
 Moreover, probabilistic programming packages supply the tools needed to perform inference with these models by taking experimental evidence (data) and fitting the distributions to the data using Bayes\textquotesingle theorem.
-The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using algorithms provided with the probabilistic programming tool.
+
+%The fit of the model to the data entails the computation of the posterior probability distribution, which is tallied using algorithms provided with the probabilistic programming tool.
 
@@
 
-The final state of the data is shown in table \ref{tab:data_model} for a single protein.
+The final state of the data is shown in table \ref{tab:data_model} for two different proteins.
 
@@
 
-When running with active sequence modelling, the peptide sequences and their neighborhood in the protein was extracted from \textit{E.
-coli} and \textit{Homo sapiens} proteome databases (\ref{subsec:database_preparation}).
-The protein neighborhood was defined by a window spanning 15 aminoacids on both sides of the peptide.
+When running with active sequence modelling, the peptide sequences were extracted from \textit{E. coli} and \textit{Homo sapiens} proteome databases (\ref{subsec:database_preparation}).
+%The protein neighborhood was defined by a window spanning 15 aminoacids on both sides of the peptide.
 It was extracted using the seqinr \cite{Charif2007} and GenomicRanges \cite{Lawrence2013} packages implemented in R and Bioconductor.
@@
 The following properties were extracted: amino-acid percentage, peptide length, molecular weight, mass/length ratio, isoelectric point, aromaticity and instability.
-
-\subsection{Hierarchical modelling}
-\label{subsec:hierarchy}
-
-The \ac{MS1} intensity measurements are affected by two main sources of variation:
-
-\begin{enumerate}
-\item The fixed effect that the researcher aims at unraveling with proteomics (treatment effect)
-\item Random noise produced by experimental procedures, sequence bias, etc, which lead to a loss of data quality and resolving power.
-\end{enumerate}
-
-A mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
-The mathematical framework implemented to model these effects was similar to the one used by MSqRob:
-
-\begin{equation}\label{eq:model}
-y_{ijkl} = \beta_{ijkl}^{0} + \beta_{ij}^{treatment} + \beta_{ik}^{peptide} + \beta_{il}^{run} + \epsilon_{i}
-\end{equation}
-
-where $y_{ijkl}$ stands for the $\log_2$-transformed intensity registered in treatment $j$, peptide $k$ and run $l$.
-This way, separate treatment, peptide and run effects are distinguished.
-$i$ stands for the protein index, which is evidently the same for all measurements belonging to peptides mapped to the same protein.
-$\epsilon$ models the noise that cannot be explained with the three effects aforementioned, and is the same for all measurements associated to the same protein.
-
-For example, the measurement $y_{P,A,p,1}$ corresponds to a measurement mapped to protein $P$.
-The measurement is impacted by a fixed effect, produced by treatment $A$, but also the peptide effect in peptide $p$ and the run effect in run $1$.
-An intercept term was also added to account for the starting intensity value.
-
-The data is thus affected by different effects acting independently, and in order to model them properly, multilevels need to be defined.
-Multilevel effects can be modelled with three approaches: pooled, unpooled, and hierarchically (see figure \ref{fig:multilevel}).
-Hierarchical modeling is used when the sampling variance is not the only source of variation among the parameters, but they still share a dependency.
-The intercept and the three effects were modeled as independent hierarchies.
-
-\begin{figure}[H]
-\centering
-\begin{subfigure}{.9\textwidth}
-\centering
-\caption*{Pooled}
-\includegraphics[width=0.85\linewidth]{model_pooled}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{.9\textwidth}
-\centering
-\caption*{Unpooled}
-\includegraphics[width=0.85\linewidth]{model_unpooled}
-\end{subfigure}
-\bigskip
-\begin{subfigure}{.9\textwidth}
-\centering
-\caption*{Hierarchical}
-\includegraphics[width=0.85\linewidth]{model_hierarchical}
-\end{subfigure}
-\caption[]{Alternative multilevel modelling approaches.
-A pooled model assumes that the parameter distribution is the same for all the datapoints.
-The unpooled model represents the opposite case, there every datapoint is assumed to be modelled by a different probability distribution of the parameter for each.
-Finally, a hierarchical model settles for a middle ground where the parameters will not be exactly the same but not completely different either.
-This is achieved by generating a global distribution from which the parameter for each datapoint is sampled \footnotemark{}.}
-\label{fig:multilevel}
-\end{figure}
-
-\footnotetext{\href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
-
-This way, independent and identically distributed (\ac{IID}) priors were set for the intercept, treatment, peptide and run effects.
-The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} prior.
-A diagram of the resulting model is displayed in figure \ref{fig:daft_model}).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=\textwidth]{graph}
-\caption[Bayesian model simplified diagram]{Diagram of the Bayesian model developed in the present work, without sequence modelling of the peptide effect.
-It is represented as a Directed Acyclic Graph (\ac{DAG}) where nodes represent hyperparameters or random variables (in a circle) and directed edges  represent the dependencies between them.
-The nodes on the top depict the hyperparameters of the model, governing the prior probability distributions represented by the nodes to which their edges lead to.
-The remaining nodes articulate the probabilistic model and all lead to a final node ($y_{ijkl}$) which represents the modelling of the observed data with the instantiated model.
-The model hierarchies are organized in boxes.
-The one corresponding to the treatment effect can be read as \textit{the treatment effect $\beta_{ij}^{t}$ observed in the data is modelled as a probability distribution conditional on the value of $\mu^t$ and $\sigma^t$, which are in turn probability distributions governed by the hyperparameters 0 and 1.}}
-\label{fig:daft_model}
-\end{figure}
 
@@
                framesep=2mm]{python}
-         betae = pm.Normal("betae", mu, sigma, n_elements)  
+         beta = pm.Normal("beta", mu, sigma, n_elements)  
 \end{minted}
@@
 \section{Results}
-
-\begin{figure}[!h]
- \centering 
- \includepdf[pages=-]{hierarch_diagram.pdf}
-\end{figure}
 
@@
 A step by step exposition of the plots and results produced by the program when ran on the proteins  P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}) will follow.
+
+%\begin{figure}[H]
+% \centering 
+% \includegraphics{hierarch_diagram.pdf}
+% \caption{}
+% label{fig:dbda}
+%\end{figure}
+\begin{figure}[!h]
+\input{hierarch_diagram}
+\caption{Kruschke-like diagram of the Bayesian model}
+\label{fig:dbda}
+\end{figure}
 

### 2018-08-01 pipeline.tex
Backup

@@
 Since the \textit{E.
-coli} proteome was mixed on a 3:1 ratio in condition H, a \ac{log2FC} of $log2(3)=1.58$ is expected for proteins coming from this organism.
+coli} proteome was mixed on a 3:1 ratio in condition H, a \ac{log2FC} of $log_2(3)=1.58$ is expected for proteins coming from this organism.
 Likewise, human proteins should have a \ac{log2FC} of 0.
@@
 coli} proteins found to be differentially abundant by each pipeline.
-\textbf{B} The log2FC estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
+\textbf{B} The \ac{log2FC} estimated in all three pipelines for the 121 shared proteins is plotted by pairs to evaluate their correlation.}
 \label{fig:venn_cor}

### 2018-08-01 conclusion.tex
Backup

@@
 
-The pipelines presented in chapter \ref{chap:pipeline} and the MaxQuant suite offered the easiest to deploy and most output rich of all the cost-free pieces of software available for the proteomics community.
+The pipelines presented in chapter \ref{chap:pipeline} offer an output-rich and easy to deploy cost-free pieces of software.
 All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools.
 More functionalities can be added, including \textit{de novo} and \ac{PTM} search, and missing data management.
-BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model} can be passed any peptide level dataset and provide estimates of the uncertainty behind traditional \ac{log2FC} estimates and nuisance effects.
+BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model} can be passed any peptide level dataset and provide estimates of the uncertainty behind frequentist \ac{log2FC} estimates and nuisance effects.
 It can be customised to expand its currently implemented model.
-The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of the competitive ionization problem.
-Recently developed tools like RawQuant could be applied in future work to extract from RAW files a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach.
+The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of its origin.
+Recently developed tools like RawQuant could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach.
 The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software, and it is encouraged to continue on that line.

### 2018-08-01 modelling.tex
Backup

@@
 A mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
-A mathematical framework implemented to model these effects, similar to the one used by MSqRob:
+A mathematical modelling of these effects (see equation \ref{eq:model}), similar to the one used by MSqRob, can be deployed to attempt to minimise the noise:
 
@@
 \end{figure}
-
 \footnotetext{\href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
@@
 The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} prior.
-A diagram of the resulting model is displayed in figure \ref{fig:daft_model}).
+A diagram of the resulting model is displayed in figure \ref{fig:daft_model}.
 
@@
 \ac{ELBO} optimisation was run for 40k iterations.
-Optimisation was validated by visual inspection of \ac{ELBO} traceplots over the 40k iterations.
+Optimisation was validated by visual inspection of \ac{ELBO} traceplots.
 Once optimisation was validated, 10k samples were drawn from the approximate distribution to simulate the true posterior.
@@
 
-The code below formalizes in PyMC3 code the bias in \ac{MS1} intensity measurements due to a random effect.
+A PyMC3 implementation of the priors set on the bias in \ac{MS1} intensity measurements due to a random effect is shown here.
 
@@
 
-The snippet above loads the program, reads in the data, selects the data for a specific protein, and builds a model that matches the number of peptides observed.
+The snippet above loads the program, reads in the data, selects a specific protein, and builds a model that matches the number of peptides observed.
 The posterior distribution can be computed using pure MCMC or VI methods, as stated in section \ref{subsec:posterior_compute}.
-The VI procedure will now be explained, due to its significantly better performance.
+The VI procedure will now be explained, due to its significantly better performance in this particular problem.
 
@@
 
-A step by step exposition of the plots and results produced by the program when ran on the proteins  P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}) will follow.
+A step by step exposition of the plots and results produced by the program when ran on the proteins  P0A818 (\textit{E. coli}) and A6NDG6 (\textit{Homo sapiens}) follows.
 
@@
 
-After the \texttt{.fit()} method is run, a plot of the maximisation of the \ac{ELBO} is produced automatically.
+A plot of the maximisation of the \ac{ELBO} is automatically produced when the texttt{.fit()} method is run.
 Even though the ELBO took extremely low (negative) values at start (see figure \ref{fig:ELBO}), it was maximised fast to values around 0, indicating that the mean-field approximation was acceptably good.
@@
 The \ac{ROPE} is a small interval considered to be essentually the same as the null value \cite{Kruschke}.
-In this and all future applications in the thesis, the null value is set to 0, and the interval is given a width of 0.4 on each side.
+In this and all future applications in the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrality set to $[-0.4, 0.4]$ based on the data.
 
@@
 A green vertical line marks the expected \ac{log2FC} estimate for \textit{E.
-coli} proteins ($\log_2(3)$).}
+coli} proteins ($\log_2(3)=1.58$).}
 \label{fig:posteriors}
@@
 The analysis indicated that P0Q8I8 (\textit{E.
-coli}) most likely has a \ac{log2FC} between 1.4 and 1.7, and it\textquotesingle significantly different from 0, as no overlap was observed between \ac{HPDI} and \ac{ROPE}.
-On the other hand, the human protein was found to acquire a \ac{log2FC} most likely between -0.1 and 0.192.
+coli}) most likely has a \ac{log2FC} between 1.4 and 1.7, and was found to be significantly different from 0, as no overlap was observed between the \ac{HPDI} and the\ac{ROPE}.
+On the other hand, the human protein was found to acquire a \ac{log2FC} most likely between -0.1 and 0.2.
 This range is fully contained within the \ac{ROPE}, which indicates that the \ac{log2FC} was practically 0 (see figure \ref{fig:posteriors}).
@@
 \caption[Posterior predictive checks]{Projection of the posterior predictive checks on the 2D plane capturing most variance.
-Simulated datapoints are shown in blue, whereas actual observations are shown in red.}
+Simulated datapoints are shown in blue, whereas actual observations are shown on a clear color.}
 \label{fig:ppc}
@@
 
-Indeed, a general trend of width reduction was observed with the increased amount of peptides.
-For example, the widest intervals are observed on proteins with just two peptides available, and the narrowest on proteins with ten.
-Moreover, the \ac{HPDI} tended to align more strongly around an agreement value as well.
+Indeed, a general trend of width reduction was observed as the amount of peptides increased.
+For example, the widest intervals are observed on proteins with just two observed peptides, and the narrowest on proteins with ten.
+Moreover, the \ac{HPDI}s tended to align more strongly around an agreement value as well.
 
@@
 where $\theta_k$ is the weight of the $k^{th}$ feature, and $x_k$ is the numerical value of the $k^{th}$ feature.
-$\sigma^{\theta}$ is the prior for the standard deviation of the Normal all feature weights are sampled from.
+$\sigma^{\theta}$ is the prior for the standard deviation of the distribution all feature weights are sampled from.
 
@@
 
-While BayesQuant performed well on the majority of proteins, it took around 30 seconds to approximate each protein\textquotesingle posterior distributions using \ac{VI}, and more than two minutes with \ac{MCMC} methods.
+While BayesQuant performed well on the majority of proteins, it took around 30 seconds to approximate each protein\textquotesingle s posterior distributions using \ac{VI}, and more than two minutes with \ac{MCMC} methods.
 This translates to a dataset of 1k proteins taking 500 minutes (>8h) to be processed.
@@
 It could be used to measure how much more likely a model taking into account a treatment effect is compared to another one where the treatment effect is considered null i.e the \ac{log2FC} is set to 0.
-The result would provide an alternative criteria for the declaration of a protein being differentially abundant.
+The result would provide an alternative criterion for the declaration of a protein being differentially abundant.
 
@@
 Unfortunately, during the development of the project, it was acknowledged that the peptide effect is a very difficult phenomenon to capture in a simple Bayesian model as the one presented in this work.
-This is due to the motley and and multilevel nature of the noise caused by the aminoacidic sequences:
+This is due to the motley and multilevel nature of the noise caused by the aminoacidic sequences:
 
@@
 This phenomenon implies that the \ac{PSM} step is more difficult as a different spectrum is expected for molecules with the same peptide sequence but different charges.
-The decreased identifications and quality of the peaks makes feature extraction algorithms like the Apex module in moFF would extract worse \ac{MS1} intensities, in turn injecting noise in the quantification data.
+The decreased identifications and quality of the peaks worsen the performance of feature extraction algorithms like the Apex module in moFF.
+In turn, this results in the injection of more noise in the quantification data.
 
-\item The characterization of powerful sequence properties that specifically capture the differential ionization efficiency is undone work, as most feature extractors are designed to solve different problems, like protein sructure prediction.
+\item The characterization of powerful sequence properties that specifically capture differential ionization efficiency is undone work, as most feature extractors are designed to solve different problems, like protein sructure prediction.
 
@@
 
-Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of proper neural networks, capable of deciphering the intrincate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
+Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of proper neural networks, capable of deciphering the intricate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
 
@@
 For example, a functional analysis program could decide to drop proteins declared differentially abundant if the uncertainty is greater than a threshold.
-A Gene Set Enrichment Analysis (\ac{GSEA}) could make use of the probabilistic information to refine the results of the enrichment.
+A Gene Set Enrichment Analysis (\ac{GSEA}) could make use of the probabilistic information to refine its results.
 
@@
 
-A Bayesian framework for the relative quantification of protein abundance rations (Fold-Change, FC) using \ac{MS1} intensity was presented in this work.
-Unlike the currently prevailing methods in the field, it provides uncertainty estimates that provide a direct interpretation about the accuracy of the quantification.
-Moreover, preliminary results on the sequence modelling of the peptide bias revealed that more complex architectures and sequnce features are required to solve this problem.
-Finally, the method can be used to investigate the presence of batch effects in \ac{MS} experiments for their minimisation in future.
+A Bayesian framework for the relative quantification of protein abundance rations (Fold-Change, FC) using \ac{MS1} intensity was presented in this chapter.
+Unlike the currently prevailing methods in the field, it gives uncertainty estimates that provide a direct interpretation about the accuracy of the quantification.
+Moreover, preliminary results on the sequence modelling of the peptide bias revealed that more complex architectures and sequnce features are required to model noise with peptide sequences.
+Finally, the method can be used to investigate the presence of batch effects in \ac{MS} experiments for their minimisation in future experiments.
 While the software can be integrated in any pipeline producing peptide-level data, further testing and improvements are probably required for optimal performance.

### 2018-08-01 pipeline.tex
Backup

@@
 
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or \textit{log2-fold-changes} (\ac{log2FC}) across condition H and condition L.
+The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) across conditions H and L.
 During quantification, it is important to take into account as many of the effects influencing protein quantification.
 In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
-The MSqRob quantification engine treats effects differently depending on whether or not they are random, or fixed.
+The MSqRob quantification engine treats effects differently depending on whether or not they are random or fixed.
 Fixed effects should have a consistent impact in the ion currents measured in the spectrometer, whereas random effects are truly random and thus unpredictable.
@@
 As explained in section \ref{sec:search_engines}, it has been shown that the combined usage of multiple search engines can increase identifications, since the different statistical frameworks implemented in each of them compensate each other\textquotesingle s caveats.
-Only MS-GF+ was used in thi work for simplicity.
+Only MS-GF+ was used in this work for simplicity.
 Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
@@
 Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it psimultaneously provides an extra random effect that difficults quantification if not handled properly.
+However, it simultaneously provides an extra random effect that difficults quantification if not handled properly.
 

### 2018-08-01 thp1.tex
Backup

@@
 
-After preprocessing and missing-data handling, a dataset of 1641 peptides was input to BayesQuant, and 269 proteins were quantified.
-A histogram of the mean of the inferred posterior distributions manifests the centrality of its distribution, with most values within the predefined \ac{ROPE} of [-0.4, 0.4].
+After preprocessing and missing-data handling, a dataset of 1641 peptides was passed to BayesQuant, and 269 proteins were quantified.
+A histogram of the mean of the inferred posterior distributions manifests its centrality, with most values within the predefined \ac{ROPE} of [-0.4, 0.4].
 In line with what was observed in the results above, the distribution was found to be skewed toward negative values (see figure \ref{fig:compomics_bq}A).
@@
 \end{tabular}
-\caption{\ac{DAP} declared by the Compomics+BayesQuant pipeline.}
+\caption{\ac{DAP}s declared by the Compomics+BayesQuant pipeline.}
 \label{tab:thp1_bq_results}
@@
 On the other hand, five terms (5, 7, 14, 15, 19) were related to the immunological response.
-The remaining terms were connected to RNA functions and the nervous system.
+The remaining terms were connected to RNA functions.
+
+These results can be better visualized by making use of a treemap (see figure \ref{fig:treemap}), which makes the interpretation of table \ref{tab:gsea} more straightforward.
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{revigo_treemap.png}
+\caption[REVIGO treemap]{\textbf{REVIGO treemap}.
+A top-level representation of the enriched GO-terms.
+Each top-level term is mapped to an area whose size is proportional to the significance of the term.
+This way, a big area indicates more enrichment \cite{Supek2011}.}
+\label{fig:treemap}
+\end{figure}
 
@@
 \section{Conclusion}
-A fine biological interpretation of the results depends on the original question that motivated the study, and thus the work on this chapter did not focus on that, but instead showed how it can be started with the outputsability of the tools presented in the previous chapters.
-The results indicated that these tools can be deployed to perform computational analyses and biological inference on protein samples with open-source software with no cost for \ac{NZ}.
+A fine biological interpretation of the results depends on the original question that motivated the study, and thus this chapter did not focus on that, but instead showed how it can be started with the output of the tools presented in the previous chapters.
+The results indicated the tools can be deployed to perform computational analyses and biological inference on protein samples with open-source software with no cost for \ac{NZ}.
 However, the absence of proper missing data handling developed into the inability to quantify hundreds of proteins, signaling what are the next steps to improve the tools.

### 2018-08-02 acknowledgements.tex
Backup and spell check

\section*{Acknowledgements}

The basis for this project stemmed from my eagerness to face new challenges in bioinformatics. For that reason, I am very grateful to Novozymes for having given me the opportunity to carry out my own project, and I really hope my results can be turned into innovative ways on how proteomics analyses are done in the company. Specially, I would like to thank Mathias Gruber for having supported and helped me throughout the journey and Anders M. Bernth Giessing for having come up with a project proposal on mass spectrometry and proteomics, a field of bioinformatics I knew little about at the project start. I would like to thank Samay Mir, Henrik M. Geertz-Hansen and Marija Majer for their suggestions and discussion.

Outside of Novozymes, I would like to thank my supervisor Thomas Hamelryck for his guidance and his feedback. Also, I would like to thank Harald Barsnes and Ludger Goeminne for their great advice and comments regarding the software implementation. Finally, I would like to mention those without whom the project would have been much harder to complete. My friends: Adham Khaled and Carlotta Porcelli and my family, thank you so much.

### 2018-08-02 conclusion.tex
Backup and spell check

@@
 All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools.
-More functionalities can be added, including \textit{de novo} and \ac{PTM} search, and missing data management.
-BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model} can be passed any peptide level dataset and provide estimates of the uncertainty behind frequentist \ac{log2FC} estimates and nuisance effects.
-It can be customised to expand its currently implemented model.
+More functionalities can be added, including \textit{de novo} and \ac{PTM} search, fractioned samples and missing data management.
+BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model}, can be passed any peptide level dataset and provide estimates of the uncertainty beyond frequentist \ac{log2FC} estimates.
+It can be customized to expand its currently implemented model.
 The lack of success in the attempt to model the peptide bias using the aminoacidic sequence reveals the complexity of its origin.
-Recently developed tools like RawQuant could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach.
+Recently developed tools  could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach.
 The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software, and it is encouraged to continue on that line.

### 2018-08-02 modelling.tex
Backup and spell check

@@
 While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and comes equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
+Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and come equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
 This is the so called Bayes\textquotesingle s theorem.
@@
 
-The data is thus said to be affected by different effects acting independently, and in order to model them properly, multilevels need to be defined.
+The data is thus said to be affected by different effects acting independently, and in order to model them properly, multi-levels need to be defined.
 It can be done via three approaches: pooled, unpooled, and hierarchically (see figure \ref{fig:multilevel}).
@@
 
-However the posterior distribution is obtained, a proper Bayesian analysis is not finished without a model checking step that looks at potential problems happenning in the fitting step, and confirms the goodness of fit of the model to the data.
+However the posterior distribution is obtained, a proper Bayesian analysis is not finished without a model checking step that looks at potential problems happening in the fitting step, and confirms the goodness of fit of the model to the data.
 
@@
 It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (ADVI) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
-Therefore, it provides the tools required to build a Bayesian model for the probabilisitc estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
+Therefore, it provides the tools required to build a Bayesian model for the probabilistic estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
 
@@
 This can be done by defining a new normal distribution where its $\mu$ and $\sigma$ are set to the random variables defined above.
-Im PyMC3 code,
+In PyMC3 code,
 
@@
                framesep=2mm]{python}
-# VI approximation: much faster and pretty accurate
+# VI approximation: much faster and accurate
 trace_advi = bayesquant.fit(model_name=p, n_draws=40000)
@@
 \input{hierarch_diagram}
-\caption{Kruschke-like diagram of the Bayesian model}
+\caption[BayesQuant Kruschke-like diagram]{\textbf{BayesQuant Kruschke-like diagram}.
+Extends figure \ref{fig:daft_model} to give more detail into the model.
+Prior probability distributions are indicated with "distrograms", and the dependencies between them via arrows.
+The hyperparameters are represented by latin capital letters.}
 \label{fig:dbda}
@@
 
-The 10k traces produced by BayesQuant from the \ac{VI} approximation function exhibited a clear random walk behaviour, which decreases the posibility that the approximation was wrong.
+The 10k traces produced by BayesQuant from the \ac{VI} approximation function exhibited a clear random walk behaviour, which decreases the possibility that the approximation was wrong.
 Moreover, the  effects caused by the different peptides and runs were estimated and separated from the overall observed noise, allowing for a better estimation of the \ac{log2FC} (see figure \ref{fig:traceplots}).
@@
 Presence or absence of overlap between the 95\% \ac{HPDI} and the Region Of Practical Equivalence (\ac{ROPE}) provides a formal way of discarding a point value of the parameter.
-The \ac{ROPE} is a small interval considered to be essentually the same as the null value \cite{Kruschke}.
-In this and all future applications in the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrality set to $[-0.4, 0.4]$ based on the data.
+The \ac{ROPE} is a small interval considered to be essentially the same as the null value \cite{Kruschke}.
+In this and all future applications in the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.
 
@@
 
-The posterior predictive checks on both proteins pinpoints that the fitted models captured a data generation process approximating what was observed (see figure \ref{fig:ppc}).
-The three observed peptides were found to be undistinguishable from the 500 simulated ones.
+The posterior predictive checks on both proteins pinpoint that the fitted models captured a data generation process approximating what was observed (see figure \ref{fig:ppc}).
+The three observed peptides were found to be indistinguishable from the 500 simulated ones.
 
@@
 The consistent observation that \textit{Homo sapiens} proteins got a \ac{log2FC} estimate around 0, and \textit{E.
-coli} proteins got an estimate which overlapped the predefinned \ac{ROPE} in only three cases (two of which on proteins with two peptides), confirmed the validity of the quantification framework.
+coli} proteins got an estimate which overlapped the predefined \ac{ROPE} in only three cases (two of which on proteins with two peptides), confirmed the validity of the quantification framework.
 
@@
 
-Given the open-source nature of the program, it is possible to build upon it and extend its functionality.
 The possibility of making use of the peptide sequences to model the peptide effect using a linear model was considered and put to practice.
@@
 \begin{equation}
-\nonumber \mu^p \sim \mathcal{N}(\sum_{k=0}^{k=K}{x_k \theta_k}, \sigma^p)
+\nonumber \mu^p \sim \mathcal{N}(\sum_{k=1}^{k=K}{x_k \theta_k}, \sigma^p)
 \end{equation}
@@
 
+A list of points to address to improve how BayesQuant works is discussed in this section.
+
 \subsection{Improving usability: parallelization}
@@
 A solid way of checking the results are not flawed is the specification of alternative priors.
-Unless precise prior knowledge is available, the selection of one prior over another prior to model the beliefs about the system should not have any impact on the results.
+Unless precise prior knowledge is available, the selection of one prior over another to model the beliefs about the system should not have great impact on the results.
 It is for that reason that models built upon different priors could be run to further validate the program.
+Moreover, a formal posterior validation beyond visual inspection of simulated data could be implemented.
 
@@
 
-Besides the predefinition of a \ac{ROPE} and the check of overlap with the 95 \% \ac{HPDI}, more advanced model comparison methods making use of the Bayes factor are available.
-The Bayes factor is defined as the ratio of the posterior probability of the models given the data \cite{Kruschke}.
+Likewise, a formal way of establishing the \ac{ROPE} in the quantification context could be enacted depending on the data.
+On the other hand, a more advanced model comparison method could be built upon the Bayes factor.
+The Bayes factor is defined as the ratio of the posterior probability of two models given the data \cite{Kruschke}.
 It could be used to measure how much more likely a model taking into account a treatment effect is compared to another one where the treatment effect is considered null i.e the \ac{log2FC} is set to 0.
@@
 
-\item The characterization of powerful sequence properties that specifically capture differential ionization efficiency is undone work, as most feature extractors are designed to solve different problems, like protein sructure prediction.
+\item The characterization of powerful sequence properties that specifically capture differential ionization efficiency is undone work, as most feature extractors are designed to solve different problems, like protein structure prediction.
 
@@
 Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of proper neural networks, capable of deciphering the intricate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
+The development of very new RAW spectra-files parsing-tools like RawQuant \cite{Kovalchik2018} could be used in a future implementation to provide the required feature-rich data.
 
@@
 
-A Bayesian framework for the relative quantification of protein abundance rations (Fold-Change, FC) using \ac{MS1} intensity was presented in this chapter.
+A Bayesian framework for the relative quantification of protein abundance ratios (Fold-Change, FC) using \ac{MS1} intensity was presented in this chapter.
 Unlike the currently prevailing methods in the field, it gives uncertainty estimates that provide a direct interpretation about the accuracy of the quantification.
-Moreover, preliminary results on the sequence modelling of the peptide bias revealed that more complex architectures and sequnce features are required to model noise with peptide sequences.
+Moreover, preliminary results on the sequence modelling of the peptide bias revealed that more complex architectures and sequence features are required to model noise with peptide sequences.
 Finally, the method can be used to investigate the presence of batch effects in \ac{MS} experiments for their minimisation in future experiments.

### 2018-08-02 msreview.tex
Backup and spell check

@@
 The bottom-up paradigm is most often used because peptides are much more suitable to analysis by mass spectrometry, as illustrated in section \ref{subsec:the_detector}.
-This review will focus on the Data-Dependent Aquisition (\ac{DDA} approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
+This review will focus on the Data-Dependent Acquisition (\ac{DDA} approach within bottom-up proteomics, which currently is the most frequent workflow in proteomics \cite{Sinitcyn2018}.
 In this approach, the equipment is configured for the analysis of one peptide at a time, with the drawback of being biased for the most abundant peptides.
@@
 An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denaturalized  so as to remove bias due to the divergent properties acquired by folded proteins.
+Proteins are then denatured  so as to remove bias due to the divergent properties acquired by folded proteins.
 Then, proteins are subjected to digestion with specific proteases i.e protein-cutting molecules, which cut the aminoacidic chains following a predictable pattern.
 Trypsin is the most frequently used protease.
-It cuts peptidic bonds whenever a positively charged residue, either Lysine (K) or Arginine (R), lies on the carboxyl side of the peptidic bond.
+It cuts peptidic bonds whenever a positively charged residue, either lysine (K) or arginine (R), lies on the carboxyl side of the peptidic bond.
 Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before.
@@
 Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function \cite{Siepen2007}.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is Proline.
+This is the case of Trypsin whenever the residue on the other side of the peptidic bond is proline.
 Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
@@
 This can be surmounted by analyzing one peptide in the sample at a time.
-The required sample separation is achieved by High-Performance Liquid Cromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
+The required sample separation is achieved by High-Performance Liquid Chromatography (\ac{HPLC}) methods, like reverse phase chromatography (separating on hydrophobicity) and strong cation exchange chromatography (separating on isoelectric point) \cite{Barsnes2008}.
 
@@
 All mass spectrometers exploit the physical properties of mass and electric charge exhibited by the analyzed components.
-Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be unobservable to the equipment.
+Ionization of the analytes is absolutely essential prior to any measurement, as analytes left uncharged will be invisible to the equipment.
 
@@
 The acquired charge yields a mass/charge (\ac{m/z}) ratio, a property that is applied in the downstream component separation and measurement steps.
-It is measured in $\frac{Da}{e}$, sometimes referred to as Th (Thomson).
+It is measured in $\frac{Da}{e}$ (Dalton over elementary charge), sometimes referred to as Th (Thomson).
 
@@
 These are Fourier Transform Ion Cyclotron Resonance (\ac{FT-ICR}) and Orbitrap.
-They both register cylotron resonance frequencies that are Fourier transformed into the spectrum space.
-Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operability \cite{Barsnes2008}.
+They both register cyclotron resonance frequencies that are Fourier transformed into the spectrum space.
+Remarkably, \ac{FT-ICR} exhibits great resolving power, at the cost of high maintenance costs and difficult operation \cite{Barsnes2008}.
 
@@
 They constitute the isotopic envelope of the ion (see figure \ref{fig:envelope}) \cite{Mirzaei2016}, and represent the signal created by peptides containing an increasing number of \ce{^{13}_{}C} atoms or other naturally occurring isotopes.
-Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the mass increases by 1 Da.
+Every time a \ce{^{12}_{}C} is replaced by \ce{^{13}_{}C}, the molecule\textquotesingle s mass increases by 1 Da.
 Even though the natural abundance of \ce{^{13}_{}C} is 1.1 \%, the sheer number of carbon atoms in a peptide makes it likely that at least one or even more carbon atoms will be \ce{^{13}_{}C}, eventually driving the pure \ce{^{12}_{}C} signal to comparatively small intensity values, and down to intensities below the background noise.
@@
 During this step, search engines attempt to match each observed spectrum to theoretical spectra based on the reference proteome.
-Theoretical spectra are obtained via (I) \textit{in silico} simulation of the spectra produced by the peptides contained in the proteome, in turn obtained by performing a virtual cleaveage of the sequences in the proteome, or (II) spectral libraries, consisting of annotated observed spectra \cite{Li2012}.
+Theoretical spectra are obtained via (I) \textit{in silico} simulation of the spectra produced by the peptides contained in the proteome, in turn obtained by performing a virtual cleavage of the sequences in the proteome, or (II) spectral libraries, consisting of annotated observed spectra \cite{Li2012}.
 
@@
 \item \textbf{Peptide inference}: infer the peptides present in the sample.
-\item \textbf{Protein inferece \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
+\item \textbf{Protein inference \textit{proper}}: based on the inferred peptides, infer what proteins generated them.
 This is not trivial as peptides are degenerate and frequently map to more than one protein \cite{Li2012}.
@@
 
-Quantification can be spectrum counting (\ac{SC}) or extractde ion currents (\ac{XIC}) based.
+Quantification can be spectrum counting (\ac{SC}) or extracted ion currents (\ac{XIC}) based.
 

### 2018-08-02 pipeline.tex
Backup and spell check

@@
 
-\begin{figure}[!h]
-\includegraphics[width=\textwidth]{pipeline}
-\caption[Schema of the presented pipeline]{\textbf{Schema of the presented pipeline}.
-.RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
-After that, spectra are searched by means of a search engine to perform the PSM step.
-Thereafter, PSMs are validated and the most likely set proteins is inferred.
-If quantative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
-The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
-\label{fig:pipeline}
-\end{figure}
-
 Several proteomics pipelines are available under different licensing conditions.
 Many are released as closed-source commercial software, where information on how the program works is kept from the user.
-This is a serious drawback as it hinders the study of the implemented models and its customisation.
+This is a serious drawback as it hinders the study of the implemented models and its customization.
 Open-source, free alternatives, like the Trans Proteomic Pipeline (\ac{TPP}) \cite{Deutsch2011} or openMS \cite{Sturm2008} are nevertheless available for the community, but they either lack good documentation, a broad user base or format exchangeability.
@@
 Nevertheless it has been extremely successfully adopted by the scientific community due to its ease of use and comprehensive pipeline.
-The organization of a pipeline with comprehensive documentation, open-source nature, and easy implementation and customisation was asserted to be lacking in the literature.
+The organization of a pipeline with comprehensive documentation, open-source nature, and easy implementation and customization was asserted to be lacking in the literature.
 
@@
 \item Fully command line, documented and Linux-supported interface for easy automation and scalability.
-\item Open-source for easy customisation and extension.
+\item Open-source for easy customization and extension.
 \item Free-licensed and cost-free, so anybody with the knowledge can run it, democratizing the analyses.
@@
 
-The proteome benchmark dataset from \cite{Cox2014} was reanalysed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+The proteome benchmark dataset from \cite{Cox2014} was reanalyzed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{Link to the PRIDE repository entry https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
 Briefly, the \textit{Homo sapiens} and \textit{E.
 coli (strain K12)} proteomes were mixed in 1:1 (condition L) and 1:3 (condition H) proportions, with 3 replicates for each combination.
-Moreover, each of the 3 replicates of the 2 conditions was analysed over 24 fractions.
+Moreover, each of the 3 replicates of the 2 conditions was analyzed over 24 fractions.
 This experimental setting thus generated a total of $2 \times 3 \times 24=144$ RAW files.
 One file was missing in the repository.
-The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.
+The ThermoRawFileParser \footnote{\href{https://github.com/compomics/ThermoRawFileParser}{ThermoRawFileParser repository https://github.com/compomics/ThermoRawFileParser}} program was used to convert RAW files to the MGF open format.
 These files were used as input for the Compomics+Rob pipeline.
 
-The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
+The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the supplemental data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
 The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
@@
 
-The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker version 1.16.22 utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{A summary of the validation filters in PeptideShaker https://github.com/compomics/peptide-shaker/issues/300}{https://github.com/compomics/peptide-shaker/issues/300}}.
+The SearchGUI results were filtered using the default built-in checks available in the PeptideShaker version 1.16.22 utility \texttt{PeptideShakerCLI} \cite{Vaudel2015} \footnote{\href{A summary of the validation filters in PeptideShaker https://github.com/compomics/peptide-shaker/issues/300}{A summary of the default filters in PeptideShaker: https://github.com/compomics/peptide-shaker/issues/300}}.
 By default, the FDR was set to 1\%.
@@
 
-\subsection{Code implementation}
+\section{Results}
 
-\section{Results}
+\subsection{Running the Compomics+MSqRob pipeline}
+
+The pipeline consists of a series of bash, Python and R scripts calling the tools published by the Computational Omics and Systems Biology group at the University of Ghent (see figure \ref{fig:pipeline}).
+
+\begin{minted}[mathescape,
+               linenos,
+               numbersep=5pt,
+               autogobble=true,
+               frame=lines,
+               framesep=2mm]{python}
+
+create_decoy_database.sh
+search_validate.sh
+mbr_apex.sh
+Rscript MSqRob.R
+
+\end{minted}
+
+\begin{figure}[!h]
+\includegraphics[width=\textwidth]{pipeline}
+\caption[Schema of the presented pipeline]{\textbf{Schema of the presented pipeline}.
+.RAW files produced by the spectrometer are converted to an open format like the Mascot Generic Format (\ac{MGF}).
+After that, spectra are searched by means of a search engine to perform the PSM step.
+Thereafter, PSMs are validated and the most likely set proteins is inferred.
+If quantitative information is to be extracted, a quantification step attempting to estimate protein quantities is executed.
+The biological interpretation of the pipeline results can be achieved by interacting with public databases using R/Bioconductor or Python thanks to their open format.}
+\label{fig:pipeline}
+\end{figure}
 
@@
 The total number of spectra per sample ranged between 5894 and 20249.
-\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and viceversa.
+\textbf{A} Percentages are encoded with a blue palette, the darker, the higher, and vice versa.
 The sixth fraction of the first replicate in condition H was missing in the PRIDE data repository.
@@
 
-The match between runs step allowed for increased identifications by transferring successful matches between replicate runs.
+The match-between-runs (\ac{MBR}) step allowed for increased identifications by transferring successful matches between replicate runs.
 The results of this process for the 13th fraction of the L condition is shown in figure \ref{fig:mbr}.
@@
 \caption[Proteome benchmark MBR results]{Match Between Runs with moFF.
-\textbf{A} Count of identifications on each run segreagated by source.
+\textbf{A} Count of identifications on each run segregated by source.
 More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
@@
 The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) across conditions H and L.
-During quantification, it is important to take into account as many of the effects influencing protein quantification.
+It is important to take into account as many of the effects influencing protein quantification as possible.
 In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.
@@
 Likewise, human proteins should have a \ac{log2FC} of 0.
-As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between proteins coming from one or the other organism.
+As such, the evaluation can be reformulated as a classification task were the quantification algorithm tries to distinguish between one of the two organisms.
 
@@
 
-Some of the most frequent evaluations of the performance of a binary classiffier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
+Some of the most frequent evaluations of the performance of a binary classifier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
 Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
 
-In the present problem, they can be used to show how good the \textit{q-val} associated to a protein group is at separating the two proteomes.
+In the present problem, they can be used to show how well the \textit{q-value} separates the two classes (organisms).
 Thus, \textit{E.
@@
 Moreover, a filter based on the \ac{log2FC} can be applied to enrich the data in positives and facilitate the classification task.
+
 In order to compare the results of the here presented pipeline (Comp+Rob) with those achieved by other available tools, the analysis was repeated using the data processing implemented in MaxQuant \cite{Cox2008} at different steps.
-The pipeline combining MaxQuant and MaxLFQ (MQ+LFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+Rob) used MaxQuant up to the fraction normalization step.
+The pipeline combining MaxQuant and MaxLFQ (MQ+LFQ) was fully carried out using the MaxQuant suite, whereas the combination of MaxQuant and MSqRob (MQ+Rob) used MaxQuant without its quantification engine MaxLFQ.
 
@@
 Venn diagrams provide a straightforward visualization of the overlap of several ensembles of proteins.
-The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detectd by MQ+Rob too.
+The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detected by MQ+Rob too.
 Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power.
@@
 
-Additionally, the accuracy with which each program assigns a \ac{log2FC} estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline, as shown in figure \ref{fig:combined_plot}.
+Additionally, the accuracy with which each program assigns a \ac{log2FC} estimate to each protein group, and how significantly different from the null value of 0 it was found to be, provides further insights on the pros and cons of each pipeline.
+
+Two separate distributions were distinguished for each organism, with the most overlap being  observed in the Comp+Rob pipeline.
+Remarkably, a ratio of 1 (\ac{log2FC} of 0) was frequently assigned by MSqRob (see figure \ref{fig:combined_plot} A).
+The significance acquired by \textit{E.
+coli} proteins was highest in the MQ+Rob pipeline.
 This analysis confirmed that the pipeline providing the best separation between organisms was MQ+Rob.
@@
 Sample fractionation allows for a deeper analysis of the peptide mix, which in turn leads to increased identifications and an increased amount of data to be analyzed.
-However, it simultaneously provides an extra random effect that difficults quantification if not handled properly.
+However, it simultaneously provides an extra random effect that hinders quantification if not handled properly.
 
-The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a ponderated average.
-One normalizing weight is determined for every sample via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
+The way the MaxLFQ engine solves the problem is by aggregating the data from the same sample across fractions using a weighted average.
+One normalizing weight is determined for every fraction via the \textit{Levenberg-Marquant minimisation of the overall proteome variation} $H(N)$ as defined in \cite{Cox2014}.
 A custom implementation was written in Python using the scipy module, but the results were not satisfactory.
+An issue regarding this is active on GitHub \footnote{\href{https://github.com/lgatto/MSnbase/issues/344}{https://github.com/lgatto/MSnbase/issues/344}}.
 
-MSqRob treats the fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit2{E.
+MSqRob treats fractions as a random effect, as explained in section \ref{subsec:quantification}.
+As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit{E.
 coli} proteins.
@@
 While no \textit{E.
-coli} proteins fell in this category in the MQ-MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.
+coli} proteins fell in this category in the MQ+MSqRob pipeline, many did in the Comp+Rob, proving room for improvement in this aspect.
 
@@
 The application of mass spectrometry to proteomics is a very hot topic in research, driving innovations in the way the different components of the spectrometer work.
-One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
+One recent example is the development of the timsTOF\texttrademark\xspace Pro (Trapped Ion Mobility Spectrometry-Time Of Flight) system by Bruker, which supports PASEF (Parallel Accumulation and SErial Fragmentation) spectrometry \footnote{\href{https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}{Press release of the PASEF Method https://www.prnewswire.com/news-releases/bruker-launches-the-timstof-pro-mass-spectrometer-to-enable-the-revolutionary-pasef-method-for-next-generation-proteomics-300520791.html}}.
 This technology adds  an extra dimension peptide separation process, leading to cleaner spectra.
@@
 
-The SearchGUI and PeptideShaker recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{https://github.com/compomics/searchgui/wiki/DatabaseHelp}}.
+The SearchGUI and PeptideShaker recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{Database formatting guide https://github.com/compomics/searchgui/wiki/DatabaseHelp}}.
 This is due to the software making use of the annotations to improve the protein inference step.
@@
 
-An open-source, cost-free, platform-agnostic and customisable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
+An open-source, cost-free, platform-agnostic and customizable label-free relative quantification pipeline assembling publicly available tools was presented in this work.
 The pipeline is capable of providing qualitative information on the protein composition of a sample using the latest proteomics search engines and validation software.
-Moreover, it can be extended to support more complex tasks, like \textit{De novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
+Moreover, it can be extended to support more complex tasks, like \textit{de novo} search, or the study of \ac{PTM}s, thanks to its non-monolithic nature and usage of open-formats.
 Likewise, the openness of the data in all steps allows for quality control checks along the way.
-However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractionated data, albeit deficiencies in this aspect should have no impact in non-fractionated datasets.
+However, the comparison against other workflows shows the room for improvement in the quantification step when dealing with fractioned data, albeit deficiencies in this aspect should have no impact in non-fractioned datasets.
 

### 2018-08-02 thp1.tex
Backup and spell check

@@
 
-A benchmark dataset was run through the programmes presented in chapters \ref{chap:pipeline} and \ref{chap:model} to showcase their performance.
-The experimental desgin attempted to capture a protein profile change reflecting a known biological process.
-Thus, a successful computational analysis of the dataset should also reflect the phenomenon.
+A second benchmark dataset was run through the programmes presented in chapters \ref{chap:pipeline} and \ref{chap:model} to showcase their performance.
+The experimental design attempted to capture a protein profile change reflecting a known biological process.
+Thus, a successful computational analysis of the dataset should reflect the same phenomenon.
 The results confirmed that the computational analysis successfully captured this response.
@@
 Cells were exposed to 500 ng/mL LPS (E.
-coli O111:B4) (positive control condition, \textit{PC}) or nothing (negative control condition, \textit{NC}) for 48 hours in 37 degrees, 5\% \ce{{CO}_2}.
+coli O111:B4) (+LPS) or nothing (-LPS) for 48 hours in 37 degrees, 5\% \ce{{CO}_2}.
 After stimulation supernatants were collected and cell pellets were washed in PBS (Gibco).
 Both supernatants and pellets were frozen (-20 degrees).
+This work was carried out by Esben G.
+W.
+Schmidt.
 Only the pellet samples were considered for the analysis.
-EGWS, Esben Gjerløff Wedebye Schmidt.
 
-\subsection{Mass Spectrometry analysis}
-
-AEGI
+%\subsection{Mass Spectrometry analysis}
 
@@
 
-The resulting .RAW files were processed with the two different pipelines fully developed in the present work.
+ deveThe resulting .RAW files were processed with the two different pipelines introduced in the present work.
 (I) Compomics+MSqRob and (II) Compomics+BQ (Compomics and BayesQuant).
@@
 
-The gProfiler R package \cite{Boyd2010} was used to run automatic \ac{GSEA} and provide a biological interpretation of the results.
+The gProfiler R package \cite{Boyd2010} was used to perform \ac{GSEA}, and its results were visualized with the REVIGO tool \cite{Supek2011}.
 
-Pathways analysis was executed by calling the ComPath Webserver \cite{Domingo2018}.
-UniprotKB IDs were transformed to "Gene name" IDs, as required by the tool using the Retrieve/ID mapping tool from Uniprot.
+Pathways analysis was executed with the ComPath Webserver \cite{Domingo2018}.
+UniprotKB IDs were transformed to "Gene name" IDs, as required by the tool, using the Retrieve/ID mapping tool from Uniprot.
 The KEGG \cite{Kanehisa2000} and Reactome \cite{Croft2014} databases were used as source.
@@
 A histogram of the \ac{log2FC} reveals that in most cases, not enough evidence was collected to declare the \ac{log2FC} different from 0 (see figure \ref{fig:compomics_rob} A).
-Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which an non-null estimate was provided were more abundant in the \textit{PC} condition.
-A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria ($qval < 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
+Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which a non-null estimate was provided were more abundant in the +LPS condition.
+A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria ($q-value < 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
 However, all were considered differentially abundant.
@@
 
-A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most overrepresented biological processes, cellular compartments and molecular functions, compared to the human background.
+A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background.
 The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.
@@
 
-The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multiprotein groups) returned by MSqRob returned a list of 51 entities.
-The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were overrepresented compared to the human background.
-50 entities were mapped to the database (only \textit{C12orf42} was unmapped).
+The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multi-protein groups) returned by MSqRob produced a list of 51 entities.
+The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were over-represented compared to the human background.
+50 entities were mapped to the database.
 In agreement with the \ac{GSEA} results, a handful of pathways linked to the immune response were found to be significantly enriched (see table \ref{tab:pathway_enrichment}) (I) Immune System, (II) Innate Immune System, (III) Necroptosis (IV) Programmed Cell Death, (V) Regulation of Apoptosis, (VI) Leukocyte transedothelial migration.
-Moreover, the finding of several RNA related pathways exposes the presence of potential cell reprogramming, essential in any change in protein profile.
+Moreover, the finding of several RNA related pathways exposed the presence of potential cell reprogramming, essential in any change in protein profile.
 
@@
 \caption{Pathway Enrichment results.
-50 proteins were mapped to pathway databses and an enrichment was performed.
+50 proteins were mapped to pathway databases and an enrichment was performed.
 The 10 most significant pathways found in Reactome and Kegg are shown arranged by decreasing significance.
@@
 
-A discussion over the major issues and strenghts detected while obtaining these results follows.
+A discussion over the major issues and strengths detected while obtaining these results follows.
 
@@
 The presence of missing data in many of the rows in the data passed to BayesQuant led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups.
-Thanks to the missing data handling implemented in MSqRob, this tool quantified 1433 proteins, enough to continue the analysis (see table \ref{tab:balance}).
+Thanks to the missing data handling implemented in MSqRob, this tool quantified 1433 proteins, enough to continue the quantitative analysis (see table \ref{tab:balance}).
 However, the fact that the number of observed groups was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.
@@
 
-\item Missing Completely At Random (MCAR): Unpredictable, caused by random errors in the data acquistion step and affecting the whole dataset.
+\item Missing Completely At Random (MCAR): Unpredictable, caused by random errors in the data acquisition step and affecting the whole dataset.
 
-\item Missing At Random (MAR): Caused by other observed variables. Citing \cite{Wei2018}: \textit{ Inaccurate peak detection and deconvolution of co-eluting compounds can be called MAR}.
+\item Missing At Random (MAR): Caused by other observed variables. \textit{ Inaccurate peak detection and deconvolution of co-eluting compounds can be called MAR} \cite{Wei2018}.
 
@@
 Thus, proteins exhibiting abundances below the \ac{LOD} due to the experimental treatment are bound to generate peptide data featuring missing values that, if not handled properly, are discarded.
-Subsequently, \ac{DAP}s or in general proteins with variant abundance become less likely to be quantified, difficulting the extraction of protein sets.
+Subsequently, \ac{DAP}s or in general proteins with variant abundance become less likely to be quantified, complicating the extraction of relevant protein sets.
 

### 2018-08-03 modelling.tex
Backup
Further proofreading

@@
 
-The current label-free quantification methods, reviewed in section \ref{sec:quantification}, rely on frequentist statistics, which return point estimates of model parameters such as the abundance ratio (fold change) across conditions.
+The current label-free quantification methods, presented in chapter \ref{chap:pipeline}, rely on frequentist statistics, which return point estimates of model parameters such as the abundance ratio (fold change) across conditions.
 However, a Bayesian  approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the same benchmark dataset from chapter \ref{chap:pipeline}).
+As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the proteome benchmark dataset.
 The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.
@@
 The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a p-value, which measures the probability of the null hypothesis.
+This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a \textit{p-value}, which measures the probability of the null hypothesis.
 The null hypothesis usually states that the true value of the parameter is 0.
-Thus, p-values do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.
+Thus, \textit{p-values} do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.
 
 The Bayesian statistical framework provides an alternative point of view by revolving the role played by the model parameters and the observed data.
-While frequentist statistics treats the data as random and the parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided data.
-Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only, and come equipped with an equation that mathematically formalizes how to introduce these two dependencies in a valid way.
-This is the so called Bayes\textquotesingle s theorem.
+While frequentist statistics treats data as random and parameters as fixed, a Bayesian framework swaps their roles and yields a probability distribution for any model parameter given the provided, fixed data.
+Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only.
+Unlike \ac{NHST}, it does not depend of the data collection process.
+The equation that mathematically formalizes how to introduce the observed data and the prior in a valid way is the so called Bayes\textquotesingle s theorem.
 
@@
 
-The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterior probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the likelihood divided by a constant}.
+The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterior probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the \textbf{likelihood} divided by a constant \textbf{(evidence)}}.
 
@@
 The posterior distribution can be approximated (inferred) by means of \ac{MCMC} and \ac{VI} methods.
-On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2017}.
+On the one hand, \ac{MCMC} methods simulate sampling from the posterior distribution by constructing an ergodic \footnote{Explanation of ergodic processes \href{https://dsp.stackexchange.com/questions/34976/what-is-a-good-example-of-an-ergodic-process}{https://dsp.stackexchange.com/questions/34976/what-is-a-good-example-of-an-ergodic-process}} Markov chain on $\theta$ whose stationary distribution is the posterior p($\theta$ | y) \cite{Blei2017}.
 Monte Carlo sampling is performed on the Markov Chain, (hence the name of the technique) to randomly explore the parameter space with some heuristics.
 The heuristics guarantee convergence of the empirical estimate on the true posterior, provided a big enough sample size.
-Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018} \footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
-The first sampling algorithms, like Metropolis-Hastings \cite{Chib1995} \footnote{A very nice tutorial on the Metropolis-Hastings algorithm \href{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}} have given way to the much more efficient sampler \ac{NUTS} \footnote{Likewise, this resource provides very illustrative explanations of Hamiltonian-MC and NUTS \href{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}} \cite{Hoffman2011}.
+Sampling convergence is defined as the status reached by the sampler when it estimates a distribution of probability that does not change anymore, regardless of how much longer the sampler runs \cite{Tran2018}.
+The first sampling algorithms, like Metropolis-Hastings \footnote{A very nice tutorial on the Metropolis-Hastings algorithm \href{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}{http://twiecki.github.io/blog/2015/11/10/mcmc-sampling/}} \cite{Chib1995} have given way to the much more efficient sampler \ac{NUTS} \footnote{Likewise, this resource provides very illustrative explanations of Hamiltonian-MC and NUTS \href{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}{http://elevanth.org/blog/2017/11/28/build-a-better-markov-chain/}} \cite{Hoffman2011}.
 
@@
 \textit{The term variational is used because the best $q$ in $Q$ is picked.
-The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation}\footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
+The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation}  \footnote{A high level explanation of \ac{VI} \href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}\label{fn:vi}}.
 
@@
 $p(x)$ is computationally intractable, thus making the computation of the exact \ac{KL} divergence impossible.
-A lower bound up to an additive constant can however be computed thanks to Jensen\textquotesingle s inequality \cite{jensen1906fonctions} \.
+A lower bound up to an additive constant can however be computed thanks to Jensen\textquotesingle s inequality \cite{jensen1906fonctions}.
 This approximate amount is the Variational or Evidence Lower BOund (\ac{ELBO}).
@@
 The mean-field family of densities is one of the most simple, as it ignores all dependencies between latent variables, and treats them as independent.
-The mean-field approximation works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}\footnote{\href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}}.
+The mean-field approximation works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}\footnoteref{fn:vi}.
 A generic mean-field family member is shown in equation \ref{eq:mean_field}.
@@
 
-A mass spectrometry-proteomics workflow aims at providing estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
+A successful mass spectrometry-proteomics workflow  provides estimates of the treatment effect, upon which the biological interpretation of the results is done, while minimising the remaining unwanted effects.
 A mathematical modelling of these effects (see equation \ref{eq:model}), similar to the one used by MSqRob, can be deployed to attempt to minimise the noise:
@@
 This way, separate treatment, peptide and run effects are distinguished.
-$i$ stands for the protein index, which is evidently the same for all measurements belonging to peptides mapped to the same protein.
+$i$ stands for the protein index, which is the same for all measurements belonging to peptides mapped to the same protein.
 $\epsilon$ models the noise that cannot be explained with the three effects aforementioned, and is the same for all measurements associated to the same protein.
 
-For example, the measurement $y_{P,A,p,1}$ corresponds to a measurement mapped to protein $P$.
+For example, the measurement $y_{P,A,p,1}$ corresponds to an observation of protein $P$.
 The measurement is impacted by a fixed effect, produced by treatment $A$, but also the peptide effect in peptide $p$ and the run effect in run $1$.
@@
 Hierarchical modelling is used when the sampling variance is not the only source of variation among the parameters, but they still share a dependency.
-The intercept and the three effects were modelled as independent hierarchies.
+The three effects were modelled as independent hierarchies.
 
@@
 \end{figure}
-\footnotetext{\href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
 
-This way, independent and identically distributed (\ac{IID}) priors were set for the intercept, treatment, peptide and run effects.
-The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} prior.
+\footnotetext{Introduction to multilevel modelling \href{https://docs.pymc.io/notebooks/multilevel\_modeling.html}{https://docs.pymc.io/notebooks/multilevel\_modeling.html}}
+
+This way, independent and identically distributed (\ac{IID}) priors were set for the intercept ($\beta^0$), treatment, peptide and run effects ($\beta$), and $\epsilon$.
+The particular effect observed on each peptide was then modelled as a value sampled from the corresponding \ac{IID} priors.
 A diagram of the resulting model is displayed in figure \ref{fig:daft_model}.
@@
 Therefore, the true posterior could be a bad reflection of the natural process being modelled and the best approximation will not fix it.
-Posterior Predictive Checks are carried out to measure how well the model captured the data properties \cite{Kruschke}.
+Posterior Predictive Checks are carried out to measure how well the model captured the data properties \cite{Kruschke} and reveal the presence of epistemic error.
 
@@
 PyMC3 is a mature Python module dedicated to support probabilistic programming.
-It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (ADVI) \footnote{\href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
+It features   intuitive model specification syntax, modern and powerful \ac{MCMC} sampling algorithms like the No-U-Turn-Sampler (\ac{NUTS}) as well as Automatic Differentiation Variational Inference (ADVI) \footnote{Presentation of PyMC3 \href{https://github.com/pymc-devs/pymc3}{https://github.com/pymc-devs/pymc3}}.
 Therefore, it provides the tools required to build a Bayesian model for the probabilistic estimation of protein \ac{log2FC} in proteomics datasets, that is, together with an estimate of its uncertainty.
@@
 
-The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the dataset published in \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant()} and \texttt{MSnSet2protdata()} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
+The peptides.txt and proteinGroups.txt files produced by MaxQuant \cite{Cox2008} after the analysis of the proteome benchmark dataset \cite{Cox2014} were used as input for the modelling algorithm when running without sequence modelling, and processed using the \texttt{preprocess\_MaxQuant()} and \texttt{MSnSet2protdata()} functions in the MSqRob \cite{Goeminne2016} package, similar to what was done in section \ref{subsec:database_preparation}.
 
-Peptides missing in any of the samples available were dropped from the analysis.
+The sample fractionation present in the dataset was aggregated to yield one column for each of the six samples. Peptides missing in any of the resulting six columns were dropped from the analysis.
 
@@
 This was condensed in the specification of Normal and Half Normal distributions for all the random stochastic variables in the model.
-The hyperparameters were selected to be as non informative as possible.
+The hyperparameters were selected to be non-informative.
 
@@
 
-\subsection{PyMC3 implementation}
+\subsection{PyMC3}
 
@@
 and is read as \textit{the observed data is modelled by a normal distribution with mean $m$ and standard deviation $\epsilon$.
-$m$ is a random deterministic variable resulting from the sum the effects defined above.
+$m$ is a random deterministic variable resulting from the sum of the effects defined above.
+$i$ indexes every effect.
 $\epsilon$ follows a new half normal distribution with standard deviation 1}.
@@
 Prior probability distributions are indicated with "distrograms", and the dependencies between them via arrows.
-The hyperparameters are represented by latin capital letters.}
+The hyperparameters are represented by latin capital letters.
+Each $\beta$ will consist of a vector of dimensions $1 \times n$ where $n$ is the number of different factors in the category (i.e 2 for treatment, 6 for run, etc).
+The corresponding $x$ variable is a matrix of dimensions equal to $n \times n_peptides$ where the $i,j$ cell indicates if the $jth$ peptide is affected by the $ith$ factor (1) or not (0).}
 \label{fig:dbda}
@@
 
-A plot of the maximisation of the \ac{ELBO} is automatically produced when the texttt{.fit()} method is run.
+A plot of the maximisation of the \ac{ELBO} is automatically produced when the \texttt{.fit()} method is run.
 Even though the ELBO took extremely low (negative) values at start (see figure \ref{fig:ELBO}), it was maximised fast to values around 0, indicating that the mean-field approximation was acceptably good.
@@
 The 95\% \ac{HPDI} is marked with a black line.
-The \ac{ROPE} defined as a 0.4 window around 0 is shown in red.
+The \ac{ROPE}, defined as a 0.4 window around 0, is shown in red.
 A green vertical line marks the expected \ac{log2FC} estimate for \textit{E.
@@
 Each pair of groups was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}).
-The more peptides, the more accurate the estimation will be as more data will be available.
+The more observed peptides, the more accurate the estimation will be as more data will be available.
 Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed (see figure \ref{fig:dumbbell}).
@@
 The intervals are represented by horizontal lines.
-The dot represents the mean of the whole distribution, and it will be centered in the interval if the distribution is symmetrical.
+The dot represents the mean of the whole distribution, and it will be centered in the interval if its probability is symmetrical.
 Vertical blue lines represent the ROPE defined as in figure \ref{fig:posteriors}, whereas the red line represents the expected estimate for bacterial proteins.}
@@
 
-While BayesQuant performed well on the majority of proteins, it took around 30 seconds to approximate each protein\textquotesingle s posterior distributions using \ac{VI}, and more than two minutes with \ac{MCMC} methods.
-This translates to a dataset of 1k proteins taking 500 minutes (>8h) to be processed.
-However, the average modern computer comes with several processors that could support the parallelization of the program in several threads, decreasing the total computing time.
+While BayesQuant performed well on the majority of proteins, it takes around 30 seconds to approximate each protein\textquotesingle s posterior distributions using \ac{VI}, and more than two minutes with \ac{NUTS}.
+This translates to a dataset of 1k proteins taking 500 minutes (>8h) to be processed with the fastest method.
+Fortunately, the average modern computer comes with several processors that could support the parallelization of the program in several threads, decreasing the total computing time.
 
@@
 The Bayes factor is defined as the ratio of the posterior probability of two models given the data \cite{Kruschke}.
-It could be used to measure how much more likely a model taking into account a treatment effect is compared to another one where the treatment effect is considered null i.e the \ac{log2FC} is set to 0.
+It could be used to measure how much more likely a model taking into account a treatment effect is compared to another one where the treatment effect is considered null i.e the \ac{log2FC} is set to 0 \cite{Goodman1999}.
 The result would provide an alternative criterion for the declaration of a protein being differentially abundant.
@@
 
-Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of proper neural networks, capable of deciphering the intricate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
+Modelling such a complex phenomenon will require accordingly complex predictive architectures and the introduction of machine learning algorithms, capable of deciphering the intricate pattern mapping sequence to noise in the spectrometer\textquotesingle s measurements.
 The development of very new RAW spectra-files parsing-tools like RawQuant \cite{Kovalchik2018} could be used in a future implementation to provide the required feature-rich data.
@@
 The characterization of the uncertainty behind a \ac{log2FC} estimate can be used to better inform downstream analysis programs and eventually help interpret the biological results.
-For example, a functional analysis program could decide to drop proteins declared differentially abundant if the uncertainty is greater than a threshold.
+For example, a functional analysis program could decide to drop proteins declared differentially abundant if the uncertainty is greater than a threshold or if the Bayes factor does not indicate a high preference for a treatment effect.
 A Gene Set Enrichment Analysis (\ac{GSEA}) could make use of the probabilistic information to refine its results.

### 2018-08-03 pipeline.tex
Further proofreading

@@
 
-A pipeline making use of the set of tools published by the Compomics  and StatOmics groups SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
+A pipeline making use of the tools SearchGUI \cite{Barsnes2018}, PeptideShaker \cite{Vaudel2015}, moFF \cite{Argentini2016} and MSqRob \cite{Goeminne2016}, was developed to support complete label-free protein quantification analyses using the most recent advances in the field with open-source software.
 The pipeline can be run on Linux computer clusters to perform (I) peptide to spectrum matching against a reference database, (II) quality control and filtering, (III) protein inference, (IV) \ac{MBR} and feature extraction, and (V) relative quantification.
@@
 
-The proteome benchmark dataset from \cite{Cox2014} was reanalyzed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{Link to the PRIDE repository entry https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
+The proteome benchmark dataset \cite{Cox2014} was reanalyzed starting at the output RAW files available at the PRIDE repository \footnote{\href{https://www.ebi.ac.uk/pride/archive/projects/PXD000279}{Link to the PRIDE repository entry https://www.ebi.ac.uk/pride/archive/projects/PXD000279}}.
 Briefly, the \textit{Homo sapiens} and \textit{E.
@@
 The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the supplemental data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
-The results of the MQ+Rob pipeline employed the Levenberg-Marquandt minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
+The results of the MQ+Rob pipeline employed the Levenberg-Marquardt-minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
 
@@
 This required passing the original RAW files, together with the Default PSM report from PeptideShaker.
-Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity in each sample.
+Output was exported to a peptide summary file, containing  one row per peptide and for every peptide, the detected apex intensity for each sample in a column
 
@@
 Significance was evaluated by means of the two-tailed Welch Two Sample t-test implemented in the \texttt{t.test()} function in R.
-P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R.
+P-values were corrected using the FDR method implemented in the \texttt{p.adjust()} function in R with the "fdr" method.
 
@@
 
-The pipeline consists of a series of bash, Python and R scripts calling the tools published by the Computational Omics and Systems Biology group at the University of Ghent (see figure \ref{fig:pipeline}).
+The pipeline consists of a series of bash, Python and R scripts calling the tools published by the Compomics and StatOmics groups at the University of Ghent (see figure \ref{fig:pipeline}).
+A parameter file must be created before running the pipeline.
 
@@
 As expected, matches to the target and decoy databases exhibited similar score distributions at low score values, while a divergence is observed at higher score values (see figure \ref{figure:qc_validation} A).
-Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1\% FDR filter.
+Likewise, the m/z error was found to be closer to 0 on validated PSMs than on those which did not pass the 1\% FDR filter (see figure \ref{figure:qc_validation} C).
 The application of this filter implied that the FNR (false negative rate) was set to 5 \%, i.e 5 out of every 100 discarded matches were estimated to be true positives.
@@
 More than 4k spectra were identified and validated by SearchGUI+peptideShaker.
-In this particular case, hundreds of new identifications were accomplished.
+In this particular case, hundreds of new identifications were accomplished via moFF.
 \textbf{B} Match visualization.
@@
 \textbf{Comp+Rob} is the pipeline presented in the previous sections.
-\textbf{MQ+LFQ} consists a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
+\textbf{MQ+LFQ} consists of a standard MaxQuant upstream analysis combined with its LFQ quantification engine.
 Finally, \textbf{MQ+Rob} blends MaxQuant upstream analysis with the MSqRob quantification engine.
@@
 
-Some of the most frequent evaluations of the performance of a binary classifier are Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
+Some of the most frequent evaluations of the performance of a binary classifier are the Receiver Operating Characteristic (ROC) and Precision-Recall (PR) curves \cite{Bradley1997}, which demonstrate the performance of the algorithm at different cutoff values of a predictor variable.
 Together, they show the False Positive Rate (FPR), the True Positive Rate (TPR) or recall, and the precision exhibited by the algorithm.
@@
 Nonetheless, the latter performed clearly better when the \ac{log2FC} filter was applied, indicating that a combined fold change and q-value criteria is best for declaring proteins as differentially abundant.
-In all cases MQ+Rob was as good as MQ+LFQ, if not much better.
+In all cases MQ+Rob was the best performing classifier.
 
@@
 The scrutiny of the counts reveals a high degree of agreement, with most proteins detected by Comp+Rob or MQ+LFQ being detected by MQ+Rob too.
-Remarkably, 517 proteins were detected by MQ+Rob alone, which further confirmed its resolving power.
+Remarkably, 517 proteins were detected by MQ+Rob alone, showcasing its resolving power.
 Analyzing the log2FC pairwise-correlation (see figure \ref{fig:venn_cor} B) reveals that the highest correlation was found between pipelines using MaxQuant as upstream spectra processor, as expected.
@@
 Only MS-GF+ was used in this work for simplicity.
-Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
+Furthermore, \textit{de novo} search engines are available and supported by SearchGUI, and the latest developments in the field, like the IdentiPy engine \cite{Levitsky2018} are progressively incorporated to the tool \footnote{Issue on how to add IdentiPy\href{https://github.com/compomics/peptide-shaker/issues/309}{https://github.com/compomics/peptide-shaker/issues/309}}.
 Their usage is predicted to improve identification rates and quality of the matches.
 
-The most important reason why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
+A key reason on why many spectra remain unidentified is the presence of post-translational modifications (\ac{PTM}s), which exponentially increase the search space, forcing most workflows to discard many of the peptides featuring a PTM.
 New approaches to the problem are emerging, mainly machine learning methods for the handling of unexpected modifications \cite{Gabriels}.
@@
 
-MSqRob treats fractions as a random effect, as explained in section \ref{subsec:quantification}.
-As a consequence, the statistical model in MSqRob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit{E.
+If fraction aggregation is not supported, MSqRob can be set to treat fractions as a random effect, as explained in section \ref{subsec:quantification}.
+However, if this is done, the statistical model in Comp+Rob can be overwhelmed by the inconsistency resulting from a 24-plex sample fractionation, leading to null \ac{log2FC} estimates for many \textit{E.
 coli} proteins.
-This is clearly manifested in figure \ref{fig:combined_plot} as the bins at \ac{log2FC} of 0, which represent protein groups for which MSqRob did not find enough evidence to discard the null hypothesis.
+This is clearly manifested in figure \ref{fig:combined_plot} B, which shows many bacterial proteins were assigned  a \ac{log2FC} of 0, indicating MSqRob did not find enough evidence to discard the null hypothesis.
 While no \textit{E.
@@
 
-The SearchGUI and PeptideShaker recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{Database formatting guide https://github.com/compomics/searchgui/wiki/DatabaseHelp}}.
-This is due to the software making use of the annotations to improve the protein inference step.
-In any case, a consistent formatting in the FASTA databases is required to used the tools.
-Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with these and many other tools.
+The SearchGUI and PeptideShaker developers recommend using Uniprot-formatted databases for optimal performance of their tools \footnote{\href{https://github.com/compomics/searchgui/wiki/DatabaseHelp}{Database formatting guide https://github.com/compomics/searchgui/wiki/DatabaseHelp}}.
+This is due to the software making use of protein annotations to improve the protein inference step.
+In any case, a consistent formatting in the FASTA databases is required to use the tools.
+Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with nost just these but many other tools too.
 

### 2018-08-03 thp1.tex
Further proofreading

@@
 
-\item Propose a function inference analytical pipeline to complete the proteomics workflow.
+\item Propose a biological inference pipeline to complete the proteomics workflow.
 
@@
 THP-1 cells were grown in RPMI1640 (Sigma) supplemented with 10\% FBS (SeraLab) and 1\% pen/strep (Gibco).
-Stimulation of THP-1 cells was done in 48-well plates (Gibco) with 2x106 cells/well.
+Stimulation of THP-1 cells was done in 48-well plates (Gibco) with $2 \times 10^6$ cells/well.
 Cells were exposed to 500 ng/mL LPS (E.
@@
 
-%\subsection{Mass Spectrometry analysis}
+\subsection{Mass Spectrometry analysis}
+
+Peptide digest of THP-1 samples were prepared using a filter assisted sample preparation protocol (FASP).
+Online desalting and nanoscale LC separation of tryptic peptides was performed with an Dionex Ultimate 3000 Dual –column nano-liquid chromatography system (Thermo Scientific) equipped with two Acclaim PepMap nano-trap columns, 20 mm, 75µ I.D., 3µ particles and two analytical Acclaim PepMap C18, 150 mm, 75µ capillary columns with 2µ particles (Thermo Scientific).
+Mobile phase A was 0.1\% formic acid (FA) in water and mobile phase B was 0.1\% FA in acetonitrile.
+Peptides were separated using a 60 min gradient from 5\% to 30\% mobile phase B at 350 nl/min and 35°C injecting samples on the two analytical columns in sequential randomized design.
+Eluting peptides were analyzed using data-dependent acquisition on an LTQ-Orbitrap Velos Pro mass spectrometer (Thermo Scientific).
+Each MS scan (400-1600 m/z) was acquired at a resolution of 30000 FWHM followed by 10 low resolution LTQ MS/MS scans triggered above an intensity of 10000 using collision induced dissociation CID (normalized collision energy 35).
+This work was carried out by David Gleerup and Anders M.
+B.
+Giessing.
+
+%In short the FASP method is a filter aided in-solution sample preparation technique comprising of protein extraction with strong chaotropic reagents such as urea or Guanidine-HCL, protein precipitation with TCA, and digestion using trypsin under denaturing conditions of reduced and alkylated protein.
+Tryptic peptides are extracted from the filter and analyzed by LC/MS.
+
+%Acquired MS data was processed and annotated using Genedata Expressionist.
+Acquired MS/MS spectra were searched against Homo sapiens reference proteome database taken using an in-house Mascot server ver.
+2.5.1 (Matrix Science).
+Searches were conducted as follows: semi-tryptic specificity allowing up to two missed cleavages, 10 ppm peptide and 0.6 Da fragment ion tolerances, with carboiodimethyl of cysteine as fixed modification as well as variable modifications consisting of oxidation of methionine and deamidation (NQ).
+The final protein list was compiled when 2 or more unique peptides per protein were identified with a with a false Discovery Rate (FDR) cutoff of 1%, estimated using a target-decoy search strategy.
 
@@
 
- deveThe resulting .RAW files were processed with the two different pipelines introduced in the present work.
+The resulting .RAW files were processed with the two different pipelines introduced in the present work.
 (I) Compomics+MSqRob and (II) Compomics+BQ (Compomics and BayesQuant).

### 2018-08-03 conclusion.tex
Final improvements

@@
 Recently developed tools  could be applied in future work to extract a feature-rich representation of the collected spectra that could be used to model this phenomenon with a Deep Learning approach.
-The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software, and it is encouraged to continue on that line.
+The results from chapter \ref{chap:benchmark} manifest that \ac{NZ} can at the same time decrease computational costs and access the state of the art in proteomics research by making use of the latest open-source software.
+In summary, this work grants exciting opportunities for developing workflows in the field.

### 2018-08-03 introduction.tex
Backup and spell check
Final improvements

@@
 
-Novozymes A/S \ac{NZ} is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes.
+Novozymes A/S (\ac{NZ}) is a company whose line of business consists of the development of products performing chemical transformations in economically relevant processes.
 The application of these products, instead of conventional solutions, has the advantage of requiring less chemical substances, potentially simplifying industrial processes, reducing their costs and their environmental impact \cite{madigan2000brock}.
@@
 The company\textquotesingle s name is inspired by the molecular agents that power these transformations, \textbf{enzymes}.
-Enzymes consist of a extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.
+Enzymes consist of an extremely diverse family of molecules catalyzing a plethora of biochemical reactions happening in living beings, from bacteria to humans \cite{Nelson2008}.
 
@@
   \caption[The 20 aminoacids]{Molecular structure of the 20 aminoacids, organized by chemical properties into five groups.
-Taken from \footnote{\href{https://cnx.org/contents/vxf032Bc@4/Proteins}{https://cnx.org/contents/vxf032Bc@4/Proteins}}.}
+Taken from \href{https://cnx.org/contents/vxf032Bc@4/Proteins}{https://cnx.org/contents/vxf032Bc@4/Proteins}.}
   \label{fig:aminoacids}

### 2018-08-03 modelling.tex
Final improvements

@@
 However, a Bayesian approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, a statistical model, named BayesQuant, was written in the probabilistic programming framework PyMC3 and tested on the proteome benchmark dataset.
+As a response to this shortcoming, the thesis focused on writing a statistical model, named BayesQuant, in the probabilistic programming framework PyMC3 and testing it on the proteome benchmark dataset.
 The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.
@@
 Thus, \textit{p-values} do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.
+Is there a framework that reports results for a particular hypothesis, based just on the seen data and our prior knowledge?
 
@@
 All is needed is the computation of the posterior probability distribution.
-Unfortunately, this is tractable analytically for simple models only.
-However, Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior.
-Both of which are now possible to run thanks to the advent of modern computing.
+Even though, this is tractable analytically for simple models only, fortunately Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior when the analytical derivation is not possible.
+Both \ac{MCMC} and \ac{VI} are now possible to run thanks to the advent of modern computing.
 
@@
 The \ac{ROPE} is a small interval considered to be essentially the same as the null value \cite{Kruschke}.
-In this and all future applications in the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.
+In this and all future applications of the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.
 
@@
 
-In order to further validate BayesQuant\textquotesingle s performance, it was tested on groups of 5 proteins, one group from each proteome.
-Each pair of groups was made by proteins with a different number of observed peptides (2, 3, 4, 6, 7, 10) (see figure \ref{fig:dumbbell}).
-The more observed peptides, the more accurate the estimation will be as more data will be available.
-Since the \ac{HPDI} reflects the uncertainty, a decrease on its width was expected to be observed (see figure \ref{fig:dumbbell}).
+Could these results be extended to other proteins in the dataset and validate the tool?
+Six groups of 5 proteins, with a different number of observed peptides (2, 3, 4, 6, 7, 10) were organized in each proteome and employed to test BayesQuant (see figure \ref{fig:dumbbell}).
+The more observed peptides, the more accurate the estimation is expected to be as more data for the same protein becomes available.
+Since the \ac{HPDI} reflects the uncertainty, a decrease on its width is thus expected to be observed (see figure \ref{fig:dumbbell}).
 

### 2018-08-03 msreview.tex
Final improvements

@@
 
-The mass spectrometer is the ensemble of pieces of equipment analyzing a peptide like as those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
+The mass spectrometer is the ensemble of pieces of equipment analyzing peptides like those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
 It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.
@@
 \includegraphics[width=\textwidth]{mass_spectrometer}
-\caption[Mass spectrometer diagram]{Schematic view of a mass spectrometer. Taken from \footnotemark{}}
+\caption[Mass spectrometer diagram]{Schematic view of a mass spectrometer.
+Taken from \href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}.}
 \label{fig:mass_spectrometer}
 \end{figure}
-
-\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
-\stepcounter{footnote}
+%
+%\footnotetext{\href{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}{https://www.slideshare.net/joachimjacob/bits-introduction-to-mass-spec-data-generation}}
+%\stepcounter{footnote}
 
@@
 
-The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly time of flight (\ac{TOF}), Ion trap (\ac{IT}) and quadrupole (\ac{Q}).
+The plethora of ion separation methods is reflected upon the range of different analyzers available, mainly Time Of Flight (\ac{TOF}), Ion Trap (\ac{IT}) and Quadrupole (\ac{Q}).
 These apply different principles to perform the same task: separation (analysis) of the ion mix by the \ac{m/z} ratio.

### 2018-08-03 pipeline.tex
Final improvements

@@
 
-Relative quantification was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob).
+Relative quantification i.e the computation of $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob).
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
@@
 
-The extracted apex MS1 intensities were used as proxy for peptide abundance to estimate protein $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) across conditions H and L.
-It is important to take into account as many of the effects influencing protein quantification as possible.
+Could the apex MS1 intensities extracted by moFF be used as proxy for peptide abundance and perform relative quantification?
+It is important to take into account as many of the effects influencing the recorded data as possible.
 In this case, besides the different H and L conditions, peptide, run and fraction effects could be distinguished.

### 2018-08-03 thp1.tex
Final improvements

@@
 Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which a non-null estimate was provided were more abundant in the +LPS condition.
-A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria ($q-value < 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
+A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria (\textit{q-value} $< 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
 However, all were considered differentially abundant.
@@
 
-A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background.
+A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) with the gProfileR \cite{gprofiler} illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background.
 The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.
@@
 
-These results can be better visualized by making use of a treemap (see figure \ref{fig:treemap}), which makes the interpretation of table \ref{tab:gsea} more straightforward.
+A Revigo \cite{Supek2011} treemap (see figure \ref{fig:treemap}) makes the interpretation of table \ref{tab:gsea} more straightforward.
 
@@
 The UniprotKB to Gene name mapping of the set of 49 \ac{DAP}s (including multi-protein groups) returned by MSqRob produced a list of 51 entities.
-The list was supplied to the ComPath Pathway Enrichment tool to highlight which cellular pathways were over-represented compared to the human background.
+The list was supplied to the ComPath Pathway Enrichment tool \cite{Domingo2018} to highlight which cellular pathways were over-represented compared to the human background.
 50 entities were mapped to the database.
@@
 \centering
-\includegraphics[width=\linewidth]{string_normal_image}
+\includegraphics[width=\linewidth]{string_normal_image_confidence}
 \caption[Protein interaction network]{\textbf{STRING interaction network}.
@@
 Evidence is either experimentally determined or predicted using diverse methods.
+The thickness of the edges represents the level of confidence in that interaction.
 Red colored nodes are included in the "immune system process" set.}
@@
 One of the most frequent issues found when analyzing proteomics datasets is the handling of missing values \cite{Lazar2016}.
-The presence of missing data in many of the rows in the data passed to BayesQuant led to the dropping of 4803 out of 6443 peptides, producing a peptide dataset of only 1641 data points that accounted for 350 protein groups.
-Thanks to the missing data handling implemented in MSqRob, this tool quantified 1433 proteins, enough to continue the quantitative analysis (see table \ref{tab:balance}).
-However, the fact that the number of observed groups was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.
+The presence of missing datapoints led BayesQuant to drop 4803 out of 6443 peptides, producing a peptide dataset of only 1641 rows that accounted for 350 protein groups.
+Thanks to the missing data handling implemented in MSqRob, the latter quantified 1433 proteins, enough to continue the quantitative analysis (see table \ref{tab:balance}).
+However, the fact that the number of observed groups in the original dataset was 2787 implies that the management of missing values could potentially be improved in both tools, and particularly in BayesQuant.
 
@@
 Missing data should be imputed at the peptide level.
-The implementation of a module performing this task prior to quantification would  improve the accuracy and throughput of the quantification process.
 Several approaches have been proposed, some within a probabilistic framework \cite{Lazar2016}.

### 2018-08-04 modelling.tex
add network figure
Fix definition of p-value

@@
 The results of this analysis will depend not only of the observed data, but also the data generation process \cite{Kruschke}.
-This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a \textit{p-value}, which measures the probability of the null hypothesis.
+This way, both tools return point estimates of the \ac{log2FC} between a pair of conditions, and a \textit{p-value}, which measures the probability of the data under the null hypothesis.
 The null hypothesis usually states that the true value of the parameter is 0.
-Thus, \textit{p-values} do not say anything about the probability of hypotheses outside $H_0$ and $H_1$.
-Is there a framework that reports results for a particular hypothesis, based just on the seen data and our prior knowledge?
+Thus, \textit{p-values} do not say anything about the probability of the hypotheses.
+Is there a framework that reports hypotheses probabilities, based just on the seen data and our prior knowledge?
 
@@
 Even though, this is tractable analytically for simple models only, fortunately Markov Chain-Monte Carlo (\ac{MCMC}) and Variational Inference (\ac{VI}) methods can be used to approximate this posterior when the analytical derivation is not possible.
-Both \ac{MCMC} and \ac{VI} are now possible to run thanks to the advent of modern computing.
 
@@
 
-On the other hand, \ac{VI} methods, very recently developed and introduced in PyMC3, provide a fast alternative to \ac{MCMC} methods, yet they are not guaranteed to asymptotically approximate the true posterior.
+On the other hand, \ac{VI} methods, very recently developed and introduced in PyMC3, provide a fast alternative to \ac{MCMC}, yet they are not guaranteed to asymptotically approximate the true posterior.
 In \ac{VI}, optimization, instead of sampling, is employed as the way to approximate the posterior \cite{Blei2017}.
@@
 
-Both because of the little knowledge on the value of the parameters governing the data generation process and to provide a prior that everyone can agree upon, non informative priors were provided to the model.
+Both because of the little knowledge on the value of the parameters governing the data generation process and to provide a prior that everyone can agree upon, weakly informative priors were provided to the model.
 This was condensed in the specification of Normal and Half Normal distributions for all the random stochastic variables in the model.
@@
 Each $\beta$ will consist of a vector of dimensions $1 \times n$ where $n$ is the number of different factors in the category (i.e 2 for treatment, 6 for run, etc).
-The corresponding $x$ variable is a matrix of dimensions equal to $n \times n_peptides$ where the $i,j$ cell indicates if the $jth$ peptide is affected by the $ith$ factor (1) or not (0).}
+The corresponding $x$ variable is a matrix of dimensions equal to $n \times n_peptides$ where the $i,j$ cell indicates if the $jth$ peptide is affected by the $ith$ factor (1) or not (0).
+Style based on \cite{Kruschke}.}
 \label{fig:dbda}
@@
 
-\subsection{Sampling from the approximation to the posterior}
+\subsection{Sampling from the approximated posterior}
 \label{subsec:basic_model}

### 2018-08-05 introduction.tex
Carlotta correct I

@@
 Proteomics is the branch of bioinformatics whose purpose is the application of algorithms and data analysis to biological studies dealing with proteins.
-With proteomics, one endeavors to infer the protein composition of a sample, and eventually quantify its protein amounts.
+With proteomics, one endeavors to infer the protein composition of a sample, and subsequently quantify its protein abundances.
 
 More concretely, proteomics studies how proteins are capable of coordinating their functions by means of complex systems of regulation.
-Proteomics aims at unraveling the complexity, understanding how its powerful properties emerge from its individual components \cite{Milo2002}.
-For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of a frequent disturbance of these regulations: cancer \cite{Saraon2012}.
-The application of proteomics to biological problems is not restricted to medical problems such as cancer, but also biotechnological processes like those mentioned above.
+Proteomics aims at unraveling this complexity, and understanding how its powerful properties emerge from its individual components \cite{Milo2002}.
+For instance, the interplay of proteomics and biotechnology helps researchers learn which proteins are key in the onset of disturbances to these regulations \cite{Saraon2012}.
+The application of proteomics to biological problems is not restricted to medical problems such as cancer, but also biotechnological processes like those carried out by \ac{NZ}.
 

### 2018-08-05 msreview.tex
Carlotta correct I

@@
 
-The main source of data in proteomics is \ac{MS}.
+The main source of data in proteomics is Mass Spectrometry (\ac{MS}).
 However, different approaches to how this technique is used make for two paradigms in proteomics analysis: top-down and bottom-up \cite{Joshi2016}.
@@
 
-MS is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for eventually thousands of sample components.
-A detailed explanation of the sample processing required prior to MS is given in section \ref{sec:sample_processing}, while an overview on mass spectrometers is given in section \ref{sec:the_mass_spectrometer}.
-The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein sample.
+\ac{MS} is performed by means of a mass spectrometer, an ensemble of pieces of equipment that can acquire mass measurements for eventually thousands of analytes.
+A detailed explanation of the sample processing required prior to \ac{MS} is given in section \ref{sec:sample_processing}, while an overview on mass spectrometers is given in section \ref{sec:the_mass_spectrometer}.
+The result of the MS analysis is a dataset that, with adequate computational analysis tools, is enough to perform the inference steps required to gather knowledge about the original protein composition in a sample.
 These inference steps can be condensed to the peptide and protein inference problems, explained in section \ref{sec:inference}.
-A third computational challenge needs to be solved if quantitative, and not just qualitative information, is to be gained from the experiment.
+A third computational challenge needs to be solved if quantitative, and not just qualitative information, has to be gained from the experiment.
 This is the quantification problem, explained in section \ref{sec:quantification}.
@@
 An \ac{MS} experiment starts with the generation of a protein sample from the biological system of interest.
-Proteins are then denatured  so as to remove bias due to the divergent properties acquired by folded proteins.
+Proteins are then denatured  so as to remove bias due to the divergent properties acquired by their folded state.
 Then, proteins are subjected to digestion with specific proteases i.e protein-cutting molecules, which cut the aminoacidic chains following a predictable pattern.
@@
 It cuts peptidic bonds whenever a positively charged residue, either lysine (K) or arginine (R), lies on the carboxyl side of the peptidic bond.
-Since roughly 1/10 residues are either R or K, the average peptide length is 10 residues, as mentioned before.
+Since roughly 2/20 residues are either R or K, the average peptide length is 10 residues, as mentioned before.
 As demonstrated in section \ref{subsec:the_detector}, this length distribution is fitted to the resolution of the MS analyzer.
-Moreover, in as much R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed in most cases to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}.
-All of these properties combined, together with its low prize, makes Trypsin the protease of choice in this step for most cases.
+Moreover, for as much as R and K are positively charged aminoacids (see figure \ref{fig:aminoacids}), the resulting peptide is guaranteed, in most cases, to be able to capture at least one charge, which is key in the \ac{MS} workflow as described in section \ref{sec:the_mass_spectrometer}.
+All of these properties combined, together with its low price, makes Trypsin the protease of choice in this step for most cases.
 
@@
 
-\item Unexpected cleavages occurring either in the wet-lab procedure of the proteolytic treatment.
+\item Unexpected cleavages occurring in the wet-lab procedure of the proteolytic treatment.
 
@@
 Missed cleavages can happen due to steric impediments or the presence of specific aminoacids that can weaken the enzyme\textquotesingle s function \cite{Siepen2007}.
-This is the case of Trypsin whenever the residue on the other side of the peptidic bond is proline.
+This is the case of Trypsin whenever the residue on the other side of the peptidic bond is a proline.
 Altogether, a variability is created in the cleavage process that, though limited, needs to be taken care of in downstream analysis, as it could introduce biases in peptide observability.
@@
 During \ac{HPLC}, the peptide mix is loaded into a column containing a stationary and a solid phase.
-These phases create an environment where peptides interact differently based on their physico-chemical properties, set by the nature of the phases.
-The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e the amount of time passed before the peptide is observed in the mass spectrometer.
+These phases create an environment where peptides interact with the column differently based on their physico-chemical properties, set by the nature of the phases.
+The output of the column, called elute, will consist of subsets or fractions of peptides leaving the column at different retention times (\ac{RT}) i.e.
+the amount of time passed before the peptide is observed in the mass spectrometer.
 Therefore, the input to the spectrometer will consist of few peptides at any given time.
@@
 
-The mass spectrometer is the ensemble of pieces of equipment analyzing peptides like those generated following the workflow enunciated in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
+The mass spectrometer is the ensemble of pieces of equipment analyzing peptides like those generated following the workflow described in sections \ref{sec:sample_processing} and \ref{subsec:peptide_separation}.
 It consists of three main parts: an ion source, a mass analyzer, and a detector (see figure \ref{fig:mass_spectrometer}) \cite{Barsnes2008}.
@@
 The most frequent ionization methods in proteomics are Matrix-Assisted Laser Desorption-Ionization (\ac{MALDI}) and Electro Spray Ionization (\ac{ESI}) \cite{Mirzaei2016}.
-Most peptides ionized by MALDI will acquire a single charge, whereas ESI can provide multiple charges (+2, +3, etc) too \cite{Joshi2016}.
-Thus, the charge exhibited by an ion is not obvious when produced via ESI.
-Moreover, ESI can be run online with the right robotic equipment, while MALDI demands waiting time for vacuum generation.
-Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW) \cite{Hessling2013}, thus introducing a bias \cite{Hessling2013}.
-Bias in ESI is less predictable.
+Most peptides ionized by \ac{MALDI} will acquire a single charge, whereas \ac{ESI} can provide multiple charges (+2, +3, etc) \cite{Joshi2016}.
+Thus, the charge exhibited by an ion is not obvious when produced via \ac{ESI}.
+Moreover, \ac{ESI} can be run online with the right robotic equipment, while \ac{MALDI} demands waiting time for vacuum generation.
+Finally, due to the chemical nature of the matrix components, MALDI ionizes more easily peptides containing aminoacids featuring aromatic rings (PYW) \cite{Hessling2013}, thus introducing a bias \cite{Hessling2013}, whereas bias in \ac{ESI} is less predictable.
 All the sources of bias introduced during ionization cause the competitive ionization problem \cite{Zhang2009} \cite{Tang2004}.
@@
 The ion\textquotesingle s \ac{m/z} ratio is known thanks to the previous mass analysis step.
-Performed for enough \ac{m/z} ratios, the detector can produce a \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
+Run on enough \ac{m/z} ratios, the detector can produce an \ac{MS} spectrum, which shows the intensity of ion current over an \ac{m/z} range.
 Some topics in signal detection in \ac{MS} need to be discussed.
 
-On the one hand, the precision of the signal measurement is given by its mass resolution It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
+On the one hand, the precision of the signal measurement is given by its mass resolution.
+It is conventionally defined as the closest distinguishable separation between two peaks of equal height and width \cite{Marshall2013}.
 The resolution decreases as the \ac{m/z} ratio increases because small increments in the \ac{m/z} ratio become negligible at high \ac{m/z} ratios.
@@
 The resolution achieved by modern equipment allows for the distinction of each individual signal in most isotopic envelopes.
-Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide with the following expression.
+Remarkably, the separation across peaks in the envelope can be used to infer the charge of the peptide with the following expression (see equation \ref{eq:envelope}):
 
-\begin{equation}
+\begin{equation}\label{eq:envelope}
 m/z = \frac{(m + z × H^+)}{z} \text{for \textit{z}} \in {1, 2,...}
@@
 
-\item The ions filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
+\item The ions (ionized peptides) filtered in the first spectrometer undergo fragmentation in the second spectrometer. %, as explained in \ref{subsec:fragmentation}.
 
@@
 Fragmentation occurs in the second spectrometer.
-But what is it, and why is it done?
-The information that can be extracted from \ac{MS1} scans consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
+What is it, and why is it done?
+The information that can be extracted from \ac{MS1} outputs consists of the \ac{m/z} ratio and retention time of the peptide ion, but not its sequence.
 Having the latter is paramount if the spectrum is to be matched to a theoretical peptide.
-Fortunately, the peptide sequence of the ion can be inferred if fragmentation is performed.
+Fortunately, the peptide sequence of the ions can be inferred when fragmentation is performed.
 
@@
 Therefore, the lists of matches needs to be somehow ranked by goodness-of-fit.
-The issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
+This issue is addressed by search engines through the deployment of statistical models that provide scoring systems.
 Assuming the correct protein is present in the database, a good scoring system should give the best score to the right peptide.
@@
 Using this method, the search engine replicates the matching process, using the same spectra, but instead against a decoy database i.e it simulates random matches.
-The decoy database is generated by reversing or more generally applying a randomization technique upon the sequences present in the original (target) database \cite{Elias2010}.
+The decoy database is generated by reversing, or more generally, by applying a randomization technique upon the sequences present in the original (target) database \cite{Elias2010}.
 
 All matches to the decoy are by definition wrong.
-Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target, the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score.
+Since the basic properties of the decoy (size, composition, etc) remain identical to those of the target, the amount of matches to the decoy exhibiting more than a given score $s$ can be regarded as an estimate of the number of false identifications ($\hat{n}_{fp}$) in the list of target results exhibiting at least the same score (see equation \ref{eq:fdr}).
 This is because the existence of shared properties entails that random matches are equally likely to happen in both databases \cite{Elias2010}.
@@
 
-The equation tells us that the FDR at any score $s$ can be computed by counting how many decoy hits have a score greater than $s$ ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
+Equation \ref{eq:fdr} tells us that the FDR at any score $s$ can be computed by counting how many decoy hits have a score greater than $s$ ($\hat{n}_{fp}$), and dividing by the length of our target hits list.
 Thus, the score that makes the FDR equal to a predefined value, frequently 0.01 or 1 \%, can be computed and used as threshold for the target hits.
@@
 
-The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e it is the smallest FDR we can allow while still keeping the PSM \cite{Nesvizhskii2010}.
+The minimal FDR at which a given PSM is considered a valid match constitutes the PSM\textquotesingle s q-value, i.e.
+it is the smallest FDR we can allow while still keeping the PSM \cite{Nesvizhskii2010}.
 Related to the q-value, the Posterior Error Probability (PEP) is an estimate of the probability of a given PSM of being an incorrect assignment (see figure \ref{fig:pep}).
@@
 
-As stated in \cite{Cox2014}, \textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
-These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}.
-For this reason, only the \ac{XIC} approach will be regarded in the rest of the manuscript.
+As stated in \cite{Cox2014}, "\textit{although the abundance of proteins and the probability of their peptides being selected for \ac{MS/MS} sequencing are correlated to some extent, \ac{XIC}-based methods should clearly be superior to \ac{SC} given sufficient resolution and optimal algorithms.
+These advantages are most prominent for low-intensity protein/peptide species, for which a continuous intensity readout is more information-rich than discrete counts of spectra}".
+For this reason, only the \ac{XIC} approach will be regarded in the rest of this manuscript.
 
@@
 
-As stated in \cite{Goeminne2015} \textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
+As stated in \cite{Goeminne2015} "\textit{peptides originating from the same protein can indeed be considered technical replicates and theoretically should lead to similar abundance estimates.
 However, the summarization of the peptide intensities into protein expression values is cumbersome, and most summarization-based methods do not correct for differences in peptide characteristics or for the between-sample differences in the number of peptides that are identified per protein.
-This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}.
+This might introduce bias and differences in uncertainty between the aggregated protein expression values, which are typically ignored in downstream data analysis steps}".
 

### 2018-08-05 pipeline.tex
Carlottas last corrections

@@
 
-The MQ+LFQ pipeline results were obtained starting at the Supplemental table 1 from the MaxLFQ paper supplemental data \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the supplemental data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
+The MQ+LFQ pipeline results were obtained starting at the supplemental table 1 from the MaxLFQ publication \footnote{\href{http://www.mcponline.org/content/13/9/2513/suppl/DC1}{Link to the data http://www.mcponline.org/content/13/9/2513/suppl/DC1}} \cite{Cox2014}, which are available as an excel file.
 The results of the MQ+Rob pipeline employed the Levenberg-Marquardt-minimised peptide intensities stored in the peptides.txt file contained in the spectraHeLaEColi.zip in the supplemental data.
@@
 
-Relative quantification i.e the computation of $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob).
+Relative quantification i.e.
+the computation of $\log_2$(ratios), or logarithm base 2 fold changes, $\log_2(FC)$ (\ac{log2FC}) was performed using the MSqRob package \cite{Goeminne2016} by passing the peptide summary file from moFF (Compomics+Rob) or the peptides.txt from MaxQuant (MQ+Rob).
 Prior to quantification, the data was preprocessed using the \texttt{preprocess\_MSnSet()} function.
@@
 
-Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the supplemental data file.
+Quantification in the MQ+LFQ pipeline was executed using the LFQ intensity columns stored in the data file.
 Intensities were $\log_2$ transformed and averaged.
@@
 
-The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence of at least 65\% (see figure \ref{figure:qc_validation} B).
+The 1\% FDR cutoff selected PSMs with a score higher than 78, which translated to a confidence (probability of the match being true) of at least 65\% (see figure \ref{figure:qc_validation} B).
 
@@
 In any case, a consistent formatting in the FASTA databases is required to use the tools.
-Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with nost just these but many other tools too.
+Thus, the development of sequence databases featuring consistent FASTA identifiers is highly advisable to make the best use of the data and improve the interplay with not just these but many other tools too.
 

### 2018-11-22 tex/conclusion.tex
Clean repo

--- 2018-08-03 conclusion.tex
@@
 
-The pipelines presented in chapter \ref{chap:pipeline} offer an output-rich and easy to deploy cost-free pieces of software.
+The pipelines presented in chapter \ref{chap:pipeline} offer output-rich and easy to deploy cost-free pieces of software.
 All of them have pros and cons, and their combined usage was found to provide the best results, underlining the importance of the development of programs implementing open data formats exchangeable across tools.
 More functionalities can be added, including \textit{de novo} and \ac{PTM} search, fractioned samples and missing data management.
-BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model}, can be passed any peptide level dataset and provide estimates of the uncertainty beyond frequentist \ac{log2FC} estimates.
+BayesQuant, the probabilistic framework for relative quantification, presented in chapter \ref{chap:model}, can be passed a peptide level dataset and provide estimates of the uncertainty beyond frequentist \ac{log2FC} estimates.
 It can be customized to expand its currently implemented model.

### 2018-11-22 tex/modelling.tex
Clean repo

--- 2018-08-04 modelling.tex
@@
 However, a Bayesian approach to this problem, that we know of, is lacking in the literature.
-As a response to this shortcoming, the thesis focused on writing a statistical model, named BayesQuant, in the probabilistic programming framework PyMC3 and testing it on the proteome benchmark dataset.
-The execution of the three steps required when doing Bayesian modelling, mainly (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.
+As a response to this shortcoming, this thesis focused on writing a statistical model, named BayesQuant, in the probabilistic programming framework PyMC3 and testing it on the proteome benchmark dataset.
+The execution of the three steps required when doing Bayesian modelling, (I) model implementation, (II) computation of posterior probabilities and (III) model checking, will be described for this particular problem in the present chapter, together with a discussion on its usability, its strengths and its weaknesses.
 
@@
 Bayesian analyses rely on the observed data and prior knowledge about the phenomenon only.
-Unlike \ac{NHST}, it does not depend of the data collection process.
+Unlike \ac{NHST}, the results obtained with a Bayesian approach do not depend of the data collection process.
 The equation that mathematically formalizes how to introduce the observed data and the prior in a valid way is the so called Bayes\textquotesingle s theorem.
@@
 
-The Bayes\textquotesingle s theorem formulated above is thus read as \textit{the posterior probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the \textbf{likelihood} divided by a constant \textbf{(evidence)}}.
+The Bayes\textquotesingle s theorem formulated above is thus read as: \textit{the posterior probability of the parameter $\theta$ is equal to the \textbf{prior} probability distribution times the \textbf{likelihood} divided by a constant \textbf{(evidence)}}.
 
@@
 
-\textit{The term variational is used because the best $q$ in $Q$ is picked.
-The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation}  \footnote{A high level explanation of \ac{VI} \href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}\label{fn:vi}}.
+"\textit{The term variational is used because the best $q$ in $Q$ is picked.
+The term derives from the "calculus of variations," which deals with optimization problems that pick} [...] \textit{a particular $q$ in $Q$, specified by setting some variational parameters i.e the knobs on $Q$ that need to be turned to get a good approximation}"  \footnote{A high level explanation of \ac{VI} \href{http://www.cs.jhu.edu/~jason/tutorials/variational.html}{http://www.cs.jhu.edu/~jason/tutorials/variational.html}\label{fn:vi}}.
 
@@
 This approximate amount is the Variational or Evidence Lower BOund (\ac{ELBO}).
-\textit{The \ac{ELBO} is the negative \ac{KL} divergence of plus $logp(x)$, which is a constant with respect to $q(z)$.
+"\textit{The \ac{ELBO} is the negative \ac{KL} divergence of plus $logp(x)$, which is a constant with respect to $q(z)$.
 Thus maximizing
-the \ac{ELBO} is equivalent to minimizing the \ac{KL} divergence} \cite{Blei2017}.
+the \ac{ELBO} is equivalent to minimizing the \ac{KL} divergence}" \cite{Blei2017}.
 
@@
 The mean-field family of densities is one of the most simple, as it ignores all dependencies between latent variables, and treats them as independent.
-The mean-field approximation works by \textit{pretending that the variables are just behaving that way "on their own." The mean-field method throws away all of the interactions}\footnoteref{fn:vi}.
+The mean-field approximation works by "\textit{pretending that the variables are just behaving that way on their own.
+The mean-field method throws away all of the interactions}"\footnoteref{fn:vi}.
 A generic mean-field family member is shown in equation \ref{eq:mean_field}.
@@
 
-\item Evidence for non-convergence must be collected. Albeit lack of evidence of non-convergence does not guarantee convergence, the existence of evidence definitely proofs it \cite{Kruschke}.
+\item Evidence for non-convergence must be collected. Albeit lack of evidence of non-convergence does not guarantee convergence, the existence of evidence definitely proves it \cite{Kruschke}.
 
@@
 The \ac{ROPE} is a small interval considered to be essentially the same as the null value \cite{Kruschke}.
-In this and all future applications of the thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.
+In this and all future applications of this thesis, the null value is set to 0, and the \ac{ROPE} is arbitrarily set to $[-0.4, 0.4]$ based on the data.
 

### 2018-11-22 tex/thp1.tex
Clean repo

--- 2018-08-03 thp1.tex
@@
 
-A second benchmark dataset was run through the programmes presented in chapters \ref{chap:pipeline} and \ref{chap:model} to showcase their performance.
+A second benchmark dataset was run through the program	s presented in chapters \ref{chap:pipeline} and \ref{chap:model} to showcase their performance.
 The experimental design attempted to capture a protein profile change reflecting a known biological process.
@@
 
-Peptide digest of THP-1 samples were prepared using a filter assisted sample preparation protocol (FASP).
+Peptide digestions of THP-1 samples were prepared using a filter assisted sample preparation protocol (FASP).
 Online desalting and nanoscale LC separation of tryptic peptides was performed with an Dionex Ultimate 3000 Dual –column nano-liquid chromatography system (Thermo Scientific) equipped with two Acclaim PepMap nano-trap columns, 20 mm, 75µ I.D., 3µ particles and two analytical Acclaim PepMap C18, 150 mm, 75µ capillary columns with 2µ particles (Thermo Scientific).
@@
 
-In the Compomics+MSqRob pipeline, a protein was declared to be differentially abundant (\ac{DAP}) between the 2 conditions assayed if the result of the student\textquotesingle s-T test implemented in MSqRob returned a q-value less than 0.05.
+In the Compomics+MSqRob pipeline, a protein was declared to be differentially abundant (\ac{DAP}) between the 2 conditions assayed if the result of the student\textquotesingle s-T test implemented in MSqRob returned a q-value lower than 0.05.
 The criteria in the Compomics+BayesQuant pipeline was the absence of overlap between the \ac{ROPE} [-0.4, 0.4] and the 95\% \ac{HPDI}.
@@
 
-The gProfiler R package \cite{Boyd2010} was used to perform \ac{GSEA}, and its results were visualized with the REVIGO tool \cite{Supek2011}.
+The gProfiler R package \cite{gprofiler} was used to perform \ac{GSEA}, and its results were visualized with the REVIGO tool \cite{Supek2011}.
 
@@
 Remarkably, the distribution was shifted toward negative values, depicting that most of the proteins for which a non-null estimate was provided were more abundant in the +LPS condition.
-A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria (\textit{q-value} $< 0.05$), displayed an absolute value of the \ac{log2FC} less than 1.
+A volcano plot (see figure \ref{fig:compomics_rob} B), reveals that the majority of the 49 protein groups passing the significance criteria (\textit{q-value} $< 0.05$), displayed an absolute value of the \ac{log2FC} lower than 1.
 However, all were considered differentially abundant.
@@
 \caption{THP-1 MSqRob results.
-32 single-protein groups were found to be differentially abundant under significance criteria of q-value less than 0.05.
+32 single-protein groups were found to be differentially abundant under significance criteria of q-value lower than 0.05.
 The protein id, the gene name, the point estimate of its \ac{log2FC} and its significance is shown}
@@
 
-A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) with the gProfileR \cite{gprofiler} illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background.
+A \ac{GSEA} performed on the list of 32 single-protein \ac{DAP}s produced by MSqRob (see table \ref{tab:thp1_rob_results}) with gProfileR \cite{gprofiler} illustrates the most over-represented biological processes, cellular compartments and molecular functions, compared to the human background.
 The 20 most significant terms found in the dataset are displayed in table \ref{tab:gsea}.
@@
 \centering
-\includegraphics[width=\linewidth]{string_normal_image_confidence}
+\includegraphics[width=\linewidth]{string_normal_image_immune_confidence}
 \caption[Protein interaction network]{\textbf{STRING interaction network}.
@@
 
-Six samples organized in two different treatments of three biological replicates without sample prefractionation each were analyzed.
+Six samples organized in two different treatments of three biological replicates without sample prefractionation each, were analyzed.
 The resulting data throughput allowed for the performance of biological inference, as it successfully captured the undergoing biological processes.
